### 2014-02-28 exception-analysis/reviews/esop2014/ESOP-REBUTTAL.txt
icfp etc.

We would like to thank the reviewers for their constructive feedback.

==REMARKS==

    CBN/CBV:

        The primary reason for choosing a call-by-name language lies in our
        ambition of scaling the analysis to the full Haskell language. This
        choice does influence several technical details (the need for an
        exception type for both argument and result type of a function, as
        well as an exception annotation on the function space constructor
        itself). Somewhat counter-intuitively the imprecise exception semantics
        complicate things for the analysis as well (we need non-emptiness
        guards). But these are details we want to handle given our ambitions.

        Some of the general ideas of the algorithm, such as the use of
        implication constraints, could be transfered to an analysis for a
        call-by-value language.

    Related work:

        Due to the page limit we had to make some concessions in the Related
        Work section. We will try to incorporate some of the suggestions,
        criticism and pointers to other related work offered by the reviewers.

    Constraint solving:

        We do not have a fully worked-out proof of correctness of the
        constraint solver, but one can at least convince oneself that the
        algorithm terminates as follows:

            During each iteration of the while-loop either W will decrease or
            \theta will increase. As long as \Lambda_\iota satisfy the
            ascending chain condition (trivially if they are finite) the
            algorithm will terminate.

        To see that the algorithm finds the least solution, note that
        constraints appearing on the left-hand side of an implication are
        monotonic when used as a predicate (as in the if-then statement
        within the while-loop). Intuitively: once a guard holds in iteration
        $n$ of the algorithm, it will never fail to hold during a later
        iteration. This property can fail to hold for more expressive
        constraint languages, as explained in Section 4.5.

==QUESTIONS==

    - Reviewer 1(iv): The expressions which are typeable in the system in
            Figure 2 are exactly those expressions which are typeable in the
            underlying type system. Furthermore, types in the analysis have
            the same shape as the underlying types. So 'f 3 4 5' is indeed not
            typeable, but this is because it would not be typeable in the
            underlying (monomorphic) type system with f : A -> B for base
            types A, B. Instead, f would be given the type

                forall a b c d. a -> b -> c -> d.

            An alternative presentation in terms of an annotated type system
            (intertwining both the underlying type system and the analysis)
            would be possible, but would in our opinion complicate the
            presentation more than it would help to clarify.

    - Reviewer 1(v): Section 6 presents an extension of the analysis. It is
            not included in Figure 2.

    - Reviewer 2: Section 6 sketches an algorithm that handles polymorphism.

    - Reviewer 3: "Isn't the type language of 4.2 missing some base type
            constructors?": No, these are all subsumed by the type variable
            \alpha. Concrete base type are part of the underlying type system,
            not the analysis.

### 2014-02-28 exception-analysis/prototype/pma-v2/Analysis.hs
prototypes

@@ -1,0 +1,1993 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+module Analysis where
+
+import Prelude hiding (($!))
+
+import Control.Monad
+import Control.Monad.State
+
+import qualified Data.Graph      as G
+import qualified Data.Graph.Util as G
+import qualified Data.List       as L
+import qualified Data.Map        as M
+import qualified Data.Map.Util   as M
+import           Data.Maybe
+import qualified Data.Set        as S
+import qualified Data.Set.Util   as S
+import qualified Data.Tree       as T
+
+import Common
+import Constr
+
+-- | Legacy 
+
+injExn :: Exn -> Exn
+injExn (Unif e) = Conc (S.singleton (ExnVar e))
+
+-- | Abstract Syntax
+
+data Expr' ann    -- annotated expression
+    = Var'  Name                                     ann
+    | Con'  Con                                      ann
+    | Abs'  Name (Expr' ann)                         ann
+    | Fix'  Name (Expr' ann)                         ann
+    | App'       (Expr' ann) (Expr' ann)             ann
+    | Let'  Name (Expr' ann) (Expr' ann)             ann
+    | If'   Lbl  (Expr' ann) (Expr' ann) (Expr' ann) ann
+    -- Operators
+    | Op'   Op   (Expr' ann) (Expr' ann)             ann
+    -- Pairs
+    | Pair'      (Expr' ann) (Expr' ann)             ann
+    | Fst'       (Expr' ann)                         ann
+    | Snd'       (Expr' ann)                         ann
+    -- List
+    | Nil'                                           ann
+    | Cons'      (Expr' ann) (Expr' ann)             ann
+    | Case' Lbl  (Expr' ann) (Maybe (Expr' ann)) (Maybe (Name, Name, Expr' ann)) ann
+    deriving (Eq, Ord, Show)
+
+fakeType :: Expr -> Expr' Ty'
+fakeType (Var x) = Var' x fake
+fakeType (Con c) = Con' c fake
+fakeType (Abs x e) = Abs' x (fakeType e) fake
[…]

### 2014-02-28 exception-analysis/prototype/pma-v2/Common.hs

@@ -1,0 +1,222 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+module Common (
+    -- Syntax
+    Name(), augment, only, Lbl,
+    Expr(..), Con(..), ExnC(..), Op(..), freshIdents, name, fromName,
+    _f, _x, _xs, _y, _ys, _z, _zs, _a, _as, _k, _id, _const,
+    -- LaTeX
+    LaTeX(..), command, maybeColor, space, align, newline, newpage,
+    mathbf, section, subsection, subsubsection,
+    gather, latexSet, latexMap, preamble, postamble
+) where
+
+import qualified Data.List  as L
+import qualified Data.Map   as M
+import           Data.Maybe
+import qualified Data.Set   as S
+
+-- | Names and Labels
+
+newtype Name = Name String deriving (Eq, Ord, Show)
+
+augment :: String -> Name -> Name
+augment prefix (Name name') = Name (prefix ++ name')
+
+only :: String -> S.Set Name -> S.Set Name
+only tag = S.filter (\(Name name) -> take (length tag) name == tag)
+
+type Lbl = String   -- FIXME: primarily used for debugging, but we will
+                    -- eventually need something like this for error reporting
+
+-- * Predefined names for Main
+
+name :: String -> Name
+name = Name
+
+fromName :: Name -> String
+fromName (Name name) = name
+
+_f, _x, _xs, _y, _ys, _z, _zs, _a, _as, _k, _id, _const :: Name
+_f     = Name "f"
+_x     = Name "x"
+_xs    = Name "xs"
+_y     = Name "y"
+_ys    = Name "ys"
+_z     = Name "z"
+_zs    = Name "zs"
+_a     = Name "a"
+_as    = Name "as"
+_k     = Name "k"
+_id    = Name "id"
+_const = Name "const"
+
+-- | Abstract Syntax
+
+data Expr
+    = Var Name
+    | Con Con
+    | Abs Name Expr
+    | Fix Name Expr
+    | App Expr Expr
+    | Let Name Expr Expr
+    | If  Lbl  Expr Expr Expr
+    -- Operators
+    | Op  Op   Expr Expr
+    -- Pairs
+    | Pair Expr Expr
+    | Fst  Expr
+    | Snd  Expr
+    -- Lists
+    | Nil
+    | Cons Expr Expr
[…]

### 2014-02-28 exception-analysis/prototype/pma-v2/Revert.hs

@@ -1,0 +1,31 @@
+theModule
+  = let risers1
+          = let r2
+                  = \ d ->
+                      case d of
+                          e : es -> (es, e)
+              in
+              let r1
+                    = \ a ->
+                        \ b ->
+                          \ c ->
+                            if b then ((c : snd a) : fst a) else ((c : []) : (snd a : fst a))
+                in
+                let rx'ry
+                      = let rx'ry_f
+                              = \ rx'ry_x ->
+                                  (\ u ->
+                                     case u of
+                                         [] -> []
+                                         v : vs -> case vs of
+                                                       [] -> ((v : []) : [])
+                                                       w : ws -> r1 (snd (rx'ry_f ()) w ws) (v <= w)
+                                                                   v,
+                                   \ x -> \ y -> r2 (fst (rx'ry_f ()) (x : y)))
+                          in rx'ry_f
+                  in fst (rx'ry ())
+      in
+      let main
+            = risers1 (1 : (3 : (5 : (8 : (5 : (3 : (6 : (3 : []))))))))
+        in main
+

### 2014-02-28 exception-analysis/prototype/pma-v2/Sample.hs

@@ -1,0 +1,34 @@
+module Sample where
+{-
+-- Prelude functions
+length []     = 0
+length (b:bs) = 1 + length bs
+
+map f []     = []
+map f (y:ys) = f y : map f ys
+            
+filter p xs = case xs of
+                [] -> []
+                (y:ys) -> if p y then y : filter p ys else filter p ys
+                
+
+main = let l = [1,3,5] in filter (\z -> z <= length l) l
+-}
+-- Risers
+
+risers1 =
+  let r1 a b c = if b then (c : snd a) : (fst a) else [c] : (snd a : fst a)
+
+      r2 d = case d of
+               (e:es) -> (es, e)
+
+      rx = \u -> case u of
+                     []     -> []
+                     (v:vs) -> case vs of
+                                 []     -> (v : []) : []
+                                 (w:ws) -> r1 (ry w ws) (v <= w) v
+      ry = \x y -> r2 (rx (x : y))
+
+  in rx
+  
+main = risers1 [1,3,5,8,5,3,6,3]

### 2014-02-28 exception-analysis/prototype/pma-v3/Analysis.hs

@@ -1,0 +1,427 @@
+{-# LANGUAGE FlexibleContexts      #-}
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+module Analysis where
+
+import Debug.Trace               as Debug
+
+import Control.Monad
+import Control.Monad.State
+
+import qualified Data.Graph      as G
+import qualified Data.Graph.Util as G
+import qualified Data.List       as L
+import qualified Data.Map        as M
+import qualified Data.Map.Util   as M
+import           Data.Maybe
+import qualified Data.Set        as S
+import qualified Data.Set.Util   as S
+import qualified Data.Tree       as T
+
+import Subst
+import Syntax
+import Fresh
+import FTV
+import Types
+import Shapes
+import Constr
+import Ref
+import Exn
+
+-- | Pattern-matching types and type schemes
+
+-- * Types
+
+type PmTy = (Sh, Constr Ref Ref, Constr Exn Ref)
+    -- FIXME: should perhaps be a datatype, as we run into overlapping instances
+
+instance FTV PmTy where
+    ftv (sh, c1, c2) = ftv sh `S.union` ftv c1 `S.union` ftv c2
+
+-- * Type schemes
+
+data PmScheme
+    = Forall (S.Set Name) {- (S.Set Name) -} PmTy
+    deriving (Eq, Show) -- FIXME: dangerous, we often want eq modulo alpha-renaming
+    
+instance FTV PmScheme where
+    ftv (Forall q pmty) = ftv pmty `S.difference` q
+    
+scheme :: PmTy -> PmScheme
+scheme = Forall S.empty
+
+-- | Instantiation
+
+inst :: PmScheme -> InferM PmTy
+inst (Forall q (sh, c1, c2))
+    = do q' <- replicateM (S.size q) (fresh :: InferM Name)
+         let s = M.fromList (zip (S.toList q) q')
+         return (s $@ sh, s $@ c1, s $@ c2)
+
+-- | Generalization
+
+-- gen :: x -> PmTy -> PmScheme -- FIXME: needs env and constr
+gen env pmty
+    = let -- polyvariant
+          q = ftv pmty `S.difference` ftv env
+          -- monovariant
+          -- q = S.empty
+       in Forall q pmty
+
+schemeFromShape :: Sh -> PmScheme
+schemeFromShape sh = Forall S.empty (sh, S.empty, S.empty)
+
[…]

### 2014-02-28 exception-analysis/prototype/pma-v3/Constr.hs

@@ -1,0 +1,285 @@
+{-# LANGUAGE FlexibleContexts      #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE ScopedTypeVariables   #-}
+
+module Constr where
+
+import qualified Data.Graph      as G
+import qualified Data.Graph.Util as G
+import qualified Data.List       as L
+import qualified Data.List.Util  as L
+import qualified Data.Map        as M
+import qualified Data.Map.Util   as M
+import           Data.Maybe
+import qualified Data.Set        as S
+import qualified Data.Set.Util   as S
+import qualified Data.Tree       as T
+
+import Names
+import FTV
+import Subst
+import Shapes
+
+-- | Subset constraints
+
+type Constr a b = S.Set (Constr' a b)
+
+data Constr' a b
+    = Sh      :<# Sh             -- "undecomposed" constraints
+    | CC (S.Set b) Name Sh Sh    -- FIXME: "undecomposed" conditionals, i.e. we
+                                 -- need to rethink our constraints
+    | LHS a b :<: Name           -- decomposed constraints
+    deriving (Eq, Ord, Show)
+    
+data LHS a b
+    = V Name
+    | C (S.Set a)
+    | I (S.Set b) Name (S.Set a) -- FIXME: need to be careful here to maintain
+                                 --        monotonicity, B should not grow?
+                                 --        (i.e. B should be variable free, or
+                                 --         the computation staged, so variables
+                                 --         can be substituted away.)
+                                 --        on the other hand, propagating a
+                                 --        conditional constraint might be sound
+                                 --        in out situation (but imprecise)
+    deriving (Eq, Ord, Show)
+    -- FIXME: - merge V and C using a general Free-purpose data type
+    --        - add a bit vector framework-like constraint
+    
+instance (ConstrElem a, ConstrElem b) => FTV (Constr' a b) where
+    ftv (sh1  :<# sh2 )
+        = ftv sh1 `S.union` ftv sh2
+    ftv (CC _ name sh1 sh2) 
+        = name `S.insert` ftv sh1 `S.union` ftv sh2
+    ftv (V v1 :<: v2  )
+        = S.fromList [v1, v2]
+    ftv (C sa :<: name) 
[…]

### 2014-02-28 exception-analysis/prototype/pma-v3/Examples.hs

@@ -1,0 +1,227 @@
+module Examples where
+
+import Names
+import Syntax
+
+-- | Examples
+
+allExamples =
+    -- basic constructs
+    [("id", exId)
+    ,("const", exConst)
+    ,("id True", exIdApp)
+    ,("const 42", exConstApp1)
+    ,("const 42 False", exConstApp2)
+    ,("\\f x. f x", exApply)
+    ,("(\\f x. f x) id", exApplyId)
+    ,("(\\f x. f x) id id", exApplyIdId)
+    ,("4", exCon1)
+    ,("-3", exCon2)
+    ,("\\k -> False", exCon3)
+    ,("let y = (\\x -> if x then False else x) in y", exLet1)
+    ,("let y = (\\x -> if x then False else x) in y True", exLet2)
+    ,("let y = (\\x -> if x then False else x) in (y True, y False)", exLet3)
+    ,("if True then 3 else -3", exIf1)
+    ,("\\x -> if x then 3 else -3", exIf2)
+    ,("(\\x -> if x then 3 else -3) True", exIf2App)
+    ,("\\x -> if x then False else x", exIf3)
+    ,("(\\x -> if x then False else x) True", exIf3App)
+    ,("5 <= 6", exOp1)
+    ,("\\x y -> x + y", exOp2)
+    -- polyvariance
+    ,("(id CRASH, id True)", exIdPoison)
+    -- conditionals
+    ,("(\\x -> if x then CRASH else True) False", ex17)
+    -- complex functions
+    ,("length", exLength)
+    ,("length [1,2,3]", exLengthApp)
+    ,("map", exMap)
+    ,("map (+1) [1,2,3] {- inlined -}", exMapPlus1_Fix)
+    ,("map (+1) [1,2,3] {- let-bound -}", exMapPlus1_Let)
+    ,("map length []", exMapLengthE)
+    ,("map length [[1,2,3],[],[4]]", exMapLengthNE)
+    ,("risers", exR0)
+    ,("risers (if False then [] else [2,3,1])", exR1)
+    ,("risers (if True then [] else [2,3,1])", exR2)
+    ,("map risers [...]", exMapRisers)
+    -- crashing
+    ,("[..,[],..]", someList)
+    ,("crashing list", crashingList)
+    ,("head []", exHeadNil)
+    ,("map head NE[]", exMapHeadNE)
[…]

### 2014-02-28 exception-analysis/prototype/pma-v3/Exn.hs

@@ -1,0 +1,35 @@
+{-# LANGUAGE MultiParamTypeClasses #-}
+
+module Exn where
+
+import qualified Data.Set as S
+
+import Subst
+import FTV
+import Syntax
+import Constr
+         
+-- * Exceptions
+
+data Exn
+    = ExnVar   Name
+    | ExnCrash Lbl
+    deriving (Eq, Ord)
+    
+instance Show Exn where
+    show (ExnVar name)  = show name
+    show (ExnCrash lbl) = "CRASH_" ++ lbl
+    
+instance FTV Exn where
+    ftv (ExnVar name) = S.singleton name
+    ftv _             = S.empty
+    
+instance ConstrElem Exn where
+    injName = ExnVar
+    
+    prjName (ExnVar n) = Just n
+    prjName _          = Nothing
+    
+instance Substitute Name Exn where
+    subst $@ ExnVar a = ExnVar (subst $@ a)
+    _     $@ exn      = exn

### 2014-02-28 exception-analysis/prototype/pma-v3/FTV.hs

@@ -1,0 +1,15 @@
+module FTV where
+
+import qualified Data.Set        as S
+import qualified Data.Set.Util   as S
+
+import Names
+
+-- * Free variables
+
+class FTV a where
+    ftv :: a -> S.Set Name
+    
+instance (FTV a, Ord a) => FTV (S.Set a) where
+    ftv s = S.unionMap ftv s
+

### 2014-02-28 exception-analysis/prototype/pma-v3/Fresh.hs

@@ -1,0 +1,25 @@
+module Fresh where
+
+import Control.Monad
+import Control.Monad.State
+
+import Names
+
+-- * Inference Monad
+
+type InferM r = State [Name] r
+
+-- * Fresh variables
+
+class Fresh a where
+    fresh :: InferM a
+
+instance (Fresh a, Fresh b) => Fresh (a, b) where
+    fresh = do x <- fresh
+               y <- fresh
+               return (x, y)
+
+instance Fresh Name where
+    fresh = do (x:xs) <- get
+               put xs
+               return x

### 2014-02-28 exception-analysis/prototype/pma-v3/Parser.disabled.hs

@@ -1,0 +1,15 @@
+module Parser where
+
+import qualified Data.Map      as M
+import qualified Data.Map.Util as M
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+import qualified Data.Graph    as G
+
+import System.Environment
+
+import qualified Language.Haskell.Exts as HSE
+import           Syntax
+
+desugar :: HSE.Exp -> Expr ()
+desugar = undefined

### 2014-02-28 exception-analysis/prototype/pma-v3/Semantics.hs

@@ -1,0 +1,470 @@
+{-# LANGUAGE FlexibleInstances    #-}
+{-# LANGUAGE TypeSynonymInstances #-}
+{-# LANGUAGE TupleSections        #-}
+
+module Semantics where
+
+import Control.Applicative
+import Control.Monad.State
+
+import Data.Traversable
+
+import qualified Data.List     as L
+import qualified Data.Map      as M
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+
+import Syntax
+import Ref
+import Exn
+
+-- | Normalisation
+
+-- * Check if bound variables are distinct
+
+boundVariablesDistinct :: Expr () -> Bool
+boundVariablesDistinct = maybe False (const True) . boundVariablesDistinct'
+    where boundVariablesDistinct' :: Expr () -> Maybe (S.Set Name)
+          boundVariablesDistinct' (Var x ())
+            = return (S.singleton x)
+          boundVariablesDistinct' (Con _ ())
+            = return S.empty
+          boundVariablesDistinct' (Abs x e ())
+            = do xs <- boundVariablesDistinct' e
+                 if x `S.member` xs then
+                    notDistinct
+                 else
+                    return (x `S.insert` xs)
+          boundVariablesDistinct' (Fix f e ())
+            = do xs <- boundVariablesDistinct' e
+                 if f `S.member` xs then
+                    notDistinct
+                 else
+                    return (f `S.insert` xs)
+          boundVariablesDistinct' (App e1 e2 ())
+            = do xs1 <- boundVariablesDistinct' e1
+                 xs2 <- boundVariablesDistinct' e2
+                 if xs1 `S.overlap` xs2 then
+                    notDistinct
+                 else
+                    return (xs1 `S.union` xs2)
+          boundVariablesDistinct' (Let x e1 e2 ())
+            = do xs1 <- boundVariablesDistinct' e1
+                 xs2 <- boundVariablesDistinct' e2
+                 if x `S.member` xs1 || x `S.member` xs2 || xs1 `S.overlap` xs2 then
+                    notDistinct
+                 else
+                    return (x `S.insert` (xs1 `S.union` xs2))
+          boundVariablesDistinct' (If _ e0 e1 e2 ())
[…]

### 2014-02-28 exception-analysis/prototype/pma-v3/Shapes.hs

@@ -1,0 +1,115 @@
+{-# LANGUAGE MultiParamTypeClasses #-}
+
+module Shapes where
+
+import qualified Data.Set as S
+
+import Names
+import Fresh
+import FTV
+import Subst
+import Types
+
+-- | Shapes
+
+data Sh
+    = ShBase       Name
+    | ShFun  Sh Sh Name
+    -- Pairs
+    | ShPair Sh Sh Name
+    -- Lists
+    | ShList Sh    Name
+    deriving (Eq, Ord)
+    
+instance Show Sh where
+    show (ShBase         x)
+        = show x
+    show (ShFun  sh1 sh2 x)
+        = "(" ++ show sh1 ++ " -> " ++ show sh2 ++ ")" ++ "^(" ++ show x ++ ")"
+    -- Pairs
+    show (ShPair sh1 sh2 x)
+        = "(" ++ show sh1 ++ ", "   ++ show sh2 ++ ")" ++ "^(" ++ show x ++ ")"
+    -- Lists
+    show (ShList sh      x)
+        = "[" ++ show sh ++ "]"                        ++ "^(" ++ show x ++ ")"
+    
+instance Fresh Sh where
+    fresh = do a <- fresh
+               return (ShBase a)
+
+instance Substitute Name Sh where
+    subst $@ ShBase         x
+        = ShBase                               (subst $@ x)
+    subst $@ ShFun  sh1 sh2 x
+        = ShFun  (subst $@ sh1) (subst $@ sh2) (subst $@ x)
+    -- Pairs
+    subst $@ ShPair sh1 sh2 x
+        = ShPair (subst $@ sh1) (subst $@ sh2) (subst $@ x)
+    -- Lists
+    subst $@ ShList sh      x
+        = ShList (subst $@ sh)                 (subst $@ x)
+        
+-- | Free variables
+
+instance FTV Sh where
+    ftv (ShBase         x)
+        = S.singleton x
+    ftv (ShFun  sh1 sh2 x)
+        = x `S.insert` ftv sh1 `S.union` ftv sh2
+    ftv (ShPair sh1 sh2 x)
+        = x `S.insert` ftv sh1 `S.union` ftv sh2
+    ftv (ShList sh      x)
+        = x `S.insert` ftv sh
+
+positiveVariables :: Sh -> S.Set Name
+positiveVariables (ShBase         x)
+    = S.singleton x
+positiveVariables (ShFun  sh1 sh2 x)
[…]

### 2014-02-28 exception-analysis/prototype/pma-v3/Subst.hs

@@ -1,0 +1,58 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+
+module Subst where
+
+import qualified Data.Map      as M
+import qualified Data.Map.Util as M
+import qualified Data.Set      as S -- for the instance only
+
+import Names
+
+-- | Substitutions
+
+type Subst t = M.Map Name t
+
+idSubst :: Subst t
+idSubst = M.empty
+
+class Substitute t e where
+    ($@) :: Subst t -> e -> e
+
+($.) :: (Substitute t t, Show t) => Subst t -> Subst t -> Subst t
+s2 $. s1 = (s2 $$ s1) $+ s2
+    where ($$), ($+) :: (Substitute t t, Show t) => Subst t -> Subst t -> Subst t
+          subst $$ tv
+            = M.map (subst $@) tv
+          tv1 $+ tv2
+            = M.unionWith overlapError tv1 tv2
+                where overlapError a b
+                        = error $ "Types.$+: overlapping substitutions:"
+                                  ++ " a = "    ++ show a
+                                  ++ ", b = "   ++ show b
+                                  ++ ", tv1 = " ++ show tv1
+                                  ++ ", tv2 = " ++ show tv2
+
+($\) :: Subst t -> S.Set Name -> Subst t
+($\) = M.restrictBySet
+                                  
+-- * Concrete instances
+
+instance Substitute Name Name where
+    subst $@ name
+        | Just name' <- M.lookup name subst = name'
+        | otherwise                         = name
+
+-- * Generic instances
+
+instance Substitute s () where
+    subst $@ () = ()
+                                  
+instance Substitute s t => Substitute s (Maybe t) where
+    subst $@ m = fmap (subst $@) m
+    
+instance Substitute s t => Substitute s (x, y, t) where
+    subst $@ (x, y, z) = (x, y, subst $@ z)
+    
+instance (Substitute s t, Ord t) => Substitute s (S.Set t) where
+    subst $@ x = S.map (subst $@) x

### 2014-02-28 exception-analysis/prototype/pma-v3/Syntax.hs

@@ -1,0 +1,123 @@
+{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+
+module Syntax
+    ( Name()
+    , augment
+    , freshIdents
+    , Lbl
+    , Expr(..)
+    -- , topLevelAnn
+    , Con(..)
+    , ExnC(..)
+    , Op(..)
+    )
+where
+
+import Names
+import Subst
+
+-- | Labels
+
+type Lbl = String   -- primarily used for debugging, but we will eventually
+                    -- need something like this for error reporting
+                    
+-- | (Annotated) Expressions
+
+data Expr ann
+    = Var  Name                                  ann
+    | Con  Con                                   ann
+    | Abs  Name (Expr ann)                       ann
+    | Fix  Name (Expr ann)                       ann
+    | App       (Expr ann) (Expr ann)            ann
+    | Let  Name (Expr ann) (Expr ann)            ann
+    | If   Lbl  (Expr ann) (Expr ann) (Expr ann) ann
+    -- Operators
+    | Op   Op   (Expr ann) (Expr ann)            ann
+    -- Pairs
+    | Pair      (Expr ann) (Expr ann)            ann
+    | Fst       (Expr ann)                       ann
+    | Snd       (Expr ann)                       ann
+    -- Lists
+    | Nil                                        ann
+    | Cons      (Expr ann) (Expr ann)            ann
+    | Case Lbl  (Expr ann)
+                (Maybe (            Expr ann))
+                (Maybe (Name, Name, Expr ann))   ann
+    deriving (Eq, Ord, Show)
+    -- FIXME: moving ann to the front would give better
+    --        partial application behaviour
+    
+instance Substitute s t => Substitute s (Expr t) where
+    subst $@ (Var x ann)
+        = Var x (subst $@ ann)
+    subst $@ (Con c ann)
+        = Con c (subst $@ ann)
+    subst $@ (Abs x e ann)
+        = Abs x (subst $@ e) (subst $@ ann)
+    subst $@ (Fix f e ann)
[…]

### 2014-02-28 exception-analysis/prototype/pma-v3/Types.hs

@@ -1,0 +1,249 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE ScopedTypeVariables   #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+-- FIXME: make InferM into a substitution monad
+
+module Types
+    ( Ty(..)
+    , TyCon(..)
+    , Env
+    , infer
+    )
+where
+
+import qualified Data.Map as M
+import qualified Data.Set as S
+
+import Fresh
+import FTV
+import Subst
+import Syntax
+
+-- | Types
+
+data Ty
+    = TyVar  Name
+    | TyCon  TyCon
+    | TyFun  Ty Ty
+    | TyPair Ty Ty
+    | TyList Ty
+    deriving (Eq, Ord)
+    
+data TyCon
+    = TyUnit
+    | TyBool
+    | TyInt
+    deriving (Eq, Ord, Show)
+
+instance Fresh Ty where
+    fresh = do a <- fresh
+               return (TyVar a)
+               
+instance FTV Ty where
+    ftv (TyVar a)
+        = S.singleton a
+    ftv (TyCon _)
+        = S.empty
+    ftv (TyFun t1 t2)
+        = ftv t1 `S.union` ftv t2
+    ftv (TyPair t1 t2)
+        = ftv t1 `S.union` ftv t2
+    ftv (TyList t)
+        = ftv t
+        
+instance Substitute Ty Ty where
+    subst $@ t@(TyVar a)
+        | Just t' <- M.lookup a subst = t'
+        | otherwise                   = t
+    _     $@ t@(TyCon _) 
+        = t
+    subst $@   (TyFun  t1 t2)
+        = TyFun  (subst $@ t1) (subst $@ t2)
+    subst $@   (TyPair t1 t2)
+        = TyPair (subst $@ t1) (subst $@ t2)
+    subst $@   (TyList t)
+        = TyList (subst $@ t)
+
+instance Show Ty where
+    show (TyVar a)
+        = show a
+    show (TyCon c) 
+        = show c
+    show (TyFun t1 t2)
+        = "(" ++ show t1 ++ " -> " ++ show t2 ++ ")"
+    show (TyPair t1 t2) 
+        = "(" ++ show t1 ++ ", "   ++ show t2 ++ ")"
+    show (TyList t)
+        = "[" ++ show t                       ++ "]"
+        
+-- | Environments
+
+type Env t = M.Map Name t
+
[…]

### 2014-02-28 exception-analysis/prototype/pma-v3/Util.hs

@@ -1,0 +1,4 @@
+module Util where
+
+fixpoint f x
+    = let fx = f x in if fx == x then x else fixpoint f fx

### 2014-02-28 exception-analysis/paper/popl2014/Bitstrings.hs
papers

@@ -1,0 +1,11 @@
+module Main where
+
+type Bitstring = [Int]
+
+add :: Bitstring -> Bitstring -> Bitstring
+add []      y      = y
+add x       []     = x
+add (0:x)   (0:y)  = 0 : add x y
+add (0:x)   (1:y)  = 1 : add x y
+add (1:x)   (0:y)  = 1 : add x y
+add (1:x)   (1:y)  = 0 : add (add [1] x) y

### 2014-02-28 exception-analysis/prototype/pma-v2/build.sh
missing stuff

@@ -1,0 +1,2 @@
+#!/bin/bash
+ghc -O3 --make Main && ./Main > output.tex && pdflatex output.tex && gv output.pdf

### 2015-04-21 higher-ranked-exception-types/prototype/src/Analysis/Common.hs
derivation tree passes checker

@@ -18,6 +18,8 @@ module Analysis.Common (
     exnNormalize,
     exnEq,
     exnTyEq,
+    isSubeffect,
+    isSubtype,
     simplifyExn,
     simplifyExnTy,
     Subst,
@@ -390,6 +392,22 @@ exnTyEq env (ExnArr t1 exn1 t2 exn2) (ExnArr t1' exn1' t2' exn2')
         && exnTyEq env t2 t2' && exnEq env exn2 exn2'
 exnTyEq env _ _
     = False
+    
+-- * Checking of subtyping and subeffecting
+
+isSubeffect :: KindEnv -> Exn -> Exn -> Bool
+isSubeffect env ann1 ann2 = exnEq env (ExnUnion ann1 ann2) ann2
+
+isSubtype :: KindEnv -> ExnTy -> ExnTy -> Bool
+isSubtype env ExnBool ExnBool = True
+isSubtype env ExnInt  ExnInt  = True
+isSubtype env (ExnArr ty1 ann1 ty2 ann2) (ExnArr ty1' ann1' ty2' ann2')
+    = isSubtype env ty1' ty1 && isSubeffect env ann1' ann1
+        && isSubtype env ty2 ty2' && isSubeffect env ann2 ann2'
+isSubtype env (ExnList ty ann) (ExnList ty' ann')
+    = isSubtype env ty ty' && isSubeffect env ann ann'
+isSubtype env (ExnForall e k ty) (ExnForall e' k' ty')
+    = k == k' && isSubtype ((e,k):env) ty (substExnTy e' e ty')
 
 instance Latex ExnTy where
     latex (ExnForall e k t)

### 2015-04-21 higher-ranked-exception-types/prototype/src/Analysis/Infer.hs

@@ -146,9 +146,12 @@ inferenceExamples = map (\(l,x) -> (l, show x, mathjax' x)) [
     "hcompose' compose" # App exHCompose' exCompose,
     -- * very high-order
     -- * crashing fixpoints
+    "fixId"     # Fix $ Abs 1 Bool (Var 1),
     "fixCrash1" # Fix $ Crash "foo" (Bool :-> Bool),
     "fixCrash2" # Fix $ Crash "foo" ((Bool :-> Bool) :-> (Bool :-> Bool)),
-    "fixCrash3" # Fix $ If (Con True) (Abs 1 Bool $ Crash "foo" Bool) (Crash "bar" (Bool :-> Bool))
+    "fixCrash3" # Fix $ If (Con True) (Abs 1 Bool $ Crash "foo" Bool) (Crash "bar" (Bool :-> Bool)),
+    "fixIfIdCrash" # Abs 1 Bool $ If (Var 1) (Fix $ Abs 2 Bool (Var 2))
+                                             (Fix $ Crash "foo" (Bool :-> Bool))
   ] where
         ex29 =
             Abs 1 (List Bool :-> List Bool) $ Abs 2 (List Bool) $

### 2015-04-21 higher-ranked-exception-types/prototype/src/Analysis/Infer/Common.hs

@@ -1,0 +1,38 @@
+module Analysis.Infer.Common (
+    checkKind
+) where
+
+import           Analysis.Common
+import           Analysis.Infer.Types
+
+-- * Kind checking of annotation
+
+checkKind :: KindEnv -> Exn -> Maybe Kind
+checkKind env ExnEmpty
+    = return EXN
+checkKind env (ExnUnion exn1 exn2)
+    = do k1 <- checkKind env exn1
+         k2 <- checkKind env exn2
+         if k1 == k2 then
+            return k1
+         else
+            error "kind (ExnUnion)"
+checkKind env (ExnCon _)
+    = return EXN
+checkKind env (ExnVar e)
+    = case lookup e env of
+        Just k -> return k
+        _      -> error "unbound (ExnVar)"
+checkKind env (ExnApp exn1 exn2)
+    = do (k2' :=> k1) <- checkKind env exn1
+         k2           <- checkKind env exn2
+         if k2 == k2' then
+            return k1
+         else
+            error "kind (ExnApp)"
+checkKind env (ExnAbs e k exn)
+    = do () <- case lookup e env of
+                 Nothing -> return ()
+                 _       -> error "shadowing (ExnAbs)"
+         k2 <- checkKind ((e,k):env) exn
+         return (k :=> k2)

### 2015-04-21 higher-ranked-exception-types/prototype/src/Analysis/Infer/Print.hs

@@ -19,12 +19,13 @@ import Analysis.Common
 import Analysis.Print
 
 import Analysis.Infer.Types
+import Analysis.Infer.Common
 
 -- | Types
 
 instance Latex (Name, (ExnTy, Exn)) where
     latex (show -> e, (latex -> ty, latex -> exn))
-        = "e_{" ++ e ++ "} : " ++ ty ++ "\\ \\&\\ " ++ exn
+        = "x_{" ++ e ++ "} : " ++ ty ++ "\\ \\&\\ " ++ exn
 
 instance Latex (Name, Exn) where
     latex (show -> e, latex -> exn)
@@ -51,9 +52,9 @@ instance ToMarkup DerivType where
     toMarkup dt@(TypeAnnApp jk dt1 jt)
         = derive (checkDerivType dt) "T-AnnApp"
             [H.toMarkup dt1, judgeKind jk] (judgeType jt)
-    toMarkup dt@(TypeFix dt1 jt)
+    toMarkup dt@(TypeFix jse1 jse2 dt1 jt)
         = derive (checkDerivType dt) "T-Fix"
-            [H.toMarkup dt1] (judgeType jt)
+            [H.toMarkup dt1, judgeSubEff jse1, judgeSubEff jse2] (judgeType jt)
     toMarkup dt@(TypeOp dt1 dt2 jt)
         = derive (checkDerivType dt) "T-Op"
             (map H.toMarkup [dt1, dt2]) (judgeType jt)
@@ -80,7 +81,7 @@ boolToColor :: Bool -> Color
 boolToColor True  = Green
 boolToColor False = error "Red"
 
-checkDerivType :: DerivType -> Color -- FIXME: could check env/kenv
+checkDerivType :: DerivType -> Color -- FIXME: could check env/kenv, tm
 checkDerivType (TypeVar (env, kenv, Var' x, exnTy, exn)) = boolToColor $
     let Just (exnTy', exn') = lookup x env
      in exnTyEq kenv exnTy exnTy' && exnEq kenv exn exn'
@@ -97,26 +98,78 @@ checkDerivType (TypeAnnAbs dt1 (env, kenv, AnnAbs e k tm
                                          , ExnForall e' k' ty, exn)) = boolToColor $
[…]

### 2015-04-21 higher-ranked-exception-types/prototype/src/Analysis/Infer/Reconstruct.hs

@@ -12,6 +12,7 @@ import           Analysis.Names
 import           Analysis.Common
 import qualified Analysis.Completion as C
 
+import           Analysis.Infer.Common
 import           Analysis.Infer.Types
 import           Analysis.Infer.Join
 import           Analysis.Infer.Match
@@ -55,8 +56,8 @@ reconstruct env kenv tm@(Abs x ty tm') -- TODO: common subexpression elimination
          let env' = (x, (t1', ExnVar exn1)) : env
          re@((dt,de,_), etm', t2', exn2) <- reconstruct env' (kenv1 ++ kenv) tm'
          let t' = C.forallFromEnv kenv1 (ExnArr t1' (ExnVar exn1) t2' exn2)
-         return $ (\_ _ _ -> typeAnnAbsFromEnv env' kenv (reverse kenv1) $
-                        TypeAbs dt (env', kenv1++kenv, Abs' x t1' (ExnVar exn1) etm'
+         return $ (\_ _ _ -> typeAnnAbsFromEnv env kenv (reverse kenv1) $
+                        TypeAbs dt (env, kenv1++kenv, Abs' x t1' (ExnVar exn1) etm'
                                    ,ExnArr t1' (ExnVar exn1) t2' exn2, ExnEmpty)
                   ,ElabAbs (kenv1++kenv, t1', ty) (kenv1++kenv, ExnVar exn1, EXN) de ##
                         (env,kenv,tm)
@@ -72,7 +73,15 @@ reconstruct env kenv tm@(App e1 e2)
          let subst = [(exn2', exn2)] <.> subst'
          let exn = ExnUnion (substExn' subst exn') exn1
          return $ (TypeApp (
-                        typeAnnAppFromEnv subst dt1
+                        let exnC = simplifyExn   kenv $ exn
+                            tyC  = simplifyExnTy kenv $
+                                        ExnArr t2 exn2 (substExnTy' subst t') exnC
+                            tyD  = simplifyExnTy kenv $ substExnTy' subst (fst ins)
+                        in TypeSub
+                            (kenv,tyD,tyC)
+                            (kenv,exn1,exnC)
+                            (typeAnnAppFromEnv kenv' subst dt1)
+                            (env,kenv,annAppFromEnv kenv' subst etm1,tyC,exnC)
[…]

### 2015-04-21 higher-ranked-exception-types/prototype/src/Analysis/Infer/Types.hs

@@ -23,21 +23,21 @@ type JudgeSubTy  = (KindEnv, ExnTy, ExnTy)
 type JudgeSubEff = (KindEnv, Exn, Exn)
 
 data DerivType
-    = TypeVar                                                         JudgeType
-    | TypeCon                                                         JudgeType
-    | TypeCrash                                                       JudgeType
-    | TypeAbs                           DerivType                     JudgeType
-    | TypeAnnAbs                        DerivType                     JudgeType
-    | TypeApp                           DerivType DerivType           JudgeType
-    | TypeAnnApp JudgeKind              DerivType                     JudgeType
-    | TypeFix                           DerivType                     JudgeType
-    | TypeOp                            DerivType DerivType           JudgeType
-    | TypeSeq                           DerivType DerivType           JudgeType
-    | TypeIf                            DerivType DerivType DerivType JudgeType
-    | TypeNil                                                         JudgeType
-    | TypeCons                          DerivType DerivType           JudgeType
-    | TypeCase   JudgeSubEff            DerivType DerivType DerivType JudgeType
-    | TypeSub    JudgeSubTy JudgeSubEff DerivType                     JudgeType
+    = TypeVar                                                          JudgeType
+    | TypeCon                                                          JudgeType
+    | TypeCrash                                                        JudgeType
+    | TypeAbs                            DerivType                     JudgeType
+    | TypeAnnAbs                         DerivType                     JudgeType
+    | TypeApp                            DerivType DerivType           JudgeType
+    | TypeAnnApp JudgeKind               DerivType                     JudgeType
+    | TypeFix    JudgeSubEff JudgeSubEff DerivType                     JudgeType
+    | TypeOp                             DerivType DerivType           JudgeType
+    | TypeSeq                            DerivType DerivType           JudgeType
+    | TypeIf                             DerivType DerivType DerivType JudgeType
+    | TypeNil                                                          JudgeType
+    | TypeCons                           DerivType DerivType           JudgeType
+    | TypeCase   JudgeSubEff             DerivType DerivType DerivType JudgeType
+    | TypeSub    JudgeSubTy  JudgeSubEff DerivType                     JudgeType
     
 getJT :: DerivType -> JudgeType
[…]

### 2015-04-21 higher-ranked-exception-types/prototype/src/Main.hs

@@ -188,7 +188,7 @@ inferencePage = msum [ viewForm, processForm ] where
             H.h2 "Elaborated expression"
             H.p $ mathjax elabTm
             
-            H.h3 "Type"
+            H.h3 "Exception type"
             H.p $ toHtml $
                 case elabTy of
                     Just (ty, ann) -> if An.exnTyEq [] ty exnTy && An.exnEq [] ann exn then "\\[\\color{green}" ++ latex ty ++ "\\ \\&\\ " ++ latex ann ++ "\\color{black}\\]" else error "inferred type does not match type of elaborated expression"

### 2015-04-22 presentations/domjudge-presentatie/demo/Demo.cs
domjudge demo

@@ -1,0 +1,23 @@
+﻿using System;
+using System.IO;
+ 
+namespace Demo
+{
+    class Program
+    {
+        static void Main(string[] args)
+        {
+            long n = long.Parse(Console.ReadLine());
+            
+            long t = n*(n+1)/2;
+            
+            /*
+            for (long i = u; i <= v; ++i) {
+                t += i;
+            }
+            */
+            
+            Console.WriteLine(t);           
+        }
+    }
+}

### 2015-04-22 presentations/domjudge-presentatie/demo/DemoTemplate.cs

@@ -1,0 +1,14 @@
+﻿using System;
+using System.IO;
+ 
+namespace Demo
+{
+    class Program
+    {
+        static void Main(string[] args)
+        {
+            long n = long.Parse(Console.ReadLine());
+           
+        }
+    }
+}

### 2015-09-01 program-analysis-implementations/NielsonNielsonAmtoft/NielsonNielsonAmtoft.tex
program-analysis-implementations

\documentclass{article}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[paperwidth=250cm,paperheight=250cm]{geometry}
\usepackage[cm]{fullpage}
\usepackage{stmaryrd}
\usepackage{xcolor}
\begin{document}

\begin{description}
\item[Substitution] \[\left[\begin{split}\alpha_{11}& \mapsto \dot\alpha_{222}\\\alpha_{121}& \mapsto \dot\alpha_{132}\\\alpha_{12}& \mapsto \dot\alpha_{205}\\\alpha_{145}& \mapsto \left(\dot\alpha_{181}\xrightarrow{\dot\beta_{160}}\left(\dot\alpha_{158} \times \dot\alpha_{159}\right)\right)\\\alpha_{14}& \mapsto \dot\alpha_{150}\\\alpha_{179}& \mapsto \left(\dot\alpha_{189} \times \dot\alpha_{190}\right)\\\alpha_{18}& \mapsto \dot\alpha_{131}\\\alpha_{1}& \mapsto \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\\alpha_{203}& \mapsto \dot\alpha_{213}\\\alpha_{20}& \mapsto \dot\alpha_{92}\\\alpha_{239}& \mapsto \left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right)\\\alpha_{23}& \mapsto \mathbf{Chan}\ \dot\alpha_{50}\\\alpha_{276}& \mapsto \left(\dot\alpha_{406}\xrightarrow{\dot\beta_{310}}\left(\left(\dot\alpha_{335}\xrightarrow{\dot\beta_{338}}\dot\alpha_{336}\right) \times \dot\alpha_{309}\right)\right)\\\alpha_{29}& \mapsto \mathbf{Chan}\ \dot\alpha_{31}\\\alpha_{36}& \mapsto \left(\dot\alpha_{72}\xrightarrow{\dot\beta_{41}}\left(\mathbf{Chan}\ \dot\alpha_{51} \times \dot\alpha_{40}\right)\right)\\\alpha_{3}& \mapsto \left(\dot\alpha_{818}\xrightarrow{\dot\beta_{568}}\dot\alpha_{567}\right)\\\alpha_{403}& \mapsto \left(\left(\dot\alpha_{451}\xrightarrow{\dot\beta_{454}}\dot\alpha_{452}\right) \times \dot\alpha_{453}\right)\\\alpha_{497}& \mapsto \dot\alpha_{541}\\\alpha_{4}& \mapsto \dot\alpha_{499}\\\alpha_{641}& \mapsto \left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\\\alpha_{642}& \mapsto \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\\\alpha_{681}& \mapsto \dot\alpha_{902}\\\alpha_{6}& \mapsto \left(\dot\alpha_{333}\xrightarrow{\dot\beta_{337}}\dot\alpha_{334}\right)\\\alpha_{70}& \mapsto \left(\mathbf{Chan}\ \dot\alpha_{803} \times \dot\alpha_{76}\right)\\\alpha_{719}& \mapsto \left(\dot\alpha_{740}\xrightarrow{\dot\beta_{745}}\dot\alpha_{741}\right)\\\alpha_{87}& \mapsto \dot\alpha_{103}\ \mathbf{Com}\ \dot\beta_{104}\\\dot\alpha_{123}& \mapsto \dot\alpha_{132}\\\dot\alpha_{126}& \mapsto \dot\alpha_{803}\\\dot\alpha_{147}& \mapsto \dot\alpha_{803}\\\dot\alpha_{157}& \mapsto \dot\alpha_{181}\\\dot\alpha_{184}& \mapsto \dot\alpha_{803}\\\dot\alpha_{208}& \mapsto \dot\alpha_{803}\\\dot\alpha_{236}& \mapsto \dot\alpha_{803}\\\dot\alpha_{240}& \mapsto \left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right)\\\dot\alpha_{243}& \mapsto \dot\alpha_{803}\\\dot\alpha_{248}& \mapsto \dot\alpha_{885}\\\dot\alpha_{249}& \mapsto \dot\alpha_{890}\\\dot\alpha_{251}& \mapsto \dot\alpha_{891}\\\dot\alpha_{278}& \mapsto \dot\alpha_{803}\\\dot\alpha_{281}& \mapsto \dot\alpha_{891}\\\dot\alpha_{286}& \mapsto \dot\alpha_{885}\\\dot\alpha_{291}& \mapsto \dot\alpha_{890}\\\dot\alpha_{292}& \mapsto \dot\alpha_{891}\\\dot\alpha_{307}& \mapsto \dot\alpha_{406}\\\dot\alpha_{308}& \mapsto \left(\dot\alpha_{335}\xrightarrow{\dot\beta_{338}}\dot\alpha_{336}\right)\\\dot\alpha_{38}& \mapsto \dot\alpha_{72}\\\dot\alpha_{39}& \mapsto \mathbf{Chan}\ \dot\alpha_{51}\\\dot\alpha_{405}& \mapsto \dot\alpha_{803}\\\dot\alpha_{413}& \mapsto \dot\alpha_{891}\\\dot\alpha_{422}& \mapsto \dot\alpha_{885}\\\dot\alpha_{431}& \mapsto \dot\alpha_{890}\\\dot\alpha_{432}& \mapsto \dot\alpha_{891}\\\dot\alpha_{500}& \mapsto \dot\alpha_{803}\\\dot\alpha_{507}& \mapsto \dot\alpha_{891}\\\dot\alpha_{516}& \mapsto \dot\alpha_{885}\\\dot\alpha_{525}& \mapsto \dot\alpha_{890}\\\dot\alpha_{526}& \mapsto \dot\alpha_{891}\\\dot\alpha_{566}& \mapsto \dot\alpha_{818}\\\dot\alpha_{612}& \mapsto \dot\alpha_{803}\\\dot\alpha_{615}& \mapsto \dot\alpha_{891}\\\dot\alpha_{620}& \mapsto \dot\alpha_{885}\\\dot\alpha_{625}& \mapsto \dot\alpha_{890}\\\dot\alpha_{626}& \mapsto \dot\alpha_{891}\\\dot\alpha_{677}& \mapsto \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\\\dot\alpha_{716}& \mapsto \dot\alpha_{902}\\\dot\alpha_{721}& \mapsto \left(\dot\alpha_{740}\xrightarrow{\dot\beta_{745}}\dot\alpha_{741}\right)\\\dot\alpha_{722}& \mapsto \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\\\dot\alpha_{729}& \mapsto \dot\alpha_{902}\\\dot\alpha_{738}& \mapsto \dot\alpha_{880}\\\dot\alpha_{743}& \mapsto \dot\alpha_{914}\\\dot\alpha_{75}& \mapsto \dot\alpha_{803}\\\dot\alpha_{806}& \mapsto \dot\alpha_{885}\\\dot\alpha_{809}& \mapsto \dot\alpha_{891}\\\dot\alpha_{819}& \mapsto \dot\alpha_{885}\\\dot\alpha_{820}& \mapsto \dot\alpha_{891}\\\dot\alpha_{827}& \mapsto \dot\alpha_{902}\\\dot\alpha_{834}& \mapsto \dot\alpha_{914}\\\dot\alpha_{847}& \mapsto \dot\alpha_{880}\\\dot\alpha_{848}& \mapsto \dot\alpha_{902}\\\dot\alpha_{849}& \mapsto \dot\alpha_{914}\\\dot\alpha_{877}& \mapsto \dot\alpha_{891}\\\dot\alpha_{89}& \mapsto \dot\alpha_{803}\\\dot\alpha_{93}& \mapsto \dot\alpha_{803}; \epsilon\end{split}\right]\]
\item[Type]         \[\left(\left(\dot\alpha_{818}\xrightarrow{\dot\beta_{252}}\dot\alpha_{891}\right)\xrightarrow{\beta_{876}}\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{745}}\dot\alpha_{880}\right)\right)\]
\item[Behaviour]    \[\left\{\right\}\]
\item[Constraints]  \[\left\{\dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{720}\right\}\subseteq \beta_{876}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\]
\end{description}
\begin{gather}
\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{snd}\\&  : \left(\left(\left(\dot\alpha_{818}\xrightarrow{\dot\beta_{568}}\dot\alpha_{567}\right) \times \dot\alpha_{499}\right)\xrightarrow{\beta_{5}}\dot\alpha_{499}\right)\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{(,)}\\&  : \left(\left(\dot\alpha_{333}\xrightarrow{\dot\beta_{337}}\dot\alpha_{334}\right)\xrightarrow{\beta_{8}}\left(\alpha_{7}\xrightarrow{\beta_{9}}\left(\left(\dot\alpha_{333}\xrightarrow{\dot\beta_{337}}\dot\alpha_{334}\right) \times \alpha_{7}\right)\right)\right)\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{True}\\&  : \mathbf{Bool}\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash f\\&  : \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{snd}\\&  : \left(\left(\dot\alpha_{222} \times \dot\alpha_{205}\right)\xrightarrow{\beta_{13}}\dot\alpha_{205}\right)\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{(,)}\\&  : \left(\dot\alpha_{150}\xrightarrow{\beta_{16}}\left(\alpha_{15}\xrightarrow{\beta_{17}}\left(\dot\alpha_{150} \times \alpha_{15}\right)\right)\right)\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{sync}\\&  : \left(\dot\alpha_{131}\ \mathbf{Com}\ \beta_{19}\xrightarrow{\beta_{19}}\dot\alpha_{131}\right)\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{send}\\&  : \left(\left(\mathbf{Chan}\ \dot\alpha_{92} \times \dot\alpha_{92}\right)\xrightarrow{\beta_{21}}\dot\alpha_{92}\ \mathbf{Com}\ \beta_{22}\right)\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{(,)}\\&  : \left(\mathbf{Chan}\ \dot\alpha_{50}\xrightarrow{\beta_{25}}\left(\alpha_{24}\xrightarrow{\beta_{26}}\left(\mathbf{Chan}\ \dot\alpha_{50} \times \alpha_{24}\right)\right)\right)\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{channel}\\&  : \left(\mathbf{()}\xrightarrow{\beta_{28}}\mathbf{Chan}\ \alpha_{27}\right)\\&  \& \left\{\left\{\mathbf{CHAN}\ \alpha_{27}\right\}\subseteq \beta_{28}\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{()}\\&  : \mathbf{()}\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\mathbf{channel}\ \mathbf{()}\right)\\&  : \mathbf{Chan}\ \dot\alpha_{31}\\&  \& \left\{\left(\mathbf{()}\xrightarrow{\beta_{28}}\mathbf{Chan}\ \alpha_{27}\right)\subseteq \left(\mathbf{()}\xrightarrow{\beta_{30}}\mathbf{Chan}\ \dot\alpha_{31}\right), \left\{\mathbf{CHAN}\ \alpha_{27}\right\}\subseteq \beta_{28}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\left(\mathbf{()}\xrightarrow{\beta_{28}}\mathbf{Chan}\ \alpha_{27}\right)\subseteq \left(\mathbf{()}\xrightarrow{\beta_{30}}\mathbf{Chan}\ \dot\alpha_{31}\right), \left\{\mathbf{CHAN}\ \alpha_{27}\right\}\subseteq \beta_{28}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{27}\subseteq \dot\alpha_{31}, \dot\alpha_{31}\subseteq \alpha_{27}, \left\{\mathbf{CHAN}\ \alpha_{27}\right\}\subseteq \beta_{28}, \left\{\beta_{28}\right\}\subseteq \beta_{30}\right\}\\ \leadsto_{\mathcal{R}} & \mathbf{Chan}\ \dot\alpha_{31}\& \left\{\left\{\mathbf{CHAN}\ \dot\alpha_{31}\right\}\subseteq \beta_{28}, \left\{\beta_{28}\right\}\subseteq \beta_{30}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\\&  : \left(\dot\alpha_{72}\xrightarrow{\dot\beta_{41}}\left(\mathbf{Chan}\ \dot\alpha_{51} \times \dot\alpha_{40}\right)\right)\\&  \& \left\{\left(\mathbf{Chan}\ \dot\alpha_{50}\xrightarrow{\beta_{25}}\left(\alpha_{24}\xrightarrow{\beta_{26}}\left(\mathbf{Chan}\ \dot\alpha_{50} \times \alpha_{24}\right)\right)\right)\subseteq \left(\mathbf{Chan}\ \dot\alpha_{31}\xrightarrow{\beta_{37}}\left(\dot\alpha_{72}\xrightarrow{\dot\beta_{41}}\left(\mathbf{Chan}\ \dot\alpha_{51} \times \dot\alpha_{40}\right)\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{31}\right\}\subseteq \beta_{28}, \left\{\beta_{28}\right\}\subseteq \beta_{30}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\left(\mathbf{Chan}\ \dot\alpha_{50}\xrightarrow{\beta_{25}}\left(\alpha_{24}\xrightarrow{\beta_{26}}\left(\mathbf{Chan}\ \dot\alpha_{50} \times \alpha_{24}\right)\right)\right)\subseteq \left(\mathbf{Chan}\ \dot\alpha_{31}\xrightarrow{\beta_{37}}\left(\dot\alpha_{72}\xrightarrow{\dot\beta_{41}}\left(\mathbf{Chan}\ \dot\alpha_{51} \times \dot\alpha_{40}\right)\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{31}\right\}\subseteq \beta_{28}, \left\{\beta_{28}\right\}\subseteq \beta_{30}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{24}\subseteq \dot\alpha_{40}, \dot\alpha_{31}\subseteq \dot\alpha_{50}, \dot\alpha_{50}\subseteq \dot\alpha_{31}, \dot\alpha_{50}\subseteq \dot\alpha_{51}, \dot\alpha_{51}\subseteq \dot\alpha_{50}, \dot\alpha_{72}\subseteq \alpha_{24}, \left\{\mathbf{CHAN}\ \dot\alpha_{31}\right\}\subseteq \beta_{28}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}\right\}\\ \leadsto_{\mathcal{R}} & \left(\dot\alpha_{72}\xrightarrow{\dot\beta_{41}}\left(\mathbf{Chan}\ \dot\alpha_{51} \times \dot\alpha_{72}\right)\right)\& \left\{\left\{\mathbf{CHAN}\ \dot\alpha_{51}\right\}\subseteq \beta_{28}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}\right\}\end{split}}\frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash y\\&  : \alpha_{2}\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\\&  : \left(\mathbf{Chan}\ \dot\alpha_{803} \times \dot\alpha_{76}\right)\\&  \& \left\{\left(\dot\alpha_{72}\xrightarrow{\dot\beta_{41}}\left(\mathbf{Chan}\ \dot\alpha_{51} \times \dot\alpha_{72}\right)\right)\subseteq \left(\alpha_{2}\xrightarrow{\beta_{71}}\left(\mathbf{Chan}\ \dot\alpha_{803} \times \dot\alpha_{76}\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{51}\right\}\subseteq \beta_{28}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\left(\dot\alpha_{72}\xrightarrow{\dot\beta_{41}}\left(\mathbf{Chan}\ \dot\alpha_{51} \times \dot\alpha_{72}\right)\right)\subseteq \left(\alpha_{2}\xrightarrow{\beta_{71}}\left(\mathbf{Chan}\ \dot\alpha_{803} \times \dot\alpha_{76}\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{51}\right\}\subseteq \beta_{28}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{72}, \dot\alpha_{51}\subseteq \dot\alpha_{803}, \dot\alpha_{72}\subseteq \dot\alpha_{76}, \dot\alpha_{803}\subseteq \dot\alpha_{51}, \left\{\mathbf{CHAN}\ \dot\alpha_{51}\right\}\subseteq \beta_{28}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \left(\mathbf{Chan}\ \dot\alpha_{803} \times \alpha_{2}\right)\& \left\{\left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\\&  : \dot\alpha_{103}\ \mathbf{Com}\ \dot\beta_{104}\\&  \& \left\{\left(\left(\mathbf{Chan}\ \dot\alpha_{92} \times \dot\alpha_{92}\right)\xrightarrow{\beta_{21}}\dot\alpha_{92}\ \mathbf{Com}\ \beta_{22}\right)\subseteq \left(\left(\mathbf{Chan}\ \dot\alpha_{803} \times \alpha_{2}\right)\xrightarrow{\beta_{88}}\dot\alpha_{103}\ \mathbf{Com}\ \dot\beta_{104}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\left(\left(\mathbf{Chan}\ \dot\alpha_{92} \times \dot\alpha_{92}\right)\xrightarrow{\beta_{21}}\dot\alpha_{92}\ \mathbf{Com}\ \beta_{22}\right)\subseteq \left(\left(\mathbf{Chan}\ \dot\alpha_{803} \times \alpha_{2}\right)\xrightarrow{\beta_{88}}\dot\alpha_{103}\ \mathbf{Com}\ \dot\beta_{104}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{92}, \dot\alpha_{803}\subseteq \dot\alpha_{92}, \dot\alpha_{92}\subseteq \dot\alpha_{103}, \dot\alpha_{92}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \dot\alpha_{803}\ \mathbf{Com}\ \dot\beta_{104}\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\\&  : \dot\alpha_{132}\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left(\dot\alpha_{131}\ \mathbf{Com}\ \beta_{19}\xrightarrow{\beta_{19}}\dot\alpha_{131}\right)\subseteq \left(\dot\alpha_{803}\ \mathbf{Com}\ \dot\beta_{104}\xrightarrow{\beta_{122}}\dot\alpha_{132}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left(\dot\alpha_{131}\ \mathbf{Com}\ \beta_{19}\xrightarrow{\beta_{19}}\dot\alpha_{131}\right)\subseteq \left(\dot\alpha_{803}\ \mathbf{Com}\ \dot\beta_{104}\xrightarrow{\beta_{122}}\dot\alpha_{132}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{131}\subseteq \dot\alpha_{132}, \dot\alpha_{803}\subseteq \dot\alpha_{131}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \dot\alpha_{803}\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\\&  : \left(\dot\alpha_{181}\xrightarrow{\dot\beta_{160}}\left(\dot\alpha_{158} \times \dot\alpha_{159}\right)\right)\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left(\dot\alpha_{150}\xrightarrow{\beta_{16}}\left(\alpha_{15}\xrightarrow{\beta_{17}}\left(\dot\alpha_{150} \times \alpha_{15}\right)\right)\right)\subseteq \left(\dot\alpha_{803}\xrightarrow{\beta_{146}}\left(\dot\alpha_{181}\xrightarrow{\dot\beta_{160}}\left(\dot\alpha_{158} \times \dot\alpha_{159}\right)\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left(\dot\alpha_{150}\xrightarrow{\beta_{16}}\left(\alpha_{15}\xrightarrow{\beta_{17}}\left(\dot\alpha_{150} \times \alpha_{15}\right)\right)\right)\subseteq \left(\dot\alpha_{803}\xrightarrow{\beta_{146}}\left(\dot\alpha_{181}\xrightarrow{\dot\beta_{160}}\left(\dot\alpha_{158} \times \dot\alpha_{159}\right)\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{15}\subseteq \dot\alpha_{159}, \alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{150}\subseteq \dot\alpha_{158}, \dot\alpha_{181}\subseteq \alpha_{15}, \dot\alpha_{803}\subseteq \dot\alpha_{150}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \left(\dot\alpha_{181}\xrightarrow{\dot\beta_{160}}\left(\dot\alpha_{803} \times \dot\alpha_{181}\right)\right)\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}\frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash x\\&  : \alpha_{10}\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\\&  : \left(\dot\alpha_{189} \times \dot\alpha_{190}\right)\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left(\dot\alpha_{181}\xrightarrow{\dot\beta_{160}}\left(\dot\alpha_{803} \times \dot\alpha_{181}\right)\right)\subseteq \left(\alpha_{10}\xrightarrow{\beta_{180}}\left(\dot\alpha_{189} \times \dot\alpha_{190}\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left(\dot\alpha_{181}\xrightarrow{\dot\beta_{160}}\left(\dot\alpha_{803} \times \dot\alpha_{181}\right)\right)\subseteq \left(\alpha_{10}\xrightarrow{\beta_{180}}\left(\dot\alpha_{189} \times \dot\alpha_{190}\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{10}\subseteq \dot\alpha_{181}, \alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{181}\subseteq \dot\alpha_{190}, \dot\alpha_{803}\subseteq \dot\alpha_{189}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \left(\dot\alpha_{803} \times \alpha_{10}\right)\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\x& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{10}\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\\&  : \dot\alpha_{213}\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left(\left(\dot\alpha_{222} \times \dot\alpha_{205}\right)\xrightarrow{\beta_{13}}\dot\alpha_{205}\right)\subseteq \left(\left(\dot\alpha_{803} \times \alpha_{10}\right)\xrightarrow{\beta_{204}}\dot\alpha_{213}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left(\left(\dot\alpha_{222} \times \dot\alpha_{205}\right)\xrightarrow{\beta_{13}}\dot\alpha_{205}\right)\subseteq \left(\left(\dot\alpha_{803} \times \alpha_{10}\right)\xrightarrow{\beta_{204}}\dot\alpha_{213}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{10}\subseteq \dot\alpha_{205}, \alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{205}\subseteq \dot\alpha_{213}, \dot\alpha_{803}\subseteq \dot\alpha_{222}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \alpha_{10}\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\lambda x.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\right)\\&  : \left(\alpha_{10}\xrightarrow{\beta_{235}}\alpha_{10}\right)\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}, \beta_{146}, \beta_{180}, \beta_{204}, \beta_{30}, \beta_{37}, \beta_{71}, \beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}, \beta_{146}, \beta_{180}, \beta_{204}, \beta_{30}, \beta_{37}, \beta_{71}, \beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \left(\alpha_{10}\xrightarrow{\beta_{235}}\alpha_{10}\right)\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \mathbf{if}\ \mathbf{True}\ \mathbf{then}\ f\ \mathbf{else}\ \left(\lambda x.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\right)\\&  : \left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right)\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \mathbf{Bool}\subseteq \mathbf{Bool}, \left(\alpha_{10}\xrightarrow{\beta_{235}}\alpha_{10}\right)\subseteq \left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right), \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\subseteq \left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \mathbf{Bool}\subseteq \mathbf{Bool}, \left(\alpha_{10}\xrightarrow{\beta_{235}}\alpha_{10}\right)\subseteq \left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right), \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\subseteq \left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{10}\subseteq \dot\alpha_{891}, \alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{250}\subseteq \alpha_{10}, \dot\alpha_{250}\subseteq \dot\alpha_{885}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right)\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{250}\subseteq \dot\alpha_{885}, \dot\alpha_{250}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\mathbf{(,)}\ \mathbf{if}\ \mathbf{True}\ \mathbf{then}\ f\ \mathbf{else}\ \left(\lambda x.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\right)\right)\\&  : \left(\dot\alpha_{406}\xrightarrow{\dot\beta_{310}}\left(\left(\dot\alpha_{335}\xrightarrow{\dot\beta_{338}}\dot\alpha_{336}\right) \times \dot\alpha_{309}\right)\right)\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{250}\subseteq \dot\alpha_{885}, \dot\alpha_{250}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left(\left(\dot\alpha_{333}\xrightarrow{\dot\beta_{337}}\dot\alpha_{334}\right)\xrightarrow{\beta_{8}}\left(\alpha_{7}\xrightarrow{\beta_{9}}\left(\left(\dot\alpha_{333}\xrightarrow{\dot\beta_{337}}\dot\alpha_{334}\right) \times \alpha_{7}\right)\right)\right)\subseteq \left(\left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right)\xrightarrow{\beta_{277}}\left(\dot\alpha_{406}\xrightarrow{\dot\beta_{310}}\left(\left(\dot\alpha_{335}\xrightarrow{\dot\beta_{338}}\dot\alpha_{336}\right) \times \dot\alpha_{309}\right)\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{250}\subseteq \dot\alpha_{885}, \dot\alpha_{250}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left(\left(\dot\alpha_{333}\xrightarrow{\dot\beta_{337}}\dot\alpha_{334}\right)\xrightarrow{\beta_{8}}\left(\alpha_{7}\xrightarrow{\beta_{9}}\left(\left(\dot\alpha_{333}\xrightarrow{\dot\beta_{337}}\dot\alpha_{334}\right) \times \alpha_{7}\right)\right)\right)\subseteq \left(\left(\dot\alpha_{250}\xrightarrow{\dot\beta_{253}}\dot\alpha_{891}\right)\xrightarrow{\beta_{277}}\left(\dot\alpha_{406}\xrightarrow{\dot\beta_{310}}\left(\left(\dot\alpha_{335}\xrightarrow{\dot\beta_{338}}\dot\alpha_{336}\right) \times \dot\alpha_{309}\right)\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \alpha_{7}\subseteq \dot\alpha_{309}, \dot\alpha_{250}\subseteq \dot\alpha_{885}, \dot\alpha_{250}\subseteq \dot\alpha_{891}, \dot\alpha_{333}\subseteq \dot\alpha_{250}, \dot\alpha_{334}\subseteq \dot\alpha_{336}, \dot\alpha_{335}\subseteq \dot\alpha_{333}, \dot\alpha_{406}\subseteq \alpha_{7}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \dot\alpha_{891}\subseteq \dot\alpha_{334}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \left(\dot\alpha_{406}\xrightarrow{\dot\beta_{310}}\left(\left(\dot\alpha_{335}\xrightarrow{\dot\beta_{338}}\dot\alpha_{891}\right) \times \dot\alpha_{406}\right)\right)\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{335}\subseteq \dot\alpha_{885}, \dot\alpha_{335}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}\frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash y\\&  : \alpha_{2}\\&  \& \left\{\right\}\end{split}}}{\displaystyle \Downarrow }}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\left(\mathbf{(,)}\ \mathbf{if}\ \mathbf{True}\ \mathbf{then}\ f\ \mathbf{else}\ \left(\lambda x.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\right)\right)\ y\right)\\&  : \left(\left(\dot\alpha_{451}\xrightarrow{\dot\beta_{454}}\dot\alpha_{452}\right) \times \dot\alpha_{453}\right)\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{335}\subseteq \dot\alpha_{885}, \dot\alpha_{335}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left(\dot\alpha_{406}\xrightarrow{\dot\beta_{310}}\left(\left(\dot\alpha_{335}\xrightarrow{\dot\beta_{338}}\dot\alpha_{891}\right) \times \dot\alpha_{406}\right)\right)\subseteq \left(\alpha_{2}\xrightarrow{\beta_{404}}\left(\left(\dot\alpha_{451}\xrightarrow{\dot\beta_{454}}\dot\alpha_{452}\right) \times \dot\alpha_{453}\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{335}\subseteq \dot\alpha_{885}, \dot\alpha_{335}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left(\dot\alpha_{406}\xrightarrow{\dot\beta_{310}}\left(\left(\dot\alpha_{335}\xrightarrow{\dot\beta_{338}}\dot\alpha_{891}\right) \times \dot\alpha_{406}\right)\right)\subseteq \left(\alpha_{2}\xrightarrow{\beta_{404}}\left(\left(\dot\alpha_{451}\xrightarrow{\dot\beta_{454}}\dot\alpha_{452}\right) \times \dot\alpha_{453}\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{406}, \alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{335}\subseteq \dot\alpha_{885}, \dot\alpha_{335}\subseteq \dot\alpha_{891}, \dot\alpha_{406}\subseteq \dot\alpha_{453}, \dot\alpha_{451}\subseteq \dot\alpha_{335}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \dot\alpha_{891}\subseteq \dot\alpha_{452}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{R}} & \left(\left(\dot\alpha_{451}\xrightarrow{\dot\beta_{454}}\dot\alpha_{891}\right) \times \alpha_{2}\right)\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{451}\subseteq \dot\alpha_{885}, \dot\alpha_{451}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\y& \mapsto \forall  . \left\{\right\} \Rightarrow \alpha_{2}\end{split}\right]\\&  \vdash \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \mathbf{if}\ \mathbf{True}\ \mathbf{then}\ f\ \mathbf{else}\ \left(\lambda x.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\right)\right)\ y\right)\right)\\&  : \dot\alpha_{541}\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{451}\subseteq \dot\alpha_{885}, \dot\alpha_{451}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left(\left(\left(\dot\alpha_{818}\xrightarrow{\dot\beta_{568}}\dot\alpha_{567}\right) \times \dot\alpha_{499}\right)\xrightarrow{\beta_{5}}\dot\alpha_{499}\right)\subseteq \left(\left(\left(\dot\alpha_{451}\xrightarrow{\dot\beta_{454}}\dot\alpha_{891}\right) \times \alpha_{2}\right)\xrightarrow{\beta_{498}}\dot\alpha_{541}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{451}\subseteq \dot\alpha_{885}, \dot\alpha_{451}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left(\left(\left(\dot\alpha_{818}\xrightarrow{\dot\beta_{568}}\dot\alpha_{567}\right) \times \dot\alpha_{499}\right)\xrightarrow{\beta_{5}}\dot\alpha_{499}\right)\subseteq \left(\left(\left(\dot\alpha_{451}\xrightarrow{\dot\beta_{454}}\dot\alpha_{891}\right) \times \alpha_{2}\right)\xrightarrow{\beta_{498}}\dot\alpha_{541}\right), \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{499}, \alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{451}\subseteq \dot\alpha_{885}, \dot\alpha_{451}\subseteq \dot\alpha_{891}, \dot\alpha_{499}\subseteq \dot\alpha_{541}, \dot\alpha_{818}\subseteq \dot\alpha_{451}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \dot\alpha_{891}\subseteq \dot\alpha_{567}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\}\\ \leadsto_{\mathcal{R}} & \alpha_{2}\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\end{split}\right]\\&  \vdash \left(\lambda y.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \mathbf{if}\ \mathbf{True}\ \mathbf{then}\ f\ \mathbf{else}\ \left(\lambda x.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\right)\right)\ y\right)\right)\right)\\&  : \left(\alpha_{2}\xrightarrow{\beta_{611}}\alpha_{2}\right)\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}, \beta_{404}, \beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}, \beta_{404}, \beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\}\\ \leadsto_{\mathcal{R}} & \left(\alpha_{2}\xrightarrow{\beta_{611}}\alpha_{2}\right)\& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\}\end{split}}\frac{\displaystyle \frac{\displaystyle \frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\id& \mapsto \forall \alpha_{2}\dot\alpha_{612}\beta_{122}\beta_{13}\beta_{146}\beta_{16}\beta_{17}\beta_{180}\beta_{19}\beta_{204}\beta_{21}\beta_{22}\beta_{235}\beta_{25}\beta_{26}\beta_{277}\beta_{28}\beta_{30}\beta_{37}\beta_{404}\beta_{498}\beta_{5}\beta_{611}\beta_{71}\beta_{88}\beta_{8}\beta_{9}\dot\beta_{104}\dot\beta_{160}\dot\beta_{253}\dot\beta_{310}\dot\beta_{337}\dot\beta_{338}\dot\beta_{41}\dot\beta_{454}\dot\beta_{568} . \left\{\alpha_{2}\subseteq \dot\alpha_{612}, \left\{\mathbf{CHAN}\ \dot\alpha_{612}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\} \Rightarrow \left(\alpha_{2}\xrightarrow{\beta_{611}}\alpha_{2}\right)\end{split}\right]\\&  \vdash id\\&  : \left(\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\xrightarrow{\beta_{663}}\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\right)\\&  \& \left\{\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\subseteq \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right), \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}\right\}\end{split}}}{\displaystyle \Downarrow }\frac{\displaystyle \frac{\displaystyle }{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\id& \mapsto \forall \alpha_{2}\dot\alpha_{612}\beta_{122}\beta_{13}\beta_{146}\beta_{16}\beta_{17}\beta_{180}\beta_{19}\beta_{204}\beta_{21}\beta_{22}\beta_{235}\beta_{25}\beta_{26}\beta_{277}\beta_{28}\beta_{30}\beta_{37}\beta_{404}\beta_{498}\beta_{5}\beta_{611}\beta_{71}\beta_{88}\beta_{8}\beta_{9}\dot\beta_{104}\dot\beta_{160}\dot\beta_{253}\dot\beta_{310}\dot\beta_{337}\dot\beta_{338}\dot\beta_{41}\dot\beta_{454}\dot\beta_{568} . \left\{\alpha_{2}\subseteq \dot\alpha_{612}, \left\{\mathbf{CHAN}\ \dot\alpha_{612}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\} \Rightarrow \left(\alpha_{2}\xrightarrow{\beta_{611}}\alpha_{2}\right)\end{split}\right]\\&  \vdash id\\&  : \left(\alpha_{680}\xrightarrow{\beta_{702}}\alpha_{680}\right)\\&  \& \left\{\alpha_{680}\subseteq \dot\alpha_{902}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}\right\}\end{split}}}{\displaystyle \Downarrow }}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\\id& \mapsto \forall \alpha_{2}\dot\alpha_{612}\beta_{122}\beta_{13}\beta_{146}\beta_{16}\beta_{17}\beta_{180}\beta_{19}\beta_{204}\beta_{21}\beta_{22}\beta_{235}\beta_{25}\beta_{26}\beta_{277}\beta_{28}\beta_{30}\beta_{37}\beta_{404}\beta_{498}\beta_{5}\beta_{611}\beta_{71}\beta_{88}\beta_{8}\beta_{9}\dot\beta_{104}\dot\beta_{160}\dot\beta_{253}\dot\beta_{310}\dot\beta_{337}\dot\beta_{338}\dot\beta_{41}\dot\beta_{454}\dot\beta_{568} . \left\{\alpha_{2}\subseteq \dot\alpha_{612}, \left\{\mathbf{CHAN}\ \dot\alpha_{612}\right\}\subseteq \beta_{28}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}\right\} \Rightarrow \left(\alpha_{2}\xrightarrow{\beta_{611}}\alpha_{2}\right)\end{split}\right]\\&  \vdash \left(id\ id\right)\\&  : \left(\dot\alpha_{740}\xrightarrow{\dot\beta_{745}}\dot\alpha_{741}\right)\\&  \& \left\{\alpha_{680}\subseteq \dot\alpha_{902}, \left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\subseteq \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right), \left(\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\xrightarrow{\beta_{663}}\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\right)\subseteq \left(\left(\alpha_{680}\xrightarrow{\beta_{702}}\alpha_{680}\right)\xrightarrow{\beta_{720}}\left(\dot\alpha_{740}\xrightarrow{\dot\beta_{745}}\dot\alpha_{741}\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{680}\subseteq \dot\alpha_{902}, \left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\subseteq \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right), \left(\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\xrightarrow{\beta_{663}}\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{744}}\dot\alpha_{739}\right)\right)\subseteq \left(\left(\alpha_{680}\xrightarrow{\beta_{702}}\alpha_{680}\right)\xrightarrow{\beta_{720}}\left(\dot\alpha_{740}\xrightarrow{\dot\beta_{745}}\dot\alpha_{741}\right)\right), \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{680}\subseteq \dot\alpha_{739}, \alpha_{680}\subseteq \dot\alpha_{902}, \dot\alpha_{739}\subseteq \dot\alpha_{741}, \dot\alpha_{739}\subseteq \dot\alpha_{914}, \dot\alpha_{740}\subseteq \dot\alpha_{880}, \dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{880}\subseteq \alpha_{680}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\\ \leadsto_{\mathcal{R}} & \left(\dot\alpha_{880}\xrightarrow{\dot\beta_{745}}\dot\alpha_{880}\right)\& \left\{\dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\end{split}}}{\displaystyle \begin{split}& \left[\begin{split}f& \mapsto \forall  . \left\{\right\} \Rightarrow \left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\end{split}\right]\\&  \vdash \mathbf{let}\ id\ \mathbf{=}\ \left(\lambda y.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \mathbf{if}\ \mathbf{True}\ \mathbf{then}\ f\ \mathbf{else}\ \left(\lambda x.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\right)\right)\ y\right)\right)\right)\ \mathbf{in}\ \left(id\ id\right)\\&  : \left(\dot\alpha_{880}\xrightarrow{\dot\beta_{745}}\dot\alpha_{880}\right)\\&  \& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\alpha_{2}\subseteq \dot\alpha_{803}, \dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\\ \leadsto_{\mathcal{R}} & \left(\dot\alpha_{880}\xrightarrow{\dot\beta_{745}}\dot\alpha_{880}\right)\& \left\{\dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\end{split}}}{\displaystyle \begin{split}& \epsilon \\&  \vdash \left(\lambda f.\ \mathbf{let}\ id\ \mathbf{=}\ \left(\lambda y.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \mathbf{if}\ \mathbf{True}\ \mathbf{then}\ f\ \mathbf{else}\ \left(\lambda x.\ \left(\mathbf{snd}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{sync}\ \left(\mathbf{send}\ \left(\left(\mathbf{(,)}\ \left(\mathbf{channel}\ \mathbf{()}\right)\right)\ y\right)\right)\right)\right)\ x\right)\right)\right)\right)\ y\right)\right)\right)\ \mathbf{in}\ \left(id\ id\right)\right)\\&  : \left(\left(\dot\alpha_{885}\xrightarrow{\dot\beta_{252}}\dot\alpha_{890}\right)\xrightarrow{\beta_{876}}\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{745}}\dot\alpha_{880}\right)\right)\\&  \& \left\{\dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{720}\right\}\subseteq \beta_{876}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\end{split}}}{\displaystyle \begin{split}& \left\{\dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{720}\right\}\subseteq \beta_{876}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\\ \leadsto_{\mathcal{F}} & \left\{\dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{885}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \dot\alpha_{890}\subseteq \dot\alpha_{891}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{720}\right\}\subseteq \beta_{876}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\\ \leadsto_{\mathcal{R}} & \left(\left(\dot\alpha_{818}\xrightarrow{\dot\beta_{252}}\dot\alpha_{891}\right)\xrightarrow{\beta_{876}}\left(\dot\alpha_{880}\xrightarrow{\dot\beta_{745}}\dot\alpha_{880}\right)\right)\& \left\{\dot\alpha_{742}\subseteq \dot\alpha_{880}, \dot\alpha_{818}\subseteq \dot\alpha_{891}, \dot\alpha_{880}\subseteq \dot\alpha_{902}, \dot\alpha_{880}\subseteq \dot\alpha_{914}, \left\{\mathbf{CHAN}\ \dot\alpha_{803}\right\}\subseteq \beta_{28}, \left\{\mathbf{CHAN}\ \dot\alpha_{902}\right\}\subseteq \beta_{696}, \left\{\mathbf{CHAN}\ \left(\dot\alpha_{742}\xrightarrow{\dot\beta_{746}}\dot\alpha_{914}\right)\right\}\subseteq \beta_{657}, \left\{\beta_{122}\right\}\subseteq \beta_{235}, \left\{\beta_{13}\right\}\subseteq \beta_{204}, \left\{\beta_{146}\right\}\subseteq \beta_{235}, \left\{\beta_{16}\right\}\subseteq \beta_{146}, \left\{\beta_{17}\right\}\subseteq \dot\beta_{160}, \left\{\beta_{180}\right\}\subseteq \beta_{235}, \left\{\beta_{19}\right\}\subseteq \beta_{122}, \left\{\beta_{204}\right\}\subseteq \beta_{235}, \left\{\beta_{21}\right\}\subseteq \beta_{88}, \left\{\beta_{22}\right\}\subseteq \dot\beta_{104}, \left\{\beta_{235}\right\}\subseteq \dot\beta_{253}, \left\{\beta_{25}\right\}\subseteq \beta_{37}, \left\{\beta_{26}\right\}\subseteq \dot\beta_{41}, \left\{\beta_{277}\right\}\subseteq \beta_{611}, \left\{\beta_{28}\right\}\subseteq \beta_{30}, \left\{\beta_{30}\right\}\subseteq \beta_{235}, \left\{\beta_{37}\right\}\subseteq \beta_{235}, \left\{\beta_{404}\right\}\subseteq \beta_{611}, \left\{\beta_{498}\right\}\subseteq \beta_{611}, \left\{\beta_{5}\right\}\subseteq \beta_{498}, \left\{\beta_{643}\right\}\subseteq \beta_{653}, \left\{\beta_{644}\right\}\subseteq \beta_{650}, \left\{\beta_{645}\right\}\subseteq \beta_{653}, \left\{\beta_{646}\right\}\subseteq \beta_{645}, \left\{\beta_{647}\right\}\subseteq \beta_{669}, \left\{\beta_{648}\right\}\subseteq \beta_{653}, \left\{\beta_{649}\right\}\subseteq \beta_{643}, \left\{\beta_{650}\right\}\subseteq \beta_{653}, \left\{\beta_{651}\right\}\subseteq \beta_{665}, \left\{\beta_{652}\right\}\subseteq \beta_{668}, \left\{\beta_{653}\right\}\subseteq \beta_{670}, \left\{\beta_{654}\right\}\subseteq \beta_{659}, \left\{\beta_{655}\right\}\subseteq \beta_{674}, \left\{\beta_{656}\right\}\subseteq \beta_{663}, \left\{\beta_{657}\right\}\subseteq \beta_{658}, \left\{\beta_{658}\right\}\subseteq \beta_{653}, \left\{\beta_{659}\right\}\subseteq \beta_{653}, \left\{\beta_{660}\right\}\subseteq \beta_{663}, \left\{\beta_{661}\right\}\subseteq \beta_{663}, \left\{\beta_{662}\right\}\subseteq \beta_{661}, \left\{\beta_{663}\right\}\subseteq \beta_{720}, \left\{\beta_{664}\right\}\subseteq \beta_{653}, \left\{\beta_{665}\right\}\subseteq \beta_{653}, \left\{\beta_{666}\right\}\subseteq \beta_{656}, \left\{\beta_{667}\right\}\subseteq \beta_{671}, \left\{\beta_{668}\right\}\subseteq \beta_{649}, \left\{\beta_{669}\right\}\subseteq \beta_{648}, \left\{\beta_{670}\right\}\subseteq \beta_{672}, \left\{\beta_{671}\right\}\subseteq \beta_{660}, \left\{\beta_{672}\right\}\subseteq \beta_{673}, \left\{\beta_{673}\right\}\subseteq \beta_{675}, \left\{\beta_{674}\right\}\subseteq \beta_{664}, \left\{\beta_{675}\right\}\subseteq \beta_{676}, \left\{\beta_{682}\right\}\subseteq \beta_{692}, \left\{\beta_{683}\right\}\subseteq \beta_{689}, \left\{\beta_{684}\right\}\subseteq \beta_{692}, \left\{\beta_{685}\right\}\subseteq \beta_{684}, \left\{\beta_{686}\right\}\subseteq \beta_{708}, \left\{\beta_{687}\right\}\subseteq \beta_{692}, \left\{\beta_{688}\right\}\subseteq \beta_{682}, \left\{\beta_{689}\right\}\subseteq \beta_{692}, \left\{\beta_{690}\right\}\subseteq \beta_{704}, \left\{\beta_{691}\right\}\subseteq \beta_{707}, \left\{\beta_{692}\right\}\subseteq \beta_{709}, \left\{\beta_{693}\right\}\subseteq \beta_{698}, \left\{\beta_{694}\right\}\subseteq \beta_{713}, \left\{\beta_{695}\right\}\subseteq \beta_{702}, \left\{\beta_{696}\right\}\subseteq \beta_{697}, \left\{\beta_{697}\right\}\subseteq \beta_{692}, \left\{\beta_{698}\right\}\subseteq \beta_{692}, \left\{\beta_{699}\right\}\subseteq \beta_{702}, \left\{\beta_{700}\right\}\subseteq \beta_{702}, \left\{\beta_{701}\right\}\subseteq \beta_{700}, \left\{\beta_{702}\right\}\subseteq \dot\beta_{744}, \left\{\beta_{703}\right\}\subseteq \beta_{692}, \left\{\beta_{704}\right\}\subseteq \beta_{692}, \left\{\beta_{705}\right\}\subseteq \beta_{695}, \left\{\beta_{706}\right\}\subseteq \beta_{710}, \left\{\beta_{707}\right\}\subseteq \beta_{688}, \left\{\beta_{708}\right\}\subseteq \beta_{687}, \left\{\beta_{709}\right\}\subseteq \beta_{711}, \left\{\beta_{710}\right\}\subseteq \beta_{699}, \left\{\beta_{711}\right\}\subseteq \beta_{712}, \left\{\beta_{712}\right\}\subseteq \beta_{714}, \left\{\beta_{713}\right\}\subseteq \beta_{703}, \left\{\beta_{714}\right\}\subseteq \beta_{715}, \left\{\beta_{71}\right\}\subseteq \beta_{235}, \left\{\beta_{720}\right\}\subseteq \beta_{876}, \left\{\beta_{88}\right\}\subseteq \beta_{235}, \left\{\beta_{8}\right\}\subseteq \beta_{277}, \left\{\beta_{9}\right\}\subseteq \dot\beta_{310}, \left\{\dot\beta_{104}\right\}\subseteq \beta_{19}, \left\{\dot\beta_{160}\right\}\subseteq \beta_{180}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{670}, \left\{\dot\beta_{252}\right\}\subseteq \beta_{709}, \left\{\dot\beta_{252}\right\}\subseteq \dot\beta_{253}, \left\{\dot\beta_{253}\right\}\subseteq \dot\beta_{337}, \left\{\dot\beta_{310}\right\}\subseteq \beta_{404}, \left\{\dot\beta_{337}\right\}\subseteq \dot\beta_{338}, \left\{\dot\beta_{338}\right\}\subseteq \dot\beta_{454}, \left\{\dot\beta_{41}\right\}\subseteq \beta_{71}, \left\{\dot\beta_{454}\right\}\subseteq \dot\beta_{568}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{745}, \left\{\dot\beta_{744}\right\}\subseteq \dot\beta_{746}\right\}\end{split}}
\end{gather}
\end{document}

### 2015-09-01 program-analysis-implementations/Data/Equiv.hs

@@ -1,0 +1,92 @@
+module Data.Equiv
+    ( Equiv(), empty, elems, insert, insert2, union, equivalent
+    , equivalenceClass, representatives                         )
+where
+
+import qualified Data.List             as L
+import qualified Data.Map              as M
+import           Data.Maybe
+import qualified Data.Set              as S
+
+import qualified Data.UnionFind.IntMap as UF               -- == union-find-0.2
+
+-- | Exported data types and functions
+
+data Equiv a = Equiv { m :: M.Map a (UF.Point a), ps :: UF.PointSupply a }
+
+-- * The empty equivalence relation
+empty :: Equiv a
+empty = Equiv { m = M.empty, ps = UF.newPointSupply }
+
+-- * The elements of the relation
+elems :: Ord a => Equiv a -> [(a,a)]
+elems e@Equiv { m = m, ps = ps }
+    = [(a,b) | a <- M.keys m, b <- equivalenceClass' e a ]
+
+-- * Insert new element into equivalence relation; must be fresh (verified)
+insert :: Ord a => a -> Equiv a -> Equiv a
+insert x Equiv { m = m, ps = ps }
+    = let (ps', p) = UF.fresh ps x
+       in Equiv { m = M.insertWith (error "duplicate element") x p m, ps = ps' }
+                                     
+-- * Insert pair into equivalence relation; may be fresh or pre-existing
+insert2 :: Ord a => a -> a -> Equiv a -> Equiv a
+insert2 x y e0@Equiv { m = m0, ps = ps0 }
+    = let e1@Equiv { m = m1, ps = ps1 }
+            = case M.lookup x m0 of
+                    Nothing -> insert x e0
+                    Just _  ->          e0
+          e2@Equiv { m = m2, ps = ps2 }
+            = case M.lookup y m1 of 
+                    Nothing -> insert y e1
+                    Just _  ->          e1
+       in union e2 x y
+
+-- * Unions two equivalence classes in the relation
+union :: Ord a => Equiv a -> a -> a -> Equiv a
+union Equiv { m = m, ps = ps } x y
+    = Equiv { m = m, ps = UF.union ps (m ?? x) (m ?? y) }
+
+-- * Test if two elements are equivalent
+equivalent :: Ord a => Equiv a -> a -> a -> Bool
[…]

### 2015-09-01 program-analysis-implementations/Data/List/Util.hs

@@ -1,0 +1,8 @@
+module Data.List.Util (pairwiseT) where
+
+pairwiseT :: (a -> a -> b) -> [a] -> [b]
+pairwiseT _ []        = []
+pairwiseT f [x]       = [f x x]
+pairwiseT f xxs@(x:_) = pairwiseT' xxs
+    where pairwiseT' [x']           = [f x' x]
+          pairwiseT' (x1:xs@(x2:_)) = f x1 x2 : pairwiseT' xs

### 2015-09-01 program-analysis-implementations/Data/Map/Util.hs

@@ -1,0 +1,30 @@
+module Data.Map.Util ((!!), restrict, mapKeysAndValuesWith, restrictBySet, unionMap) where
+
+import           Prelude       hiding ((!!))
+
+import qualified Data.Map as M
+import qualified Data.Set as S
+
+-- * Lookup with additional error reporting
+(!!) :: (Ord k, Show k, Show a) => M.Map k a -> k -> a
+m !! k | Just v <- M.lookup k m = v
+       | otherwise              = error $ "Data.Map.Util.!!: key '" ++ show k ++ " not found in " ++ show m
+
+restrict :: Ord k => M.Map k a -> [k] -> M.Map k a
+restrict = foldr M.delete
+
+mapKeysAndValues :: (Ord k1, Ord k2) =>
+    (k1 -> k2) -> (a1 -> a2) -> M.Map k1 a1 -> M.Map k2 a2
+mapKeysAndValues f g = M.mapKeys f . M.map g
+
+mapKeysAndValuesWith :: (Ord k1, Ord k2) =>
+    (a2 -> a2 -> a2) -> (k1 -> k2) -> (a1 -> a2) -> M.Map k1 a1 -> M.Map k2 a2
+mapKeysAndValuesWith c f g = M.mapKeysWith c f . M.map g
+
+-- | Map + Set
+
+restrictBySet :: Ord k => M.Map k a -> S.Set k -> M.Map k a
+restrictBySet = S.foldr M.delete
+
+unionMap :: (Ord k, Ord b) => (a -> S.Set b) -> M.Map k a -> S.Set b
+unionMap f = S.unions . M.elems . M.map f

### 2015-09-01 program-analysis-implementations/Data/Set/Util.hs

@@ -1,0 +1,13 @@
+module Data.Set.Util (overlap, overlaps, unionMap) where
+
+import qualified Data.List.Util as L
+import qualified Data.Set       as S
+
+overlap :: Ord a => S.Set a -> S.Set a -> Bool
+overlap s t = not (S.null (S.intersection s t))
+
+overlaps :: Ord a => [S.Set a] -> Bool
+overlaps = or . L.pairwiseT overlap
+
+unionMap :: (Ord a, Ord b) => (a -> S.Set b) -> S.Set a -> S.Set b
+unionMap f = S.unions . map f . S.toList

### 2015-09-01 program-analysis-implementations/DussartHengleinMossin1995.hs

@@ -1,0 +1,632 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE TupleSections         #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+module DussartHengleinMossin1995 where
+
+-- TODO: more type classes (unify, isMonomorphic)
+-- TODO: more parameterized types (Constr)
+-- TODO: use light-weight generic programming (isMonomorphic, subst)
+-- TODO: substitution monad
+-- TODO: make unification and inference a Maybe
+
+import Control.Applicative
+import Control.Monad
+import Control.Monad.Identity
+import Control.Monad.State
+
+import qualified Data.Graph    as G
+import qualified Data.List     as L
+import qualified Data.Map      as M
+import qualified Data.Map.Util as M
+import           Data.Maybe
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+import qualified Data.Tree     as T
+
+import qualified Debug.Trace as Debug
+
+-- | Examples
+
+test example
+    = let (s1, t1, e1') = evalInferM (wTy M.empty example)
+       in evalInferM (wBt M.empty e1')
+
+-- * Henglein & Mossin: Ackermann
+
+exAck = Fix a (Lam i (Lam j (If (COp (Var i) EQL (Con $ Int 0))
+                                (AOp (Var j) ADD (Con $ Int 1))
+                                (If (COp (Var j) EQL (Con $ Int 0))
+                                    (App (App (Var a)
+                                              (AOp (Var i) SUB (Con $ Int 1))
+                                         )
+                                         (Con $ Int 1)
+                                    )
+                                    (App (App (Var a)
+                                              (AOp (Var i) SUB (Con $ Int 1))
+                                         )
+                                         (App (App (Var a) (Var i))
[…]

### 2015-09-01 program-analysis-implementations/Mycroft1984.hs

@@ -1,0 +1,191 @@
+{-# LANGUAGE FlexibleInstances    #-}
+{-# LANGUAGE TupleSections        #-}
+{-# LANGUAGE TypeSynonymInstances #-}
+
+module Mycroft1984 where
+
+import Control.Applicative
+import Control.Monad
+import Control.Monad.Identity
+import Control.Monad.State
+
+import qualified Data.Graph    as G
+import qualified Data.List     as L
+import qualified Data.Map      as M
+import qualified Data.Map.Util as M
+import           Data.Maybe
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+import qualified Data.Tree     as T
+
+-- | Syntax
+
+type Name = String
+
+data Expr
+    = Var Name
+    | App Expr Expr
+    | Lam Name Expr
+    | Fix Name Expr
+    | Let Name Expr Expr
+    deriving (Eq, Ord, Show)
+    
+-- | Static Semantics
+
+-- * Types
+
+data Ty
+    = TyVar Name
+    | TyCon TyCon
+    | TyFun Ty Ty
+    deriving (Eq, Ord, Show)
+
+data TyCon
+    = TyBool
+    | TyInt
+    deriving (Eq, Ord, Show)
+
+data TyScheme = Forall [Name] Ty deriving (Eq, Ord, Show)
+    -- FIXME: Eq should be modulo alpha-renaming
+
+type TyEnv = M.Map Name TyScheme
+
+-- * Substitutions
+
+infixr 0 $@
+infixr 9 $.
+
+type Subst = M.Map Name Ty
+
+idSubst :: Subst
+idSubst = M.empty
+
+($.) :: Subst -> Subst -> Subst
+subst2 $. subst1 = (subst2 $$ subst1) $+ subst2
+    where ($$), ($+) :: Subst -> Subst -> Subst
+          subst1 $$ subst2
+            = M.map (subst1 $@) subst2
+          subst1 $+ subst2
+            = M.unionWith (error "overlapping substitutions") subst1 subst2
+
+class Substitute t where
+    ($@) :: Subst -> t -> t
+
+instance Substitute Ty where
+    subst $@ t@(TyVar a)     = M.findWithDefault t a subst
+    _     $@ t@(TyCon _)     = t
+    subst $@   (TyFun t1 t2) = TyFun (subst $@ t1) (subst $@ t2)
+
+instance Substitute TyScheme where
[…]

### 2015-09-01 program-analysis-implementations/NielsonNielsonAmtoft.hs

@@ -1,0 +1,1203 @@
+{-# LANGUAGE TypeSynonymInstances #-}
+{-# LANGUAGE FlexibleInstances    #-}
+
+module Main (main) where
+
+-- FIXME: Seems to generate more annotation constraints than the paper claims.
+-- FIXME: Additional reduction rules not yet implemented.
+
+-- PERFORMANCE: commutativity and associativity of S.union (bigset `S.union` smallset)
+--              M.findWithDefault
+--              TyScheme, quantified variables as list or set?
+
+import           Prelude                hiding (cycle)
+
+import           Control.Monad
+import           Control.Monad.Identity
+import           Control.Monad.State
+
+import qualified Data.Graph             as G
+import qualified Data.List              as L
+import qualified Data.Map               as M
+import           Data.Maybe
+import qualified Data.Set               as S
+import qualified Data.Tree              as T
+
+import qualified Data.Equiv             as E               -- => union-find-0.2
+
+-- | Syntax
+
+type Name  = String
+type Ident = Name
+
+data Expr
+    = Con Con
+    | Var Ident
+    | Fn  Ident Expr
+    | App Expr  Expr
+    | Let Ident Expr  Expr
+    | Rec Ident Ident Expr
+    | If  Expr  Expr  Expr
+    deriving (Eq, Ord, Show)
+    
+_seq :: Expr -> Expr -> Expr
+_seq e1 e2 = Con Snd `App` (Con Pair `App` e1 `App` e2)
+
+_pair :: Expr -> Expr -> Expr
+_pair e1 e2 = Con Pair `App` e1 `App` e2
+    
+data Con
+    = Unit
+    | Bool Bool
+    | Int  Int
+    | Plus | Times   | Eq
+    | Pair | Fst     | Snd
+    | Nil  | Cons    | Hd   | Tl      | IsNil
+    | Send | Receive | Sync | Channel | Fork
+    deriving (Eq, Ord, Show)
+
+-- | Static semantics
+
+data Ty
+    = TyVar  Ident
+    | TyUnit
+    | TyInt
+    | TyBool
+    | TyPair Ty Ty
+    | TyList Ty
+    | TyFun  Ty Be Ty
+    | TyChan Ty
+    | TyCom  Ty Be
+    deriving (Eq, Ord, Show)
+
+data Be
+    = BeUnif Name
+    | Be     (S.Set Be')
[…]

### 2015-09-01 program-analysis-implementations/SetConstraints/AikenWimmer1992.hs

@@ -1,0 +1,62 @@
+module AikenWimmers1992 where
+
+import qualified Data.Set as S
+
+-- | Syntax
+
+type Name = String
+
+data SetExpr
+    = O
+    | I
+    | Var Name
+    | Not SetExpr
+    | SetExpr :|: SetExpr
+    | SetExpr :&: SetExpr
+    | Con Con [SetExpr]
+    deriving (Eq, Ord, Show)
+
+data SetConstr
+    = SetExpr :<: SetExpr
+    deriving (Eq, Ord, Show)
+    
+data SetEquality
+    = SetExpr :=: SetExpr
+    deriving (Eq, Ord, Show)
+
+type System           = S.Set SetConstr
+type SolvedFormSystem = S.Set SetEquality
+
+-- | Algorithms
+
+oneLevelSystem :: System -> FreshM System
+oneLevelSystem = S.mapM (\(a :<: b) -> oneLevel (a :&: Not b) :<: O)
+
+oneLevel :: SetExpr -> FreshM SetExpr
+oneLevel = undefined
+
+-- | Freshness Monad
+
+type FreshM r = State [Name] r
+
+class Fresh t where
+    fresh :: InferM t
+
+instance Fresh Name where
+    fresh = do (x:xs) <- get
+               put xs
+               return x
+
+-- | Examples
+
+-- * Constructors (should be a parameter)
+
+data Con = A | B | C deriving (Eq, Ord, Show)
+    
+a :: Con -> Int
+a A = 0
+a B = 1
+a C = 2
+
+
+

### 2015-09-01 program-analysis-implementations/TalpinJouvelot.hs

@@ -1,0 +1,478 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE Rank2Types            #-}
+{-# LANGUAGE ScopedTypeVariables   #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+module TalpinJouvelot where
+
+import Control.Monad
+import Control.Monad.State
+
+import qualified Data.List as L
+import qualified Data.Map  as M
+import qualified Data.Set  as S
+
+-- | Syntax
+
+type Ident    = String
+
+data Expr
+    = Var Ident
+    | App Expr  Expr
+    | Abs Ident Expr
+    | Let Ident Expr Expr
+    | New Expr
+    | Get Expr
+    | Set Expr  Expr
+    deriving (Eq, Show)
+
+-- | Dynamic Semantics
+
+data Value
+    = Unit
+    | Ref     Ident
+    | Closure Ident Expr Env
+    deriving (Eq, Show)
+
+type Env   = M.Map Ident Value
+type Store = M.Map Ident Value
+
+eval :: Store -> Env -> Expr -> State [Ident] (Value, Store)
+eval s env (Var x)      | Just v <- M.lookup x env = return (v, s)
+                        | otherwise = error "unbound variable"
+eval s env (Abs x e)    = return (Closure x e env, s)
+eval s env (Let x e e') = do (v, s') <- eval s env e
+                             (v', s'') <- eval s' (M.insert x v env) e'
+                             return (v', s'')
+eval s env (App e e')   = do (Closure x e'' env', s') <- eval s env e
+                             (v', s'') <- eval s' env e'
+                             (v'', s''') <- eval s'' (M.insert x v' env') e''
+                             return (v'', s''')
+eval s env (New e)      = do l <- fresh
+                             (v, s') <- eval s env e
+                             return (Ref l, M.insert l v s')
+eval s env (Get e)      = do (Ref l, s') <- eval s env e
+                             let Just v = M.lookup l s'
+                             return (v, s')
+eval s env (Set e e')   = do (Ref l, s') <- eval s env e
+                             (v, s'') <- eval s' env e'
[…]

### 2015-09-01 data-util/Graph/Util.hs
data-util

@@ -1,0 +1,7 @@
+{-# LANGUAGE StandaloneDeriving #-}
+
+module Data.Graph.Util where
+
+import Data.Graph
+
+deriving instance Show a => Show (SCC a)

### 2015-09-01 data-util/List/Util.hs

@@ -1,0 +1,13 @@
+module Data.List.Util (pairwiseT, worklist) where
+
+pairwiseT :: (a -> a -> b) -> [a] -> [b]
+pairwiseT _ []        = []
+pairwiseT f [x]       = [f x x]
+pairwiseT f xxs@(x:_) = pairwiseT' xxs
+    where pairwiseT' [x']           = [f x' x]
+          pairwiseT' (x1:xs@(x2:_)) = f x1 x2 : pairwiseT' xs
+
+-- worklist :: Foldable m => (a -> b -> (b, m a)) -> b -> m a -> b
+worklist :: (a -> b -> (b, [a])) -> b -> [a] -> b
+worklist (#) i []     = i
+worklist (#) i (m:ms) = let (j, ns) = m # i in worklist (#) j (ms ++ ns)

### 2015-09-01 data-util/Map/Util.hs

@@ -1,0 +1,34 @@
+module Data.Map.Util ((!!), restrict, mapKeysAndValuesWith, restrictBySet, unionMap) where
+
+import           Prelude       hiding ((!!))
+
+import qualified Data.Map as M
+import qualified Data.Set as S
+
+-- * Lookup with additional error reporting
+(!!) :: (Ord k, Show k, Show a) => M.Map k a -> k -> a
+m !! k | Just v <- M.lookup k m = v
+       | otherwise              = error $ "Data.Map.Util.!!: key '" ++ show k ++ " not found in " ++ show m
+
+-- FIXME: should be renames: restrict should 'keep'
+--        only the given elements, not delete them
+restrict :: Ord k => M.Map k a -> [k] -> M.Map k a
+restrict = foldr M.delete
+
+mapKeysAndValues :: (Ord k1, Ord k2) =>
+    (k1 -> k2) -> (a1 -> a2) -> M.Map k1 a1 -> M.Map k2 a2
+mapKeysAndValues f g = M.mapKeys f . M.map g
+
+mapKeysAndValuesWith :: (Ord k1, Ord k2) =>
+    (a2 -> a2 -> a2) -> (k1 -> k2) -> (a1 -> a2) -> M.Map k1 a1 -> M.Map k2 a2
+mapKeysAndValuesWith c f g = M.mapKeysWith c f . M.map g
+
+-- | Map + Set
+
+-- FIXME: should be renames: restrict should 'keep'
+--        only the given elements, not delete them
+restrictBySet :: Ord k => M.Map k a -> S.Set k -> M.Map k a
+restrictBySet = S.foldr M.delete
+
+unionMap :: (Ord k, Ord b) => (a -> S.Set b) -> M.Map k a -> S.Set b
+unionMap f = S.unions . M.elems . M.map f

### 2015-09-01 data-util/Set/Util.hs

@@ -1,0 +1,21 @@
+module Data.Set.Util (all, overlap, overlaps, unionMap, showSet) where
+
+import Prelude hiding (all)
+
+import qualified Data.List.Util as L
+import qualified Data.Set       as S
+
+all :: Ord a => (a -> Bool) -> S.Set a -> Bool
+all p = S.foldr (\x r -> p x && r) True
+
+overlap :: Ord a => S.Set a -> S.Set a -> Bool
+overlap s t = not (S.null (S.intersection s t))
+
+overlaps :: Ord a => [S.Set a] -> Bool
+overlaps = or . L.pairwiseT overlap
+
+unionMap :: (Ord a, Ord b) => (a -> S.Set b) -> S.Set a -> S.Set b
+unionMap f = S.unions . map f . S.toList
+
+showSet :: Show a => S.Set a -> String
+showSet s = "{" ++ drop 10 (init (show s)) ++ "}"

### 2015-09-01 deprecated/pma-v2/Constr.hs
pma-v2

@@ -1,0 +1,112 @@
+module Constr where
+
+-- FIXME: can be generalized from a set to a lattice solver (join/meet)
+-- FIXME: a lot of variables are still called ref*
+
+import qualified Data.Graph      as G
+import qualified Data.Graph.Util as G
+import qualified Data.Map        as M
+import qualified Data.Map.Util   as M
+import qualified Data.Set        as S
+import qualified Data.Set.Util   as S
+
+import Common
+
+-- | Subset constraints
+
+type Constr a = S.Set (Constr' a)
+
+data Constr' a
+    = Elem a :<: Elem a     -- FIXME: LHS a :<: Name?
+    deriving (Eq, Ord, Show)
+    
+data Elem a  -- FIXME: needs a better name
+    = Unif Name
+    | Conc (S.Set a)        -- FIXME: needs a better name
+    deriving (Eq, Ord, Show)
+    
+class (Ord a, Show a) => ConstrElem a where
+    demote :: a -> Elem a
+    injVar :: Name -> a
+    prjVar :: a -> Name
+    
+injUnif :: ConstrElem a => Elem a -> Elem a       -- FIXME: call injVar directly?
+injUnif (Unif u) = Conc (S.singleton (injVar u))
+
+prjUnif :: ConstrElem a => a -> Elem a
+prjUnif x = Unif (prjVar x)
+
+-- | Simplification
+
+-- * Decomposition into singleton (atomic?) constraints
+
+decompose :: ConstrElem a => Constr a -> Constr a
+decompose c
+    = S.foldr (\c' r -> r `S.union` decompose' c') S.empty c
+
+decompose' :: ConstrElem a => Constr' a -> Constr a
+decompose' (Conc refs :<: b@(Unif _))
+    = S.unionMap (\ref' -> S.singleton (Conc (S.singleton ref') :<: b)) refs
+decompose' (Unif b1 :<: b2@(Unif _)) -- FIXME: we shouldn't (but do) generate these
+    = S.singleton (Conc (S.singleton (injVar b1)) :<: b2)
+   
+-- * Remove reflexive constraints
+
+removeReflexive :: ConstrElem a => Constr a -> Constr a
+removeReflexive
+    = S.foldr (\c'@(l :<: r) cr -> if l == injUnif r then cr else S.insert c' cr)
+              S.empty
+
+-- * Constract cyclic constraints (strongly connected components)
+
+-- FIXME: extremely dubious code
+
[…]

### 2015-09-01 deprecated/pma-v2/Data/Map/Util.hs

@@ -1,0 +1,27 @@
+module Data.Map.Util ((!!), restrict, mapKeysAndValuesWith, unionMap) where
+
+import           Prelude       hiding ((!!))
+
+import qualified Data.Map as M
+import qualified Data.Set as S
+
+-- * Lookup with additional error reporting
+(!!) :: (Ord k, Show k, Show a) => M.Map k a -> k -> a
+m !! k | Just v <- M.lookup k m = v
+       | otherwise              = error $ "Data.Map.Util.!!: key '" ++ show k ++ " not found in " ++ show m
+
+restrict :: Ord k => M.Map k a -> [k] -> M.Map k a
+restrict = foldr M.delete
+
+mapKeysAndValues :: (Ord k1, Ord k2) =>
+    (k1 -> k2) -> (a1 -> a2) -> M.Map k1 a1 -> M.Map k2 a2
+mapKeysAndValues f g = M.mapKeys f . M.map g
+
+mapKeysAndValuesWith :: (Ord k1, Ord k2) =>
+    (a2 -> a2 -> a2) -> (k1 -> k2) -> (a1 -> a2) -> M.Map k1 a1 -> M.Map k2 a2
+mapKeysAndValuesWith c f g = M.mapKeysWith c f . M.map g
+
+-- | Map + Set
+
+unionMap :: (Ord k, Ord b) => (a -> S.Set b) -> M.Map k a -> S.Set b
+unionMap f = S.unions . M.elems . M.map f

### 2015-09-01 deprecated/pma-v2/Data/Tuple/Instances.hs

@@ -1,0 +1,21 @@
+{-# LANGUAGE TupleSections #-}
+
+module Data.Tuple.Instances where
+
+import Control.Applicative
+
+import Data.Foldable
+import Data.Traversable
+
+import Data.Monoid
+
+-- | 3-tuple
+
+instance Functor ((,,) a b) where
+    fmap f (x,y,z) = (x, y, f z)
+    
+instance Foldable ((,,) a b) where
+    foldMap f (x, y, z) = f z `mappend` mempty
+
+instance Traversable ((,,) a b) where
+    traverse f (x, y, z) = (x, y,) <$> f z

### 2015-09-01 deprecated/pma-v2/Frontend.hs

@@ -1,0 +1,361 @@
+module Frontend where
+
+import qualified Data.Map      as M
+import qualified Data.Map.Util as M
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+import qualified Data.Graph    as G
+
+import System.Environment
+
+import Language.Haskell.Exts
+import qualified Common as PMA
+
+main = do
+    args <- getArgs
+    let fileName = if null args then "Sample.hs" else head args
+    main' fileName
+
+main' fileName = do
+    contents <- readFile fileName
+    let parseResult = parseModuleWithMode defaultParseMode contents
+    case parseResult of
+        ParseOk hModule            
+            -> do putStrLn $ header "Source"
+                  putStrLn $ prettyPrint hModule
+                  putStrLn $ header "AST (haskell-src-exts)"
+                  putStrLn $ show hModule
+                  putStrLn $ header "Pre-Processing"
+                  let ppr = preProcess hModule
+                  putStrLn $ prettyPrint ppr
+                  putStrLn $ header "AST (PMA)"
+                  let ast = convert ppr
+                  putStrLn $ show ast
+                  putStrLn $ header "Reverted AST"
+                  putStrLn $ prettyPrint (revert' ast)
+        ParseFailed srcLoc message
+            -> do print srcLoc
+                  print message
+
+-- | Conversion
+
+convert :: Exp -> PMA.Expr
+convert (Var (UnQual (Ident name)))
+    = PMA.Var (PMA.name name )
+convert (Con (UnQual (Ident "True")))
+    = PMA.Con (PMA.Bool True )
+convert (Con (UnQual (Ident "False")))
+    = PMA.Con (PMA.Bool False)
+convert (Con (Special UnitCon))
+    = PMA.Con PMA.Unit
+convert (Lit (Int n))
+    = PMA.Con (PMA.Int (fromInteger n))
+convert (InfixApp e1 (QConOp (Special Cons)) e2)
+    = PMA.Cons (convert e1) (convert e2)
+convert (InfixApp e1 (QVarOp (UnQual (Symbol "<="))) e2)
+    = PMA.Op PMA.LEQ (convert e1) (convert e2)
+convert (InfixApp e1 (QVarOp (UnQual (Symbol "+"))) e2)
+    = PMA.Op PMA.ADD (convert e1) (convert e2)
+convert (App e1 e2)
+    = PMA.App (convert e1) (convert e2)
+convert (Lambda _ pats e)
[…]

### 2015-09-01 deprecated/pma-v2/Main.hs

@@ -1,0 +1,625 @@
+module Main (main) where
+
+import System.IO
+
+import Control.Monad.State
+
+import qualified Data.Map      as M
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+import qualified Data.Tree     as T
+
+import           Common
+import qualified Sem
+import qualified Analysis as A
+import qualified Constr   as C
+
+-- | Main
+
+main :: IO ()
+main
+    = do -- Preamble
+         putStrLn $ preamble
+
+         mapM_ testcase [ exV1,  exV2,  exV3,  exV4,  exV5,  exV6
+                        , ex1 ,  ex2 ,  ex3 ,  ex4 ,  ex5 ,  ex6 ,  ex7 ,  ex8
+                        , ex9 ,  ex10,  ex11,  ex12,  ex13,  ex14,  ex15,  ex16
+                        , ex17  -- interesting example
+                        , exT1,  exT2,  exT3,  exT4,  exT5
+                        , exL1,  exL2,  exL3,  exL4,  exL5,  exL6,  exL7
+                        , exC1,  exC2,  exC3,  exC4,  exC5,  exC6,  exC7,  exC8
+                        , exC9,  exC10, exC11, exC12, exC13, exC14, exC15
+                        , exF1,  exF2,  exF3,  exF4,  exF5,  exF6,  exF7,  exF8
+                        , exQ1, exQ2
+                        , exCS0, exCS1
+                        , exR0, exR0a, exR0b
+                        , exR1  -- interesting example
+                        , exR1Nil
+                        , exLength, exLengthApp, exLengthBroken, exLengthBrokenApp
+                        , exRminimal, exRminimal0, exRminimal1, exRminimal2
+                        ]
+
+         -- Postamble
+         putStrLn $ postamble
+
+testcase :: Expr -> IO ()
+testcase example
+    = do putStrLn $ newpage
+    
+         -- * Expression
+         putStrLn $ section "Expression"
+         putStrLn $ gather example
+
+         -- * Evaluation
+         putStrLn $ section "Evaluation"
+
+         let e' = evalState (Sem.transform example)
+                            (map (augment "\\zeta") freshIdents)
+
+         -- putStrLn $ subsection "Normal form"
+         -- putStrLn $ gather e'
+
+         putStrLn $ subsection "Reduction (WHNF)"
[…]

### 2015-09-01 deprecated/pma-v2/Sem.hs

@@ -1,0 +1,407 @@
+{-# LANGUAGE FlexibleInstances    #-}
+{-# LANGUAGE TypeSynonymInstances #-}
+{-# LANGUAGE TupleSections        #-}
+
+module Sem where
+
+import Control.Applicative
+import Control.Monad.State
+
+import Data.Traversable
+
+import qualified Data.List     as L
+import qualified Data.Map      as M
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+
+import Common
+
+-- | Normalisation
+
+-- * Check if bound variables are distinct
+
+boundVariablesDistinct :: Expr -> Bool      -- FIXME: actually call this from somewhere
+boundVariablesDistinct = maybe False (const True) . boundVariablesDistinct'
+    where boundVariablesDistinct' :: Expr -> Maybe (S.Set Name)
+          boundVariablesDistinct' (Var x)
+            = return (S.singleton x)
+          boundVariablesDistinct' (Con _)
+            = return S.empty
+          boundVariablesDistinct' (Abs x e)
+            = do xs <- boundVariablesDistinct' e
+                 if x `S.member` xs then
+                    notDistinct
+                 else
+                    return (x `S.insert` xs)
+          boundVariablesDistinct' (Fix f e)
+            = do xs <- boundVariablesDistinct' e
+                 if f `S.member` xs then
+                    notDistinct
+                 else
+                    return (f `S.insert` xs)
+          boundVariablesDistinct' (App e1 e2)
+            = do xs1 <- boundVariablesDistinct' e1
+                 xs2 <- boundVariablesDistinct' e2
+                 if xs1 `S.overlap` xs2 then
+                    notDistinct
+                 else
+                    return (xs1 `S.union` xs2)
+          boundVariablesDistinct' (Let x e1 e2)
+            = do xs1 <- boundVariablesDistinct' e1
+                 xs2 <- boundVariablesDistinct' e2
+                 if x `S.member` xs1 || x `S.member` xs2 || xs1 `S.overlap` xs2 then
+                    notDistinct
+                 else
+                    return (x `S.insert` (xs1 `S.union` xs2))
+          boundVariablesDistinct' (If _ e0 e1 e2)
+            = do xs0 <- boundVariablesDistinct' e0
[…]

### 2015-09-01 deprecated/pma-v3/Main.hs
pma-v3

@@ -1,0 +1,157 @@
+{-# LANGUAGE MultiWayIf #-}
+
+module Main where
+
+import System.IO
+import System.Process
+
+import Control.Monad
+import Control.Monad.State
+
+import qualified Data.Graph      as G
+import qualified Data.Graph.Util as G
+import qualified Data.List       as L
+import qualified Data.Map        as M
+import qualified Data.Map.Util   as M
+import           Data.Maybe
+import qualified Data.Set        as S
+import qualified Data.Set.Util   as S
+import qualified Data.Tree       as T
+import           Data.Tuple.Util
+
+import Subst
+import Syntax
+import Semantics
+import Types     hiding (main)
+import Shapes
+import Constr
+import Analysis
+import Ref                          -- unused
+import Exn
+import Examples
+
+-- | Main
+
+-- * Type inference
+
+testTypes :: IO ()
+testTypes = do
+    let ((s, t, e'), _) = runState (infer M.empty exR1) freshIdents
+    putStrLn $ "t = " ++ show t
+    putStrLn $ "e = " ++ show e'
+
+-- * Analysis
+
+main :: IO ()
+main = do
+    mapM_ (\(desc,e) -> testcase desc e) allExamples
+
+testcase :: String -> Expr () -> IO ()
+testcase desc e = do
+
+    putStrLn $ "== " ++ desc ++ " " ++ replicate (76 - length desc) '='
+
+    -- SEMANTICS
+    
+    let (eT, freshIdents2) = runState (transform e) freshIdents
+    -- putStrLn $ show e
+    -- putStrLn $ show eT
+    -- putStrLn $ "boundVariablesDistinct e = " ++ show (boundVariablesDistinct e)
+    -- putStrLn $ "boundVariablesDistinct eT = " ++ show (boundVariablesDistinct eT)
+    let (_heap1, z1) = evalState (reduce        10000 M.empty eT) freshIdents2
+    let (_heap2, z2) = evalState (reduceFurther 10000 M.empty eT) freshIdents2
+    let ev1@(evRef1, evExn1) = postProcessEval1 z1
+    let ev2@(evRef2, evExn2) = postProcessEval2 z2
+    putStrLn $ "ev1 = " ++ show ev1
+    putStrLn $ "ev2 = " ++ show ev2
+
+    -- TYPE INFERENCE
[…]

### 2015-09-01 deprecated/pma-v3/Names.hs

@@ -1,0 +1,16 @@
+module Names where
+
+-- | Names
+
+newtype Name = Name String deriving (Eq, Ord)
+
+instance Show Name where
+    show (Name x) = x
+
+augment :: String -> Name -> Name
+augment prefix (Name name') = Name (prefix ++ name')
+
+-- | Fresh Names
+
+freshIdents :: [Name]
+freshIdents = map (\n -> Name $ "?" ++ show n) [(1::Int)..]

### 2015-09-01 deprecated/pma-v3/Ref.hs

@@ -1,0 +1,63 @@
+{-# LANGUAGE MultiParamTypeClasses #-}
+
+module Ref where
+
+import qualified Data.Set as S
+
+import Subst
+import FTV
+import Syntax
+import Constr
+
+-- * Refinements
+
+data Ref
+    = RefVar  Name
+    | RefTop
+    | RefBool Bool
+    | RefInt  Int'
+    -- Lists
+    | RefList List'
+    deriving (Eq, Ord)
+    
+instance Show Ref where
+    show (RefVar  x) = show x
+    show (RefTop   ) = "T"
+    show (RefBool b) = show b
+    show (RefInt  n) = show n
+    show (RefList l) = show l
+    
+instance FTV Ref where
+    ftv (RefVar name) = S.singleton name
+    ftv _             = S.empty
+
+instance ConstrElem Ref where
+    injName = RefVar
+
+    prjName (RefVar n) = Just n
+    prjName _          = Nothing
+    
+instance Substitute Name Ref where
+    subst $@ RefVar a = RefVar (subst $@ a)
+    _     $@ ref      = ref        
+    
+data Int'
+    = Neg
+    | Zero
+    | Pos
+    deriving (Eq, Ord)
+    
+instance Show Int' where
+    show Neg  = "-"
+    show Zero = "0"
+    show Pos  = "+"
+
+data List'
+    = E
+    | NE
+    deriving (Eq, Ord, Show)
+
+injInt :: Int -> Int'
+injInt n | n < 0     = Neg
+         | n > 0     = Pos
+         | otherwise = Zero

### 2015-09-01 deprecated/exception-analysis/src/Dynamics.hs
exception-analysis

@@ -1,0 +1,118 @@
+{-# LANGUAGE ViewPatterns #-}
+
+module Dynamics where
+
+import Data.Set as S
+
+-- | Utility
+
+iterateUntil :: (a -> Bool) -> (a -> a) -> a -> [a]
+iterateUntil p f x = takeWhile (not . p) (iterate f x)
+
+-- | Names
+
+type Name = String
+
+-- | Exception labels
+
+data ExnLbl = DivByZero | PMF deriving (Eq, Ord, Show)
+
+-- | Environments
+
+data Env range = E
+               | Env range :> (Name, range)
+               deriving Show
+
+type ExprEnv = Env Expr
+
+(#) :: Name -> ExprEnv -> (ExprEnv, Expr)
+y # (env :> (x, e))
+    | x == y    = (env, e)
+    | otherwise = let (env', e) = y # env in (env :> (x, e), e)
+    
+-- | Operators
+
+data OP = ADD deriving (Eq, Ord, Show)
+
+-- | Expressions
+
+data Expr = Var Name
+          | Bool Bool
+          | Int Int
+          | Exn (S.Set ExnLbl)
+          | Nil
+          | Close Expr ExprEnv
+          | Abs Name Expr
+          | Fix Name Expr
+          | Expr :@: Expr
+          | Let Name Expr Expr
+          | If Expr Expr Expr
+          | Op Expr OP Expr
+          | Expr :*: Expr
+          | Fst Expr
+          | Snd Expr
+          | Expr ::: Expr
+          | Case Expr Expr (Name, Name, Expr)
+          | Bind ExprEnv Expr
+          deriving Show
+
+-- | Values and normal forms
+
+isValue :: Expr -> Bool
+isValue (Bool  _  ) = True
+isValue (Int   _  ) = True
+isValue (Exn   _  ) = True
+isValue  Nil        = True
+isValue (Close _ _) = True
+isValue _           = False
+
+-- | Reduction relation
+
+reduce :: ExprEnv -> Expr -> Expr
+reduce env e = reduce' e env
+    where
+        reduce' :: Expr -> ExprEnv -> Expr
+        reduce' (Var x) ((x#) -> (env, e))
+            = Bind env e
+        reduce' (Fix f e) env
+            = Bind (env :> (f, Bind env $ Fix f e)) e
+        reduce' e@(e1 ::: e2) env
+            = Close e env
[…]

### 2015-09-01 deprecated/pattern-match-analysis/thesis-proposal/thesis-proposal.tex
pattern-match-analysis

\documentclass[a4paper]{report}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{alltt}

\usepackage{tikz}

\begin{document}

\title{Thesis Proposal}
\author{Ruud Koot}
\maketitle

% Also see:
% http://www.cs.uu.nl/wiki/bin/view/Thesis/WritingAThesisProposal
% mail from Jur
% Tamar's/Eric's proposal

\chapter*{Introduction}

This document is a proposal for a thesis on answering the research question:
\begin{quote}
Can we use a type and effect system in combination with refinement types to develop a pattern-match analysis for a non-strict higher-order functional language that is both performant and precise enough to be of practical use?
\end{quote}

In Chapter \ref{secmotivation} we present a number of examples to demonstrate why this is an interesting problem. In Chapter \ref{seccontext} we give a short introduction to the relevant concepts of the research question: higher-order functional languages, type and effect systems and refinement types. In Chapter \ref{secrelatedwork} we present work related to our research question and discuss which aspects of that work are relevant to or different from our proposed system. In Chapter \ref{secapproach}, \ref{secdeliverables} and \ref{secevaluation} we will discuss our envisioned approach to answering the research question, the set of deliverables and how we will evaluate whether the research question has been answered successfully or not.

\chapter{Motivation}\label{secmotivation}

In 1978, Robin Milner \cite{Milner78atheory} famously wrote (and proved) that:
\begin{quotation}
Well-typed programs cannot ``go wrong''. 
\end{quotation}
This observation is sometimes
%\footnote{Especially by professors wishing to impress their audience of undergraduate students with the power of types.}
optimistically overstated as ``if your Haskell program type checks and compiles it will work.'' Even if we ignore logic errors, such a strong interpretation of Milner's statement does not hold, as anyone who has seen the dreaded:
\begin{verbatim}
*** Exception: PM.hs:1:1-13: Non-exhaustive patterns in function f
\end{verbatim}
can attest to. To correctly interpret Milner we need to distinguish between three kinds of ``wrongness'':
\begin{description}
\item[Getting stuck] If a program tries to evaluate a non-sensical expression -- such as {\tt 3 + true} -- it cannot possibly make any further progress and is said to be ``stuck''. This is the ``go wrong'' Milner referred to. He proved that a sound type system can statically guarantee such expressions will never occur or need to be evaluated at run-time.
\item[Diverging] If we have a function definition {\tt f = f}, the evaluation of {\tt f} $\leadsto$ {\tt f} $\leadsto ...$ will fail to terminate. We might be making progress in a technical sense, but it will have no useful observable result.
\item[Undefinedness] Another source of ``wrongness'' are partial functions: functions which are not defined on all the elements of their domain. Prime examples are case-statements with missing constructors and functions defined by pattern matching but which do not cover all possible patterns. 
\end{description}

Unlike Milner, who spoke about the first kind of wrongness and the work on termination checking, which concerns itself with the second, we shall focus on the third: a pattern-match analysis which determines if functions are only invoked on values for which they are defined.

% split error and partial definedness?
% loop and crash

\section{Examples}\label{secexamples} % test-driven development

\paragraph{Partial functions} Haskell programmers often work with partial functions, the most common one possibly being {\tt head}:

\begin{verbatim}
1  main = let xs = if length "foo" > 5 then [1,2,3] else []
2          in head xs
3
4  head (x : xs) = x
\end{verbatim}

This program is guaranteed to crash at run-time, so we would like to be warned beforehand by the compiler or a tool:

\begin{verbatim}
On line 2 you applied the function "head" to the empty list "xs".
The function "head" expects a non-empty list as its first argument.
\end{verbatim}

If the guard of the if-statement had read {\tt length "foo" < 5} the program would have run without crashing and we would like the compiler or tool not to warn us spuriously. In case it is not possible to determine statically whether or not a program will crash, a warning should still be raised.

\paragraph{Compiler construction}
Compilers work with large and complex data types to represent the abstract syntax tree. These data structures must be able to represent all syntactic constructs the parser is able to recognize. This results in an abstract syntax tree that is unnecessarily complex and too cumbersome for the later stages of the compiler -- such as the optimizer -- to work with. This problem is resolved by \emph{desugaring} the original abstract syntax tree into a simpler -- but semantically equivalent --  abstract syntax tree than does not use all of the constructors available in the original abstract syntax tree.

The compiler writer now has a choice between two different options: either write a desugaring stage {\tt desugar :: ComplexAST -> SimpleAST} -- duplicating most of the data type representing and functions operating on the abstract syntax tree -- or take the easy route {\tt desugar :: AST -> AST} and assume certain constructors will no longer be present in the abstract syntax tree at stages of the compiler executed after the desugaring step. The former has all the usual downsides of code duplication -- such as having to manually keep the two or more data types synchronized -- while that latter forgoes many of the advantages of strong typing and type safety: if the compiler pipeline is restructured and one of the stages that was originally assumed to run only after the desugaring suddenly runs before that point the error might only be detected at run-time by a pattern match failure. A pattern match analysis should be able to detect such errors statically.

\paragraph{Maintaining invariants}
Many algorithms and data structures maintain invariants that cannot be easily encoded into their type. These invariants often ensure that certain incomplete case-statements are guaranteed not to cause a pattern match failure. An example is the {\tt risers} function from \cite{DBLP:conf/sfp/MitchellR05}, calculating monotonically increasing segments of a list (e.g., {\tt risers [1,3,5,1,2]} $\leadsto$ {\tt [[1,3,5],[1,2]]}):

\begin{verbatim}
1  risers :: Ord a => [a] -> [[a]]
2  risers []         = []
3  risers [x]        = [[x]]
4  risers (x1:x2:xs) = let (s:ss) = risers (x2:xs)
5                       in if x <= y then (x:s):ss else [x]:(s:ss)
\end{verbatim}

The let-binding in the third alternative of {\tt risers} expects the recursive call to return a non-empty list. A naive analysis might raise a warning here. If we think a bit longer, however, we see that we also pass the recursive call to {\tt risers} a non-empty list. This means we will end up in either the second or third alternative in the recursive call. Both the second alternative and both branches of the if-statement in the third alternative result in a non-empty list, satisfying the assumption we made earlier.

Another example might be a collection of mathematical operations working on bitstrings (integers encoded as lists of binary digits without leading zeroes):
\begin{verbatim}
1  type Bitstring = [Int]
2
3  add :: Bitstring -> Bitstring -> Bitstring
4  add [] y = y
5  add x [] = x
6  add (0:x) (0:y) = 0 : add x y
7  add (0:x) (1:y) = 1 : add x y
8  add (1:x) (0:y) = 1 : add x y
9  add (1:x) (1:y) = 0 : add (add [1] x) y
\end{verbatim}

The patterns in {\tt add} are far from complete, but maintain the invariant if passed arguments that satisfy the invariant. So if we are careful to only pass valid bitstrings into a complex mathematical expression of bitstring-operations it will result in a valid bitstring without crashing due to a pattern match failure.

% Examples from Refinement Types for ML:
% * Binary encoded numbers
% * Variable substitutions in Boolean expressions

\chapter{Context}\label{seccontext}

% halting problem
% well-typed programs can't go wrong
% smaller types log form presentation

\section{Haskell}

Haskell is statically-typed functional programming language with non-strict evaluation semantics \cite{Haskell98}. As a functional language it has first-class and higher-order functions and features a rich type system supporting parametric polymorphism and type classes. Programmers can define custom types in the form of algebraic data types and write functions over them using pattern matching.

Like most functional languages, Haskell can be easily translated into a typed $\lambda$-calculus (System $F_C$). From the point-of-view of the programmer it offers a wealth of syntactic sugar over a plain $\lambda$-calculus -- such as guards and list comprehensions -- allowing programs to be expressed concisely and in a readable fashion.

The following example demonstrates the the syntax of Haskell and some features mentioned previously:

\begin{verbatim}
1  data Tree a = Branch (Tree a) (Tree a)
2              | Leaf a
3  
4  mapTree :: (a -> b) -> Tree a -> Tree b
5  mapTree f (Branch t1 t2) = Branch (mapTree f t1) (mapTree f t2)
6  mapTree f (Leaf a)       = Leaf (f a)
7
8  instance Functor Tree where
9      fmap = mapTree
\end{verbatim}
Lines 1--2 define an algebraic data type (ADT) representing a tree. A tree can be constructed by either a {\tt Branch} containing a left and a right subtree or by a {\tt Leaf} containing a value. The ADT is parameterized by a type {\tt a} giving the type of the value stored in the {\tt Leaf}.

Lines 4--5 define a higher-order function {\tt mapTree} which applies a given function {\tt f :: a -> b} to all the values stored in all the leaves of a given {\tt Tree} resulting in a new {\tt Tree}. As the domain and the range of {\tt f} need not coincided the type parameter of the tree changes as well. The body of the function is defined by pattern matching over the constructors of {\tt Tree} and performs recursion in the case of a {\tt Branch} and applies the argument {\tt f} to the field of type {\tt a} in {\tt Leaf} otherwise.

Lines 8--9 declare the type constructor {\tt Tree} to be an instance of the type class {\tt Functor}, expressing it to be an endofunctor in the category of data types with its mapping on morphisms given by {\tt mapTree}.

\subsection{Pattern matching}

Particularly relevant to our analysis are Haskell's pattern matching abilities. We demonstrate some of them in the following function:

\begin{verbatim}
rotate (Branch b@(Branch (Leaf x) (Leaf y)) (Leaf z))
           | z < x     = Branch (Leaf z) b
           | z < y     = Branch (Branch (Leaf x) (Leaf z)) (Leaf y)
           | otherwise = undefined
\end{verbatim}

The function will rotate a tree {\tt (Branch (Branch (Leaf 2) (Leaf 3)) (Leaf 1))} into {\tt Branch (Leaf 1) (Branch (Leaf 2) (Leaf 3))}. It accomplishes this through the use of \emph{nested pattern matching}, \emph{as-patterns} and \emph{guards}.

This function only works on trees of a very specific shape. Feeding the function a tree of a different shape will cause a pattern match failure. Similarly, an error will be raised if the function reaches the {\tt otherwise} branch, containing an {\tt undefined} value. This would also happen if the branch had called {\tt error} or the branch and its guard were left out entirely.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Type and effect systems}

Type and effect systems are approaches to program analysis suitable for typed languages \cite{Nielson:1999:PPA:555142}. In this formalism we take the the \emph{underlying type system} of the language and (conservatively) extend it by adding \emph{annotations} to the types (base, function and other) of the system.

\begin{figure}[p]
\[\frac{}{\Gamma \vdash_{\mathrm{UL}} c : \tau_c}\ \mathrm{[con]} \quad\quad\quad\quad
\frac{\Gamma(x) = \tau}{\Gamma \vdash_{\mathrm{UL}} x : \tau}\ \mathrm{[var]}\]
\[\frac{\Gamma[x \mapsto \tau_1] \vdash_{\mathrm{UL}} e : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \lambda x. e : \tau_1 \to \tau_2}\ \mathrm{[abs]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_2 \to \tau_1 \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} e_1\ e_2 : \tau_1}\ \mathrm{[app]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_1 \quad \Gamma[x\mapsto \tau_1] \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \mathrm{let\ } x = e_1 \mathrm{\ in\ } e_2 : \tau_2}\ \mathrm{[let]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau}{\Gamma \vdash \mathrm{if\ } e_0 \mathrm{\ then\ } e_1 \mathrm{\ else \ } e_2 : \tau}\ \mathrm{[if]}\]
\caption{Underlying type system}
\label{underlying}
\end{figure}

\begin{figure}[p]
\[\frac{}{\Gamma \vdash_{\mathrm{AN}} n : \mathrm{int}^{\{n\}}}\ \mathrm{[con]} \quad\quad\quad\quad
\frac{\Gamma(x) = \widehat\tau}{\Gamma \vdash_{\mathrm{AN}} x : \widehat\tau}\ \mathrm{[var]}\]
\[\frac{\Gamma[x \mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \lambda x. e : \widehat\tau_1 \to \widehat\tau_2}\ \mathrm{[abs]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_2 \to \widehat\tau_1 \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} e_1\ e_2 : \widehat\tau_1}\ \mathrm{[app]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_1 \quad \Gamma[x\mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \mathrm{let\ } x = e_1 \mathrm{\ in\ } e_2 : \widehat\tau_2}\ \mathrm{[let]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_2} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_3}}{\Gamma \vdash_{\mathrm{AN}} \mathrm{if\ } e_0 \mathrm{\ then\ } e_1 \mathrm{\ else \ } e_2 : \tau^{\varphi_2 \cup \varphi_3}}\ \mathrm{[if]}\]
\[\frac{
\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi}
}{
\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi \cup \varphi'}
}\ \mathrm{[sub]}\]
\caption{Annotated type system}
\label{annotated}
\end{figure}

\subsection{Example}

As an example we give an analysis which determines, in the form of an annotation, which values an expression can evaluate to. The language is the simply-typed $\lambda$-calculus with let-bindings and an if-statement. The underlying type systems is standard and given in Figure \ref{underlying}.

The annotated type system is given in Figure \label{annotated}. It does little more than collecting all constants in the annotations. The only interesting rule is for the if-statement. We need to take the union over the values collected in both branches of the if-statement. We could have chosen to make the rule depend on the annotation of the guard expression, passing on only the values collected in the then-branch along if we knew it could only be {\tt True} or only the values collected in the else-branch if we knew it could only be {\tt False} from its annotation. This would make the analysis more precise.

We have also added an inference rule for \emph{subeffecting}. This allows us to make the annotation less precise. This is sometimes necessary when applying a value about which we have very accurate information to a function which requires a more general argument.

\subsection{Type inference}

An analysis specified as an inference system does not allow us to compute the result of the analysis directly, as the premises of several of the inference rules usually require us to guess the correct type annotations for the type and annotation variables (e.g., in the presence of subeffecting). To overcome this problem we need a \emph{type reconstruction algorithm}.

If the underlying type system is Hindley--Milner, then Algorithm W can be used to infer the types. Algorithm W works bottom-up and can compute types in a single pass by unifying type variables. This may not always be possible for the annotations in an annotated type system, however. Here we can use a variant of Algorithm W to gather a set of constraints on the annotation variables and solve these constraints with a worklist algorithm in a second phase.

\subsection{Polyvariance}

% 1) precision (poisoning)
% 2) modularity

% annotation polymorphism

Consider the following program:
\begin{verbatim}
f = let id x = x
     in (id 1, id 2)
\end{verbatim}

When performing the analysis given above we would assign the type $\mathtt{int} \to \mathtt{int}^{\{1, 2\}}$ to {\tt id} and as a result the type $\mathtt{int}^{\{1,2\}} \times \mathtt{int}^{\{1,2\}}$ to {\tt f}. Clearly, this result is not optimal. This loss in precision has been caused by the two separate calls to {\tt id} \emph{poisoning} each other.

A similar situation occurs at the type-level in the function:
\begin{verbatim}
f = let id x = x
     in (id 1, id true)
\end{verbatim}
Hindley--Milner manages to avoid problems in this situation by a mechanism called \emph{let-generalization}. The binding {\tt id} has the type $\alpha \to \alpha$. Instead of trying -- and failing -- to unify $\alpha$ with both {\tt int} and {\tt bool} in the body of the let-binding, the type of {\tt id} is generalized to the polymorphic type $\forall \alpha. \alpha \to \alpha$. In the body of the let-binding, {\tt id} can be instantiated twice: once to $\mathtt{int} \to \mathtt{int}$ and once to $\mathtt{bool} \to \mathtt{bool}$.

A similar trick can be used for the annotated type system, except that instead of quantifying over type variables, we quantify over annotation variables.

\section{Refinement types}

\begin{figure}[t]
\centering
\begin{tikzpicture}
    \tikzstyle{all nodes}=[inner sep=4pt]
    \draw node(top)at(1,3){\tt List a}
          node(sorn)at(1,2){{\tt Singleton a} $\lor$ {\tt Nil a}}
          node(s)at(0,1){\tt Singleton a} node(n)at(2,1){\tt Nil a}
          node(bot)at(1,0){$\bot$};
    \draw[-](top)--(sorn);
    \draw[-](sorn)--(s);
    \draw[-](sorn)--(n);
    \draw[-](s)--(bot);
    \draw[-](n)--(bot);
\end{tikzpicture}
\caption{A lattice of {\tt List} and its {\tt rectype}s.}
\label{listlattice}
\end{figure}

Refinement types are a form of subtyping that allow us to state more accurately what values a variable of a particular type can contain. For example, a refinement type {\tt \{ a :: Int | a < 0 \}} states that {\tt a} is a variable of type {\tt Int} than can only contain negative numbers.

Freeman and Pfenning \cite{Freeman:1991:RTM:113445.113468,Freeman94refinementtypes} give a formal development of refinement types for the ML programming language that can support recursive types.

Their key contribution is allowing the programmer to define {\tt rectype}s. Besides defining a recursive type for lists:

\begin{verbatim}
data List a = Nil | Cons a (List a)
\end{verbatim}

The programmer can also define a {\tt rectype} describing singleton lists:

\begin{verbatim}
rectype Singleton a = Cons a Nil
\end{verbatim}

Together with an automatically derived {\tt rectype} for the non-recurisive constructor {\tt Nil} (and a union type contructor) the compiler can construct -- using known algorithms on regular tree grammars \cite{DBLP:books/others/tree1984} -- a finite lattice of refinement types for {\tt List a} given in Figure \ref{listlattice}.

\subsection{Union and Intersection Types}

Refinement types are constructed from regular types, including function types, as well as union and intersection types.

A union type such as {\tt Int $\lor$ Bool}, {\tt List a $\lor$ Singleton a} or {\tt Int -> Int $\lor$ Bool -> Bool} state a value can be either of the type on the left or of the type on the right, but we do not know which. In the case of {\tt List a $\lor$ Singleton a} we are able to simplify this type to {\tt List a}, as {\tt Singleton a} is a subtype of {\tt List a}.

An intersection type such as {\tt Int $\land$ Bool}, {\tt List a $\land$ Singleton a} or {\tt Int -> Int $\land$ Bool -> Bool} state the a value has both the type on the left and the type on the right. Compared to union types these are more interesting. The type {\tt List a $\land$ Singleton a} can still be simplified, but now to {\tt Singleton a}. A value cannot be of both type {\tt Int} and {\tt Bool}, so we can simply the type of such a value to the bottom or empty type. There do exist functions that are of both type {\tt Int -> Int} and {\tt Bool -> Bool}, for example {\tt id}. In fact, intersection types can in the limit be viewed as a form of parametric polymorphism.

\subsection{Constructors and Pattern Matching}

To infer refinement types more accurate types need to be given to constructors. This is done by using a restricted\footnote{We only take intersections of subtypes of the same data type.} form of intersection types. For example, the {\tt Cons} constructor is given the type:

\begin{verbatim}
Cons :: a -> Nil       a -> Singleton a
     /\ a -> Singleton a -> List      a
     /\ a -> List      a -> List      a
\end{verbatim}

This type can be derived automatically from the {\tt rectype} declarations. Types of functions can also be inferred automatically, although there may be a loss of precision when higher-order functions or polymorphism are involved.

Type inference for intersection types is undecidable in general \cite{Pierce91programmingwith,Reynolds96designof}. Because the lattice of types is finite the algorithm is effectively able to do an exhaustive search over all possible types, however. Higher-order functions can cause the size of the type to blow up exponentially, each pairing of their range and domain needs to be included in the intersection type.

A case-statement

\begin{verbatim}
case G of
  Nil       -> E1
  Cons a as -> E2
\end{verbatim}

can be seen as a call to a higher-order function

\begin{verbatim}
case_List G E1 (\a as -> E2)
\end{verbatim}

where {\tt case\_list} has the refinement type

\begin{verbatim}
case_List :: forall a. forallR (r1 :: a). forallR (r2 :: a).
       Nil a       -> r1 -> (List a -> r2) -> r1
    /\ Singleton a -> r1 -> (List a -> r2) -> r2
    /\ List a      -> r1 -> (List a -> r2) -> (r1 \/ r2)
\end{verbatim}

\chapter{Related Work}\label{secrelatedwork}

% table comparing the approaches
% * laziness
% * polymorphism
% * annotations (required / supported)
% * decidability
% * errors generated

\section{Neil Mitchell's \emph{Catch}}

Neil Mitchel's \emph{Catch} (``CAse Totality CHecker'') \cite{Mitchell} is a tool specialized in finding potential pattern match failures in Haskell programs.

\subsection{Overview}

Catch works by calculates preconditions on functions. The preconditions are given in a constraint language. By varying the constraint language trade-offs between performance and precision can be made. If the calculated precondition for \texttt{main} is True, the program does not crash.

Preconditions are calculated iteratively. Precondition start out as \texttt{True}, except for \texttt{error}, whose precondition is \texttt{False}.\footnote{Note that {\tt undefined} is defined in terms of {\tt error}.}

\paragraph{Example} Following the example given by Mitchell, given the function:
\begin{verbatim}
safeTail xs = case null xs of
                True  -> []
                False -> tail xs
\end{verbatim}
The computed precondition will be:
\[ \textrm{Precondition}(\texttt{null xs}) \land (\texttt{null xs} \in \{\texttt{True}\} \lor \textrm{Precondition}(\texttt{tail xs})), \]
stating the necessary precondition for \texttt{null xs} not to crash mush be fulfilled and either \texttt{null xs} must evaluate to \texttt{True} or the necessary precondition for \texttt{tail xs} not to crash must be fulfilled.

As \texttt{null xs} cannot crash we find its precondition to be simply True and we can deduce the precondition for \texttt{tail xs} to be $\texttt{xs}\in\{(:)\}$. Substituting these into our precondition we find
\[ \textrm{True} \land (\texttt{null xs} \in \{\texttt{True}\} \lor \texttt{xs}\in\{(:)\}), \]
which simplifies to
\[ \texttt{null xs} \in \{\texttt{True}\} \lor \texttt{xs}\in\{(:)\}. \]
The subexpression  $\texttt{null xs} \in \{\texttt{True}\}$ forms a postcondition on \texttt{null xs}. From this postcondition Catch computes the required precondition that satisfies it, which in this particular case should turn out to be $\texttt{xs}\in\{[]\}$. Substituting, we find
\[ \texttt{xs}\in\{[]\} \lor \texttt{xs}\in\{(:)\}, \]
which, using our knowledge of the list data type, turns out the be a tautology, i.e. the (trivial) precondition for \texttt{safeTail} is True.

\subsection{Constraint systems}
The example in the previous section used a simple constraint language, specifying to which set of head constructors an expression should evaluate. Mitchell developed two more expressive constraint systems: regular expression constraints and multipattern constraints.

\paragraph{Regular expression constraints} Constraints are formed by a regular expression over an alphabet of a number of ad-hoc \emph{selectors} (e.g. \texttt{hd} and \texttt{tl} for lists.) A precondition for \texttt{map head xs} using regular expression constraints reads $\texttt{xs} \in (\texttt{tl}^{*} \cdot \texttt{hd} \leadsto \{(:)\})$. It should be interpreted as ``after applying zero or more \texttt{tl}s to \texttt{xs} and then applying a \texttt{hd} we should find a \texttt{(:)} constructor.''

Accoriding to Mitchell regular expression constraints tend to scale badly as the program increase in size, although he could not identify more specific condition under which this problem manifests.\footnote{Personal communication}

\paragraph{Multipattern constraints} Multipatterns are of the form $\alpha \star \rho$, where $\alpha$ gives the set of constructors that are valid at the head of the value, while $\rho$ gives the set of constructors that can appear at the recursive positions (if any) of the constructor at the head of the value. The elements of these sets can again be multipatterns. To specify that \texttt{xs} in \texttt{map head xs} should be a list of non-empty lists, we use the multipattern:
\begin{align*}
\{[]&, (:) (\{(:)\ \texttt{Any} \} \star \{ [], (:)\ \texttt{Any}\})\} \\
&\star \\
\{[]&, (:) (\{(:)\ \texttt{Any} \} \star \{ [], (:)\ \texttt{Any}\})\}
\end{align*}
\texttt{Any} is a wildcard that matches any constructor.

\subsection{Discussion}
% should have been mentioned earlier?
The analysis of Catch works on a first-order language. The input program needs to be defunctionalized before it can be analyzed. The defunctionalization algorithm employed by Catch is not complete.

Calculated preconditions are unnecessarily restrictive in the presence of laziness (Mitchell, p. 142).

\section{Static Contract Checking}
\subsection{Overview}
Dana Xu's work on static contract checking for Haskell \cite{Xu09staticcontract} (including the ESC/Haskell system \cite{Xu:2006:ESC:1159842.1159849}) allows checking of arbitrary programmer supplied pre- and postconditions on functions and is able to detect pattern match failures in the process.

Programmers define pre- and postconditions in the form of a contract (a refinement type). An appropriate contract for the \texttt{head} function defined above might be:
\begin{verbatim}
{-# CONTRACT head :: {s | not (null s)) -> {x | True} #-}
\end{verbatim}
This contract states that if head is given a list \texttt{s} for which the predicate \texttt{not (null s)} holds, i.e. it is a non-empty list, the function will not crash\footnote{The system only guarantees partial correctness, so the term might still diverge.}, indicated by the trivial postcondition \texttt{True} on the return value.

\begin{figure}
\centering
\begin{tabular}{lclr}
t &$::=$& $\{ x | p \}$     & Predicate          \\
  & $|$ & $x : t_1 \to t_2$ & Dependent function \\
  & $|$ & $(t_1, t_2)$      & Constructor         \\
  & $|$ & $\mathtt{Any}$    & Polymorphic ``Any''
\end{tabular}
\caption{Syntax of Xu's contracts}
\label{xu-contract-syntax}
\end{figure}

Contracts can be constructed from predicates, a dependent function space constructor, arbitrary constructors and a polymorphic \texttt{Any} contract that is always satisfied, including by functions which crash (Fig. \ref{xu-contract-syntax}). Any \texttt{Bool}-valued Haskell expression can be used as a predicate.

Dependent functions are helpful when declaring a contract, e.g. for the \texttt{reverse} function:
\begin{verbatim}
{-# CONTRACT reverse :: {xs | True} -> {rs | length xs == length rs} #-}
\end{verbatim}
The predicate on the return value depends on the the input. Constructors and the \texttt{Any} contract are useful when declaring a contract for \texttt{fst}:
\begin{verbatim}
{-# CONTRACT fst :: (Ok, Any) -> Ok #-}
\end{verbatim}
Here \texttt{Ok} is used as a shorthand for \texttt{\{$x$ | True\}}. As \texttt{fst} discards the value on the right in the tuple we should not care if it is an expression that crashes, so we can not use \texttt{Ok}.

\subsection{Contract checking}
Haskell's syntax is extended with two \emph{exception values} -- \texttt{BAD} and \texttt{UNR} -- which are only used internally by the checker. \texttt{BAD} signifies an expression which crashes and \texttt{UNR} an expression which is either unreachable or diverges.

In the program that is going to be verified all crashing expressions (including \texttt{error} and \texttt{undefined}), as well as missing patterns in case-expressions are replaced by \texttt{BAD} exceptions.

The verification of the contracts is based on a translation of the contracts into Findler--Felleisen wrappers \cite{Findler:2002:CHF:581478.581484}, which will cause a function to crash (using \texttt{BAD}) if it disobeys its contract or diverge if it called in a way that is not permitted. While technically interesting it is not directly relevant to the detection of pattern match failures so we shall not discuss it in more depth.

The actual verification process continues by \emph{symbolic evaluation}, basically applying various simplifications to the resulting program including $\beta$-reductions. Any code deemed to be unreachable is pruned from the program.

The presence of recursive functions in the contracts might cause the symbolic evaluation to diverge if care is not taken to limit the number of evaluation steps. By setting an upper bound on the number of simplification steps we lose accuracy, but gain decidability of the verification process.

Arithmetical expressions in case-expressions (e.g. \texttt{case x*x >= 0 of \{ True -> ...; False -> BAD \}}) cannot be handled directly by the symbolic evaluator. These are collected in a table and send to an external constraint or SMT solver. If the solver determines these expressions are inconsistent the code is unreachable and can be pruned.

After the symbolic evaluation has terminated the checker will tell the programmer if the program is ``definitely satisfies the contract'' (if no \texttt{BAD} exceptions remain anywhere in the program), ``definitely does not satisfy the contract'' or ``don't know.''

\subsection{Discussion}
Unlike our envisioned system, Xu's static contract checking requires programmer supplied contracts on nearly all functions (contracts on trivial functions can be omitted and are handled by inlining their definition when called.) Compared to Mitchell's Catch it can handle higher-order functions natively and has a more (too?) expressive contract language.

\section{Dependent ML}
DML($\mathcal{L}$) is an extension of the ML programming language which enriches ML's type system with a limited form of dependent types over a (decidable) constraint domain (or index language) $\mathcal{L}$ \cite{DML-JFP07}.

Xi's initial example -- recast in a more Haskell-like syntax -- declares a {\tt List} data type dependent on an integer corresponding to the length of the list and a function to concatenate two such lists:
\begin{verbatim}
1  type Nat = {a :: Int | a >= 0}
2
3  data List<Int> a = Nil<0>
4                   | {n :: Nat} Cons<n+1> a (List<n> a)
5
6  (++) :: {m :: Nat} {n :: Nat} List<m> a -> List<n> a -> List<m+n> a
7  (++) Nil         ys = ys
8  (++) (Cons x xs) ys = Cons x (append xs ys)
\end{verbatim}

For many functions producing a list, the exact length might not be derived by such a trivial calculation ({\tt m+n}) on the lengths of the input lists. A prominent example is the {\tt filter} function:
\begin{verbatim}
1  filter :: (a -> Bool) -> {m :: Nat} List<m> a -> [n :: Nat | n <= m] List<n> a
2  filter p Nil         = Nil
3  filter p (Cons x xs) | p x       = Cons x (filter p xs)
4                       | otherwise =         filter p xs
\end{verbatim}
here we only know that the resulting list is equal or shorter in length than the input list ({\tt n <= m}). This is expressed as the dependent sum {\tt [n :: Nat | n <= m]}. In a full-fledged dependently-typed language we would also return a proof object stating the predicate {\tt p} holds for all the elements in the output list, but this is beyond the expressive power of DML.

What we gain is a relief from the need to provide proofs for trivial arithmetical (in)equalities. Imagine a {\tt zip} function which requires the two list to be zipped together to be of equal length. When calling {\tt zip (xs ++ ys) (ys ++ xs)} this preconditions seems to be intuitively satisfied. From the point-of-view of the compiler the first list has length $m+n$ and the second lists $n+m$, however. In most dependently-typed languages we will now have to invoke a lemma or tactic proving the commutativity of addition. DML can simply send the constraint $m + n = n + n$ to an ILP solver and conclude the constraint is satisfiable.

\subsection{Discussion}
Compared to Xu's static contract checking, Xi's DML constrains the constraint system to a decidable theory. The constraints and indices -- such as {\tt m+n} and {\tt n~<=~m} -- may superficially look like ordinary Haskell expression, but in fact belong to a much smaller index language. While type checking is decidable, type inference is not and as a result DML still requires type annotations.

To be applicable to a pattern match analysis we must try to infer a type like:
\begin{verbatim}
head :: {n :: Int | n >= 1} [a]<n> -> a
\end{verbatim}
for the {\tt head} function. We also have to implicitly extend the list data type with an index representing its length.

This seems significantly more challenging than inferring the Catch-like type
\begin{alltt}
head :: \{xs :: [a] | xs \(\in\) (:) \} -> a
\end{alltt}
The correspondence between the the type and the pattern matching happening in the definition of {\tt head} is much more direct.

%\begin{small}
%There is an indirect correspondence, however. The value of the index of the list is given by the {\tt length} function and by analyzing the pattern matching happening in that function we could derive that $\{n :: \mathtt{Int} | n >= 1\}\ \left[a\right]\left<n\right> \simeq \{xs :: [a] | xs \in (:) \}$.
%
%Instead of adding the an index of type {\tt Int} to {\tt List} to create {\tt List<Int>}, we add the abstraction {\tt length :: List -> Int}. This generalizes to arbitrary data types $F$ and constraint systems over $\tau$, by adding an abstraction $F \to \tau$ to create $F\left<\tau\right>$. (But can we use arbitrary functions of type $F\to\tau$, or does {\tt lenght} have some particularly nice properties? Certainly if viewed as a catamorphism (e.g. $(0, (\lambda n. 1+n))$) it should consist of only expressions in the constraint language $\mathcal{L}$ -- they correspond exactly to the annotations on {\tt Nil<0>} and {\tt \{n :: Int\} Cons<n+1>} -- but is this sufficient? What if we look at a {\tt Tree}? Adding a property {\tt minDepth :: Tree -> Int} would guarantee we can pattern match on a certain number of {\tt Branch}es before encountering a {\tt Leaf}, but what does {\tt maxDepth :: Tree -> Int} buy us? More importantly, while we might be able to infer an appropriate type for {\tt head}, inferring the type given to {\tt (++)} and {\tt filter} above still seems out of reach.)
%\end{small}

\section{Compiling Pattern Matching}
Case-statements in toy languages often have a very simple \emph{decision tree} semantics. Case-statements in Haskell have a more complex \emph{backtracking automaton} semantics. There is a body of work on compiling the latter into the former. Maranget \cite{DBLP:journals/jfp/Maranget07} gives an algorithm for determining whether a case-statement is exhaustive and whether all patterns are useful. As the analysis does not consider any dataflow information it is much too imprecise for our purpose. It does turn out that the analysis closely follows, but can at places be simplified with respect to, the manner in which pattern matching is compiled. This indicates might also be able to follow such an approach when analyzing case-statements in our pattern-match analysis.

\chapter{Approach}\label{secapproach}

% type and effect system
% refinement types / subtyping / widening (compare to multi-patterns)
% how much precision do we need?

\section{Incremental development}
We will incrementally develop our analysis for the various constructs in the Haskell language.

\paragraph{Variables, constants, abstraction, application, if-statements and simple let-bindings} We will start out by implementing the basic constructs of the simply typed $\lambda$-calculus. This includes simple let-bindings, i.e. non-recursive let-bindings in which no pattern matching or generalization occurs.

At this stage we will only allow booleans and integers as constants. Due to their primitive and infinity nature integers will need to be treated specially by the analysis in any case. Booleans and if-statements can be handles in a more general fashion during later stages.
\paragraph{Recursion} The next elementary construct is recursion. This will require us to implement a form of \emph{widening} for the integers and lists to ensure termination of the analysis. As the exact form of widening used can have a big impact on the precision of the analysis we do not want to commit to a specific lattice here. Various widening strategies should be easily pluggable into the source code and perhaps eventually be able to be specified at a higher-level by the user of the tool.
\paragraph{Simple case-statements} The most critical part of the analysis will be how it handles case-statements. In this state we will add simple case-statements (i.e., non-nested, non-backtracking, guardless case-statements) and primitive list constants, our first recursive data type.
\paragraph{Complex case-statements and guards} To be able to handle the full Haskell 98 language our analysis must be able to deal with the complex semantics of case-statements. This includes nested-patterns, guards and backtracking. We would like to be able to analyze any boolean-valued expression that is used as a guard.
\paragraph{Algebraic data types} Next we extend the analysis to work on arbitrary programmer-defined data types instead of only on our primitive list data type. This will mostly involve generalizing out analysis to work on with a representation of arbitrary algebraic data types. Data types containing fields with function types may also prove interesting.
\paragraph{Polymorphism} Polymorphism can generally be added to an analysis as a relatively orthogonal feature. However, there might be some interesting interactions with our refinement and intersection types here.
\paragraph{Full let-bindings} We extend the analysis for let-bindings to handle (mutual) recursion and pattern matching.
\paragraph{Type classes} An interesting feature of Haskell's type system are type classes. We currently do not foresee any specific difficulties with respect to be able to handle them in out analysis.
\paragraph{Syntactic sugar} As the final stage we would add constructs which are only syntactic sugar and therefore would not pose any significant problem to our analysis. This includes for example list-comprehensions and do-notation.

\section{Refinement types}

Compared to Freeman and Pfenning's approach to refinement types we will initially not rely on programmer supplied, but instead build our lattice of types from finite unrollings of data types. We hope this makes calculating the lattice easier.

Freeman's inference algorithm for refinement type assign a single, albeit complex, type to each binding. This makes it a monovariant analysis. We will develop a polyvariant analysis, making it sensitive to the sites where a binding is used. We hope this will improve the performance of the analysis.

\section{Parsing}

We will develop our tool in the Haskell programming language. A number of libraries for parsing input programs are available. The most obvious choice seem to be the {\tt haskell-src-exts} package, which is also used by the HLint tool. If this proves to be a problematic choice, other possibilities are the parser from GHC or UHC.

\chapter{Deliverables}\label{secdeliverables}

There will be two deliverables of this project:
\begin{enumerate}
\item A \emph{tool} implementing the pattern match analysis for Haskell 98. This tool is intended to be practically usable and not a mere research prototype.
\item A \emph{thesis} describing the theory underlying the tool and documenting the implementation choices made in the tool.
\end{enumerate}

\chapter{Evaluation}\label{secevaluation}

The evaluate whether we were able to answer research question we first need to evaluate whether the developed tool implements a practically usable pattern-match analysis. We can do so by trying it out on the examples sketched in Section \ref{secexamples}, the literature and for example the {\tt nofib} testsuite.

If we can answer this question with a ``yes'', and assuming the tool indeed implements the analysis using a type and effect system and refinement types, we can also answer out research question with a ``yes''.

In case of a ``no'' we will need to argument whether this is due to a failure of our implementation or because there are good theoretical reasons why such an analysis cannot be implemented.

\chapter{Planning}\label{secplanning}

\begin{description}
\item[Implement framework (April)] Implement a constraint-based version of Hindley--Milney for Haskell (the underlying the system of the analysis). This is known work \cite{Odersky:1999:TIC:308181.308203,DBLP:journals/entcs/HageH09} and should not prove to be difficult. This implementation will provide a guiding framework for the analysis, highlight possible issues we might run into when implementing a constraint-based analysis for full Haskell and allow us to decide which parsing library and abstract syntax representation to use.
\item[Implement analysis (April--July)] The largest portion of time is assigned to incrementally developing and building the pattern-match analysis. The likely order will be:
\begin{enumerate}
\item Variables, constants, abstraction, application, if-statements and simple let-bindings (1.5 weeks)
\item Recursion (1.5 weeks)
\item Simple case-statements (2 weeks)
\item Complex case-statements and guards (2 weeks)
\item Algebraic data types (1.5 weeks)
\item Polymorphism (1.5 weeks)
\item Full let-bindings (1 week)
\item Type classes (1.5 weeks)
\item Syntactic sugar (1.5 weeks)
\end{enumerate}
\item[Evaluation and thesis writing (July--August)] The final phase of this project will include an evaluation of the tool and finishing of the thesis. A significant potion of the thesis is expected to be written in parallel with the implementation, after a particular subtask has been implemented successfully.
\end{description}

\chapter{Future Work}

Non-goals of the thesis project are:
\begin{enumerate}
\item A formal soundness proof of the analysis, and
\item Supporting Haskell extensions beyond Haskell 98.
\end{enumerate}
Both of these would be candidates for future work.

\bibliography{thesis-proposal}
\bibliographystyle{plain}

\end{document}

### 2015-09-01 deprecated/pattern-match-analysis/thesis/abstract.tex
pattern-match-analysis

\chapter*{\centering \begin{normalsize}Abstract\end{normalsize}}
\begin{quotation}
\noindent
In this thesis we attempt to answer the research question:
\begin{quote}
Can we use a type and effect system in combination with refinement types to develop a pattern-match analysis for a non-strict higher-order functional language that is both performant and precise enough to be of practical use?
\end{quote}

In Chapter \ref{chapintroduction} we present a number of examples to demonstrate why this is an interesting problem. In Chapter \ref{chapcontext} we give a short introduction to the relevant concepts of the research question: higher-order functional languages and type and effect systems. In Chapter \ref{chappma} we give an overview of the pattern match analysis we developed and give a detailed description of the constraint generation and constraint solving phases in respectively Chapter \ref{chapcg} and Chapter \ref{chapcs}. In Chapter \ref{chapimplementation} we discussion the implementation of the analysis we have built. We evaluate the effectiveness and limitations of our analysis in Chapter \ref{chapevaluation}. In Chapter \ref{chaprelatedwork} we present work related to our research question and discuss which aspects of that work are relevant to or different from our proposed system. Finally, in Chapter \ref{chapfurtherresearch}, we propose a number of directions for further research and future implementation work to improve the precision and applicability of the analysis.

\end{quotation}
\clearpage

### 2015-09-01 deprecated/pattern-match-analysis/thesis/context.tex
pattern-match-analysis

\chapter{Context}\label{chapcontext}

% halting problem
% well-typed programs can't go wrong
% smaller types log form presentation

\section{Haskell}

Haskell is statically-typed functional programming language with non-strict evaluation semantics \cite{Haskell98}. As a functional language it has first-class and higher-order functions and features a rich type system supporting parametric polymorphism and type classes. Programmers can define custom types in the form of algebraic data types and write functions over them using pattern matching.

Like most functional languages, Haskell can be easily translated into a typed $\lambda$-calculus (System $F_C$). From the point-of-view of the programmer it offers a wealth of syntactic sugar over a plain $\lambda$-calculus---such as guards and list comprehensions---allowing programs to be expressed concisely and in a readable fashion.

The following example demonstrates the syntax of Haskell and some features mentioned previously:

\begin{code}
data Tree a  =  Branch (Tree a) (Tree a)
             |  Leaf a

mapTree :: (a -> b) -> Tree a -> Tree b
mapTree f (Branch t1 t2)  = Branch (mapTree f t1) (mapTree f t2)
mapTree f (Leaf a)        = Leaf (f a)

instance Functor Tree where
    fmap = mapTree
\end{code}
Lines 1--2 define an algebraic data type (ADT) representing a tree. A tree can be constructed by either a |Branch| containing a left and a right subtree or by a |Leaf| containing a value. The ADT is parameterized by a type |a| giving the type of the value stored in the |Leaf|.

Lines 4--5 define a higher-order function |mapTree| which applies a given function |f :: a -> b| to all the values stored in all the leaves of a given |Tree| resulting in a new |Tree|. As the domain and the range of |f| need not coincide, the type parameter of the tree changes as well. The body of the function is defined by pattern matching over the constructors of |Tree| and recurses in the case of a |Branch| and applies the argument |f| to the field of type |a| in |Leaf| otherwise.

Lines 8--9 declare the type constructor |Tree| to be an instance of the type class |Functor|, expressing it to be an endofunctor in the category of data types with its mapping on morphisms given by |mapTree|.

\subsection{Pattern Matching}

Particularly relevant to our analysis are Haskell's pattern matching abilities. We demonstrate some of them in the following function:

\begin{code}
rotate (Branch b@(Branch (Leaf x) (Leaf y)) (Leaf z))
           | z < x      = Branch (Leaf z) b
           | z < y      = Branch (Branch (Leaf x) (Leaf z)) (Leaf y)
           | otherwise  = undefined
\end{code}

The function will rotate a tree |(Branch (Branch (Leaf 2) (Leaf 3)) (Leaf 1))| into |Branch (Leaf 1) (Branch (Leaf 2) (Leaf 3))|. It accomplishes this through the use of \emph{nested pattern matching}\index{pattern matching!nested}, \emph{as-patterns}\index{as-pattern} and \emph{guards}\index{guard}.

This function only works on trees of a very specific shape. Feeding the function a tree of a different shape will cause a pattern match failure. Similarly, an error will be raised if the function reaches the |otherwise| branch, containing an |undefined| value. This would also happen if the branch had called |error| or the branch and its guard were left out entirely.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Type and Effect Systems}

Type and effect systems are approaches to program analysis suitable for typed languages \cite{Nielson:1999:PPA:555142}. In this formalism we take the \emph{underlying type system}\index{underlying type!underlying type system} of the language and (conservatively) extend it by adding \emph{annotations}\index{annotation} to the types (base, function and other) of the system.

\begin{figure}[p]
\[\frac{}{\Gamma \vdash_{\mathrm{UL}} c : \tau_c}\ \textrm{[T-Con]} \quad\quad\quad\quad
\frac{\Gamma(x) = \tau}{\Gamma \vdash_{\mathrm{UL}} x : \tau}\ \textrm{[T-Var]}\]
\[\frac{\Gamma[x \mapsto \tau_1] \vdash_{\mathrm{UL}} e : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \lambda x. e : \tau_1 \to \tau_2}\ \textrm{[T-Abs]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_2 \to \tau_1 \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} e_1\ e_2 : \tau_1}\ \textrm{[T-App]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_1 \quad \Gamma[x\mapsto \tau_1] \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \mathbf{let\ } x = e_1 \mathbf{\ in\ } e_2 : \tau_2}\ \textrm{[T-Let]}\]
\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau}{\Gamma \vdash \mathbf{if\ } e_0 \mathbf{\ then\ } e_1 \mathbf{\ else \ } e_2 : \tau}\ \textrm{[T-If]}\]
\caption{Underlying type system}
\label{underlying}
\end{figure}

\begin{figure}[p]
\[\frac{}{\Gamma \vdash_{\mathrm{AN}} n : \mathrm{Int}^{\{n\}}}\ \textrm{[T-Con]} \quad\quad\quad\quad
\frac{\Gamma(x) = \widehat\tau}{\Gamma \vdash_{\mathrm{AN}} x : \widehat\tau}\ \textrm{[T-Var]}\]
\[\frac{\Gamma[x \mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \lambda x. e : \widehat\tau_1 \to \widehat\tau_2}\ \textrm{[T-Abs]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_2 \to \widehat\tau_1 \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} e_1\ e_2 : \widehat\tau_1}\ \textrm{[T-App]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_1 \quad \Gamma[x\mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \mathbf{let\ } x = e_1 \mathbf{\ in\ } e_2 : \widehat\tau_2}\ \textrm{[T-Let]}\]
\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_2} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_3}}{\Gamma \vdash_{\mathrm{AN}} \mathbf{if\ } e_0 \mathbf{\ then\ } e_1 \mathbf{\ else \ } e_2 : \tau^{\varphi_2 \cup \varphi_3}}\ \textrm{[T-If]}\]
\[\frac{
\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi}
}{
\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi \cup \varphi'}
}\ \textrm{[T-Sub]}\]
\caption{Annotated type system}
\label{annotated}
\end{figure}

\subsection{Example}

As an example we give an analysis which determines, in the form of an annotation, which values an expression can evaluate to. The language is the simply-typed $\lambda$-calculus with let-bindings and an if-statement. The underlying type systems is standard and given in Figure \ref{underlying}.

The annotated type system is given in Figure \ref{annotated}. It does little more than collecting all constants in the annotations. The only interesting rule is for the if-statement. We need to take the union over the values collected in both branches of the if-statement. We could have chosen to make the rule depend on the annotation of the guard expression, passing on only the values collected in the then-branch along if we knew it could only be |True| or only the values collected in the else-branch if we knew it could only be |False| from its annotation. This would make the analysis more precise.

We have also added an inference rule for \emph{subeffecting}\index{subeffecting}. This allows us to make the annotation less precise. This is sometimes necessary when applying a value about which we have very accurate information to a function which requires a more general argument.

\subsection{Type Inference}

An analysis specified as an inference system does not allow us to compute the result of the analysis directly, as the premises of several of the inference rules usually require us to guess the correct type annotations for the type and annotation variables (e.g., in the presence of subeffecting). To overcome this problem we need a \emph{type reconstruction algorithm}\index{type reconstruction!algorithm}.

If the underlying type system is Hindley--Milner, then Algorithm W can be used to infer the types. Algorithm W works bottom-up and can compute types in a single pass by unifying type variables. This may not always be possible for the annotations in an annotated type system, however. Here we can use a variant of Algorithm W to gather a set of constraints on the annotation variables and solve these constraints with a worklist algorithm in a second phase.

\subsection{Polyvariance}

% 1) precision (poisoning)
% 2) modularity

% annotation polymorphism

Consider the following program:
\begin{code}
main =  let id x = x
        in (id 1, id 2)
\end{code}

When performing the analysis given above we would assign the type $\mathit{Int} \to \mathit{Int}^{\{1, 2\}}$ to |id| and as a result the type $\mathit{Int}^{\{1,2\}} \times \mathit{Int}^{\{1,2\}}$ to |f|. Clearly, this result is not optimal. This loss in precision has been caused by the two separate calls to |id| \emph{poisoning}\index{poisoning} each other.

A similar situation occurs at the type-level in the function:
\begin{code}
main =  let id x = x
        in (id 1, id true)
\end{code}
Hindley--Milner manages to avoid problems in this situation by a mechanism called \emph{let-generalization}\index{generalization!let-generalization}. The binding |id| has the type $\alpha \to \alpha$. Instead of trying---and failing---to unify $\alpha$ with both |Int| and |Bool| in the body of the let-binding, the type of |id| is generalized to the polymorphic type $\forall \alpha. \alpha \to \alpha$. In the body of the let-binding, |id| can be instantiated twice: once to |Int -> Int| and once to |Bool -> Bool|.

A similar trick can be used for the annotated type system, except that instead of quantifying over type variables, we quantify over annotation variables.

### 2015-09-01 deprecated/pattern-match-analysis/thesis/evaluation.tex
pattern-match-analysis

\chapter{Evaluation}\label{chapevaluation}

\section{Examples}

\subsection{Higher-order functions}

Given the program
\begin{code}
main = \b -> \phi ->  if b
                      then  if  phi 42  then  100  else  200
                      else  if  phi 43  then  300  else  400
\end{code}
the analysis will infer the type $\forall \alpha\beta.\ \{\alpha \subseteq \{\FFalse,\TTrue\}, \beta \subseteq \{\FFalse, \TTrue\}\} \Rightarrow (\Bool^{\alpha} \overset{\top}{\to} ((\Int^{\{\mathbf{+}\}} \overset{\top}{\to} \Bool^{\beta}) \overset{\top}{\to} \Int^{\{\mathbf{+}\}}))$.

\subsection{Detecting pattern match failures}

Given the program
\begin{code}
main =  let  f =  \b  -> \phi ->  if b
                                  then  if  phi  [42]  then  100  else  200
                                  else  if  phi  [43]  then  300  else  400
             g =  \x  -> case  x of
                               []     -> True
             h =  \x  -> case  x of
                               (a:as) -> True
        in f True g
\end{code}
the analysis will report that a pattern match failure occurs. Changing |f True g| into |f True h| will make the pattern match failure---and the reporting of it by the analysis---go away.

\subsection{|filter|}
We given the program
\begin{code}
main = let filter = \p -> \ys -> case  ys of
                                       []      ->  []
                                       (x:xs)  ->  let g = p x
                                                   in if g  then x :  filter p xs
                                                            else      filter p xs
       in filter (\x -> x) [True, False]
\end{code}
the analysis will infer the type $\ListOf{\Bool^{\{\FFalse,\TTrue\}}}^\top$.

\section{Limitations}

\subsection{Context-insensitivity}

The analysis is \emph{context-insensitive}. Given the program:

\begin{code}
main =  let f = \b ->  let tail = \xss -> case  xss of
                                                (x:xs) -> xs
                       in tail (if b then [] else [1,2,3])
         in f False
\end{code}
a pattern match failure will be reported. One does not occur practice, as the value $\FFalse$ passed to $b$ will prevent an empty list being returned from the if-then-else expression to |tail|. Section \ref{ctxsens} discusses how to alleviate this limitation.

\subsection{Structurally recursive functions}

Our analysis suffer from another limitation: the inferred type for a number of structurally recursive function such as |map| and |foldr| is not precise enough.

While the inferred type for |tail|
\[ \forall \alpha, (\beta \subseteq \Conss{\gamma}).\ \ListOf{\alpha}^\beta \to \ListOf{\alpha}^\gamma \]
expresses everything we can say about the function---the values of the elements in the list are preserved and the lenght is output list is one element shorten than the input---this is not the case for |map|:
\[ \forall \alpha, \beta, \gamma.\ (\alpha \to \beta) \to \ListOf{\alpha}^\gamma \to \ListOf{\beta}^\top \]
Our constraint solver is not able to infer from the constraints generated by the pattern matching on the input list and subsequent cons of the modified head and tail that |map| preserves the length of the list. I.e., we would have expected the type
\[ \forall \alpha, \beta, \gamma.\ (\alpha \to \beta) \to \ListOf{\alpha}^\gamma \to \ListOf{\beta}^\gamma \]
This will be a severe limitation in practice, as even innocent functions such as |head (map id [1..9])| will result in a pattern match warning.

### 2015-09-01 deprecated/pattern-match-analysis/thesis/furtherresearch.tex
pattern-match-analysis

\chapter{Conclusions and Further Research}\label{chapfurtherresearch}

% undefiend, exceptions

\section{Conclusions}

Using our analysis we are able to statically detect pattern match failures---or the absence of them, where a more naive totality checker would have given us spurious warnings. Using a type and effect system to do pattern match analysis certainly appears to be a viable approach.

As we demonstrated in Chapter \ref{chapevaluation}, there are a number of limitations to our current analysis that may result in reduced accuracy, that is, an increase in the number of false positives. We will present a number of suggestion for further research that may help overcome these limitations.

\section{Improving Precision}

\subsection{Intersection Types}

The analysis sketched so far falls short if we want to analyze functions using common operators such as |(+)|  or |(<=)|. To see the problem here we should ask ourself what type we would like to assign to such expressions. The operator |(<=)| can take any integer arguments---which does not cause a pattern match failure when evaluated---and returns either $\mathbf{True}$ or $\mathbf{False}$ without causing a pattern match failure itself. The most reasonable annotated type for this operator thus seems to be: \[ (\leq) : \Int^\top \to \Int^\top \to \Bool^\top\]

Using the operator in the following situation:

\begin{code}
f x = if 0 <= 1 then x else crash
\end{code}

will produce a spurious warning about a pattern-match error occurring. The guard of the if-then-else expression will always evaluate to $\TTrue$, but because of the type we have assigned to $(\leq)$ the analyzer will believe it might evaluate to $\FFalse$ as well.

The root of the problem is that we can assign several different valid refinement types to |(<=)|:
\begin{eqnarray*}
(\leq) &:& \Int^{\{-\}} \to \Int^\top \to \Bool^{\{\TTrue\}} \\
(\leq) &:& \Int^{\{0,+\}} \to \Int^{\{-\}} \to \Bool^{\{\FFalse\}} \\
(\leq) &:& ...
\end{eqnarray*}
neither of which is a \emph{principal type}---that is having as its refinements a strictly minimal set of values on types in a covariant positions and a stricly maximal set of values on types in a contravariant position.

A solution would be to assign this expression a more accurate \emph{intersection type} (see Section \ref{secit}). For example, when using a sign abstraction for integers: \[ (\leq) : \Int^{\{\mathbf{-}\}} \to \Int^\top \to \Bool^{\{\TTrue\}} \land \Int^{\{\mathbf{0}, \mathbf{+}\}} \to \Int^{\{\mathbf{-}\}} \to \Bool^{\{\FFalse\}} \land ... \]

Note that---for our purposes at least---if we assign some expression $e$ the intersection type $\tauhat_1 \land \tauhat_2 \land ... \land \tauhat_n$ then any two annotated types $\tauhat_i$ and $\tauhat_j$ will have the same underlying type $(\lfloor\tauhat_i\rfloor = \lfloor\tauhat_j\rfloor)$ and the two types will thus only differ in their annotations.

To represent intersection types we therefore only have to keep track of the various possible assignments of the annotation variables. We can integrate this concept neatly into our type system by modifying the typing relation $\TypeRel{e}{\tauhat}{C}{R}$ to a typing relation \[ \TypeRel{e}{\tauhat}{C}{\bar R} \] where $\bar R$ is a \emph{set of constraint sets} over the refinements, expressing the various possible assignments of annotations; one constraint set for each components of the intersection type.

All typing judgments will need to be modified to properly propagate the sets of constraint sets. During simplification any constraint sets that are inconsistent can be removed.

\subsection{Implication Constraints}\label{ctxsens}

In $k$-CFA data-flow is used to improve the precision of the control-flow analysis \cite{Shivers:1991:CAH:124950}. Conversely, we can try to improve the precision of our data-flow anlysis by considering the control-flow.

Consider again the program we saw in Section \ref{sec31}:

\begin{code}
main b f =  if b then
                if  f 42  then  100  else  200
            else
                if  f 43  then  300  else  400
\end{code}

We assigned this the type:
\begin{equation}\label{orig9}
\Bool^{\{\TTrue, \FFalse\}} \to (\Int^{\{42, 43\}}\to\Bool^{\{\TTrue,\FFalse\}})\to \Int^{\{100,200,300,400\}}
\end{equation}

If we have more knowledge about the exact value of |b|, e.g. |b| being always $\TTrue$ or $\FFalse$, we can infer the more precise types
\begin{equation}\label{new91}
\Bool^{\{\TTrue\}} \to (\Int^{\{42\}}\to\Bool^{\{\TTrue,\FFalse\}})\to \Int^{\{100,200\}}
\end{equation}
and
\begin{equation}\label{new92}
\Bool^{\{\FFalse\}} \to (\Int^{\{43\}}\to\Bool^{\{\TTrue,\FFalse\}})\to \Int^{\{300,400\}}
\end{equation}

This knowledge can be expressed using \emph{implication constraints} \[ \rho_1 \subseteq \rho_2 \models \rho_3 \subseteq \rho_4 \]
where the constraint $\rho_3 \subseteq \rho_4$ on the right-hand side only has to hold if the constraint $\rho_1 \subseteq \rho_2$ on the left-hand side holds.

The original type (\ref{orig9}) and additional types (\ref{new91}) and (\ref{new92}) can together be represented as a single type using implication constraints:
\begin{align*}
\forall \alpha, \beta, \gamma.\ \{&\{\TTrue\} \subseteq \alpha \models \{42\} \subseteq \beta, \{\TTrue\} \subseteq \alpha \models \{100, 200\} \subseteq \gamma,\\ &\{\FFalse\} \subseteq \alpha \models \{43\} \subseteq \beta, \{\FFalse\} \subseteq \alpha \models \{300,400\} \subseteq \gamma,\\ &\alpha \subseteq \{\TTrue, \FFalse\}\} \Rightarrow \Bool^{\alpha} \to (\Int^{\beta}\to\Bool^{\{\TTrue,\FFalse\}})\to \Int^{\gamma}
\end{align*}

% HOW TO SOLVE GOES HERE

It is not yet clear if implications constraints would be orthogonal to intersection types. We could represent the type equivalently as:
\begin{align*}
\Bool^{\{\TTrue, \FFalse\}} &\to (\Int^{\{42, 43\}}&\to\Bool^{\{\TTrue,\FFalse\}})&\to \Int^{\{100,200,300,400\}} \\
\land\ \Bool^{\{\TTrue\}} &\to (\Int^{\{42\}}&\to\Bool^{\{\TTrue,\FFalse\}})&\to \Int^{\{100,200\}} \\
\land\ \Bool^{\{\FFalse\}} &\to (\Int^{\{43\}}&\to\Bool^{\{\TTrue,\FFalse\}})&\to \Int^{\{300,400\}}
\end{align*}

\subsection{Set-based Constraints}

It is worth investigating if the generated constraints can be formulated in terms of \emph{set-based constraints} \cite{aiken,Heintze94setconstraints}. More elaborate solvers (such as BANE) have been developed for this type of constraints and may give more accurate results than our current solver.

\section{Improving Applicability}

\subsection{Handling |undefined|}

While our analysis can detect pattern match failures due to missing alternatives in case-expressions, we would also like to detect failures occurring due to evaluating |undefined| or |error|.

A straightforward approach would seem to add an additional value $\bot$---not to be confused with the least element of a lattice---to the language of refinements $\varphi$, representing a value the diverges if evaluated. The type rules for constructs that perform pattern matching already, such as T-If and T-App, already ensure that such values would not be allowed as the scrutinee of a pattern match.

There are still two hurdles that need to be overcome, however:
\begin{enumerate}
\item While the type constructors for lists, tuples and functions already come prepared with an annotation in which the diverging value $\bot$ can be stored, |undefined| itself has the polymorphic type $\forall \alpha.\ \alpha$. Type variables, however, do not come equipped with an annotation, so we cannot assign it the obvious type $\forall \alpha.\ \alpha^{\{\bot\}}$.
\item Without context-sensitivity in the type rules T-If and T-Case, they will simply propagate the diverging value forward if it occurs in any one of the branches of an if-then-else or case-expression. Adding context-sensitivity as described Section \ref{ctxsens} would be a prerequisite.
\end{enumerate}

%\section{Improving Feedback}
%
%\section{Improving Applicability}
%
%\begin{enumerate}
%    \item Guards
%    \item Algebraic data types
%\end{enumerate}
%
%\section{Evaluation}

### 2015-09-01 deprecated/pattern-match-analysis/thesis/implementation.tex
pattern-match-analysis

\chapter{Implementation}\label{chapimplementation}

An implementation of the analysis has been implemented in Haskell. It has been built on top of the \texttt{haskell-src-exts} library, which is used for both parsing of the input and representation of the abstract syntax tree.

Refinements and other extensions on Haskell's type system that are required for the analysis are represented using the annotation fields available in the abstract syntax of \texttt{Language.Haskell.Exts.Annotated}.

An overview of the modules that comprise the implementation can be found in Figure \ref{moduleoverview}.

Additionally, we have build a collection of example functions that where used for regression testing purposes during the development of the analysis and its implementation.

A demonstration of the implementation can be found in Appendix \ref{appx}.

\vspace{5em}

\begin{figure}[h]
\centering
\begin{tabular}{lrl}
Module & Lines & Description \\
\hline
\texttt{Main}          &  134 & Main \\
\texttt{TypeInference} &  391 & Constraint generation \\
\texttt{PatternTyping} &  153 & Typing of patterns \\
\texttt{Solver}        &  263 & Constraint solving \\
\texttt{Common}        &  342 & Substitution and unification \\
\texttt{Refinements}   &  152 & Refinements (representation and operations) \\
\texttt{Abstract.*}    &  355 & Abstract values (unit, booleans, integers, lists)\\
\texttt{Util.*}        &   48 & Miscellaneous
\end{tabular}
\caption{Module overview}
\label{moduleoverview}
\end{figure}

### 2015-09-01 deprecated/pattern-match-analysis/thesis/introduction.tex
pattern-match-analysis

--- 2015-09-01 deprecated/pattern-match-analysis/thesis-proposal/thesis-proposal.tex
@@
-\documentclass[a4paper]{report}
+\chapter{Introduction}\label{chapintroduction}
 
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\usepackage{alltt}
-
-\usepackage{tikz}
-
-\begin{document}
-
-\title{Thesis Proposal}
-\author{Ruud Koot}
-\maketitle
-
-% Also see:
-% http://www.cs.uu.nl/wiki/bin/view/Thesis/WritingAThesisProposal
-% mail from Jur
-% Tamar's/Eric's proposal
-
-\chapter*{Introduction}
-
-This document is a proposal for a thesis on answering the research question:
-\begin{quote}
-Can we use a type and effect system in combination with refinement types to develop a pattern-match analysis for a non-strict higher-order functional language that is both performant and precise enough to be of practical use?
-\end{quote}
-
-In Chapter \ref{secmotivation} we present a number of examples to demonstrate why this is an interesting problem.
-In Chapter \ref{seccontext} we give a short introduction to the relevant concepts of the research question: higher-order functional languages, type and effect systems and refinement types.
-In Chapter \ref{secrelatedwork} we present work related to our research question and discuss which aspects of that work are relevant to or different from our proposed system.
-In Chapter \ref{secapproach}, \ref{secdeliverables} and \ref{secevaluation} we will discuss our envisioned approach to answering the research question, the set of deliverables and how we will evaluate whether the research question has been answered successfully or not.
-
-\chapter{Motivation}\label{secmotivation}
-
-In 1978, Robin Milner \cite{Milner78atheory} famously wrote (and proved) that:
+In 1978, Robin Milner \cite{Milner78atheory} famously wrote---and proved---that:
 \begin{quotation}
@@
 \begin{verbatim}
-*** Exception: PM.hs:1:1-13: Non-exhaustive patterns in function f
+*** Exception: Non-exhaustive patterns in function f
 \end{verbatim}
@@
 \begin{description}
-\item[Getting stuck] If a program tries to evaluate a non-sensical expression -- such as {\tt 3 + true} -- it cannot possibly make any further progress and is said to be ``stuck''.
+\item[Getting stuck] If a program tries to evaluate a non-sensical expression---such as |3 + true|---it cannot possibly make any further progress and is said to be ``stuck''.
 This is the ``go wrong'' Milner referred to.
 He proved that a sound type system can statically guarantee such expressions will never occur or need to be evaluated at run-time.
-\item[Diverging] If we have a function definition {\tt f = f}, the evaluation of {\tt f} $\leadsto$ {\tt f} $\leadsto ...$ will fail to terminate.
+\item[Diverging] If we have a function definition |f = f|, the evaluation of |f| $\leadsto$ |f| $\leadsto ...$ will fail to terminate.
 We might be making progress in a technical sense, but it will have no useful observable result.
@@
 
-\paragraph{Partial functions} Haskell programmers often work with partial functions, the most common one possibly being {\tt head}:
+\paragraph{Partial functions} Haskell programmers often work with partial functions, the most common one possibly being |head|:
 
-\begin{verbatim}
-1  main = let xs = if length "foo" > 5 then [1,2,3] else []
-2          in head xs
-3
-4  head (x : xs) = x
-\end{verbatim}
+\begin{code}
+main =  let xs = if length "foo" > 5 then [1,2,3] else []
+        in head xs
+
+head (x : xs) = x
+\end{code}
 
@@
 \begin{verbatim}
-On line 2 you applied the function "head" to the empty list "xs".
-The function "head" expects a non-empty list as its first argument.
+    On line 2 you applied the function "head" to the empty
+    list "xs". The function "head" expects a non-empty list
+    as its first argument.
 \end{verbatim}
 
-If the guard of the if-statement had read {\tt length "foo" < 5} the program would have run without crashing and we would like the compiler or tool not to warn us spuriously.
+If the guard of the if-statement had read |length "foo" < 5| the program would have run without crashing and we would like the compiler or tool not to warn us spuriously.
 In case it is not possible to determine statically whether or not a program will crash, a warning should still be raised.
@@
 These data structures must be able to represent all syntactic constructs the parser is able to recognize.
-This results in an abstract syntax tree that is unnecessarily complex and too cumbersome for the later stages of the compiler -- such as the optimizer -- to work with.
-This problem is resolved by \emph{desugaring} the original abstract syntax tree into a simpler -- but semantically equivalent --  abstract syntax tree than does not use all of the constructors available in the original abstract syntax tree.
+This results in an abstract syntax tree that is unnecessarily complex and too cumbersome for the later stages of the compiler---such as the optimizer---to work with.
+This problem is resolved by \emph{desugaring} the original abstract syntax tree into a simpler---but semantically equivalent---abstract syntax tree than does not use all of the constructors available in the original abstract syntax tree.
 
-The compiler writer now has a choice between two different options: either write a desugaring stage {\tt desugar :: ComplexAST -> SimpleAST} -- duplicating most of the data type representing and functions operating on the abstract syntax tree -- or take the easy route {\tt desugar :: AST -> AST} and assume certain constructors will no longer be present in the abstract syntax tree at stages of the compiler executed after the desugaring step.
-The former has all the usual downsides of code duplication -- such as having to manually keep the two or more data types synchronized -- while that latter forgoes many of the advantages of strong typing and type safety: if the compiler pipeline is restructured and one of the stages that was originally assumed to run only after the desugaring suddenly runs before that point the error might only be detected at run-time by a pattern match failure.
+The compiler writer now has a choice between two different options: either write a desugaring stage |desugar :: ComplexAST -> SimpleAST|---duplicating most of the data type representing and functions operating on the abstract syntax tree---or take the easy route |desugar :: AST -> AST| and assume certain constructors will no longer be present in the abstract syntax tree at stages of the compiler executed after the desugaring step.
+The former has all the usual downsides of code duplication---such as having to manually keep the two or more data types synchronized---while the latter forgoes many of the advantages of strong typing and type safety: if the compiler pipeline is restructured and one of the stages that was originally assumed to run only after the desugaring suddenly runs before that point the error might only be detected at run-time by a pattern match failure.
 A pattern match analysis should be able to detect such errors statically.
@@
 \paragraph{Maintaining invariants}
-Many algorithms and data structures maintain invariants that cannot be easily encoded into their type.
+Many algorithms and data structures maintain invariants that cannot easily be encoded into their type.
 These invariants often ensure that certain incomplete case-statements are guaranteed not to cause a pattern match failure.
-An example is the {\tt risers} function from \cite{DBLP:conf/sfp/MitchellR05}, calculating monotonically increasing segments of a list (e.g., {\tt risers [1,3,5,1,2]} $\leadsto$ {\tt [[1,3,5],[1,2]]}):
+An example is the |risers| function from \cite{DBLP:conf/sfp/MitchellR05}, calculating monotonically increasing segments of a list (e.g., |risers [1,3,5,1,2]| $\leadsto$ |[[1,3,5],[1,2]]|):
 
-\begin{verbatim}
-1  risers :: Ord a => [a] -> [[a]]
-2  risers []         = []
-3  risers [x]        = [[x]]
-4  risers (x1:x2:xs) = let (s:ss) = risers (x2:xs)
-5                       in if x <= y then (x:s):ss else [x]:(s:ss)
-\end{verbatim}
+\begin{code}
+risers :: Ord a => [a] -> [[a]]
+risers []            =  []
+risers [x]           =  [[x]]
+risers (x_1:x_2:xs)  =  let (s:ss) = risers (x_2:xs)
+                        in if x_1 <= x_2 then (x_1:s):ss else  [x_1]:(s:ss)
+\end{code}
 
-The let-binding in the third alternative of {\tt risers} expects the recursive call to return a non-empty list.
+The let-binding in the third alternative of |risers| expects the recursive call to return a non-empty list.
 A naive analysis might raise a warning here.
-If we think a bit longer, however, we see that we also pass the recursive call to {\tt risers} a non-empty list.
+If we think a bit longer, however, we see that we also pass the recursive call to |risers| a non-empty list.
 This means we will end up in either the second or third alternative in the recursive call.
@@
 
-Another example might be a collection of mathematical operations working on bitstrings (integers encoded as lists of binary digits without leading zeroes):
-\begin{verbatim}
-1  type Bitstring = [Int]
-2
-3  add :: Bitstring -> Bitstring -> Bitstring
-4  add [] y = y
-5  add x [] = x
-6  add (0:x) (0:y) = 0 : add x y
-7  add (0:x) (1:y) = 1 : add x y
-8  add (1:x) (0:y) = 1 : add x y
-9  add (1:x) (1:y) = 0 : add (add [1] x) y
-\end{verbatim}
+Another example might be a collection of mathematical operations working on bitstrings (integers encoded as lists of binary digits):
+\begin{code}
+type Bitstring = [Int]
 
-The patterns in {\tt add} are far from complete, but maintain the invariant if passed arguments that satisfy the invariant.
+add :: Bitstring -> Bitstring -> Bitstring
+add []      y      = y
+add x       []     = x
+add (0:x)   (0:y)  = 0 : add x y
+add (0:x)   (1:y)  = 1 : add x y
+add (1:x)   (0:y)  = 1 : add x y
+add (1:x)   (1:y)  = 0 : add (add [1] x) y
+\end{code}
+
+The patterns in |add| are far from complete, but maintain the invariant if passed arguments that satisfy the invariant.
 So if we are careful to only pass valid bitstrings into a complex mathematical expression of bitstring-operations it will result in a valid bitstring without crashing due to a pattern match failure.
@@
 % * Variable substitutions in Boolean expressions
-
-\chapter{Context}\label{seccontext}
-
-% halting problem
-% well-typed programs can't go wrong
-% smaller types log form presentation
-
-\section{Haskell}
-
-Haskell is statically-typed functional programming language with non-strict evaluation semantics \cite{Haskell98}.
-As a functional language it has first-class and higher-order functions and features a rich type system supporting parametric polymorphism and type classes.
-Programmers can define custom types in the form of algebraic data types and write functions over them using pattern matching.
-
-Like most functional languages, Haskell can be easily translated into a typed $\lambda$-calculus (System $F_C$).
-From the point-of-view of the programmer it offers a wealth of syntactic sugar over a plain $\lambda$-calculus -- such as guards and list comprehensions -- allowing programs to be expressed concisely and in a readable fashion.
-
-The following example demonstrates the the syntax of Haskell and some features mentioned previously:
-
-\begin{verbatim}
-1  data Tree a = Branch (Tree a) (Tree a)
-2              | Leaf a
-3  
-4  mapTree :: (a -> b) -> Tree a -> Tree b
-5  mapTree f (Branch t1 t2) = Branch (mapTree f t1) (mapTree f t2)
-6  mapTree f (Leaf a)       = Leaf (f a)
-7
-8  instance Functor Tree where
-9      fmap = mapTree
-\end{verbatim}
-Lines 1--2 define an algebraic data type (ADT) representing a tree.
-A tree can be constructed by either a {\tt Branch} containing a left and a right subtree or by a {\tt Leaf} containing a value.
-The ADT is parameterized by a type {\tt a} giving the type of the value stored in the {\tt Leaf}.
-
-Lines 4--5 define a higher-order function {\tt mapTree} which applies a given function {\tt f :: a -> b} to all the values stored in all the leaves of a given {\tt Tree} resulting in a new {\tt Tree}.
-As the domain and the range of {\tt f} need not coincided the type parameter of the tree changes as well.
-The body of the function is defined by pattern matching over the constructors of {\tt Tree} and performs recursion in the case of a {\tt Branch} and applies the argument {\tt f} to the field of type {\tt a} in {\tt Leaf} otherwise.
-
-Lines 8--9 declare the type constructor {\tt Tree} to be an instance of the type class {\tt Functor}, expressing it to be an endofunctor in the category of data types with its mapping on morphisms given by {\tt mapTree}.
-
-\subsection{Pattern matching}
-
-Particularly relevant to our analysis are Haskell's pattern matching abilities. We demonstrate some of them in the following function:
-
-\begin{verbatim}
-rotate (Branch b@(Branch (Leaf x) (Leaf y)) (Leaf z))
-           | z < x     = Branch (Leaf z) b
-           | z < y     = Branch (Branch (Leaf x) (Leaf z)) (Leaf y)
-           | otherwise = undefined
-\end{verbatim}
-
-The function will rotate a tree {\tt (Branch (Branch (Leaf 2) (Leaf 3)) (Leaf 1))} into {\tt Branch (Leaf 1) (Branch (Leaf 2) (Leaf 3))}.
-It accomplishes this through the use of \emph{nested pattern matching}, \emph{as-patterns} and \emph{guards}.
-
-This function only works on trees of a very specific shape.
-Feeding the function a tree of a different shape will cause a pattern match failure.
-Similarly, an error will be raised if the function reaches the {\tt otherwise} branch, containing an {\tt undefined} value.
-This would also happen if the branch had called {\tt error} or the branch and its guard were left out entirely.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Type and effect systems}
-
-Type and effect systems are approaches to program analysis suitable for typed languages \cite{Nielson:1999:PPA:555142}.
-In this formalism we take the the \emph{underlying type system} of the language and (conservatively) extend it by adding \emph{annotations} to the types (base, function and other) of the system.
-
-\begin{figure}[p]
-\[\frac{}{\Gamma \vdash_{\mathrm{UL}} c : \tau_c}\ \mathrm{[con]} \quad\quad\quad\quad
-\frac{\Gamma(x) = \tau}{\Gamma \vdash_{\mathrm{UL}} x : \tau}\ \mathrm{[var]}\]
-\[\frac{\Gamma[x \mapsto \tau_1] \vdash_{\mathrm{UL}} e : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \lambda x. e : \tau_1 \to \tau_2}\ \mathrm{[abs]}\]
-\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_2 \to \tau_1 \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} e_1\ e_2 : \tau_1}\ \mathrm{[app]}\]
-\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_1 \quad \Gamma[x\mapsto \tau_1] \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \mathrm{let\ } x = e_1 \mathrm{\ in\ } e_2 : \tau_2}\ \mathrm{[let]}\]
-\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau}{\Gamma \vdash \mathrm{if\ } e_0 \mathrm{\ then\ } e_1 \mathrm{\ else \ } e_2 : \tau}\ \mathrm{[if]}\]
-\caption{Underlying type system}
-\label{underlying}
-\end{figure}
-
-\begin{figure}[p]
-\[\frac{}{\Gamma \vdash_{\mathrm{AN}} n : \mathrm{int}^{\{n\}}}\ \mathrm{[con]} \quad\quad\quad\quad
-\frac{\Gamma(x) = \widehat\tau}{\Gamma \vdash_{\mathrm{AN}} x : \widehat\tau}\ \mathrm{[var]}\]
-\[\frac{\Gamma[x \mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \lambda x. e : \widehat\tau_1 \to \widehat\tau_2}\ \mathrm{[abs]}\]
-\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_2 \to \widehat\tau_1 \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} e_1\ e_2 : \widehat\tau_1}\ \mathrm{[app]}\]
-\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_1 \quad \Gamma[x\mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \mathrm{let\ } x = e_1 \mathrm{\ in\ } e_2 : \widehat\tau_2}\ \mathrm{[let]}\]
-\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_2} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_3}}{\Gamma \vdash_{\mathrm{AN}} \mathrm{if\ } e_0 \mathrm{\ then\ } e_1 \mathrm{\ else \ } e_2 : \tau^{\varphi_2 \cup \varphi_3}}\ \mathrm{[if]}\]
-\[\frac{
-\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi}
-}{
-\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi \cup \varphi'}
-}\ \mathrm{[sub]}\]
-\caption{Annotated type system}
-\label{annotated}
-\end{figure}
-
-\subsection{Example}
-
-As an example we give an analysis which determines, in the form of an annotation, which values an expression can evaluate to.
-The language is the simply-typed $\lambda$-calculus with let-bindings and an if-statement.
-The underlying type systems is standard and given in Figure \ref{underlying}.
-
-The annotated type system is given in Figure \label{annotated}.
-It does little more than collecting all constants in the annotations.
-The only interesting rule is for the if-statement.
-We need to take the union over the values collected in both branches of the if-statement.
-We could have chosen to make the rule depend on the annotation of the guard expression, passing on only the values collected in the then-branch along if we knew it could only be {\tt True} or only the values collected in the else-branch if we knew it could only be {\tt False} from its annotation.
-This would make the analysis more precise.
-
-We have also added an inference rule for \emph{subeffecting}.
-This allows us to make the annotation less precise.
-This is sometimes necessary when applying a value about which we have very accurate information to a function which requires a more general argument.
-
-\subsection{Type inference}
-
-An analysis specified as an inference system does not allow us to compute the result of the analysis directly, as the premises of several of the inference rules usually require us to guess the correct type annotations for the type and annotation variables (e.g., in the presence of subeffecting).
-To overcome this problem we need a \emph{type reconstruction algorithm}.
-
-If the underlying type system is Hindley--Milner, then Algorithm W can be used to infer the types.
-Algorithm W works bottom-up and can compute types in a single pass by unifying type variables.
-This may not always be possible for the annotations in an annotated type system, however.
-Here we can use a variant of Algorithm W to gather a set of constraints on the annotation variables and solve these constraints with a worklist algorithm in a second phase.
-
-\subsection{Polyvariance}
-
-% 1) precision (poisoning)
-% 2) modularity
-
-% annotation polymorphism
-
-Consider the following program:
-\begin{verbatim}
-f = let id x = x
-     in (id 1, id 2)
-\end{verbatim}
-
-When performing the analysis given above we would assign the type $\mathtt{int} \to \mathtt{int}^{\{1, 2\}}$ to {\tt id} and as a result the type $\mathtt{int}^{\{1,2\}} \times \mathtt{int}^{\{1,2\}}$ to {\tt f}.
-Clearly, this result is not optimal.
-This loss in precision has been caused by the two separate calls to {\tt id} \emph{poisoning} each other.
-
-A similar situation occurs at the type-level in the function:
-\begin{verbatim}
-f = let id x = x
-     in (id 1, id true)
-\end{verbatim}
-Hindley--Milner manages to avoid problems in this situation by a mechanism called \emph{let-generalization}.
-The binding {\tt id} has the type $\alpha \to \alpha$.
-Instead of trying -- and failing -- to unify $\alpha$ with both {\tt int} and {\tt bool} in the body of the let-binding, the type of {\tt id} is generalized to the polymorphic type $\forall \alpha.
-\alpha \to \alpha$.
-In the body of the let-binding, {\tt id} can be instantiated twice: once to $\mathtt{int} \to \mathtt{int}$ and once to $\mathtt{bool} \to \mathtt{bool}$.
-
-A similar trick can be used for the annotated type system, except that instead of quantifying over type variables, we quantify over annotation variables.
-
-\section{Refinement types}
-
-\begin{figure}[t]
-\centering
-\begin{tikzpicture}
-    \tikzstyle{all nodes}=[inner sep=4pt]
-    \draw node(top)at(1,3){\tt List a}
-          node(sorn)at(1,2){{\tt Singleton a} $\lor$ {\tt Nil a}}
-          node(s)at(0,1){\tt Singleton a} node(n)at(2,1){\tt Nil a}
-          node(bot)at(1,0){$\bot$};
-    \draw[-](top)--(sorn);
-    \draw[-](sorn)--(s);
-    \draw[-](sorn)--(n);
-    \draw[-](s)--(bot);
-    \draw[-](n)--(bot);
-\end{tikzpicture}
-\caption{A lattice of {\tt List} and its {\tt rectype}s.}
-\label{listlattice}
-\end{figure}
-
-Refinement types are a form of subtyping that allow us to state more accurately what values a variable of a particular type can contain.
-For example, a refinement type {\tt \{ a :: Int | a < 0 \}} states that {\tt a} is a variable of type {\tt Int} than can only contain negative numbers.
-
-Freeman and Pfenning \cite{Freeman:1991:RTM:113445.113468,Freeman94refinementtypes} give a formal development of refinement types for the ML programming language that can support recursive types.
-
-Their key contribution is allowing the programmer to define {\tt rectype}s. Besides defining a recursive type for lists:
-
-\begin{verbatim}
-data List a = Nil | Cons a (List a)
-\end{verbatim}
-
-The programmer can also define a {\tt rectype} describing singleton lists:
-
-\begin{verbatim}
-rectype Singleton a = Cons a Nil
-\end{verbatim}
-
-Together with an automatically derived {\tt rectype} for the non-recurisive constructor {\tt Nil} (and a union type contructor) the compiler can construct -- using known algorithms on regular tree grammars \cite{DBLP:books/others/tree1984} -- a finite lattice of refinement types for {\tt List a} given in Figure \ref{listlattice}.
-
-\subsection{Union and Intersection Types}
-
-Refinement types are constructed from regular types, including function types, as well as union and intersection types.
-
-A union type such as {\tt Int $\lor$ Bool}, {\tt List a $\lor$ Singleton a} or {\tt Int -> Int $\lor$ Bool -> Bool} state a value can be either of the type on the left or of the type on the right, but we do not know which.
-In the case of {\tt List a $\lor$ Singleton a} we are able to simplify this type to {\tt List a}, as {\tt Singleton a} is a subtype of {\tt List a}.
-
-An intersection type such as {\tt Int $\land$ Bool}, {\tt List a $\land$ Singleton a} or {\tt Int -> Int $\land$ Bool -> Bool} state the a value has both the type on the left and the type on the right.
-Compared to union types these are more interesting.
-The type {\tt List a $\land$ Singleton a} can still be simplified, but now to {\tt Singleton a}.
-A value cannot be of both type {\tt Int} and {\tt Bool}, so we can simply the type of such a value to the bottom or empty type.
-There do exist functions that are of both type {\tt Int -> Int} and {\tt Bool -> Bool}, for example {\tt id}.
-In fact, intersection types can in the limit be viewed as a form of parametric polymorphism.
-
-\subsection{Constructors and Pattern Matching}
-
-To infer refinement types more accurate types need to be given to constructors.
-This is done by using a restricted\footnote{We only take intersections of subtypes of the same data type.} form of intersection types.
-For example, the {\tt Cons} constructor is given the type:
-
-\begin{verbatim}
-Cons :: a -> Nil       a -> Singleton a
-     /\ a -> Singleton a -> List      a
-     /\ a -> List      a -> List      a
-\end{verbatim}
-
-This type can be derived automatically from the {\tt rectype} declarations.
-Types of functions can also be inferred automatically, although there may be a loss of precision when higher-order functions or polymorphism are involved.
-
-Type inference for intersection types is undecidable in general \cite{Pierce91programmingwith,Reynolds96designof}.
-Because the lattice of types is finite the algorithm is effectively able to do an exhaustive search over all possible types, however.
-Higher-order functions can cause the size of the type to blow up exponentially, each pairing of their range and domain needs to be included in the intersection type.
-
-A case-statement
-
-\begin{verbatim}
-case G of
-  Nil       -> E1
-  Cons a as -> E2
-\end{verbatim}
-
-can be seen as a call to a higher-order function
-
-\begin{verbatim}
-case_List G E1 (\a as -> E2)
-\end{verbatim}
-
-where {\tt case\_list} has the refinement type
-
-\begin{verbatim}
-case_List :: forall a. forallR (r1 :: a). forallR (r2 :: a).
-       Nil a       -> r1 -> (List a -> r2) -> r1
-    /\ Singleton a -> r1 -> (List a -> r2) -> r2
-    /\ List a      -> r1 -> (List a -> r2) -> (r1 \/ r2)
-\end{verbatim}
-
-\chapter{Related Work}\label{secrelatedwork}
-
-% table comparing the approaches
-% * laziness
-% * polymorphism
-% * annotations (required / supported)
-% * decidability
-% * errors generated
-
-\section{Neil Mitchell's \emph{Catch}}
-
-Neil Mitchel's \emph{Catch} (``CAse Totality CHecker'') \cite{Mitchell} is a tool specialized in finding potential pattern match failures in Haskell programs.
-
-\subsection{Overview}
-
-Catch works by calculates preconditions on functions.
-The preconditions are given in a constraint language.
-By varying the constraint language trade-offs between performance and precision can be made.
-If the calculated precondition for \texttt{main} is True, the program does not crash.
-
-Preconditions are calculated iteratively.
-Precondition start out as \texttt{True}, except for \texttt{error}, whose precondition is \texttt{False}.\footnote{Note that {\tt undefined} is defined in terms of {\tt error}.}
-
-\paragraph{Example} Following the example given by Mitchell, given the function:
-\begin{verbatim}
-safeTail xs = case null xs of
-                True  -> []
-                False -> tail xs
-\end{verbatim}
-The computed precondition will be:
-\[ \textrm{Precondition}(\texttt{null xs}) \land (\texttt{null xs} \in \{\texttt{True}\} \lor \textrm{Precondition}(\texttt{tail xs})), \]
-stating the necessary precondition for \texttt{null xs} not to crash mush be fulfilled and either \texttt{null xs} must evaluate to \texttt{True} or the necessary precondition for \texttt{tail xs} not to crash must be fulfilled.
-
-As \texttt{null xs} cannot crash we find its precondition to be simply True and we can deduce the precondition for \texttt{tail xs} to be $\texttt{xs}\in\{(:)\}$.
-Substituting these into our precondition we find
-\[ \textrm{True} \land (\texttt{null xs} \in \{\texttt{True}\} \lor \texttt{xs}\in\{(:)\}), \]
-which simplifies to
-\[ \texttt{null xs} \in \{\texttt{True}\} \lor \texttt{xs}\in\{(:)\}. \]
-The subexpression  $\texttt{null xs} \in \{\texttt{True}\}$ forms a postcondition on \texttt{null xs}.
-From this postcondition Catch computes the required precondition that satisfies it, which in this particular case should turn out to be $\texttt{xs}\in\{[]\}$.
-Substituting, we find
-\[ \texttt{xs}\in\{[]\} \lor \texttt{xs}\in\{(:)\}, \]
-which, using our knowledge of the list data type, turns out the be a tautology, i.e. the (trivial) precondition for \texttt{safeTail} is True.
-
-\subsection{Constraint systems}
-The example in the previous section used a simple constraint language, specifying to which set of head constructors an expression should evaluate.
-Mitchell developed two more expressive constraint systems: regular expression constraints and multipattern constraints.
-
-\paragraph{Regular expression constraints} Constraints are formed by a regular expression over an alphabet of a number of ad-hoc \emph{selectors} (e.g.
-\texttt{hd} and \texttt{tl} for lists.) A precondition for \texttt{map head xs} using regular expression constraints reads $\texttt{xs} \in (\texttt{tl}^{*} \cdot \texttt{hd} \leadsto \{(:)\})$.
-It should be interpreted as ``after applying zero or more \texttt{tl}s to \texttt{xs} and then applying a \texttt{hd} we should find a \texttt{(:)} constructor.''
-
-Accoriding to Mitchell regular expression constraints tend to scale badly as the program increase in size, although he could not identify more specific condition under which this problem manifests.\footnote{Personal communication}
-
-\paragraph{Multipattern constraints} Multipatterns are of the form $\alpha \star \rho$, where $\alpha$ gives the set of constructors that are valid at the head of the value, while $\rho$ gives the set of constructors that can appear at the recursive positions (if any) of the constructor at the head of the value.
-The elements of these sets can again be multipatterns.
-To specify that \texttt{xs} in \texttt{map head xs} should be a list of non-empty lists, we use the multipattern:
-\begin{align*}
-\{[]&, (:) (\{(:)\ \texttt{Any} \} \star \{ [], (:)\ \texttt{Any}\})\} \\
-&\star \\
-\{[]&, (:) (\{(:)\ \texttt{Any} \} \star \{ [], (:)\ \texttt{Any}\})\}
-\end{align*}
-\texttt{Any} is a wildcard that matches any constructor.
-
-\subsection{Discussion}
-% should have been mentioned earlier?
-The analysis of Catch works on a first-order language. The input program needs to be defunctionalized before it can be analyzed. The defunctionalization algorithm employed by Catch is not complete.
-
-Calculated preconditions are unnecessarily restrictive in the presence of laziness (Mitchell, p. 142).
-
-\section{Static Contract Checking}
-\subsection{Overview}
-Dana Xu's work on static contract checking for Haskell \cite{Xu09staticcontract} (including the ESC/Haskell system \cite{Xu:2006:ESC:1159842.1159849}) allows checking of arbitrary programmer supplied pre- and postconditions on functions and is able to detect pattern match failures in the process.
-
-Programmers define pre- and postconditions in the form of a contract (a refinement type). An appropriate contract for the \texttt{head} function defined above might be:
-\begin{verbatim}
-{-# CONTRACT head :: {s | not (null s)) -> {x | True} #-}
-\end{verbatim}
-This contract states that if head is given a list \texttt{s} for which the predicate \texttt{not (null s)} holds, i.e.
-it is a non-empty list, the function will not crash\footnote{The system only guarantees partial correctness, so the term might still diverge.}, indicated by the trivial postcondition \texttt{True} on the return value.
-
-\begin{figure}
-\centering
-\begin{tabular}{lclr}
-t &$::=$& $\{ x | p \}$     & Predicate          \\
-  & $|$ & $x : t_1 \to t_2$ & Dependent function \\
-  & $|$ & $(t_1, t_2)$      & Constructor         \\
-  & $|$ & $\mathtt{Any}$    & Polymorphic ``Any''
-\end{tabular}
-\caption{Syntax of Xu's contracts}
-\label{xu-contract-syntax}
-\end{figure}
-
-Contracts can be constructed from predicates, a dependent function space constructor, arbitrary constructors and a polymorphic \texttt{Any} contract that is always satisfied, including by functions which crash (Fig.
-\ref{xu-contract-syntax}).
-Any \texttt{Bool}-valued Haskell expression can be used as a predicate.
-
-Dependent functions are helpful when declaring a contract, e.g. for the \texttt{reverse} function:
-\begin{verbatim}
-{-# CONTRACT reverse :: {xs | True} -> {rs | length xs == length rs} #-}
-\end{verbatim}
-The predicate on the return value depends on the the input. Constructors and the \texttt{Any} contract are useful when declaring a contract for \texttt{fst}:
-\begin{verbatim}
-{-# CONTRACT fst :: (Ok, Any) -> Ok #-}
-\end{verbatim}
-Here \texttt{Ok} is used as a shorthand for \texttt{\{$x$ | True\}}.
-As \texttt{fst} discards the value on the right in the tuple we should not care if it is an expression that crashes, so we can not use \texttt{Ok}.
-
-\subsection{Contract checking}
-Haskell's syntax is extended with two \emph{exception values} -- \texttt{BAD} and \texttt{UNR} -- which are only used internally by the checker.
-\texttt{BAD} signifies an expression which crashes and \texttt{UNR} an expression which is either unreachable or diverges.
-
-In the program that is going to be verified all crashing expressions (including \texttt{error} and \texttt{undefined}), as well as missing patterns in case-expressions are replaced by \texttt{BAD} exceptions.
-
-The verification of the contracts is based on a translation of the contracts into Findler--Felleisen wrappers \cite{Findler:2002:CHF:581478.581484}, which will cause a function to crash (using \texttt{BAD}) if it disobeys its contract or diverge if it called in a way that is not permitted.
-While technically interesting it is not directly relevant to the detection of pattern match failures so we shall not discuss it in more depth.
-
-The actual verification process continues by \emph{symbolic evaluation}, basically applying various simplifications to the resulting program including $\beta$-reductions.
-Any code deemed to be unreachable is pruned from the program.
-
-The presence of recursive functions in the contracts might cause the symbolic evaluation to diverge if care is not taken to limit the number of evaluation steps.
-By setting an upper bound on the number of simplification steps we lose accuracy, but gain decidability of the verification process.
-
-Arithmetical expressions in case-expressions (e.g.
-\texttt{case x*x >= 0 of \{ True -> ...; False -> BAD \}}) cannot be handled directly by the symbolic evaluator.
-These are collected in a table and send to an external constraint or SMT solver.
-If the solver determines these expressions are inconsistent the code is unreachable and can be pruned.
-
-After the symbolic evaluation has terminated the checker will tell the programmer if the program is ``definitely satisfies the contract'' (if no \texttt{BAD} exceptions remain anywhere in the program), ``definitely does not satisfy the contract'' or ``don't know.''
-
-\subsection{Discussion}
-Unlike our envisioned system, Xu's static contract checking requires programmer supplied contracts on nearly all functions (contracts on trivial functions can be omitted and are handled by inlining their definition when called.) Compared to Mitchell's Catch it can handle higher-order functions natively and has a more (too?) expressive contract language.
-
-\section{Dependent ML}
-DML($\mathcal{L}$) is an extension of the ML programming language which enriches ML's type system with a limited form of dependent types over a (decidable) constraint domain (or index language) $\mathcal{L}$ \cite{DML-JFP07}.
-
-Xi's initial example -- recast in a more Haskell-like syntax -- declares a {\tt List} data type dependent on an integer corresponding to the length of the list and a function to concatenate two such lists:
-\begin{verbatim}
-1  type Nat = {a :: Int | a >= 0}
-2
-3  data List<Int> a = Nil<0>
-4                   | {n :: Nat} Cons<n+1> a (List<n> a)
-5
-6  (++) :: {m :: Nat} {n :: Nat} List<m> a -> List<n> a -> List<m+n> a
-7  (++) Nil         ys = ys
-8  (++) (Cons x xs) ys = Cons x (append xs ys)
-\end{verbatim}
-
-For many functions producing a list, the exact length might not be derived by such a trivial calculation ({\tt m+n}) on the lengths of the input lists.
-A prominent example is the {\tt filter} function:
-\begin{verbatim}
-1  filter :: (a -> Bool) -> {m :: Nat} List<m> a -> [n :: Nat | n <= m] List<n> a
-2  filter p Nil         = Nil
-3  filter p (Cons x xs) | p x       = Cons x (filter p xs)
-4                       | otherwise =         filter p xs
-\end{verbatim}
-here we only know that the resulting list is equal or shorter in length than the input list ({\tt n <= m}).
-This is expressed as the dependent sum {\tt [n :: Nat | n <= m]}.
-In a full-fledged dependently-typed language we would also return a proof object stating the predicate {\tt p} holds for all the elements in the output list, but this is beyond the expressive power of DML.
-
-What we gain is a relief from the need to provide proofs for trivial arithmetical (in)equalities.
-Imagine a {\tt zip} function which requires the two list to be zipped together to be of equal length.
-When calling {\tt zip (xs ++ ys) (ys ++ xs)} this preconditions seems to be intuitively satisfied.
-From the point-of-view of the compiler the first list has length $m+n$ and the second lists $n+m$, however.
-In most dependently-typed languages we will now have to invoke a lemma or tactic proving the commutativity of addition.
-DML can simply send the constraint $m + n = n + n$ to an ILP solver and conclude the constraint is satisfiable.
-
-\subsection{Discussion}
-Compared to Xu's static contract checking, Xi's DML constrains the constraint system to a decidable theory.
-The constraints and indices -- such as {\tt m+n} and {\tt n~<=~m} -- may superficially look like ordinary Haskell expression, but in fact belong to a much smaller index language.
-While type checking is decidable, type inference is not and as a result DML still requires type annotations.
-
-To be applicable to a pattern match analysis we must try to infer a type like:
-\begin{verbatim}
-head :: {n :: Int | n >= 1} [a]<n> -> a
-\end{verbatim}
-for the {\tt head} function. We also have to implicitly extend the list data type with an index representing its length.
-
-This seems significantly more challenging than inferring the Catch-like type
-\begin{alltt}
-head :: \{xs :: [a] | xs \(\in\) (:) \} -> a
-\end{alltt}
-The correspondence between the the type and the pattern matching happening in the definition of {\tt head} is much more direct.
-
-%\begin{small}
-%There is an indirect correspondence, however.
-The value of the index of the list is given by the {\tt length} function and by analyzing the pattern matching happening in that function we could derive that $\{n :: \mathtt{Int} | n >= 1\}\ \left[a\right]\left<n\right> \simeq \{xs :: [a] | xs \in (:) \}$.
-%
-%Instead of adding the an index of type {\tt Int} to {\tt List} to create {\tt List<Int>}, we add the abstraction {\tt length :: List -> Int}.
-This generalizes to arbitrary data types $F$ and constraint systems over $\tau$, by adding an abstraction $F \to \tau$ to create $F\left<\tau\right>$.
-(But can we use arbitrary functions of type $F\to\tau$, or does {\tt lenght} have some particularly nice properties?
-Certainly if viewed as a catamorphism (e.g.
-$(0, (\lambda n.
-1+n))$) it should consist of only expressions in the constraint language $\mathcal{L}$ -- they correspond exactly to the annotations on {\tt Nil<0>} and {\tt \{n :: Int\} Cons<n+1>} -- but is this sufficient?
-What if we look at a {\tt Tree}?
-Adding a property {\tt minDepth :: Tree -> Int} would guarantee we can pattern match on a certain number of {\tt Branch}es before encountering a {\tt Leaf}, but what does {\tt maxDepth :: Tree -> Int} buy us?
-More importantly, while we might be able to infer an appropriate type for {\tt head}, inferring the type given to {\tt (++)} and {\tt filter} above still seems out of reach.)
-%\end{small}
-
-\section{Compiling Pattern Matching}
-Case-statements in toy languages often have a very simple \emph{decision tree} semantics.
-Case-statements in Haskell have a more complex \emph{backtracking automaton} semantics.
-There is a body of work on compiling the latter into the former.
-Maranget \cite{DBLP:journals/jfp/Maranget07} gives an algorithm for determining whether a case-statement is exhaustive and whether all patterns are useful.
-As the analysis does not consider any dataflow information it is much too imprecise for our purpose.
-It does turn out that the analysis closely follows, but can at places be simplified with respect to, the manner in which pattern matching is compiled.
-This indicates might also be able to follow such an approach when analyzing case-statements in our pattern-match analysis.
-
-\chapter{Approach}\label{secapproach}
-
-% type and effect system
-% refinement types / subtyping / widening (compare to multi-patterns)
-% how much precision do we need?
-
-\section{Incremental development}
-We will incrementally develop our analysis for the various constructs in the Haskell language.
-
-\paragraph{Variables, constants, abstraction, application, if-statements and simple let-bindings} We will start out by implementing the basic constructs of the simply typed $\lambda$-calculus.
-This includes simple let-bindings, i.e.
-non-recursive let-bindings in which no pattern matching or generalization occurs.
-
-At this stage we will only allow booleans and integers as constants.
-Due to their primitive and infinity nature integers will need to be treated specially by the analysis in any case.
-Booleans and if-statements can be handles in a more general fashion during later stages.
-\paragraph{Recursion} The next elementary construct is recursion.
-This will require us to implement a form of \emph{widening} for the integers and lists to ensure termination of the analysis.
-As the exact form of widening used can have a big impact on the precision of the analysis we do not want to commit to a specific lattice here.
-Various widening strategies should be easily pluggable into the source code and perhaps eventually be able to be specified at a higher-level by the user of the tool.
-\paragraph{Simple case-statements} The most critical part of the analysis will be how it handles case-statements.
-In this state we will add simple case-statements (i.e., non-nested, non-backtracking, guardless case-statements) and primitive list constants, our first recursive data type.
-\paragraph{Complex case-statements and guards} To be able to handle the full Haskell 98 language our analysis must be able to deal with the complex semantics of case-statements.
-This includes nested-patterns, guards and backtracking.
-We would like to be able to analyze any boolean-valued expression that is used as a guard.
-\paragraph{Algebraic data types} Next we extend the analysis to work on arbitrary programmer-defined data types instead of only on our primitive list data type.
-This will mostly involve generalizing out analysis to work on with a representation of arbitrary algebraic data types.
-Data types containing fields with function types may also prove interesting.
-\paragraph{Polymorphism} Polymorphism can generally be added to an analysis as a relatively orthogonal feature.
-However, there might be some interesting interactions with our refinement and intersection types here.
-\paragraph{Full let-bindings} We extend the analysis for let-bindings to handle (mutual) recursion and pattern matching.
-\paragraph{Type classes} An interesting feature of Haskell's type system are type classes. We currently do not foresee any specific difficulties with respect to be able to handle them in out analysis.
-\paragraph{Syntactic sugar} As the final stage we would add constructs which are only syntactic sugar and therefore would not pose any significant problem to our analysis.
-This includes for example list-comprehensions and do-notation.
-
-\section{Refinement types}
-
-Compared to Freeman and Pfenning's approach to refinement types we will initially not rely on programmer supplied, but instead build our lattice of types from finite unrollings of data types.
-We hope this makes calculating the lattice easier.
-
-Freeman's inference algorithm for refinement type assign a single, albeit complex, type to each binding.
-This makes it a monovariant analysis.
-We will develop a polyvariant analysis, making it sensitive to the sites where a binding is used.
-We hope this will improve the performance of the analysis.
-
-\section{Parsing}
-
-We will develop our tool in the Haskell programming language.
-A number of libraries for parsing input programs are available.
-The most obvious choice seem to be the {\tt haskell-src-exts} package, which is also used by the HLint tool.
-If this proves to be a problematic choice, other possibilities are the parser from GHC or UHC.
-
-\chapter{Deliverables}\label{secdeliverables}
-
-There will be two deliverables of this project:
-\begin{enumerate}
-\item A \emph{tool} implementing the pattern match analysis for Haskell 98. This tool is intended to be practically usable and not a mere research prototype.
-\item A \emph{thesis} describing the theory underlying the tool and documenting the implementation choices made in the tool.
-\end{enumerate}
-
-\chapter{Evaluation}\label{secevaluation}
-
-The evaluate whether we were able to answer research question we first need to evaluate whether the developed tool implements a practically usable pattern-match analysis.
-We can do so by trying it out on the examples sketched in Section \ref{secexamples}, the literature and for example the {\tt nofib} testsuite.
-
-If we can answer this question with a ``yes'', and assuming the tool indeed implements the analysis using a type and effect system and refinement types, we can also answer out research question with a ``yes''.
-
-In case of a ``no'' we will need to argument whether this is due to a failure of our implementation or because there are good theoretical reasons why such an analysis cannot be implemented.
-
-\chapter{Planning}\label{secplanning}
-
-\begin{description}
-\item[Implement framework (April)] Implement a constraint-based version of Hindley--Milney for Haskell (the underlying the system of the analysis).
-This is known work \cite{Odersky:1999:TIC:308181.308203,DBLP:journals/entcs/HageH09} and should not prove to be difficult.
-This implementation will provide a guiding framework for the analysis, highlight possible issues we might run into when implementing a constraint-based analysis for full Haskell and allow us to decide which parsing library and abstract syntax representation to use.
-\item[Implement analysis (April--July)] The largest portion of time is assigned to incrementally developing and building the pattern-match analysis. The likely order will be:
-\begin{enumerate}
-\item Variables, constants, abstraction, application, if-statements and simple let-bindings (1.5 weeks)
-\item Recursion (1.5 weeks)
-\item Simple case-statements (2 weeks)
-\item Complex case-statements and guards (2 weeks)
-\item Algebraic data types (1.5 weeks)
-\item Polymorphism (1.5 weeks)
-\item Full let-bindings (1 week)
-\item Type classes (1.5 weeks)
-\item Syntactic sugar (1.5 weeks)
-\end{enumerate}
-\item[Evaluation and thesis writing (July--August)] The final phase of this project will include an evaluation of the tool and finishing of the thesis.
-A significant potion of the thesis is expected to be written in parallel with the implementation, after a particular subtask has been implemented successfully.
-\end{description}
-
-\chapter{Future Work}
-
-Non-goals of the thesis project are:
-\begin{enumerate}
-\item A formal soundness proof of the analysis, and
-\item Supporting Haskell extensions beyond Haskell 98.
-\end{enumerate}
-Both of these would be candidates for future work.
-
-\bibliography{thesis-proposal}
-\bibliographystyle{plain}
-
-\end{document}

### 2015-09-01 deprecated/pattern-match-analysis/thesis/relatedwork.tex
pattern-match-analysis

--- 2015-09-01 deprecated/pattern-match-analysis/thesis-proposal/thesis-proposal.tex
@@
-\documentclass[a4paper]{report}
+\chapter{Related Work}\label{chaprelatedwork}
 
-\usepackage{amsmath}
-\usepackage{amssymb}
+% table comparing the approaches
+% * laziness
+% * polymorphism
+% * annotations (required / supported)
+% * decidability
+% * errors generated
 
-\usepackage{alltt}
+\section{Neil Mitchell's \emph{Catch}}
 
-\usepackage{tikz}
+Neil Mitchel's \emph{Catch} (``CAse Totality CHecker'') \cite{Mitchell} is a tool specialized in finding potential pattern match failures in Haskell programs.
 
-\begin{document}
+\subsection{Overview}
 
-\title{Thesis Proposal}
-\author{Ruud Koot}
-\maketitle
+Catch works by calculates preconditions on functions.
+The preconditions are given in a constraint language.
+By varying the constraint language trade-offs between performance and precision can be made.
+If the calculated precondition for |main| is True, the program does not crash.
 
-% Also see:
-% http://www.cs.uu.nl/wiki/bin/view/Thesis/WritingAThesisProposal
-% mail from Jur
-% Tamar's/Eric's proposal
+Preconditions are calculated iteratively. Precondition start out as |True|, except for |error|, whose precondition is |False|.\footnote{Note that |undefined| is defined in terms of |error|.}
 
-\chapter*{Introduction}
+\paragraph{Example} Following the example given by Mitchell, given the function:
+\begin{code}
+safeTail xs =  case null xs of
+                  True   -> []
+                  False  -> tail xs
+\end{code}
+the computed precondition will be:
+\[ \textrm{Precondition}(|null xs|) \land (|null xs| \in \{|True|\} \lor \textrm{Precondition}(|tail xs|)), \]
+stating the necessary precondition for |null xs| not to crash must be fulfilled and either |null xs| must evaluate to |True| or the necessary precondition for |tail xs| not to crash must be fulfilled.
 
-This document is a proposal for a thesis on answering the research question:
-\begin{quote}
-Can we use a type and effect system in combination with refinement types to develop a pattern-match analysis for a non-strict higher-order functional language that is both performant and precise enough to be of practical use?
-\end{quote}
+As |null xs| cannot crash we find its precondition to be simply True and we can deduce the precondition for |tail xs| to be $|xs|\in\{(:)\}$. Substituting these into our precondition we find
+\[ \textrm{True} \land (|null xs| \in \{|True|\} \lor |xs|\in\{(:)\}), \]
+which simplifies to
+\[ |null xs| \in \{|True|\} \lor |xs|\in\{(:)\}. \]
+The subexpression  $|null xs| \in \{|True|\}$ forms a postcondition on |null xs|.
+From this postcondition Catch computes the required precondition that satisfies it, which in this particular case should turn out to be $|xs|\in\{[]\}$.
+Substituting, we find
+\[ |xs|\in\{[]\} \lor |xs|\in\{(:)\}, \]
+which, using our knowledge of the list data type, turns out the be a tautology, i.e. the (trivial) precondition for |safeTail| is True.
 
-In Chapter \ref{secmotivation} we present a number of examples to demonstrate why this is an interesting problem.
-In Chapter \ref{seccontext} we give a short introduction to the relevant concepts of the research question: higher-order functional languages, type and effect systems and refinement types.
-In Chapter \ref{secrelatedwork} we present work related to our research question and discuss which aspects of that work are relevant to or different from our proposed system.
-In Chapter \ref{secapproach}, \ref{secdeliverables} and \ref{secevaluation} we will discuss our envisioned approach to answering the research question, the set of deliverables and how we will evaluate whether the research question has been answered successfully or not.
+\subsection{Constraint Systems}
+The example in the previous section used a simple constraint language, specifying to which set of head constructors an expression should evaluate.
+Mitchell developed two more expressive constraint systems: regular expression constraints and multipattern constraints.
 
-\chapter{Motivation}\label{secmotivation}
+\paragraph{Regular expression constraints} Constraints are formed by a regular expression over an alphabet of a number of ad-hoc \emph{selectors} (e.g.
+|hd| and |tl| for lists.) A precondition for |map head xs| using regular expression constraints reads $|xs| \in (|tl|^{*} \cdot |hd| \leadsto \{(:)\})$.
+It should be interpreted as ``after applying zero or more |tl|s to |xs| and then applying a |hd| we should find a |(:)| constructor.''
 
-In 1978, Robin Milner \cite{Milner78atheory} famously wrote (and proved) that:
-\begin{quotation}
-Well-typed programs cannot ``go wrong''. 
-\end{quotation}
-This observation is sometimes
-%\footnote{Especially by professors wishing to impress their audience of undergraduate students with the power of types.}
-optimistically overstated as ``if your Haskell program type checks and compiles it will work.'' Even if we ignore logic errors, such a strong interpretation of Milner's statement does not hold, as anyone who has seen the dreaded:
+According to Mitchell regular expression constraints tend to scale badly as the program increase in size, although he could not identify more specific condition under which this problem manifests.\footnote{Personal communication}
+
+\paragraph{Multipattern constraints} Multipatterns are of the form $\alpha \star \rho$, where $\alpha$ gives the set of constructors that are valid at the head of the value, while $\rho$ gives the set of constructors that can appear at the recursive positions (if any) of the constructor at the head of the value.
+The elements of these sets can again be multipatterns.
+To specify that |xs| in |map head xs| should be a list of non-empty lists, we use the multipattern:
+\begin{align*}
+\{[]&, (:) (\{(:)\ |Any| \} \star \{ [], (:)\ |Any|\})\} \\
+&\star \\
+\{[]&, (:) (\{(:)\ |Any| \} \star \{ [], (:)\ |Any|\})\}
+\end{align*}
+|Any| is a wildcard that matches any constructor.
+
+\subsection{Discussion}
+% should have been mentioned earlier?
+The analysis of Catch works on a first-order language. The input program needs to be defunctionalized before it can be analyzed. The defunctionalization algorithm employed by Catch is not complete.
+
+Calculated preconditions are unnecessarily restrictive in the presence of laziness (Mitchell, p. 142).
+
+\section{Static Contract Checking}
+\subsection{Overview}
+Dana Xu's work on static contract checking for Haskell \cite{Xu09staticcontract} (including the ESC/Haskell system \cite{Xu:2006:ESC:1159842.1159849}) allows checking of arbitrary programmer supplied pre- and postconditions on functions and is able to detect pattern match failures in the process.
+
+Programmers define pre- and postconditions in the form of a contract (a refinement type). An appropriate contract for the |head| function defined above might be:
 \begin{verbatim}
-*** Exception: PM.hs:1:1-13: Non-exhaustive patterns in function f
+{-# CONTRACT head :: {s | not (null s)) -> {x | True} #-}
 \end{verbatim}
-can attest to. To correctly interpret Milner we need to distinguish between three kinds of ``wrongness'':
-\begin{description}
-\item[Getting stuck] If a program tries to evaluate a non-sensical expression -- such as {\tt 3 + true} -- it cannot possibly make any further progress and is said to be ``stuck''.
-This is the ``go wrong'' Milner referred to.
-He proved that a sound type system can statically guarantee such expressions will never occur or need to be evaluated at run-time.
-\item[Diverging] If we have a function definition {\tt f = f}, the evaluation of {\tt f} $\leadsto$ {\tt f} $\leadsto ...$ will fail to terminate.
-We might be making progress in a technical sense, but it will have no useful observable result.
-\item[Undefinedness] Another source of ``wrongness'' are partial functions: functions which are not defined on all the elements of their domain.
-Prime examples are case-statements with missing constructors and functions defined by pattern matching but which do not cover all possible patterns.
-\end{description}
+This contract states that if head is given a list |s| for which the predicate |not (null s)| holds, i.e.
+it is a non-empty list, the function will not crash\footnote{The system only guarantees partial correctness, so the term might still diverge.}, indicated by the trivial postcondition |True| on the return value.
 
-Unlike Milner, who spoke about the first kind of wrongness and the work on termination checking, which concerns itself with the second, we shall focus on the third: a pattern-match analysis which determines if functions are only invoked on values for which they are defined.
+\begin{figure}
+\centering
+\begin{tabular}{lclr}
+t &$::=$& $\{ x || p \}$     & Predicate          \\
+  & $||$ & $x : t_1 \to t_2$ & Dependent function  \\
+  & $||$ & $(t_1, t_2)$      & Constructor         \\
+  & $||$ & $Any$             & Polymorphic ``Any''
+\end{tabular}
+\caption{Syntax of Xu's contracts}
+\label{xu-contract-syntax}
+\end{figure}
 
-% split error and partial definedness?
-% loop and crash
+Contracts can be constructed from predicates, a dependent function space constructor, arbitrary constructors and a polymorphic |Any| contract that is always satisfied, including by functions which crash (Fig.
+\ref{xu-contract-syntax}).
+Any |Bool|-valued Haskell expression can be used as a predicate.
 
-\section{Examples}\label{secexamples} % test-driven development
+Dependent functions are helpful when declaring a contract, e.g. for the |reverse| function:
+\begin{verbatim}
+{-# CONTRACT reverse
+      :: {xs | True} -> {rs | length xs == length rs} #-}
+\end{verbatim}
+The predicate on the return value depends on the input. Constructors and the |Any| contract are useful when declaring a contract for |fst|:
+\begin{verbatim}
+{-# CONTRACT fst :: (Ok, Any) -> Ok #-}
+\end{verbatim}
+Here |Ok| is used as a shorthand for |{x || True}|. As |fst| discards the value on the right in the tuple we should not care if it is an expression that crashes, so we can not use |Ok|.
 
-\paragraph{Partial functions} Haskell programmers often work with partial functions, the most common one possibly being {\tt head}:
+\subsection{Contract Checking}
+Haskell's syntax is extended with two \emph{exception values}---|BAD| and |UNR|---which are only used internally by the checker.
+|BAD| signifies an expression which crashes and |UNR| an expression which is either unreachable or diverges.
 
+In the program that is going to be verified all crashing expressions (including |error| and |undefined|), as well as missing patterns in case-expressions are replaced by |BAD| exceptions.
+
+The verification of the contracts is based on a translation of the contracts into Findler--Felleisen wrappers \cite{Findler:2002:CHF:581478.581484}, which will cause a function to crash (using |BAD|) if it disobeys its contract or diverge if it called in a way that is not permitted.
+While technically interesting it is not directly relevant to the detection of pattern match failures so we shall not discuss it in more depth.
+
+The actual verification process continues by \emph{symbolic evaluation}, basically applying various simplifications to the resulting program including $\beta$-reductions.
+Any code deemed to be unreachable is pruned from the program.
+
+The presence of recursive functions in the contracts might cause the symbolic evaluation to diverge if care is not taken to limit the number of evaluation steps.
+By setting an upper bound on the number of simplification steps we lose accuracy, but gain decidability of the verification process.
+
+Arithmetical expressions in case-expressions (e.g.
+|case x * x >= 0 of { True -> ...; False -> BAD }|) cannot be handled directly by the symbolic evaluator.
+These are collected in a table and send to an external constraint or SMT solver.
+If the solver determines these expressions are inconsistent the code is unreachable and can be pruned.
+
+After the symbolic evaluation has terminated the checker will tell the programmer if the program is ``definitely satisfies the contract'' (if no |BAD| exceptions remain anywhere in the program), ``definitely does not satisfy the contract'' or ``don't know.''
+
+\subsection{Discussion}
+Unlike our envisioned system, Xu's static contract checking requires programmer supplied contracts on nearly all functions (contracts on trivial functions can be omitted and are handled by inlining their definition when called.) Compared to Mitchell's Catch it can handle higher-order functions natively and has a more (too?) expressive contract language.
+
+\section{Dependent ML}
+DML($\mathcal{L}$) is an extension of the ML programming language which enriches ML's type system with a limited form of dependent types over a (decidable) constraint domain (or index language) $\mathcal{L}$ \cite{DML-JFP07}.
+
+Xi's initial example---recast in a more Haskell-like syntax---declares a |List| data type dependent on an integer corresponding to the length of the list and a function to concatenate two such lists:
 \begin{verbatim}
-1  main = let xs = if length "foo" > 5 then [1,2,3] else []
-2          in head xs
-3
-4  head (x : xs) = x
+type Nat = {a :: Int | a >= 0}
+
+data List<Int> a = Nil<0>
+                 | {n :: Nat} Cons<n+1> a (List<n> a)
+
+(++) :: {m :: Nat} {n :: Nat} List<m> a -> List<n> a -> List<m+n> a
+(++) Nil         ys = ys
+(++) (Cons x xs) ys = Cons x (append xs ys)
 \end{verbatim}
 
-This program is guaranteed to crash at run-time, so we would like to be warned beforehand by the compiler or a tool:
+For many functions producing a list, the exact length might not be derived by such a trivial calculation (|m+n|) on the lengths of the input lists. A prominent example is the |filter| function:
+\begin{verbatim}
+filter :: (a -> Bool) -> {m :: Nat} List<m> a
+                      -> [n :: Nat | n <= m] List<n> a
+filter p Nil         = Nil
+filter p (Cons x xs) | p x       = Cons x (filter p xs)
+                     | otherwise =         filter p xs
+\end{verbatim}
+here we only know that the resulting list is equal or shorter in length than the input list (|n <= m|).
+This is expressed as the dependent sum |[n :: Nat || n <= m]|.
+In a full-fledged dependently-typed language we would also return a proof object stating the predicate |p| holds for all the elements in the output list, but this is beyond the expressive power of DML.
 
+What we gain is a relief from the need to provide proofs for trivial arithmetical (in)equalities.
+Imagine a |zip| function which requires the two list to be zipped together to be of equal length.
+When calling |zip (xs ++ ys) (ys ++ xs)| this precondition seems to be intuitively satisfied.
+From the point-of-view of the compiler the first list has length $m+n$ and the second lists $n+m$, however.
+In most dependently-typed languages we will now have to invoke a lemma or tactic proving the commutativity of addition.
+DML can simply send the constraint $m + n = n + m$ to an ILP solver and conclude the constraint is satisfiable.
+
+\subsection{Discussion}
+Compared to Xu's static contract checking, Xi's DML constrains the constraint system to a decidable theory.
+The constraints and indices---such as |m+n| and |n <= m|---may superficially look like ordinary Haskell expression, but in fact belong to a much smaller index language.
+While type checking is decidable, type inference is not and as a result DML still requires type annotations.
+
+To be applicable to a pattern match analysis we must try to infer a type like:
 \begin{verbatim}
-On line 2 you applied the function "head" to the empty list "xs".
-The function "head" expects a non-empty list as its first argument.
+head :: {n :: Int || n >= 1} [a]<n> -> a
 \end{verbatim}
+for the |head| function. We also have to implicitly extend the list data type with an index representing its length.
 
-If the guard of the if-statement had read {\tt length "foo" < 5} the program would have run without crashing and we would like the compiler or tool not to warn us spuriously.
-In case it is not possible to determine statically whether or not a program will crash, a warning should still be raised.
+This seems significantly more challenging than inferring the Catch-like type
+\begin{alltt}
+head :: \{xs :: [a] || xs \(\in\) (:) \} -> a
+\end{alltt}
+The correspondence between the type and the pattern matching happening in the definition of |head| is much more direct.
 
-\paragraph{Compiler construction}
-Compilers work with large and complex data types to represent the abstract syntax tree.
-These data structures must be able to represent all syntactic constructs the parser is able to recognize.
-This results in an abstract syntax tree that is unnecessarily complex and too cumbersome for the later stages of the compiler -- such as the optimizer -- to work with.
-This problem is resolved by \emph{desugaring} the original abstract syntax tree into a simpler -- but semantically equivalent --  abstract syntax tree than does not use all of the constructors available in the original abstract syntax tree.
-
-The compiler writer now has a choice between two different options: either write a desugaring stage {\tt desugar :: ComplexAST -> SimpleAST} -- duplicating most of the data type representing and functions operating on the abstract syntax tree -- or take the easy route {\tt desugar :: AST -> AST} and assume certain constructors will no longer be present in the abstract syntax tree at stages of the compiler executed after the desugaring step.
-The former has all the usual downsides of code duplication -- such as having to manually keep the two or more data types synchronized -- while that latter forgoes many of the advantages of strong typing and type safety: if the compiler pipeline is restructured and one of the stages that was originally assumed to run only after the desugaring suddenly runs before that point the error might only be detected at run-time by a pattern match failure.
-A pattern match analysis should be able to detect such errors statically.
-
-\paragraph{Maintaining invariants}
-Many algorithms and data structures maintain invariants that cannot be easily encoded into their type.
-These invariants often ensure that certain incomplete case-statements are guaranteed not to cause a pattern match failure.
-An example is the {\tt risers} function from \cite{DBLP:conf/sfp/MitchellR05}, calculating monotonically increasing segments of a list (e.g., {\tt risers [1,3,5,1,2]} $\leadsto$ {\tt [[1,3,5],[1,2]]}):
-
-\begin{verbatim}
-1  risers :: Ord a => [a] -> [[a]]
-2  risers []         = []
-3  risers [x]        = [[x]]
-4  risers (x1:x2:xs) = let (s:ss) = risers (x2:xs)
-5                       in if x <= y then (x:s):ss else [x]:(s:ss)
-\end{verbatim}
-
-The let-binding in the third alternative of {\tt risers} expects the recursive call to return a non-empty list.
-A naive analysis might raise a warning here.
-If we think a bit longer, however, we see that we also pass the recursive call to {\tt risers} a non-empty list.
-This means we will end up in either the second or third alternative in the recursive call.
-Both the second alternative and both branches of the if-statement in the third alternative result in a non-empty list, satisfying the assumption we made earlier.
-
-Another example might be a collection of mathematical operations working on bitstrings (integers encoded as lists of binary digits without leading zeroes):
-\begin{verbatim}
-1  type Bitstring = [Int]
-2
-3  add :: Bitstring -> Bitstring -> Bitstring
-4  add [] y = y
-5  add x [] = x
-6  add (0:x) (0:y) = 0 : add x y
-7  add (0:x) (1:y) = 1 : add x y
-8  add (1:x) (0:y) = 1 : add x y
-9  add (1:x) (1:y) = 0 : add (add [1] x) y
-\end{verbatim}
-
-The patterns in {\tt add} are far from complete, but maintain the invariant if passed arguments that satisfy the invariant.
-So if we are careful to only pass valid bitstrings into a complex mathematical expression of bitstring-operations it will result in a valid bitstring without crashing due to a pattern match failure.
-
-% Examples from Refinement Types for ML:
-% * Binary encoded numbers
-% * Variable substitutions in Boolean expressions
-
-\chapter{Context}\label{seccontext}
-
-% halting problem
-% well-typed programs can't go wrong
-% smaller types log form presentation
-
-\section{Haskell}
-
-Haskell is statically-typed functional programming language with non-strict evaluation semantics \cite{Haskell98}.
-As a functional language it has first-class and higher-order functions and features a rich type system supporting parametric polymorphism and type classes.
-Programmers can define custom types in the form of algebraic data types and write functions over them using pattern matching.
-
-Like most functional languages, Haskell can be easily translated into a typed $\lambda$-calculus (System $F_C$).
-From the point-of-view of the programmer it offers a wealth of syntactic sugar over a plain $\lambda$-calculus -- such as guards and list comprehensions -- allowing programs to be expressed concisely and in a readable fashion.
-
-The following example demonstrates the the syntax of Haskell and some features mentioned previously:
-
-\begin{verbatim}
-1  data Tree a = Branch (Tree a) (Tree a)
-2              | Leaf a
-3  
-4  mapTree :: (a -> b) -> Tree a -> Tree b
-5  mapTree f (Branch t1 t2) = Branch (mapTree f t1) (mapTree f t2)
-6  mapTree f (Leaf a)       = Leaf (f a)
-7
-8  instance Functor Tree where
-9      fmap = mapTree
-\end{verbatim}
-Lines 1--2 define an algebraic data type (ADT) representing a tree.
-A tree can be constructed by either a {\tt Branch} containing a left and a right subtree or by a {\tt Leaf} containing a value.
-The ADT is parameterized by a type {\tt a} giving the type of the value stored in the {\tt Leaf}.
-
-Lines 4--5 define a higher-order function {\tt mapTree} which applies a given function {\tt f :: a -> b} to all the values stored in all the leaves of a given {\tt Tree} resulting in a new {\tt Tree}.
-As the domain and the range of {\tt f} need not coincided the type parameter of the tree changes as well.
-The body of the function is defined by pattern matching over the constructors of {\tt Tree} and performs recursion in the case of a {\tt Branch} and applies the argument {\tt f} to the field of type {\tt a} in {\tt Leaf} otherwise.
-
-Lines 8--9 declare the type constructor {\tt Tree} to be an instance of the type class {\tt Functor}, expressing it to be an endofunctor in the category of data types with its mapping on morphisms given by {\tt mapTree}.
-
-\subsection{Pattern matching}
-
-Particularly relevant to our analysis are Haskell's pattern matching abilities. We demonstrate some of them in the following function:
-
-\begin{verbatim}
-rotate (Branch b@(Branch (Leaf x) (Leaf y)) (Leaf z))
-           | z < x     = Branch (Leaf z) b
-           | z < y     = Branch (Branch (Leaf x) (Leaf z)) (Leaf y)
-           | otherwise = undefined
-\end{verbatim}
-
-The function will rotate a tree {\tt (Branch (Branch (Leaf 2) (Leaf 3)) (Leaf 1))} into {\tt Branch (Leaf 1) (Branch (Leaf 2) (Leaf 3))}.
-It accomplishes this through the use of \emph{nested pattern matching}, \emph{as-patterns} and \emph{guards}.
-
-This function only works on trees of a very specific shape.
-Feeding the function a tree of a different shape will cause a pattern match failure.
-Similarly, an error will be raised if the function reaches the {\tt otherwise} branch, containing an {\tt undefined} value.
-This would also happen if the branch had called {\tt error} or the branch and its guard were left out entirely.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Type and effect systems}
-
-Type and effect systems are approaches to program analysis suitable for typed languages \cite{Nielson:1999:PPA:555142}.
-In this formalism we take the the \emph{underlying type system} of the language and (conservatively) extend it by adding \emph{annotations} to the types (base, function and other) of the system.
-
-\begin{figure}[p]
-\[\frac{}{\Gamma \vdash_{\mathrm{UL}} c : \tau_c}\ \mathrm{[con]} \quad\quad\quad\quad
-\frac{\Gamma(x) = \tau}{\Gamma \vdash_{\mathrm{UL}} x : \tau}\ \mathrm{[var]}\]
-\[\frac{\Gamma[x \mapsto \tau_1] \vdash_{\mathrm{UL}} e : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \lambda x. e : \tau_1 \to \tau_2}\ \mathrm{[abs]}\]
-\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_2 \to \tau_1 \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} e_1\ e_2 : \tau_1}\ \mathrm{[app]}\]
-\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \tau_1 \quad \Gamma[x\mapsto \tau_1] \vdash_{\mathrm{UL}} e_2 : \tau_2}{\Gamma \vdash_{\mathrm{UL}} \mathrm{let\ } x = e_1 \mathrm{\ in\ } e_2 : \tau_2}\ \mathrm{[let]}\]
-\[\frac{\Gamma \vdash_{\mathrm{UL}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau \quad \Gamma \vdash_{\mathrm{UL}} e_2 : \tau}{\Gamma \vdash \mathrm{if\ } e_0 \mathrm{\ then\ } e_1 \mathrm{\ else \ } e_2 : \tau}\ \mathrm{[if]}\]
-\caption{Underlying type system}
-\label{underlying}
-\end{figure}
-
-\begin{figure}[p]
-\[\frac{}{\Gamma \vdash_{\mathrm{AN}} n : \mathrm{int}^{\{n\}}}\ \mathrm{[con]} \quad\quad\quad\quad
-\frac{\Gamma(x) = \widehat\tau}{\Gamma \vdash_{\mathrm{AN}} x : \widehat\tau}\ \mathrm{[var]}\]
-\[\frac{\Gamma[x \mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \lambda x. e : \widehat\tau_1 \to \widehat\tau_2}\ \mathrm{[abs]}\]
-\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_2 \to \widehat\tau_1 \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} e_1\ e_2 : \widehat\tau_1}\ \mathrm{[app]}\]
-\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \widehat\tau_1 \quad \Gamma[x\mapsto \widehat\tau_1] \vdash_{\mathrm{AN}} e_2 : \widehat\tau_2}{\Gamma \vdash_{\mathrm{AN}} \mathrm{let\ } x = e_1 \mathrm{\ in\ } e_2 : \widehat\tau_2}\ \mathrm{[let]}\]
-\[\frac{\Gamma \vdash_{\mathrm{AN}} e_1 : \mathrm{bool} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_2} \quad \Gamma \vdash_{\mathrm{AN}} e_2 : \tau^{\varphi_3}}{\Gamma \vdash_{\mathrm{AN}} \mathrm{if\ } e_0 \mathrm{\ then\ } e_1 \mathrm{\ else \ } e_2 : \tau^{\varphi_2 \cup \varphi_3}}\ \mathrm{[if]}\]
-\[\frac{
-\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi}
-}{
-\Gamma \vdash_{\mathrm{AN}} e : \tau^{\varphi \cup \varphi'}
-}\ \mathrm{[sub]}\]
-\caption{Annotated type system}
-\label{annotated}
-\end{figure}
-
-\subsection{Example}
-
-As an example we give an analysis which determines, in the form of an annotation, which values an expression can evaluate to.
-The language is the simply-typed $\lambda$-calculus with let-bindings and an if-statement.
-The underlying type systems is standard and given in Figure \ref{underlying}.
-
-The annotated type system is given in Figure \label{annotated}.
-It does little more than collecting all constants in the annotations.
-The only interesting rule is for the if-statement.
-We need to take the union over the values collected in both branches of the if-statement.
-We could have chosen to make the rule depend on the annotation of the guard expression, passing on only the values collected in the then-branch along if we knew it could only be {\tt True} or only the values collected in the else-branch if we knew it could only be {\tt False} from its annotation.
-This would make the analysis more precise.
-
-We have also added an inference rule for \emph{subeffecting}.
-This allows us to make the annotation less precise.
-This is sometimes necessary when applying a value about which we have very accurate information to a function which requires a more general argument.
-
-\subsection{Type inference}
-
-An analysis specified as an inference system does not allow us to compute the result of the analysis directly, as the premises of several of the inference rules usually require us to guess the correct type annotations for the type and annotation variables (e.g., in the presence of subeffecting).
-To overcome this problem we need a \emph{type reconstruction algorithm}.
-
-If the underlying type system is Hindley--Milner, then Algorithm W can be used to infer the types.
-Algorithm W works bottom-up and can compute types in a single pass by unifying type variables.
-This may not always be possible for the annotations in an annotated type system, however.
-Here we can use a variant of Algorithm W to gather a set of constraints on the annotation variables and solve these constraints with a worklist algorithm in a second phase.
-
-\subsection{Polyvariance}
-
-% 1) precision (poisoning)
-% 2) modularity
-
-% annotation polymorphism
-
-Consider the following program:
-\begin{verbatim}
-f = let id x = x
-     in (id 1, id 2)
-\end{verbatim}
-
-When performing the analysis given above we would assign the type $\mathtt{int} \to \mathtt{int}^{\{1, 2\}}$ to {\tt id} and as a result the type $\mathtt{int}^{\{1,2\}} \times \mathtt{int}^{\{1,2\}}$ to {\tt f}.
-Clearly, this result is not optimal.
-This loss in precision has been caused by the two separate calls to {\tt id} \emph{poisoning} each other.
-
-A similar situation occurs at the type-level in the function:
-\begin{verbatim}
-f = let id x = x
-     in (id 1, id true)
-\end{verbatim}
-Hindley--Milner manages to avoid problems in this situation by a mechanism called \emph{let-generalization}.
-The binding {\tt id} has the type $\alpha \to \alpha$.
-Instead of trying -- and failing -- to unify $\alpha$ with both {\tt int} and {\tt bool} in the body of the let-binding, the type of {\tt id} is generalized to the polymorphic type $\forall \alpha.
-\alpha \to \alpha$.
-In the body of the let-binding, {\tt id} can be instantiated twice: once to $\mathtt{int} \to \mathtt{int}$ and once to $\mathtt{bool} \to \mathtt{bool}$.
-
-A similar trick can be used for the annotated type system, except that instead of quantifying over type variables, we quantify over annotation variables.
-
-\section{Refinement types}
+\section{Refinement Types}\label{secrt}
 
@@
     \tikzstyle{all nodes}=[inner sep=4pt]
-    \draw node(top)at(1,3){\tt List a}
-          node(sorn)at(1,2){{\tt Singleton a} $\lor$ {\tt Nil a}}
-          node(s)at(0,1){\tt Singleton a} node(n)at(2,1){\tt Nil a}
+    \draw node(top)at(1,3){|List a|}
+          node(sorn)at(1,2){{|Singleton a|} $\lor$ {|Nil a|}}
+          node(s)at(0,1){|Singleton a|} node(n)at(2,1){|Nil a|}
           node(bot)at(1,0){$\bot$};
@@
 \end{tikzpicture}
-\caption{A lattice of {\tt List} and its {\tt rectype}s.}
+\caption{A lattice of {|List|} and its {|rectype|}s.}
 \label{listlattice}
@@
 Refinement types are a form of subtyping that allow us to state more accurately what values a variable of a particular type can contain.
-For example, a refinement type {\tt \{ a :: Int | a < 0 \}} states that {\tt a} is a variable of type {\tt Int} than can only contain negative numbers.
+For example, a refinement type {\tt \{ a :: Int || a < 0 \}} states that |a| is a variable of type |Int| that can only contain negative numbers.
 
@@
 
-Their key contribution is allowing the programmer to define {\tt rectype}s. Besides defining a recursive type for lists:
+Their key contribution is allowing the programmer to define |rectype|s. Besides defining a recursive type for lists:
 
 \begin{verbatim}
-data List a = Nil | Cons a (List a)
+data List a = Nil || Cons a (List a)
 \end{verbatim}
 
-The programmer can also define a {\tt rectype} describing singleton lists:
+the programmer can also define a |rectype| describing singleton lists:
 
@@
 
-Together with an automatically derived {\tt rectype} for the non-recurisive constructor {\tt Nil} (and a union type contructor) the compiler can construct -- using known algorithms on regular tree grammars \cite{DBLP:books/others/tree1984} -- a finite lattice of refinement types for {\tt List a} given in Figure \ref{listlattice}.
+Together with an automatically derived |rectype| for the non-recurisive constructor |Nil| (and a union type contructor) the compiler can construct---using known algorithms on regular tree grammars \cite{DBLP:books/others/tree1984}---a finite lattice of refinement types for |List a| given in Figure \ref{listlattice}.
 
-\subsection{Union and Intersection Types}
+\subsection{Union and Intersection Types}\label{secit}
 
@@
 
-A union type such as {\tt Int $\lor$ Bool}, {\tt List a $\lor$ Singleton a} or {\tt Int -> Int $\lor$ Bool -> Bool} state a value can be either of the type on the left or of the type on the right, but we do not know which.
+A union type such as {\tt Int $\lor$ Bool}, {\tt List a $\lor$ Singleton a} or {\tt Int -> Int $\lor$ Bool -> Bool} state that a value can be either of the type on the left or of the type on the right, but we do not know which.
 In the case of {\tt List a $\lor$ Singleton a} we are able to simplify this type to {\tt List a}, as {\tt Singleton a} is a subtype of {\tt List a}.
@@
 Cons  ::  a  -> Nil        a  -> Singleton  a
-     /\ a -> Singleton a -> List      a
-     /\ a -> List      a -> List      a
+      &&  a  -> Singleton  a  -> List       a
+      &&  a  -> List       a  -> List       a
 \end{verbatim}
 
-This type can be derived automatically from the {\tt rectype} declarations.
+This type can be derived automatically from the |rectype| declarations.
 Types of functions can also be inferred automatically, although there may be a loss of precision when higher-order functions or polymorphism are involved.
@@
 
-where {\tt case\_list} has the refinement type
+where |case_List| has the refinement type
 
@@
 
-\chapter{Related Work}\label{secrelatedwork}
-
-% table comparing the approaches
-% * laziness
-% * polymorphism
-% * annotations (required / supported)
-% * decidability
-% * errors generated
-
-\section{Neil Mitchell's \emph{Catch}}
-
-Neil Mitchel's \emph{Catch} (``CAse Totality CHecker'') \cite{Mitchell} is a tool specialized in finding potential pattern match failures in Haskell programs.
-
-\subsection{Overview}
-
-Catch works by calculates preconditions on functions.
-The preconditions are given in a constraint language.
-By varying the constraint language trade-offs between performance and precision can be made.
-If the calculated precondition for \texttt{main} is True, the program does not crash.
-
-Preconditions are calculated iteratively.
-Precondition start out as \texttt{True}, except for \texttt{error}, whose precondition is \texttt{False}.\footnote{Note that {\tt undefined} is defined in terms of {\tt error}.}
-
-\paragraph{Example} Following the example given by Mitchell, given the function:
-\begin{verbatim}
-safeTail xs = case null xs of
-                True  -> []
-                False -> tail xs
-\end{verbatim}
-The computed precondition will be:
-\[ \textrm{Precondition}(\texttt{null xs}) \land (\texttt{null xs} \in \{\texttt{True}\} \lor \textrm{Precondition}(\texttt{tail xs})), \]
-stating the necessary precondition for \texttt{null xs} not to crash mush be fulfilled and either \texttt{null xs} must evaluate to \texttt{True} or the necessary precondition for \texttt{tail xs} not to crash must be fulfilled.
-
-As \texttt{null xs} cannot crash we find its precondition to be simply True and we can deduce the precondition for \texttt{tail xs} to be $\texttt{xs}\in\{(:)\}$.
-Substituting these into our precondition we find
-\[ \textrm{True} \land (\texttt{null xs} \in \{\texttt{True}\} \lor \texttt{xs}\in\{(:)\}), \]
-which simplifies to
-\[ \texttt{null xs} \in \{\texttt{True}\} \lor \texttt{xs}\in\{(:)\}. \]
-The subexpression  $\texttt{null xs} \in \{\texttt{True}\}$ forms a postcondition on \texttt{null xs}.
-From this postcondition Catch computes the required precondition that satisfies it, which in this particular case should turn out to be $\texttt{xs}\in\{[]\}$.
-Substituting, we find
-\[ \texttt{xs}\in\{[]\} \lor \texttt{xs}\in\{(:)\}, \]
-which, using our knowledge of the list data type, turns out the be a tautology, i.e. the (trivial) precondition for \texttt{safeTail} is True.
-
-\subsection{Constraint systems}
-The example in the previous section used a simple constraint language, specifying to which set of head constructors an expression should evaluate.
-Mitchell developed two more expressive constraint systems: regular expression constraints and multipattern constraints.
-
-\paragraph{Regular expression constraints} Constraints are formed by a regular expression over an alphabet of a number of ad-hoc \emph{selectors} (e.g.
-\texttt{hd} and \texttt{tl} for lists.) A precondition for \texttt{map head xs} using regular expression constraints reads $\texttt{xs} \in (\texttt{tl}^{*} \cdot \texttt{hd} \leadsto \{(:)\})$.
-It should be interpreted as ``after applying zero or more \texttt{tl}s to \texttt{xs} and then applying a \texttt{hd} we should find a \texttt{(:)} constructor.''
-
-Accoriding to Mitchell regular expression constraints tend to scale badly as the program increase in size, although he could not identify more specific condition under which this problem manifests.\footnote{Personal communication}
-
-\paragraph{Multipattern constraints} Multipatterns are of the form $\alpha \star \rho$, where $\alpha$ gives the set of constructors that are valid at the head of the value, while $\rho$ gives the set of constructors that can appear at the recursive positions (if any) of the constructor at the head of the value.
-The elements of these sets can again be multipatterns.
-To specify that \texttt{xs} in \texttt{map head xs} should be a list of non-empty lists, we use the multipattern:
-\begin{align*}
-\{[]&, (:) (\{(:)\ \texttt{Any} \} \star \{ [], (:)\ \texttt{Any}\})\} \\
-&\star \\
-\{[]&, (:) (\{(:)\ \texttt{Any} \} \star \{ [], (:)\ \texttt{Any}\})\}
-\end{align*}
-\texttt{Any} is a wildcard that matches any constructor.
-
-\subsection{Discussion}
-% should have been mentioned earlier?
-The analysis of Catch works on a first-order language. The input program needs to be defunctionalized before it can be analyzed. The defunctionalization algorithm employed by Catch is not complete.
-
-Calculated preconditions are unnecessarily restrictive in the presence of laziness (Mitchell, p. 142).
-
-\section{Static Contract Checking}
-\subsection{Overview}
-Dana Xu's work on static contract checking for Haskell \cite{Xu09staticcontract} (including the ESC/Haskell system \cite{Xu:2006:ESC:1159842.1159849}) allows checking of arbitrary programmer supplied pre- and postconditions on functions and is able to detect pattern match failures in the process.
-
-Programmers define pre- and postconditions in the form of a contract (a refinement type). An appropriate contract for the \texttt{head} function defined above might be:
-\begin{verbatim}
-{-# CONTRACT head :: {s | not (null s)) -> {x | True} #-}
-\end{verbatim}
-This contract states that if head is given a list \texttt{s} for which the predicate \texttt{not (null s)} holds, i.e.
-it is a non-empty list, the function will not crash\footnote{The system only guarantees partial correctness, so the term might still diverge.}, indicated by the trivial postcondition \texttt{True} on the return value.
-
-\begin{figure}
-\centering
-\begin{tabular}{lclr}
-t &$::=$& $\{ x | p \}$     & Predicate          \\
-  & $|$ & $x : t_1 \to t_2$ & Dependent function \\
-  & $|$ & $(t_1, t_2)$      & Constructor         \\
-  & $|$ & $\mathtt{Any}$    & Polymorphic ``Any''
-\end{tabular}
-\caption{Syntax of Xu's contracts}
-\label{xu-contract-syntax}
-\end{figure}
-
-Contracts can be constructed from predicates, a dependent function space constructor, arbitrary constructors and a polymorphic \texttt{Any} contract that is always satisfied, including by functions which crash (Fig.
-\ref{xu-contract-syntax}).
-Any \texttt{Bool}-valued Haskell expression can be used as a predicate.
-
-Dependent functions are helpful when declaring a contract, e.g. for the \texttt{reverse} function:
-\begin{verbatim}
-{-# CONTRACT reverse :: {xs | True} -> {rs | length xs == length rs} #-}
-\end{verbatim}
-The predicate on the return value depends on the the input. Constructors and the \texttt{Any} contract are useful when declaring a contract for \texttt{fst}:
-\begin{verbatim}
-{-# CONTRACT fst :: (Ok, Any) -> Ok #-}
-\end{verbatim}
-Here \texttt{Ok} is used as a shorthand for \texttt{\{$x$ | True\}}.
-As \texttt{fst} discards the value on the right in the tuple we should not care if it is an expression that crashes, so we can not use \texttt{Ok}.
-
-\subsection{Contract checking}
-Haskell's syntax is extended with two \emph{exception values} -- \texttt{BAD} and \texttt{UNR} -- which are only used internally by the checker.
-\texttt{BAD} signifies an expression which crashes and \texttt{UNR} an expression which is either unreachable or diverges.
-
-In the program that is going to be verified all crashing expressions (including \texttt{error} and \texttt{undefined}), as well as missing patterns in case-expressions are replaced by \texttt{BAD} exceptions.
-
-The verification of the contracts is based on a translation of the contracts into Findler--Felleisen wrappers \cite{Findler:2002:CHF:581478.581484}, which will cause a function to crash (using \texttt{BAD}) if it disobeys its contract or diverge if it called in a way that is not permitted.
-While technically interesting it is not directly relevant to the detection of pattern match failures so we shall not discuss it in more depth.
-
-The actual verification process continues by \emph{symbolic evaluation}, basically applying various simplifications to the resulting program including $\beta$-reductions.
-Any code deemed to be unreachable is pruned from the program.
-
-The presence of recursive functions in the contracts might cause the symbolic evaluation to diverge if care is not taken to limit the number of evaluation steps.
-By setting an upper bound on the number of simplification steps we lose accuracy, but gain decidability of the verification process.
-
-Arithmetical expressions in case-expressions (e.g.
-\texttt{case x*x >= 0 of \{ True -> ...; False -> BAD \}}) cannot be handled directly by the symbolic evaluator.
-These are collected in a table and send to an external constraint or SMT solver.
-If the solver determines these expressions are inconsistent the code is unreachable and can be pruned.
-
-After the symbolic evaluation has terminated the checker will tell the programmer if the program is ``definitely satisfies the contract'' (if no \texttt{BAD} exceptions remain anywhere in the program), ``definitely does not satisfy the contract'' or ``don't know.''
-
-\subsection{Discussion}
-Unlike our envisioned system, Xu's static contract checking requires programmer supplied contracts on nearly all functions (contracts on trivial functions can be omitted and are handled by inlining their definition when called.) Compared to Mitchell's Catch it can handle higher-order functions natively and has a more (too?) expressive contract language.
-
-\section{Dependent ML}
-DML($\mathcal{L}$) is an extension of the ML programming language which enriches ML's type system with a limited form of dependent types over a (decidable) constraint domain (or index language) $\mathcal{L}$ \cite{DML-JFP07}.
-
-Xi's initial example -- recast in a more Haskell-like syntax -- declares a {\tt List} data type dependent on an integer corresponding to the length of the list and a function to concatenate two such lists:
-\begin{verbatim}
-1  type Nat = {a :: Int | a >= 0}
-2
-3  data List<Int> a = Nil<0>
-4                   | {n :: Nat} Cons<n+1> a (List<n> a)
-5
-6  (++) :: {m :: Nat} {n :: Nat} List<m> a -> List<n> a -> List<m+n> a
-7  (++) Nil         ys = ys
-8  (++) (Cons x xs) ys = Cons x (append xs ys)
-\end{verbatim}
-
-For many functions producing a list, the exact length might not be derived by such a trivial calculation ({\tt m+n}) on the lengths of the input lists.
-A prominent example is the {\tt filter} function:
-\begin{verbatim}
-1  filter :: (a -> Bool) -> {m :: Nat} List<m> a -> [n :: Nat | n <= m] List<n> a
-2  filter p Nil         = Nil
-3  filter p (Cons x xs) | p x       = Cons x (filter p xs)
-4                       | otherwise =         filter p xs
-\end{verbatim}
-here we only know that the resulting list is equal or shorter in length than the input list ({\tt n <= m}).
-This is expressed as the dependent sum {\tt [n :: Nat | n <= m]}.
-In a full-fledged dependently-typed language we would also return a proof object stating the predicate {\tt p} holds for all the elements in the output list, but this is beyond the expressive power of DML.
-
-What we gain is a relief from the need to provide proofs for trivial arithmetical (in)equalities.
-Imagine a {\tt zip} function which requires the two list to be zipped together to be of equal length.
-When calling {\tt zip (xs ++ ys) (ys ++ xs)} this preconditions seems to be intuitively satisfied.
-From the point-of-view of the compiler the first list has length $m+n$ and the second lists $n+m$, however.
-In most dependently-typed languages we will now have to invoke a lemma or tactic proving the commutativity of addition.
-DML can simply send the constraint $m + n = n + n$ to an ILP solver and conclude the constraint is satisfiable.
-
-\subsection{Discussion}
-Compared to Xu's static contract checking, Xi's DML constrains the constraint system to a decidable theory.
-The constraints and indices -- such as {\tt m+n} and {\tt n~<=~m} -- may superficially look like ordinary Haskell expression, but in fact belong to a much smaller index language.
-While type checking is decidable, type inference is not and as a result DML still requires type annotations.
-
-To be applicable to a pattern match analysis we must try to infer a type like:
-\begin{verbatim}
-head :: {n :: Int | n >= 1} [a]<n> -> a
-\end{verbatim}
-for the {\tt head} function. We also have to implicitly extend the list data type with an index representing its length.
-
-This seems significantly more challenging than inferring the Catch-like type
-\begin{alltt}
-head :: \{xs :: [a] | xs \(\in\) (:) \} -> a
-\end{alltt}
-The correspondence between the the type and the pattern matching happening in the definition of {\tt head} is much more direct.
-
-%\begin{small}
-%There is an indirect correspondence, however.
-The value of the index of the list is given by the {\tt length} function and by analyzing the pattern matching happening in that function we could derive that $\{n :: \mathtt{Int} | n >= 1\}\ \left[a\right]\left<n\right> \simeq \{xs :: [a] | xs \in (:) \}$.
-%
-%Instead of adding the an index of type {\tt Int} to {\tt List} to create {\tt List<Int>}, we add the abstraction {\tt length :: List -> Int}.
-This generalizes to arbitrary data types $F$ and constraint systems over $\tau$, by adding an abstraction $F \to \tau$ to create $F\left<\tau\right>$.
-(But can we use arbitrary functions of type $F\to\tau$, or does {\tt lenght} have some particularly nice properties?
-Certainly if viewed as a catamorphism (e.g.
-$(0, (\lambda n.
-1+n))$) it should consist of only expressions in the constraint language $\mathcal{L}$ -- they correspond exactly to the annotations on {\tt Nil<0>} and {\tt \{n :: Int\} Cons<n+1>} -- but is this sufficient?
-What if we look at a {\tt Tree}?
-Adding a property {\tt minDepth :: Tree -> Int} would guarantee we can pattern match on a certain number of {\tt Branch}es before encountering a {\tt Leaf}, but what does {\tt maxDepth :: Tree -> Int} buy us?
-More importantly, while we might be able to infer an appropriate type for {\tt head}, inferring the type given to {\tt (++)} and {\tt filter} above still seems out of reach.)
-%\end{small}
-
 \section{Compiling Pattern Matching}
@@
 It does turn out that the analysis closely follows, but can at places be simplified with respect to, the manner in which pattern matching is compiled.
-This indicates might also be able to follow such an approach when analyzing case-statements in our pattern-match analysis.
-
-\chapter{Approach}\label{secapproach}
-
-% type and effect system
-% refinement types / subtyping / widening (compare to multi-patterns)
-% how much precision do we need?
-
-\section{Incremental development}
-We will incrementally develop our analysis for the various constructs in the Haskell language.
-
-\paragraph{Variables, constants, abstraction, application, if-statements and simple let-bindings} We will start out by implementing the basic constructs of the simply typed $\lambda$-calculus.
-This includes simple let-bindings, i.e.
-non-recursive let-bindings in which no pattern matching or generalization occurs.
-
-At this stage we will only allow booleans and integers as constants.
-Due to their primitive and infinity nature integers will need to be treated specially by the analysis in any case.
-Booleans and if-statements can be handles in a more general fashion during later stages.
-\paragraph{Recursion} The next elementary construct is recursion.
-This will require us to implement a form of \emph{widening} for the integers and lists to ensure termination of the analysis.
-As the exact form of widening used can have a big impact on the precision of the analysis we do not want to commit to a specific lattice here.
-Various widening strategies should be easily pluggable into the source code and perhaps eventually be able to be specified at a higher-level by the user of the tool.
-\paragraph{Simple case-statements} The most critical part of the analysis will be how it handles case-statements.
-In this state we will add simple case-statements (i.e., non-nested, non-backtracking, guardless case-statements) and primitive list constants, our first recursive data type.
-\paragraph{Complex case-statements and guards} To be able to handle the full Haskell 98 language our analysis must be able to deal with the complex semantics of case-statements.
-This includes nested-patterns, guards and backtracking.
-We would like to be able to analyze any boolean-valued expression that is used as a guard.
-\paragraph{Algebraic data types} Next we extend the analysis to work on arbitrary programmer-defined data types instead of only on our primitive list data type.
-This will mostly involve generalizing out analysis to work on with a representation of arbitrary algebraic data types.
-Data types containing fields with function types may also prove interesting.
-\paragraph{Polymorphism} Polymorphism can generally be added to an analysis as a relatively orthogonal feature.
-However, there might be some interesting interactions with our refinement and intersection types here.
-\paragraph{Full let-bindings} We extend the analysis for let-bindings to handle (mutual) recursion and pattern matching.
-\paragraph{Type classes} An interesting feature of Haskell's type system are type classes. We currently do not foresee any specific difficulties with respect to be able to handle them in out analysis.
-\paragraph{Syntactic sugar} As the final stage we would add constructs which are only syntactic sugar and therefore would not pose any significant problem to our analysis.
-This includes for example list-comprehensions and do-notation.
-
-\section{Refinement types}
-
-Compared to Freeman and Pfenning's approach to refinement types we will initially not rely on programmer supplied, but instead build our lattice of types from finite unrollings of data types.
-We hope this makes calculating the lattice easier.
-
-Freeman's inference algorithm for refinement type assign a single, albeit complex, type to each binding.
-This makes it a monovariant analysis.
-We will develop a polyvariant analysis, making it sensitive to the sites where a binding is used.
-We hope this will improve the performance of the analysis.
-
-\section{Parsing}
-
-We will develop our tool in the Haskell programming language.
-A number of libraries for parsing input programs are available.
-The most obvious choice seem to be the {\tt haskell-src-exts} package, which is also used by the HLint tool.
-If this proves to be a problematic choice, other possibilities are the parser from GHC or UHC.
-
-\chapter{Deliverables}\label{secdeliverables}
-
-There will be two deliverables of this project:
-\begin{enumerate}
-\item A \emph{tool} implementing the pattern match analysis for Haskell 98. This tool is intended to be practically usable and not a mere research prototype.
-\item A \emph{thesis} describing the theory underlying the tool and documenting the implementation choices made in the tool.
-\end{enumerate}
-
-\chapter{Evaluation}\label{secevaluation}
-
-The evaluate whether we were able to answer research question we first need to evaluate whether the developed tool implements a practically usable pattern-match analysis.
-We can do so by trying it out on the examples sketched in Section \ref{secexamples}, the literature and for example the {\tt nofib} testsuite.
-
-If we can answer this question with a ``yes'', and assuming the tool indeed implements the analysis using a type and effect system and refinement types, we can also answer out research question with a ``yes''.
-
-In case of a ``no'' we will need to argument whether this is due to a failure of our implementation or because there are good theoretical reasons why such an analysis cannot be implemented.
-
-\chapter{Planning}\label{secplanning}
-
-\begin{description}
-\item[Implement framework (April)] Implement a constraint-based version of Hindley--Milney for Haskell (the underlying the system of the analysis).
-This is known work \cite{Odersky:1999:TIC:308181.308203,DBLP:journals/entcs/HageH09} and should not prove to be difficult.
-This implementation will provide a guiding framework for the analysis, highlight possible issues we might run into when implementing a constraint-based analysis for full Haskell and allow us to decide which parsing library and abstract syntax representation to use.
-\item[Implement analysis (April--July)] The largest portion of time is assigned to incrementally developing and building the pattern-match analysis. The likely order will be:
-\begin{enumerate}
-\item Variables, constants, abstraction, application, if-statements and simple let-bindings (1.5 weeks)
-\item Recursion (1.5 weeks)
-\item Simple case-statements (2 weeks)
-\item Complex case-statements and guards (2 weeks)
-\item Algebraic data types (1.5 weeks)
-\item Polymorphism (1.5 weeks)
-\item Full let-bindings (1 week)
-\item Type classes (1.5 weeks)
-\item Syntactic sugar (1.5 weeks)
-\end{enumerate}
-\item[Evaluation and thesis writing (July--August)] The final phase of this project will include an evaluation of the tool and finishing of the thesis.
-A significant potion of the thesis is expected to be written in parallel with the implementation, after a particular subtask has been implemented successfully.
-\end{description}
-
-\chapter{Future Work}
-
-Non-goals of the thesis project are:
-\begin{enumerate}
-\item A formal soundness proof of the analysis, and
-\item Supporting Haskell extensions beyond Haskell 98.
-\end{enumerate}
-Both of these would be candidates for future work.
-
-\bibliography{thesis-proposal}
-\bibliographystyle{plain}
-
-\end{document}
+This indicates we might also be able to follow such an approach when analyzing case-statements in our pattern-match analysis.

### 2015-09-01 deprecated/pattern-match-analysis/thesis/typesystem.tex
pattern-match-analysis

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% PATTERN MATCH ANALYSIS %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Pattern Match Analysis}\label{chappma}

\section{Overview}

The key idea behind the analysis is to keep track of the possible values each variable and expression can have. When pattern matching with an if-then-else or case-expression, we verify that the values the scrutinee can have are covered by the patterns in each of the alternatives.

The set of possible values associated with each variable and expression is called a \emph{refinement}\index{refinement} and placed as an annotation on the type of that variable or expression. We call the type without such an annotation the \emph{underlying type}\index{underlying type} of an expression and the type with such an annotation the \emph{annotated type}\index{annotated type} or \emph{refinement type}\index{refinement type}.

\paragraph{Examples}
\begin{eqnarray}
\TTrue &:& \Bool^{\{\TTrue\}} \\
42 &:& \Int^{\{42\}} \\
\TupleOf{7}{\FFalse} &:& \TupleOf{\Int^{\{7\}}}{\Bool^{\{\FFalse\}}}^\top \\
\textbf{[}3,2,1\textbf{]} &:& \textbf{[}\Int^{\{1,2,3\}}\textbf{]}^{\{\textbf{(\_\,:\,\_\,:\,\_\,:\,[])}\}} \\
\lambda x.\ x + 1 &:& \Int^\top \overset{\top}{\to} \Int^\top
\end{eqnarray}

Similarly, by analyzing the patterns of case-statements we infer the maximal set of values each variable can have without causing a pattern-match failure. Thus, if the set of values a variable can have is not a subset of the values it is allowed to have, we have statically detected a potential source of pattern-match failures.

\section{Higher-Order Functions}\label{sec31}

The analysis should be able to handle higher-order functions. Let us take a moment to review what the refinements type should look like for higher-order functions and how they should be interpreted.

Consider the function:

\begin{code}
main b f =  if b then
                if  f 42  then  100  else  200
            else
                if  f 43  then  300  else  400
\end{code}

The function |main| takes a boolean and a function as its arguments. The argument |f| is applied to an integer constant in the body of main and---because it is used as the scrutinee of an if-then-else expression---should return a boolean as its result. Depending on the values of |b| and |f|, the latter after having been applied to either |42| or |43|, the function |main| will return one of four different integer values.

From this we can infer the underlying type of |main| should be: \[ \Bool \to (\Int \to \Bool) \to \Int \]

Under what conditions is this function guaranteed not to crash due to a pattern match failure? The argument |b| and function |f|, after having been applied to an integer, can evaluate to arbitrary boolean values. The function |f|, however, might also cause a pattern match failure while being evaluated. We need to be sure it does not do so if passed one of the values |42| or |43|. An appropriate choice for the refined type would thus be: \[ \Bool^{\{\TTrue, \FFalse\}} \to (\Int^{\{42, 43\}}\to\Bool^{\{\TTrue,\FFalse\}})\to \Int^{\{100,200,300,400\}}\]

Note the difference in meaning of the refinement depending on whether it appears on an underlying type in a \emph{covariant}\index{covariant} or \emph{contravariant}\index{contravariant} position. For clarity, we give the type once more with variance annotations (where + means covariant and - means contravariant):
\[ \Bool^{\{\TTrue, \FFalse\}}_{-} \to (\Int^{\{42, 43\}}_{+}\to\Bool^{\{\TTrue,\FFalse\}}_{-})\to \Int^{\{100,200,300,400\}}_{+}\]

Refinements appearing on types in a covariant position give either the minimal set of values that a function can return, or---in the case of arguments passed to a functional parameter---a minimal set of values that function should be able to accept without failing due to a pattern match. In all cases the set of values appearing in the refinement can safely be enlarged.

Refinements appearing on types in a contravariant position give either the maximal set of values that can be passed to an argument, or---in the case of a return value of a functional parameter---a maximal set of values which that function is allowed to return without causing a pattern match failure in the caller. In all cases the set of values appearing in the refinement can safely be reduced.

Another valid refinement type for |main| would be: \[ \Bool^{\{\TTrue\}} \to (\Int^{\{41, 42, 43\}}\to\Bool^{\{\FFalse\}})\to \Int^{\{100,200,300,400,500\}}\]

This type is less desirable than the previous refinement type: it accepts fewer values for the parameter |b|, the functions satisfying the type $\Int^{\{41, 42, 43\}}\to\Bool^{\{\FFalse\}}$ form a subset of the functions satisfying $\Int^{\{42, 43\}}\to\Bool^{\{\TTrue,\FFalse\}}$ and the larger refinement on the result type may prevent |main| from being passed as an argument to another higher-order function.

The analysis should strive to infer refinement types with the smallest possible refinements on types in covariant positions and the largest possible refinement on types in contravariant positions.

\section{Analysis}\label{sec33}

The analysis is \emph{constraint-based}\index{constraint-based} and \emph{type-directed}\index{type-directed}, meaning it is formulated as a two part process.

The first is a type system that generates a set of constraints from the abstract syntax tree. It generates two distinct set of constraints: one set allows us to infer the underlying type of expressions, the other allows us to construct the refinements belonging those types.

The second part consists of solving the generated constraints. The first set of constraints---representing the underlying types---can be solved using unification. The second set requires a more complex solver based on a worklist algorithm.

The generation of constraints and the unification algorithm are described in Chapter \ref{chapcg}. The constraint solver for the second constraint set is described in Chapter \ref{chapcs}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% CONSTRAINT GENERATION %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Constraint Generation}\label{chapcg}

\section{Overview}

The constraint generation phase of the analysis is built on top of a constraint-based version of the Hindley--Milner type system \cite{Milner78atheory,Pierce:2002:TPL:509043} extended to handle annotations and generate constraints on and between those annotations.

As mentioned in Section \ref{sec33} the type system generating the constraints produces two distinct sets of constraints. The first is a set of equality constraints between \emph{simply annotated types}\index{simply annotated type} as given in Figure \ref{AnnotatedTypes}. Simply annotated types are annotated, except that all annotations are restricted to be plain annotation variables. The reason for this is restriction is to make annotated types into a free algebra, so the constraint set can be solved by a unification algorithm presented in Section \ref{unifi}. The result of the unification algorithm is a substitution from type variables to simply annotated types, that is applied to both the inferred type and the second constraint set. The second constraint set consists of subtype constraints (Figure \ref{figsubtype}) between annotations (Figure \ref{Annotations}).

\section{Types and Annotations}

Throughout this and the next chapter we will assume the following naming conventions for variables:

\begin{figure}[h!t]
\begin{align*}
        x, f    & \in          \mathbf{Var}   &&\textrm{variables} \\
%       e       & \in          \mathbf{Expr}  &&\textrm{expressions} \\
\widehat\tau    & \in \widehat{\mathbf{Type}} &&\textrm{annotated types} \\
        \alpha  & \in          \mathbf{TVar}  &&\textrm{type variables} \\
\widehat\Gamma  & \in \widehat{\mathbf{TEnv}} &&\textrm{annotated type environments} \\
        \varphi & \in          \mathbf{Ann}   &&\textrm{annotations/refinements} \\
        \beta   & \in          \mathbf{AVar}  &&\textrm{annotation variables}
\end{align*}
\caption{Naming conventions}
\label{NonTerminals}
\end{figure}

An \emph{annotated type}\index{annotated type} (Figure \ref{AnnotatedTypes}) can be a plain type variable $\alpha$, a type constant such as $\Bool$ or $\Int$ with an annotation $\varphi$, a tuple of two annotated types, a list with elements of an annotated type, or a function with the domain and range being given by annotated types.

\begin{figure}[h!t]
\begin{eqnarray*}
\widehat\tau &::=& \alpha                                       \\
             &||&  \textbf{Bool}^\varphi                        \\
             &||&  \textbf{Int}^\varphi                         \\
             &||&  \TupleOf{\widehat\tau}{\widehat\tau}^\varphi \\
             &||&  \ListOf{\widehat\tau}^\varphi                \\
             &||&  \tauhat_1 \overset{\varphi}{\to} \tauhat_2
\end{eqnarray*}
\caption{Annotated types}
\label{AnnotatedTypes}
\end{figure}

There is no annotation on the type variable as it gains one when it is substituted by a type constant or constructor. In any other case, we are dealing with a polymorphic argument of a polymorphic function. As we are not able to pattern match on such values, it is not useful to keep track of the values such a variable can have.

The annotations on tuples and functions are not genuinely used, in the sense that in the current implementation of the analysis their elements are always picked from the trivial one element lattice. We have included them here for uniformity and to make the framework more general for future work.

Lists indirectly have an annotation both on the type of its elements and on the list as a whole. This allows for more flexible ways of specifying the possible values a list may have. Recall the |add| example from Section \ref{secexamples}. While we could attempt to precisely represent all possible lists that might occur---remembering both the shape of the list and which values each element at each each position in the list can have---this approach is unlikely to scale in practice. The only information necessary for determining that the |add| function does not fail due to a pattern match is that its inputs should be lists of integers where each elements in the lists is an integer in the set $\{0, 1\}$. We can also infer it will always produce such a list. An appropriate type for |add| would thus be \[\ListOf{\Int^{\{0, 1\}}}^{\{\star\}} \to \ListOf{\Int^{\{0, 1\}}}^{\{\star\}} \to \ListOf{\Int^{\{0, 1\}}}^{\{\star\}}\]
where the refinement $\star$ represents a list of any shape.

When analyzing polymorphic functions working on lists of any type of elements, we are likely only interested in the shape of the list and not in the values of its elements. For example, an appropriate type for |head| would be: \[\ListOf{\alpha}^{\{\Conss{\star}\}} \to \alpha\]

\begin{figure}[h!t]
\begin{eqnarray*}
\widehat\sigma &::=& \forall \overline{\alpha} \overline{\beta}.\ R \Rightarrow \tauhat
\end{eqnarray*}
\caption{Annotated type schemes}
\label{AnnotatedTypeSchemes}
\end{figure}

Like in the Hindley--Milner type system, we represent polymorphic types by a \emph{type scheme}\index{type scheme} (Figure \ref{AnnotatedTypeSchemes}). The difference here is that we are also able to quantify over, not only type variables, but also over annotation variables. Furthermore, we associate a set $R$ of subtype constraints (Figure \ref{figsubtype}) with each type scheme. These are constraints the constraint solver was unable to discharge when the type was generalized and should be verified to hold when the type is later instantiated.

Type schemes can be associated with variables using a type environment (Figure \ref{AnnotatedTypeEnv}).

\begin{figure}[h!t]
\begin{eqnarray*}
\widehat\Gamma &::=& \epsilon\ ||\ \widehat\Gamma,\ x : \widehat\sigma
\end{eqnarray*}
\caption{Annotated type environments}
\label{AnnotatedTypeEnv}
\end{figure}

\emph{Annotations}\index{annotation} hold the refinements belonging to the underlying type. Refinements form a set of annotation variables and abstract values (Section \ref{secabstractvalues}).

\begin{figure}[h!t]
\begin{eqnarray*}
\varphi &::=& \beta                 \\
     &||&  \{\pi_\tau\}             \\
     &||&  \varphi_1 \cup \varphi_2
\end{eqnarray*}
\caption{Annotations/refinements}
\label{Annotations}
\end{figure}

The type system generates a set of \emph{equality constraints}\index{constraint!equality} between annotated types and a set of \emph{subset constraints}\index{constraint!subset} between refinements:
\begin{figure}[h!t]
\begin{eqnarray*}
C &::=& \{\tauhat_1 = \tauhat_2\} \\
  &||&  C_1 \cup C_2
\end{eqnarray*}
\caption{Equality constraints}
\label{figeqc}
\end{figure}

\begin{figure}[h!t]
\begin{eqnarray*}
R &::=& \{\varphi_1 \subseteq \varphi_2\} \\
  &||&  R_1 \cup R_2
\end{eqnarray*}
\caption{Subset constraints}
\label{figsubtype}
\end{figure}

\section{Refinements and Lattices}\label{secabstractvalues}

In the examples we have seen so far, we have used the actual values of a type as refinements. This approach will at best fail to scale up when analyzing larger programs, but worse, will likely cause the constraint solver to diverge: while the refinements naturally form a lattice under subset inclusion with the meet and join operations given by intersection and union, it will not satisfy the ascending chain condition for infinite types, such as integers and list. As a result the refinements can grow infinitely large.

To overcome this problem we abstract from the concrete values of a type in its refinement and apply widening to deal with recursive data types.

\subsection{Booleans}

Booleans are a finite type and thus a set of booleans forms finite lattice under subset inclusion. Finite lattices always satisfy the ascending chain condition, thus there is little need to abstract the values of a boolean in a refinement.

\begin{figure}[h!t]
\begin{eqnarray*}
\pi_\mathbf{Bool} &::=& \textbf{True}  \\
                  &||&  \textbf{False}
\end{eqnarray*}
\caption{Trivial abstraction for $\Bool$}
\label{LatticeBool}
\end{figure}

\subsection{Integers}

Integers form an infinite set, so will need to abstract from them. Possible choices include taking only the sign of an integer into account:

\begin{figure}[h!t]
\begin{eqnarray*}
\pi_\mathbf{Int} &::=& \textbf{+} \\
                 &||&  \textbf{0} \\
                 &||&  \textbf{-}
\end{eqnarray*}
\caption{Sign abstraction for $\Int$}
\label{LatticeIntSign}
\end{figure}

or only its parity:

\begin{figure}[h!t]
\label{LatticeIntParity}
\begin{eqnarray*}
\pi_\mathbf{Int} &::=& \textbf{Odd}  \\
                 &||&  \textbf{Even}
\end{eqnarray*}
\caption{Parity abstraction for $\Int$}
\end{figure}

More complicated alternatives are possible. We could take the product of the sign and parity lattices or partition the integers into a finite number of intervals with the cuts determined by concrete integers appearing in the program text.

The type system producing the constraints and constraint solver are agnostic towards the actual choice. They only require a number of operations to be defined: abstracting a concrete value (\ref{r41}), computing the greatest lower and least upper bound of two refinements (\ref{r42} and \ref{r43}) and determining whether one refinement is contained in (smaller than) another (\ref{r43}).

\begin{figure}[h!t]
\begin{eqnarray}
\iota       &:& \tau \to \pi_\tau               \\ \label{r41}
\sqcup      &:& \varphi \to \varphi \to \varphi \\ \label{r42}
\sqcap      &:& \varphi \to \varphi \to \varphi \\ \label{r43}
\sqsubseteq &:& \varphi \to \varphi \to \Bool      \label{r44}
\end{eqnarray}
\caption{Operations on refinements}
\label{LatticeOperations}
\end{figure}

\subsection{Lists}

Lists, being a recursive data type, require a more complicated abstraction. We might choose to abstract to only the shape of the list, remembering the nesting of nil and cons constructors, but forgetting the elements of the list stored in the cons constructors:

\begin{figure}[h!t]
\begin{eqnarray*}
\pi_\mathbf{List} &::=& \beta                             \\
                  &||&  \Nill                             \\
                  &||&  \textbf{(\ \_\ :\ }\pi_\mathbf{List}\textbf{)} \\
                  &||&  \star
\end{eqnarray*}
\caption{Shape abstraction for lists}
\label{LatticeList}
\end{figure}

In order to represent constraints generated by pattern-matching and applying a cons constructor to another expression, we need to be able to store annotation variables ($\beta$) in the recursive positions of the cons constructor.

As lists can be of infinite length, this is not yet sufficient to guarantee that the lattice formed by sets of abstracted lists satisfies the ascending chain condition. The constraint solver will thus use widening operators $\nabla : \varphi \to \varphi \to \varphi$ and $\Delta : \varphi \to \varphi \to \varphi$ to calculate the greatest lower and least upper bounds after instantiating annotation variables in an abstracted lists (see Section \ref{r532}), that will limit the depth of an abstract list by replacing the recursive position of the cons constructor at a user-specified depth by the wildcard $\star$. %%% TODO: explain semantics of \star, examples %%%

\section{Type System}

The analysis is defined by the \emph{constraint typing relation}\index{constraint typing relation} \[ \TypeRel{e}{\tauhat}{C}{R}. \] The relations says that expression $e$ has the annotated type $\tauhat$ in the typing environment $\widehat\Gamma$ if the type constraints $C$ and refinements constraints $R$ are satisfied.

The relation is defined by the typing rules given in Figure \ref{DeclarativeTypeSystem} and \ref{DeclarativeTypeSystem2}.

\paragraph{Implementation notes} The implementation of this analysis is also capable of handling simultaneous and mutually recursive functions and bindings. For each binding group, we build a graph with a vertex for each binding and edges from each usage of a particular binding to its definition. We determine the strongly connected components of this graph---each component being a group of mutually recursive bindings---and topologically sort the condensed graph to determine the order in which the let-bindings should be nested. Each mutually recursive group
\begin{code}
let  x_1 =  ldots  x_i  ldots  x_j  ldots
     x_2 =  ldots  x_i  ldots  x_j  ldots
     ldots
in   ldots
\end{code}
is then treated as a single recursive binding
\begin{code}
letrec  xr = (ldots pi_i xr ldots pi_j xr ldots, ldots pi_i xr ldots pi_j xr ldots, ldots)
in ldots
\end{code}
where $\pi_k$ projects the $k$th element from an $n$-tuple.

The implementation also supports tuples of arbitrary arity and list literals. Their type rules subsume T-Tuple and T-Nil.
%\clearpage

\section{Instantiation and Generalization}

As in the Hindley--Milner type system, we need to instantiate type schemes in the T-Var type rule and generalize types to a type scheme in the T-Let and \mbox{T-LetRec} type rules. This last process is called \emph{let-generalization}\index{generalization!let-generalization} and is what makes our analysis polyvariant. The addition of constraints---and necessity to quantify over them---makes the instantiation and generalization slight more complicated than it is in a plain Hindley--Milner type system.

When instantiating a type scheme we replace all type variables and annotation variables in a type scheme and which are bound by a quantifier with fresh variables and drop the quantifiers from the type scheme, resulting in a tuple containing an annotated type and its associated constraint set.

\begin{figure}[h!t]
\[ \frac{\theta = [\overline{\alpha \mapsto \alpha'}][\overline{\beta \mapsto \beta'}] \quad \overline{\alpha'},\overline{\beta'} \fresh }{inst(\forall \overline{\alpha}\overline{\beta}.\ R \Rightarrow \tauhat) = (\theta\tauhat, \theta R)} \]
\caption{Instantiation}
\end{figure}

During generalization we will unify the constraint set $C$ and apply the resulting substitution $\theta_\mathrm{C}$ to both $\tauhat$ and $R$ (as some of the annotation variables present in constraints in $R$ may have been unified as well.) Next, we invoke the constraint solver to solve or simplify a part of the constraint set $\theta_\mathrm{C} R$. We quantify over all annotation variables and type variables that are free in the environment and store the remaining constraints from $\theta_\mathrm{C} R$---those which were not solved---in the type scheme.

The essential detail here is choosing which constraints in $\theta_\mathrm{C} R$ to solve. We would like our analysis to be as polyvariant as possible and thus not prematurely fix the values of any annotation variables about which we can learn more later---where ``later'' should be taken to mean ``at the call site of a let-bound function,'' as we gain more information by applying arguments to it.

We thus exclude any constraints that directly or indirectly---through any transitive constraints, as determined by the dependency analysis (Section~\ref{depanal})---depend on annotation variables appearing in $\theta_\mathrm{C} \tauhat$ from being solved and store them in the type scheme instead, to be solved only at their call sites after having been instantiated again. We will call such constraints \emph{input-dependent constraints}.

\mathchardef\mhyphen="2D

\begin{figure}[h!t]
\[ \frac{\begin{gathered}\theta_\mathrm{C} = \widehat{\mathcal{U}}(C) \\ R' = input\mhyphen dependent(\theta_\mathrm{C} R, \theta_\mathrm{C} \tauhat)\quad \theta_\mathrm{R} = solve(\theta_\mathrm{C}R - R') \\ \overline{\alpha} = ftv(\theta_\mathrm{R}\theta_\mathrm{C}\tauhat) - ftv(\widehat\Gamma) \quad \overline{\beta} = fav(\theta_\mathrm{R}\theta_\mathrm{C}\tauhat) \cup fav(\theta_\mathrm{R}R') \end{gathered}}{gen(\widehat\Gamma, \tauhat, C, R) = \forall \overline{\alpha}\overline{\beta}.\ \theta_\mathrm{R}R' \Rightarrow \theta_\mathrm{R}\theta_\mathrm{C}\tauhat} \]
\caption{Generalization}
\end{figure}

\clearpage

\begin{figure}[h!t]
\begin{gather*}
\frac
    {\widehat\Gamma(x) = \widehat\sigma\quad inst(\widehat\sigma)=(\tauhat, R)}
    {\TypeRel{x}{\tauhat}{\emptyset}{R}}
    \mbox{\ [T-Var]} \\ \\  %%%%%%%%%%%%% CHECKED AGAINST CODE
\frac
    {\TypeRel[\widehat\Gamma, x:\alpha]{e}{\tauhat}{C}{R} \quad \alpha, \beta \fresh}
    {\TypeRel{\lambda x.\ e}{\alpha \overset{\beta}{\to} \tauhat}{C}{R\cup\{\top \subseteq \beta\}}}
    \mbox{\ [T-Abs]}\\ \\   %%%%%%%%%%%%  CHECKED AGAINST CODE
\frac
    {\TypeRel{f}{\tauhat_\mathrm{f}}{C_\mathrm{f}}{R_\mathrm{f}}\quad
     \TypeRel{x}{\tauhat_\mathrm{x}}{C_\mathrm{x}}{R_\mathrm{x}}\quad
     \alpha_\mathrm{r}, \beta \fresh}
    {\TypeRel{f\ x}{\alpha_\mathrm{r}}{C_\mathrm{f} \cup C_\mathrm{x} \cup \{\tauhat_\mathrm{f} = \tauhat_\mathrm{x} \overset{\beta}{\to} \alpha_\mathrm{r}\}}{R_\mathrm{f} \cup R_\mathrm{x} \cup \{\beta \subseteq \top\}}}
    \mbox{\ [T-App]}\\ \\  %%%%%%% CHECKED AGAINST CODE (MORE ELEGANT)
\frac
    {\beta \fresh}
    {\TypeRel{n}{\textbf{Int}^\beta}{\emptyset}{\{\{\iota(n)\} \subseteq \beta\}}}
    \mbox{\ [T-Int]} \\ \\ %%%%%%% CHECKED AGAINST CODE 
\frac
    {\beta \fresh}
    {\TypeRel{\textbf{True}}{\textbf{Bool}^\beta}{\emptyset}{\{\{\iota(\mathbf{True})\} \subseteq \beta\}}}
    \mbox{\ [T-True]} \\ \\%%%%%%% CHECKED AGAINST CODE 
\frac
    {\beta \fresh}
    {\TypeRel{\textbf{False}}{\textbf{Bool}^\beta}{\emptyset}{\{\{\iota(\mathbf{False})\} \subseteq \beta\}}}
    \mbox{\ [T-False]}\\ \\ %%%%%%% CHECKED AGAINST CODE 
\frac
    {\begin{gathered}\TypeRel{g}{\tauhat_\mathrm{g}}{C_\mathrm{g}}{R_\mathrm{g}} \quad     
     \beta \fresh \\
     \TypeRel{e_1}{\tauhat_1}{C_1}{R_1} \quad
     \TypeRel{e_2}{\tauhat_2}{C_2}{R_2} \\
     C = C_\mathrm{g} \cup C_1 \cup C_2 \cup \{\tauhat_\mathrm{g} = \Bool^\beta, \tauhat_1 = \tauhat_2\} \\
     R = R_\mathrm{g} \cup R_1 \cup R_2 \cup \{\beta \subseteq \{\TTrue, \FFalse\}\}
     \end{gathered}}
    {\TypeRel{\textbf{if\ }g\textbf{\ then\ }e_1\textbf{\ else\ }e_2}{\tauhat_1}{C}{R}}
    \mbox{\ [T-If]}\\ \\ %%%%%%% CHECKED AGAINST CODE 
\frac
    {\TypeRel{e_1}{\tauhat_1}{C_1}{R_1} \quad \TypeRel{e_2}{\tauhat_2}{C_2}{R_2} \quad \beta \fresh}
    {\TypeRel{\TupleOf{e_1}{e_2}}{\TupleOf{\tauhat_1}{\tauhat_2}^\beta}{C_1 \cup C_2}{R_1 \cup R_2 \cup \{ \Top \subseteq \beta \}}}
    \mbox{\ [T-Tuple]}\\ \\  %%%%%%% CHECKED AGAINST CODE 
\frac
    {\TypeRel{e}{\TupleOf{\tauhat_1}{\tauhat_2}^\beta}{C}{R}}
    {\TypeRel{\textbf{fst}\ e}{\tauhat_1}{C}{R \cup \{ \beta \subseteq \Top \}}}
    \mbox{\ [T-Fst]}\\ \\ %%%%%%% CHECKED (MORE ELOBARATE IN  CODE) 
\frac
    {\TypeRel{e}{\TupleOf{\tauhat_1}{\tauhat_2}^\beta}{C}{R}}
    {\TypeRel{\textbf{snd}\ e}{\tauhat_2}{C}{R \cup \{ \beta \subseteq \Top \}}}
    \mbox{\ [T-Snd]}  %%%%%%% CHECKED (MORE ELOBARATE IN  CODE)
\end{gather*}
\vfill
\caption{Type system (part 1)}
\label{DeclarativeTypeSystem}
\end{figure}
\begin{figure}[h!t]
\begin{gather*}
\frac
    {\alpha \fresh  \quad  \beta \fresh}
    {\TypeRel{\textbf{[]}}{\textbf{[}\alpha\textbf{]}^\beta}{\emptyset}{\{\{\textbf{[]}\}\subseteq\beta\}}}
    \mbox{\ [T-Nil]}\\ \\ %%%%%%% CHECKED (MORE ELEBORATE IN CODE, LIST LITERALS)
\frac
    {\begin{gathered}\TypeRel{e_\mathrm{h}}{\tauhat_\mathrm{h}}{C_\mathrm{h}}{R_\mathrm{h}} \quad \TypeRel{e_\mathrm{t}}{\tauhat_\mathrm{t}}{C_\mathrm{t}}{R_\mathrm{t}} \quad \beta_1, \beta_2 \fresh\\C = C_\mathrm{h} \cup C_\mathrm{t} \cup \{\ListOf{\tauhat_\mathrm{h}}^{\beta_2} = \tauhat_\mathrm{t}\}\quad R = R_\mathrm{h} \cup R_\mathrm{t} \cup \{ \{\Conss{\beta_2}\} \subseteq \beta_1 \}\end{gathered}}
    {\TypeRel{e_\mathrm{h} \textbf{\ :\ } e_\mathrm{t}}{\ListOf{\tauhat_\mathrm{h}}^{\beta_1}}{C}{R}}
    \mbox{\ [T-Cons]}\\ \\  %%%%%%%%%%%% CHECKED AGAINST CODE
%\frac
%    {\begin{gathered}\TypeRel{g}{\ListOf{\alpha}^\varphi}{C_\mathrm{g}}{R_\mathrm{g}} \quad \overline{\TypeRel[\widehat\Gamma,\overline{x_{ij}:\kappa_{ij}}]{e_i}{\tauhat_i}{C_i}{R_i}}\\C = C_\mathrm{g} \cup \left(\bigcup C_i\right)\cup\left(\bigcup \{\tauhat = \tauhat_i\}\right)\quad R = R_\mathrm{g} \cup \left(\bigcup R_i\right)\cup \{ \varphi \subseteq \bigcup K_i(\overline{x_{ij}})\}\end{gathered}}
%    {\TypeRel{|case|\ g\ |of|\ \{ \overline{K_i(\overline{x_{ij}}) \to e_i} \}}{\tauhat}{C}{R}}
%    \mbox{\ [T-Case]}\\ \\
\frac
    {\begin{gathered}\TypeRel{g}{\tauhat_\mathrm{g}}{C_\mathrm{g}}{R_\mathrm{g}}\quad \TypeRel{e_\mathrm{n}}{\tauhat_\mathrm{n}}{C_\mathrm{n}}{R_\mathrm{n}} \quad \alpha, \beta_1 \fresh \\ C = C_\mathrm{g} \cup C_\mathrm{n}\cup \{\tauhat_\mathrm{g} = \ListOf{\alpha}^{\beta_1}\} \quad R = R_\mathrm{g} \cup R_\mathrm{n} \cup \{ \beta_1 \subseteq \{\Nill\} \}\end{gathered}}
    {\TypeRel{|case|\ g\ |of|\ \{ \Nill \to e_\mathrm{n}\}}{\tauhat_\mathrm{n}}{C}{R}}
    \mbox{\ [T-Case-N]}\\ \\
\frac
    {\begin{gathered}\alpha, \beta_1, \beta_2 \fresh\\\TypeRel{g}{\tauhat_\mathrm{g}}{C_\mathrm{g}}{R_\mathrm{g}} \quad \TypeRel[\widehat\Gamma, h:\alpha,\,t:\ListOf{\alpha}^{\beta_2}]{e_\mathrm{c}}{\tauhat_\mathrm{c}}{C_\mathrm{c}}{R_\mathrm{c}} \\ C = C_\mathrm{g} \cup C_\mathrm{c}\cup \{\tauhat_\mathrm{g} = \ListOf{\alpha}^{\beta_1}\} \quad R = R_\mathrm{g} \cup R_\mathrm{c} \cup \{ \beta_1 \subseteq \{\Conss{\beta_2}\} \}\end{gathered}}
    {\TypeRel{|case|\ g\ |of|\ \{ \textbf{(\,}h\textbf{\,:\,}t\textbf{\,)} \to e_\mathrm{c} \}}{\tauhat_\mathrm{c}}{C}{R}}
    \mbox{\ [T-Case-C]}\\ \\
\frac
    {\begin{gathered}\TypeRel{g}{\tauhat_\mathrm{g}}{C_\mathrm{g}}{R_\mathrm{g}} \quad \alpha, \beta_1, \beta_2 \fresh \\\TypeRel{e_\mathrm{n}}{\tauhat_\mathrm{n}}{C_\mathrm{n}}{R_\mathrm{n}} \quad \TypeRel[\widehat\Gamma, h:\alpha,\,t:\ListOf{\alpha}^{\beta_2}]{e_\mathrm{c}}{\tauhat_\mathrm{c}}{C_\mathrm{c}}{R_\mathrm{c}}\\C = C_\mathrm{g} \cup C_\mathrm{n}\cup C_\mathrm{c}\cup \{\tauhat_\mathrm{g} = \ListOf{\alpha}^{\beta_1}, \tauhat_\mathrm{n} = \tauhat_\mathrm{c}\} \\ R = R_\mathrm{g} \cup R_\mathrm{n} \cup R_\mathrm{c} \cup \{ \beta_1 \subseteq \{\Nill, \Conss{\beta_2}\} \}\end{gathered}}
    {\TypeRel{|case|\ g\ |of|\ \{ \Nill \to e_\mathrm{n}; \textbf{(\,}h\textbf{\,:\,}t\textbf{\,)} \to e_\mathrm{c} \}}{\tauhat_\mathrm{n}}{C}{R}}
    \mbox{\ [T-Case-NC]}\\ \\
\frac
    {\begin{gathered}\TypeRel{e_1}{\tauhat_1}{C_1}{R_1}\quad gen(\widehat\Gamma, \tauhat_1, C_1, R_1) = \widehat\sigma \\ \TypeRel[\widehat\Gamma, x:\widehat\sigma]{e_2}{\tauhat_2}{C_2}{R_2}\end{gathered}}
    {\TypeRel{|let|\ x = e_1\ |in|\ e_2}{\tauhat_2}{C_2}{R_2}}
    \mbox{\ [T-Let]}\\ \\
\frac
    {\begin{gathered}\TypeRel{\lambda f.\ e_1}{\tauhat_\mathrm{r}}{C_1}{R_1}\quad gen(\widehat\Gamma, \tauhat_1, C_1\cup\{\tauhat_1\to\tauhat_1=\tauhat_\mathrm{r}\}, R_1) = \widehat\sigma \\ \TypeRel[\widehat\Gamma, x:\widehat\sigma]{e_2}{\tauhat_2}{C_2}{R_2}\end{gathered}}
    {\TypeRel{|let|\ f = e_1\ |in|\ e_2}{\tauhat_2}{C_2}{R_2}}
    \mbox{\ [T-LetRec]}\\ \\
\end{gather*}
\caption{Type system (part 2)}
\label{DeclarativeTypeSystem2}
\end{figure}

\clearpage

\section{Unification}\label{unifi}

To solve the constraint set $C$ we use a unification algorithm $\widehat{\mathcal{U}}$ that either computes a substitution $\theta$, such that $\theta C$ contains only trivial equalities of the form $\tauhat = \tauhat$ or fails if the constraint set is unsolvable and the program is type incorrect.

\begin{eqnarray*}
\unify &:& \mathcal{P}(\textbf{Constr}) \to \textbf{Subst} \\
\unify (\emptyset) &=& \id \\
\unify (\{\alpha = \tauhat\} \cup C) &=& \leet \theta =
    \begin{dcases*}
            \left[\alpha \mapsto \tauhat\right] & if $\alpha \notin \ftv(\tauhat)$ \\ 
            \textbf{fail} & otherwise
    \end{dcases*}\\
    & & \iin \unify(\theta C) \circ \theta \\
\unify (\{\tauhat = \alpha\} \cup C) &=& \leet \theta =
    \begin{dcases*}
            \left[\alpha \mapsto \tauhat\right] & if $\alpha \notin \ftv(\tauhat)$ \\ 
            \textbf{fail} & otherwise
    \end{dcases*}\\
    & & \iin \unify(\theta C) \circ \theta \\
\unify (\{\tauhat_1 \overset{\varphi_1}{\to} \tauhat_2 = \tauhat_1' \overset{\varphi_2}{\to} \tauhat_2'\} \cup C) &=&
    \leet \theta = \uniphi(\varphi_1 = \varphi_2)\\
& & \iin \unify(\theta\{\tauhat_1 = \tauhat_1', \tauhat_2 = \tauhat_2'\} \cup \theta C) \\    
\unify (\{\Bool^{\varphi_1} = \Bool^{\varphi_2}\} \cup C) &=& \leet \theta = \mathcal{V}(\varphi_1 = \varphi_2)\\& & \iin \unify(\theta C) \circ \theta \\
\unify (\{\Int^{\varphi_1} = \Int^{\varphi_2}\} \cup C) &=& \leet \theta = \mathcal{V}(\varphi_1 = \varphi_2)\\& & \iin \unify(\theta C) \circ \theta \\
\unify (\{\TupleOf{\tauhat_1}{\tauhat_2}^{\varphi_1} = \TupleOf{\tauhat_1'}{\tauhat_2'}^{\varphi_2}\} \cup C) &=&
    \leet \theta = \uniphi(\varphi_1 = \varphi_2)\\
& & \iin \unify(\theta\{\tauhat_1 = \tauhat_1', \tauhat_2 = \tauhat_2'\} \cup \theta C) \\
\unify (\{\ListOf{\tauhat}^{\varphi_1} = \ListOf{\tauhat'}^{\varphi_2}\} \cup C) &=&
    \leet \theta = \uniphi(\varphi_1 = \varphi_2)\\
& & \iin \unify(\theta\{\tauhat = \tauhat'\} \cup \theta C)\\
\mbox{otherwise} &=& \textbf{fail}
\end{eqnarray*}

The unification algorithms $\unify$ relies on a unification algorithm $\uniphi$ that unifies annotations.

\begin{eqnarray*}
\mathcal{V} (\varphi_1 = \varphi_2) =
    \begin{dcases*}
        \id                                      & if $\varphi_1 = \varphi_2$ \\
        \left[\varphi_1 \mapsto \varphi_2\right] & otherwise
    \end{dcases*}
\end{eqnarray*}

This unification algorithm $\uniphi$ is straightforward, because the constraints in $C$ only concern simply annotated types---remember that these are annotated types where the annotations can only be annotation variables and nothing else.

The resulting substitution can contain both mappings $\left[\alpha \mapsto \tauhat\right]$ from type variables to annotated types and mappings $\left[\beta_1 \mapsto \beta_2\right]$ from annotation variables to annotation variables.

%\subsection{Algorithm $\widehat{\mathcal{W}}$}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% CONSTRAINT SOLVING %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Constraint Solving}\label{chapcs}

\section{Overview}

The constraint generation phase of the analysis generates two constraint sets: a constraint set $C$ containing equality constraints between simply annotated types and a constraint set $R$ containing subtype constraints between the annotations. As we have seen the constraint set $C$ can be solved by a slightly modified, but fairly traditional unification algorithm, resulting in a substitution $\theta_C$ from type variables to annotated types and from annotation variables to annotation variables.

The first step in solving the constraints set $R$ is to apply the substitution $\theta_C$ to it, making sure any annotation variables that have been unified while solving $C$ are also unified in $R$.

The constraints in $R$ can have various forms, for example: producing a $\TTrue$ will introduce a constraint \[ \{\TTrue\} \subseteq \beta \] for a fresh annotation variable $\beta$. Pattern matching on an empty list (and nothing else) will generate a constraint \[ \beta \subseteq \{\Nill\} \]

Recursive data types, such as lists, give rise to a more complicated form of \emph{transitive constraints}. For example, pattern matching on a list will generate a constraint \[ \beta_1 \subseteq \{\Nill, \Conss{\beta_2}\} \] where $\beta_1$ is the annotation variable on the scrutinee of the case-expression and $\beta_2$ the annotation on the tail of the list. Intuitively, if we learn something about $\beta_1$ this might influence what we know about $\beta_2$ and vice versa.

To solve the constraint set $R$ we need to find for each annotation variable $\beta$ on a type $\tau$ an (as large as possible) interval $I = (l, u) \in \mathcal{L}^2$ of the lattice $\mathcal{L}$ abstracting the underlying type of the annotation variable that is consistent with the constraint set. As the starting interval we take the whole of the lattice $(\bot_\mathcal{L}, \top_\mathcal{L})$. We then apply a worklist algorithm that will have constraints of the form $\{\TTrue\} \subseteq \beta_1$ push the lowerbound $l$ of the interval $I_1$ belonging to $\beta_1$ upwards and constraints of the form $\beta_2 \subseteq \{\Nill\}$ the upperbound $u$ of the interval $I_2$ belonging to $\beta_2$ downwards. Transitive constraints such as $\beta_1 \subseteq \{\Nill, \Conss{\beta_2}\}$ will affect the upper- and lowerbounds of both $\beta_1$ and $\beta_2$.

\section{Dependency Analysis}\label{depanal}

\emph{Transitive constraints}\index{constraint!transitive}---such as $\beta_1 \subseteq \beta_2$---which contain annotation variables on both the left- and right-hand side of the constraint introduce dependencies between constraints.

The dependency analysis phase of the analysis determines the dependencies between subset constraints and annotation variables by building two dependency graphs. One between annotation variables and one from annotation variables to subset constraints. These two graphs are used when determining the input-independent constraints and by the worklist algorithm described next.

\section{Worklist Algorithm}

\subsection{Generic Worklist Algorithm}

The constraints solver for the constraint set $R$ is built on top of a generic worklist algorithm:

\begin{code}
worklist :: (a -> b -> ([a], b)) -> b -> [a] -> b
worklist _  r  []      = r
worklist f  r  (x:xs)  = let (ys, r') = f x r in worklist f r' (xs ++ ys)
\end{code}

The function argument |f| takes a takes a constraint |x :: a| and intermediate result |r :: b| and returns a new intermediate result |r' :: b| and an additional set of constraints |ys :: [a]| the worklist algorithm has to add to its current working set.

In our application of this generic worklist algorithm these additional constraints will be constraints that have been previously processed, but for which the dependency checker has determined they may need to be looked at again, because one or more of their dependent variables have been updated.

While the generic worklist algorithm does not necessarily terminate, it will in our application. Abstract values are modelled by lattices satisfying the ascending chain condition, meaning the dependency analysis will eventually stop feeding constraints into the worklist when all dependent variables have reached their fixed points.

\subsection{Resolving Individual Constraints}\label{r532}

The function we pass as the argument |f| to |worklist| will---after having assisted with dependency checking and updating---invoke the function |resolveConstr|, which updates the intermediate result by making it consistent with an individual constraint.

In its most generic shape such a constraint will be of the form \[\beta_1 \cup \beta_2 \cup ... \cup \beta_m \cup \{K_1,K_2,...,K_n\} \subseteq \beta'_1 \cup\beta'_2 \cup ... \cup \beta'_{m'} \cup \{K'_1,K'_2,...,K'_{n'}\} \] where $\beta_i, \beta_{i'}$ are annotation variables and $K_j, K_{j'}$ are elements of some lattice $\pi_\tau$---they can be thought of as constructors---possibly containing more annotation variables within them.

In order to make the intermediate result consistent with this constraint we may need to do two things:
\begin{enumerate}
    \item Raise the lowerbound of refinement variables present---either directly as one of the $\beta_{i'}$s or inside one of the constructors $K'_{j'}$---in the right-hand side of the constraint;
    \item Lower the upperbound of refinement variables present in the left-hand side of the constraint.
\end{enumerate}

In order to accomplish the first |resolveConstr| will:
\begin{enumerate}
    \item Instantiate all refinement variables on the left-hand side of the constraint---including those inside constructors---with the current lowerbound recorded in the intermediate result. Call this $L$. (For constructors we assume a widening operator is applied during instantiation if this is required to maintain the ascending chain condition of their corresponding lattice.)
    \item For each refinement variable $\beta'_{i'}$ on the right-hand side |resolveConstr| will update $\theta_\mathrm{L}(\beta'_{i'})$, the lowerbound for that variable as recorded in the intermediate result, to $\theta_\mathrm{L}(\beta'_{i'}) \sqcup L$.
    \item For each constructor $K'_{j'}$ on the right-hand side |resolveConstr| will make new constraints \[ \Pi_{K_{j'}}^p(L) \subseteq \Pi_{K_{j'}}^p(\{K'_{j'}\}) \] where $\Pi_{K}^p$ projects out the $p$th field of a $K$-constructor or gives an empty result if its input is not a $K$-constructor. This whole procedure will then be recursively applied to the newly generated constraints.
\end{enumerate}

In order to accomplish the second we can apply the same procedure \emph{mutatis mutandis}; in particular the updated value of the intermediate result in the second step should now become $\theta_\mathrm{U}(\beta_i) \sqcap R$.

\paragraph{Example} Assume $\theta(\beta_1) = (\{\Nill, \Conss{\Nill}, \Conss{\Conss{\Nill}}\}, \top)$ and $\theta(\beta_2) = (\bot, \top)$. Given the constraint
\begin{eqnarray}
\{\beta_1\} \subseteq \{\Conss{\beta_2}\}\label{eqco}
\end{eqnarray}
|resolveConstr| will instantiate the left-hand side to 
\begin{eqnarray}
\{\Nill, \Conss{\Nill}, \Conss{\Conss{\Nill}}\} \subseteq \{\Conss{\beta_2}\}
\end{eqnarray}

As there is both a $\Nill$-constructor and a $\Conss{\cdot}$-constructor on the right-hand side, |resolveConstr| will partition the constraint into two constraints
\begin{eqnarray}
\{ \Nill \} &\subseteq& \emptyset \label{eq1} \\
\{ \Conss{\Nill}, \Conss{\Conss{\Nill}} \} &\subseteq& \{ \Conss{\beta_2} \} \label{eq2}
\end{eqnarray}
As the right-hand side of (\ref{eq1}) is empty this constraint is discarded. Both left and right-hand side of (\ref{eq2}) have top-level $\Conss{\cdot}$-constructors. We project out their fields, resulting in the constraint
\begin{eqnarray}
\{ \Nill, \Conss{\Nill} \} &\subseteq& \{ \beta_2 \} \label{eq3}
\end{eqnarray}

To solve this constraint |resolveConstr| will be recursively called on it. As there is only a refinement variable on the right-hand side of (\ref{eq3}) |resolveConstr| will update $\theta_\mathrm{L}(\beta_2)$ to \[\bot \sqcup \{\Conss{\Nill}, \Conss{\Conss{\Nill}}\} = \{\Conss{\Nill}, \Conss{\Conss{\Nill}}\}\]

We have now ``pushed up'' the lowerbound of $\beta_2$. Next, we ``push down'' the upperbound of $\beta_1$. We begin by instantiating the right-hand side of (\ref{eqco}) to
\begin{eqnarray}
\{\beta_1\} &\subseteq& \{\Conss{\Nill}, \Conss{\Conss{\Nill}}, ..., \Conss{...\Conss{\Nill}...},  \Conss{...\Conss{\star}...}\}
\end{eqnarray}
We immediately end up with a constraint of the form $\beta \subseteq \{ ... \}$, so we do not need to recurse and can update $\theta_\mathrm{U}(\beta_1)$ to
\begin{multline*}
\top \sqcap \{\Conss{\Nill}, \Conss{\Conss{\Nill}}, ..., \Conss{...\Conss{\Nill}...},  \Conss{...\Conss{\star}...}\} \\ = \{\Conss{\Nill}, \Conss{\Conss{\Nill}}, ..., \Conss{...\Conss{\Nill}...},  \Conss{...\Conss{\star}...}\}
\end{multline*}

The resulting intervals are now
\begin{eqnarray*}
\theta(\beta_1) &=& (\{\Nill, \Conss{\Nill}, \Conss{\Conss{\Nill}}\}, \{\Conss{\Nill}, \Conss{\Conss{\Nill}}, ...,  \Conss{...\Conss{\star}...}\}) \\
\theta(\beta_2) &=& (\{\Conss{\Nill}, \Conss{\Conss{\Nill}}\}, \top)
\end{eqnarray*}

As the lowerbound of $\beta_1$ is no longer contained in its upperbound---the lowerbound includes a $\Nill$-constructor, while the upperbound does not---we have detected a possible pattern match failure.

% Note that {[]} <= {(_:[])} does not occur

### 2015-09-01 deprecated/pattern-match-analysis/notes/Bitstring.hs

@@ -1,0 +1,11 @@
+module Main where
+
+type Bitstring = [Int]
+
+add :: Bitstring -> Bitstring -> Bitstring
+add [] y = y
+add x [] = x
+add (0:x) (0:y) = 0 : add x y
+add (0:x) (1:y) = 1 : add x y
+add (1:x) (0:y) = 1 : add x y
+add (1:x) (1:y) = 0 : add (add [1] x) y

### 2015-09-01 deprecated/pattern-match-analysis/notes/Desugar.hs

@@ -1,0 +1,62 @@
+module Main where
+
+pmfail = error "pattern-match failure"
+
+f1 (a1:as)    | a1 >= 20 = 'A'
+f1 (a1:a2:as) | a2 >= 30 = 'B'
+f1 (a1:as)    | a1 >= 10 = 'C'
+
+f2 (a1:as)    | a1 >= 20 = 'A'
+f2 (a1:as)    | a1 >= 10 = 'C'
+f2 (a1:a2:as) | a2 >= 30 = 'B'
+
+test = f1 [10,30] == f2 [10,30]
+
+f' a = case a of
+        (a1:as) -> if a1 >= 20
+                     then 'a'
+                     else case as of
+                            []       -> case a of
+                                            (a1x:asx) -> if a1x >= 10
+                                                            then 'c'
+                                                            else pmfail
+                            (a2:as') -> if a2 >= 30
+                                            then 'b'
+                                            else pmfail
+
+
+{-
+
+f P1 | G1 = E1
+f P2 | G2 = E3
+f P3 | G3 = E3
+
+    ~~>
+
+f x = let C1 = case x of
+                 P1 -> if G1 then E1 else C2
+                 _  -> C2
+          C2 = case x of
+                 P2 -> if G2 then E2 else C3
+                 _  -> C3
+          C3 = case x of
+                 P3 -> if G3 then E3 else C4
+                 _  -> C4
+          C4 = pmfail
+       in C1
+
+    ~~>
+    
+f x = let C4 = pmfail
+       in let C3 = case x of
+                     P3 -> if G3 then E3 else C4
+                     _  -> C4
+           in let C2 = case x of
+                         P2 -> if G2 then E2 else C3
+                         _  -> C3
+               in let C1 = case x of
+                             P1 -> if G1 then E1 else C2
+                             _  -> C2
+                   in C1
+
+-}

### 2015-09-01 deprecated/pattern-match-analysis/notes/RefTy.hs

@@ -1,0 +1,39 @@
+module RefTy where
+
+import Data.List
+
+type ConName   = String
+type TyName    = String
+type RefTyName = String
+type TyVar     = String
+type RefTyVar  = String
+
+data DataCon  = DataCon ConName [TyName]
+data DataType = DataType TyName [DataCon]
+
+instance Show DataCon where
+    show (DataCon conName tns) = conName ++ " " ++ concat (intersperse " " tns)
+    
+instance Show DataType where
+    show (DataType tyName dcs) = "data " ++ tyName ++ " = " ++ concat (intersperse " | " (map show dcs))
+
+data RefTy = RefTy :/\: RefTy   -- intersection type
+           | RefTy :\/: RefTy   -- union type
+           | RefTy :->: RefTy   -- function type
+           | Bottom             -- bottom
+           | A TyName [RefTy]
+           | B RefTyName [RefTy]
+           | RefTyVar ::: TyVar
+
+instance Show RefTy where
+    show (t1 :/\: t2)   = "(" ++ show t1 ++ " ∧ " ++ show t2 ++ ")"
+    show (t1 :\/: t2)   = "(" ++ show t1 ++ " ∨ " ++ show t2 ++ ")"
+    show (t1 :->: t2)   = "(" ++ show t1 ++ " → " ++ show t2 ++ ")"
+    show (Bottom    )   = "⊥"
+    show (A tyName rts) = tyName ++ " " ++ concat (intersperse " " (map show rts))
+    show (B refTyName rts) = refTyName ++ " " ++ concat (intersperse " " (map show rts))
+    show (rtv ::: tv)   = rtv ++ " :: " ++ tv
+    
+-- | Examples
+
+dataBitstr = DataType "Bitstr" [DataCon "E" [], DataCon "Z" ["Bitstr"], DataCon "O" ["Bitstr"]]

### 2015-09-01 deprecated/pattern-match-analysis/presentation/build.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+lhs2TeX presentation.tex > presentation.lhs2TeX.tex
+pdflatex presentation.lhs2TeX.tex && pdflatex presentation.lhs2TeX.tex && mv presentation.lhs2TeX.pdf presentation.pdf && evince presentation.pdf

### 2015-09-01 deprecated/pattern-match-analysis/sample/Sample.hs

@@ -1,0 +1,365 @@
+-- {-# LANGUAGE NoImplicitPrelude #-}
+-- {-# LANGUAGE NoMonomorphismRestriction #-}
+
+module Test where
+{-
+data Bool
+    = True
+    | False
+-}
+{-
+data Either a b
+    = Left a
+    | Right b
+-}
+{-
+data Pair a b
+    = Pair { fst :: a, snd :: b }
+-}
+
+-- | Patterns
+
+-- * Real
+{- head
+main = \x -> case x of
+                (a:as) -> as
+-}
+
+-- * Complex
+{-
+main = \x -> case x of
+                Left [a]     -> [a,True]
+                Right (a, 4) -> [a,a]
+-}
+{-
+main = \x -> case x of
+                Left [a]     -> 1
+                Right (a, 4) -> 2
+-}
+
+-- | Either
+
+-- * Case-statement
+
+{-
+main = \x -> case x of
+                Left  a -> Right a
+                Right b -> Left  b
+-}
+{-
+main = \x -> case x of
+                Left  a -> a
+                Right b -> b
+-}
+{-
+main = \x -> case x of
+                Left  a -> 10
+                Right b -> 20
+-}      
+{-
+main = case Left True of
+        Left  a -> 10
+        Right b -> 20
+-}
+
+-- * App
+
+-- main = \x -> Left x
+
+-- * Type error
+
+-- main = \x -> if x then Left True else Left 42
+
+-- * Non-app
+
+-- main = \x -> if x then Left True else Left False
+-- main = \x -> if x then Left True else Right 42
+-- main = \x -> if x then Left True else Right False
+-- main = Left True
+
+-- | Undefined
+
+-- main = \x -> error x
+-- main = error
+-- main = undefined
+
+-- | Let-bindings
+
+-- * Mutually recursive
+{-
+main = let f = g
+           g = h
+           h = f
+        in h
+-}
+{-
+main = let f = g
+           g = f
+        in f
+-}
+
+-- * Simultaneous
+{-
+main = let a = b
+           b = 42
+        in b
+-}
+{-
+main = let a = b
+           b = 42
+        in a
+-}
+{-
+main = let a = 0
+           b = 1
+        in b
+-}
+{-
+main = let a = 0
+           b = 1
+        in a
+--}
+-- * Recursion & Generalization
+{-
+main = let di = \a -> let b = di a in a
+        in (di True, di 4, di)
[…]

### 2015-09-01 deprecated/pattern-match-analysis/sample/SampleFilter.hs

@@ -1,0 +1,178 @@
+{-# LANGUAGE NoMonomorphismRestriction #-}
+
+module Test where
+
+-- | Neil Mitchell
+
+-- * Risers
+{-
+main = let risers = \x -> case x of 
+                            [] -> []
+                            (y:ys) -> case ys of
+                                        [] -> (y : []) : []
+                                        (z:zs) -> risers2 (risers3 z zs) (y <= z) y
+           risers2 = \x -> \y -> \z -> if y then
+                                        (z : snd x) : (fst x)
+                                       else
+                                        (z : []) : (snd x : fst x)
+           risers3 = \x -> \y -> risers4 (risers (x : y))
+           risers4 = \x -> case x of
+                            (y:ys) -> (ys, y)
+        in risers
+-}
+
+-- | Prelude
+
+-- * Filter
+{-
+
+filter :: (a -> Bool) -> [a] -> [b]
+filter _ []     = []
+filter p (x:xs) | p x       = x : filter p xs
+                | otherwise =     filter p xs
+
+---}
+
+{-  <<<<<<<<<<<<<<<<<<
+-- main :: (a -> Bool) -> [a] -> [a]
+main = let filter = \p -> \ys -> case ys of
+                                   [] -> []
+                                   (x:xs) -> let g = p x
+                                              in if g then x : filter p xs
+                                                      else     filter p xs
+       in filter (\x -> x) [True, False]
+-}
+
+{-
+-- main :: (a -> Bool) -> [a] -> [a]
+main = let filter = \p -> \ys -> case ys of
+                                   [] -> []
+                                   (x:xs) -> let g = p x
+                                              in if g then x : filter p xs
+                                                      else     filter p xs
+       in filter
+-}
+
+
+{-
+-- main :: (a -> Bool) -> [a] -> [a]
+main = let filter = \p -> \ys -> case ys of
+                                   [] -> []
+                                   (x:xs) -> let g = p x
+                                              in if g then x : xs
+                                                      else     xs
+       in filter
+-}
+
+{-
+-- main :: (a -> Bool) -> [a] -> [a]
+main = \p -> \ys -> case ys of
+                        [] -> []
+                        (x:xs) -> let g = p x
[…]

### 2015-09-01 deprecated/pattern-match-analysis/sample/SampleS.hs

@@ -1,0 +1,29 @@
+{-# LANGUAGE NoImplicitPrelude #-}
+-- {-# LANGUAGE NoMonomorphismRestriction #-}
+
+module Test where
+
+data Bool
+    = True
+    | False
+
+data Either a b
+    = Left a
+    | Right b
+    
+-- | rt-unit
+
+-- main = ()
+
+-- | rt-con
+
+-- main = True
+-- main = 42
+
+-- | rt-var
+
+-- main = let x = 13 in x 
+
+-- | rt-fun
+
+main = let id = (\x -> x) in id

### 2015-09-01 deprecated/pattern-match-analysis/sample/Thesis.hs

@@ -1,0 +1,76 @@
+{-
+main = \b -> \phi -> if b
+                     then if phi 42 then 100 else 200
+                     else if phi 43 then 300 else 400
+-}
+
+--{-
+main = let f = \b -> \phi -> if b
+                             then if phi [42] then 100 else 200
+                             else if phi [43] then 300 else 400
+           g = \x -> case x of
+                        []     -> True
+           h = \x -> case x of
+                        (a:as) -> True
+        in f True g
+---}
+
+{-
+main = let pm = \a -> case a of
+                         -- []      -> [0]
+                         (a:as)  -> case as of
+                                      []      -> [0,1]
+                                      (b:bs)  -> [1,2,3]
+        in pm
+-}
+
+{-
+main = let head = \xss -> case xss of
+                            (x:xs) -> x
+        in head
+-}
+
+{-
+main = let tail = \xss -> case xss of
+                            (x:xs) -> xs
+        in tail
+-}
+
+{- (not precise enough)
+main = let f = \b -> let tail = \xss -> case xss of
+                                           (x:xs) -> xs
+                      in tail (if b then [] else [1,2,3])
+        in f False
+-}
+
+{-
+main = let map4 = \f -> \xss -> case xss of
+                                 []     -> []
+                                 (x:xs) -> f x : map4 f xs
+        in map4 (\x -> x) [1,2,3]
+-}
+
+{-
+main = let filter = \p -> \ys -> case ys of
+                                   [] -> []
+                                   (x:xs) -> let g = p x
+                                              in if g then x : filter p xs
+                                                      else     filter p xs
+       in filter (\x -> x) [True, False]
+-}
+
+{-
+main = let risers = \x -> case x of 
+                            [] -> []
+                            (y:ys) -> case ys of
+                                        [] -> (y : []) : []
+                                        (z:zs) -> risers2 (risers3 z zs) (y <= z) y
+           risers2 = \x -> \y -> \z -> if y then
+                                        (z : snd x) : (fst x)
+                                       else
+                                        (z : []) : (snd x : fst x)
+           risers3 = \x -> \y -> risers4 (risers (x : y))
+           risers4 = \x -> case x of
+                            (y:ys) -> (ys, y)
+        in risers
+-}

### 2015-09-01 deprecated/pattern-match-analysis/src/Abstract/Bool.hs

@@ -1,0 +1,40 @@
+module Abstract.Bool where
+
+import Data.List (intersperse)
+import Data.Set
+
+newtype AbsBool
+    = AbsBool (Set Bool)
+    deriving (Eq, Ord)
+    
+instance Show AbsBool where
+    show (AbsBool s) = "{" ++ concat (intersperse "," (fmap show (toList s))) ++ "}"
+
+injectBool :: Bool -> AbsBool
+injectBool = AbsBool . singleton
+
+mergeAbsBool :: AbsBool -> AbsBool -> AbsBool
+mergeAbsBool (AbsBool s) (AbsBool t) = AbsBool (s `union` t)
+
+meetAbsBool :: AbsBool -> AbsBool -> AbsBool
+meetAbsBool (AbsBool s) (AbsBool t) = AbsBool (s `intersection` t)
+
+subsetAbsBool :: AbsBool -> AbsBool -> Bool
+subsetAbsBool (AbsBool a) (AbsBool b) = a `isSubsetOf` b
+
+topAbsBool :: AbsBool
+topAbsBool = injectBool True `mergeAbsBool` injectBool False
+
+botAbsBool :: AbsBool
+botAbsBool = AbsBool empty
+
+-- | "Primitive" operators
+
+absBoolNot :: AbsBool -> AbsBool
+absBoolNot (AbsBool xs) = AbsBool (fromList [not x | x <- toList xs])
+
+absBoolAnd :: AbsBool -> AbsBool -> AbsBool
+absBoolAnd (AbsBool xs) (AbsBool ys) = AbsBool (fromList [x && y | x <- toList xs, y <- toList ys])
+
+absBoolOr :: AbsBool -> AbsBool -> AbsBool
+absBoolOr (AbsBool xs) (AbsBool ys) = AbsBool (fromList [x || y | x <- toList xs, y <- toList ys])

### 2015-09-01 deprecated/pattern-match-analysis/src/Abstract/Either.hs

@@ -1,0 +1,23 @@
+module Abstract.Either (AbsEither, mkAbsEither, joinAbsEither, absEitherContains) where
+
+import Prelude
+import Data.List (intersperse)
+import Data.Set hiding (map)
+
+import Abstract.Bool
+
+newtype AbsEither = AbsEither (Set (Either AbsBool AbsBool))
+    deriving (Eq, Ord)
+
+mkAbsEither :: Either Bool Bool -> AbsEither
+mkAbsEither (Left  b) = AbsEither (singleton (Left  (mkAbsBool b)))
+mkAbsEither (Right b) = AbsEither (singleton (Right (mkAbsBool b)))
+
+joinAbsEither :: AbsEither -> AbsEither -> AbsEither
+joinAbsEither (AbsEither a) (AbsEither b) = AbsEither (a `union` b)
+
+absEitherContains :: AbsEither -> AbsEither -> Bool
+absEitherContains (AbsEither a) (AbsEither b) = a `isSubsetOf` b
+
+instance Show AbsEither where
+    show (AbsEither a) = "{" ++ concat (intersperse "," (map show (toList a))) ++ "}"

### 2015-09-01 deprecated/pattern-match-analysis/src/Abstract/Int.hs

@@ -1,0 +1,134 @@
+module Abstract.Int {- (AbsInt, injectInt, mergeAbsInt, subsetAbsInt) -} where
+
+import Data.List (intersperse)
+import Data.Set hiding (map)
+
+import Abstract.Bool
+
+{- -- | Set-abstracted integer
+
+newtype AbsInt = AbsInt (Set Integer)
+    deriving (Eq, Ord)
+
+mkAbsInt :: Integer -> AbsInt
+mkAbsInt n = AbsInt (singleton n)
+
+joinAbsInt :: AbsInt -> AbsInt -> AbsInt
+joinAbsInt (AbsInt a) (AbsInt b) = AbsInt (a `union` b)
+
+instance Show AbsInt where
+    show (AbsInt a) = "{" ++ concat (intersperse "," (map show (toList a))) ++ "}"
+-}
+
+-- | Sign-abstracted integer
+
+data Sign = Neg | Zero | Pos
+    deriving (Eq, Ord)
+
+instance Show Sign where
+    show Neg  = "-"
+    show Zero = "0"
+    show Pos  = "+"
+
+newtype AbsInt = AbsInt (Set Sign)
+    deriving (Eq, Ord)
+    
+instance Show AbsInt where
+    show (AbsInt s) = "{" ++ concat (intersperse "," (map show (toList s))) ++ "}"
+
+injectInt :: Integer -> AbsInt
+injectInt = AbsInt . singleton . int2sign
+
+int2sign :: Integer -> Sign
+int2sign n = case signum n of
+                -1 -> Neg
+                0  -> Zero
+                1  -> Pos
+                
+mergeAbsInt :: AbsInt -> AbsInt -> AbsInt
+mergeAbsInt (AbsInt a) (AbsInt b) = AbsInt (a `union` b)
+
+subsetAbsInt :: AbsInt -> AbsInt -> Bool
+subsetAbsInt (AbsInt a) (AbsInt b) = a `isSubsetOf` b
+
+topAbsInt :: AbsInt
+topAbsInt = injectInt (-1) `mergeAbsInt` injectInt 0 `mergeAbsInt` injectInt 1
+
+-- | Primitive operators
+
+absIntPlus :: AbsInt -> AbsInt -> AbsInt
+absIntPlus (AbsInt xs) (AbsInt ys) = AbsInt (unions [signPlus x y | x <- toList xs, y <- toList ys])
+    where signPlus :: Sign -> Sign -> Set Sign
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/Abstract/List.hs

@@ -1,0 +1,135 @@
+{-# LANGUAGE ScopedTypeVariables #-}
+
+-- FIXME: Cons (ListShape var) ~> Cons (AbsList var)
+-- FIXME: Move ListVar into AbsList
+-- FIXME: Reconsider ListVar vs. RefVar?
+
+module Abstract.List {-(AbsList, ListShape, injectList, mergeAbsList, meetAbsList, absCons, substAbsList, instAbsList, headAbsList, getVarsAbsList, subsetAbsList, botAbsList, topAbsList, nonNilAbsList, consAbsList)-} where
+
+import Data.List (intersperse)
+import Data.Map (Map)
+import qualified Data.Map as Map
+import Data.Set hiding (map, null)
+import qualified Data.Set as Set
+
+import Language.Haskell.Exts.Pretty
+
+import Util
+
+-- | Widening configuration
+
+maxDepth = 5
+
+-- | List shapes
+
+data ListShape var -- var is needed to avoid mutual recursion between data types
+                   -- I'd rather keep in separate modules... (but, "Ord var =>")
+    = Nil
+    | Cons (ListShape var)
+    | Star
+--  | Infinite (= Star\Cons)
+    | ListVar var
+    deriving (Eq, Ord)
+    
+instance Pretty var => Show (ListShape var) where
+    show Nil            = "[]"
+    show (Cons ls)      = "_:" ++ show ls
+    show Star           = "*"
+    show (ListVar name) = prettyPrint name
+    
+-- | Abstract lists
+    
+newtype AbsList var
+    = AbsList (Set (ListShape var))
+    deriving (Eq, Ord)
+
+instance (Ord var, Pretty var) => Show (AbsList var) where
+    show l@(AbsList a) | l == topAbsList = "T_List"
+                       | otherwise       = "{" ++ concat (intersperse "," (map show (toList a))) ++ "}"
+
+injectList :: [a] -> AbsList var
+injectList xs = AbsList (singleton (l2ls xs))
+
+l2ls :: [a] -> ListShape var
+l2ls = l2ls' maxDepth
+    where
+        l2ls' 0 _      = Star
+        l2ls' n []     = Nil
+        l2ls' n (x:xs) = Cons (l2ls' (n-1) xs)
+
+mergeAbsList :: Ord var => AbsList var -> AbsList var -> AbsList var
+mergeAbsList (AbsList a) (AbsList b) = AbsList (a `union` b)
+
+meetAbsList :: Ord var => AbsList var -> AbsList var -> AbsList var
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/Abstract/Unit.hs

@@ -1,0 +1,23 @@
+module Abstract.Unit  where
+
+import Data.Set
+import Data.List (intersperse)
+
+newtype AbsUnit
+    = AbsUnit (Set ())
+    deriving (Eq, Ord)
+    
+instance Show AbsUnit where
+    show (AbsUnit s) = "{" ++ concat (intersperse "," (fmap show (toList s))) ++ "}"
+
+injectUnit :: () -> AbsUnit
+injectUnit = AbsUnit . singleton
+
+mergeAbsUnit :: AbsUnit -> AbsUnit -> AbsUnit
+mergeAbsUnit (AbsUnit a) (AbsUnit b) = AbsUnit (a `union` b)
+
+subsetAbsUnit :: AbsUnit -> AbsUnit -> Bool
+subsetAbsUnit (AbsUnit a) (AbsUnit b) = a `isSubsetOf` b
+
+topAbsUnit :: AbsUnit
+topAbsUnit = injectUnit ()

### 2015-09-01 deprecated/pattern-match-analysis/src/AbstractInterpretation.hs

@@ -1,0 +1,75 @@
+{-# LANGUAGE FunctionalDependencies #-}
+
+module AbstractInterpretation where
+
+import Data.List  as List
+import Data.Map   as Map   hiding (singleton)
+import Data.Maybe as Maybe
+import Data.Set   as Set
+
+import           Language.Haskell.Exts
+import qualified Language.Haskell.Exts.Annotated          as Ann
+import           Language.Haskell.Exts.Annotated.Simplify
+
+-- | Abstract values    
+
+data Sign  = Pos | Zero | Neg
+    deriving (Bounded, Enum, Eq, Ord, Show)
+
+toSign :: Integer -> Sign
+toSign n | n <  0 = Neg
+         | n == 0 = Zero
+         | n >  0 = Pos
+
+data Shape = Nil | Singleton | Cons
+    deriving (Bounded, Enum, Eq, Ord, Show)
+
+data AbsVal
+    = AbsPoly
+    | AbsBool (Set Bool)
+    | AbsInt  (Set Sign)
+    | AbsList (Set Shape)
+    deriving (Eq, Ord, Show)
+
+enumerateAll :: Type -> [AbsVal]
+enumerateAll (TyVar _)
+    = [AbsPoly]
+enumerateAll (TyCon (UnQual (Ident "Bool")))
+    = enumerateAll' AbsBool
+enumerateAll (TyCon (UnQual (Ident "Int")))
+    = enumerateAll' AbsInt
+
+enumerateAll' :: (Bounded a, Enum a, Ord a) => (Set a -> b) -> [b]
+enumerateAll' x
+    = fmap (x . Set.fromList) (subsequences (enumFromTo minBound maxBound))
+
+type Env = Map QName AbsVal
+
+-- | Main
+
+run :: Ann.Exp Type -> String
+run = show . abstractInterpret Map.empty
+
+-- | Abstract interpretation
+
+abstractInterpret :: Env -> Ann.Exp Type -> AbsVal
+
+-- * Variables
+abstractInterpret gamma (Ann.Var _ qname)
+    = fromMaybe (error "abstractInterpret: Var") (Map.lookup (sQName qname) gamma)
+    
+-- * Constructors
+abstractInterpret gamma (Ann.Con _ (Ann.UnQual _ (Ann.Ident _ "True")))
+    = AbsBool (singleton True)
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/Common.hs

@@ -1,0 +1,342 @@
+module Common  where
+
+import Data.List (intersperse)
+import qualified Data.Map as Map
+import Data.Map (Map, insertWith)
+import Data.Set (Set, (\\), delete, empty, member, singleton, toList, union)
+import qualified Data.Set as Set
+
+import Util.List
+
+import Language.Haskell.Exts.Annotated
+
+import PrintType
+import Refinement
+
+import Abstract.List hiding (Cons)
+
+-- | Basic types
+
+-- tyUnit = TyCon NoPhi (Special NoPhi (UnitCon NoPhi))
+tyBool ref = TyCon (Phi ref) (UnQual NoPhi (Ident NoPhi "Bool"))
+-- tyInt  = TyCon NoPhi (UnQual NoPhi (Ident NoPhi "Int"))
+-- (-->)  = TyFun NoPhi
+
+-- | Annotation-related
+
+-- FIXME: Change type of Env to [(forall l. Name l, Type Phi)] instead?
+
+stripNameAnn :: Name l -> Name Phi
+stripNameAnn (Ident  _ s) = Ident  NoPhi s
+stripNameAnn (Symbol _ s) = Symbol NoPhi s
+
+unstripNameAnn :: Name Phi -> Name l
+unstripNameAnn (Ident  NoPhi s) = Ident (error "unstripNameAnn: Ident")  s
+unstripNameAnn (Symbol NoPhi s) = Ident (error "unstripNameAnn: Symbol") s
+
+stripQNameAnn :: QName l -> QName Phi
+stripQNameAnn (Qual    _ _ _ ) = error "stripQNameAnn: qualified name"
+stripQNameAnn (UnQual  _ name) = UnQual  NoPhi (stripNameAnn name)
+stripQNameAnn (Special _ con ) = Special NoPhi (stripSpecialConAnn con)
+
+stripSpecialConAnn :: SpecialCon l -> SpecialCon Phi
+stripSpecialConAnn (UnitCon  _            ) = UnitCon  NoPhi
+stripSpecialConAnn (ListCon  _            ) = ListCon  NoPhi
+stripSpecialConAnn (TupleCon _ boxed arity) = TupleCon NoPhi boxed arity
+stripSpecialConAnn (Cons     _            ) = Cons     NoPhi
+
+stripTyVarBindAnn :: TyVarBind l -> TyVarBind Phi
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/Latex.hs

@@ -1,0 +1,6 @@
+module Latex where
+
+class Latex a where
+    latex :: a -> String
+
+

### 2015-09-01 deprecated/pattern-match-analysis/src/Main.hs

@@ -1,0 +1,136 @@
+module Main where
+
+import Data.List (intersperse, nub)
+import Data.Set (toList)
+
+import Util
+import Util.List
+
+import Language.Haskell.Exts.Annotated
+
+import Common
+import TypeInference
+import PatternTyping
+import Refinement -- TODO: export from Common?
+import Solver
+import qualified AbstractInterpretation as AI
+
+import Latex
+import PrintType
+
+                               
+-- | Main
+
+main = do let fileName = "../sample/"
+                          -- ++ "Sample"
+                          -- ++ "SampleFilter"
+                          ++ "Thesis"
+                          ++ ".hs"
+          contents <- readFile fileName
+          let parseResult = parseModuleWithMode defaultParseMode contents
+          case parseResult of
+            ParseOk hModule            
+                -> do --print "== ABSTRACT SYNTAX TREE =========================="
+                      --print hModule
+                      
+                      let R { _ty = ty
+                            , _cs = c
+                            , _rc = rc
+                            } = analyzeMain hModule
+
+                      print "== REFINEMENT TYPES =============================="
+                      -- putStrLn $ prettyPrint ty
+                      -- putStrLn "---------------------------"
+                      putStrLn $ printType ty
+                      putStrLn "---------------------------"
+                      putStrLn $ show c
+                      putStrLn "---------------------------"
+                      putStr   $ listPerLine rc
+                      putStrLn "---------------------------"
+                      let subst = unify c
+                      putStrLn $ show subst
+                      putStrLn "---------------------------"
+                      let checkCS = subst $@@ c
+                      -- putStrLn $ show checkCS
+                      putStr "Sanity check: "
+                      putStrLn (if and (map (\((:=:!) _ a b) -> a == b) checkCS) then "\x1b[0;32;49mPASSED\x1b[0;30;49m" else "\x1b[0;31;49mFAILED\x1b[0;30;49m")
+                      putStrLn "-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-"
+                      let ty' = subst $@ ty
+                      -- putStrLn $ prettyPrint $ ty'
+                      -- putStrLn "---------------------------"
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/PatternTyping.hs

@@ -1,0 +1,153 @@
+module PatternTyping where
+
+import Data.Maybe
+import Util.List
+
+import Language.Haskell.Exts.Annotated
+
+import Common
+import Refinement
+
+import Abstract.Int
+
+-- | Pattern typing
+
+-- TODO: Consider using a simpler reprentation for the type of constructors:
+-- type Env     = [(Name, ConType)]
+-- type ConType = ([Type], Type)
+
+patternType :: Env -> Fresh -> Pat Phi -> (Type Phi, Env)
+patternType conenv fresh pat
+    = let (ty, env, cs) = patternType' conenv fresh pat
+          subst         = unify cs
+       in (subst $@ ty, subst $&& env)
+
+-- *
+
+patternType' :: Env -> Fresh -> Pat Phi -> (Type Phi, Env, [Constr])
+
+patternType' env (f1:f2:_) (PVar _ name)
+    = let freshTyVar = TyVar (Phi (RefVar f2)) f1
+       in (freshTyVar, [(UnQual NoPhi name, freshTyVar)], [])
+    
+patternType' env fresh (PLit _ lit)
+    = (literalType lit, [], [])
+    
+patternType' env fresh (PNeg _ lit)
+    = patternType' env fresh lit
+    
+patternType' env fresh (PApp _ con ps)
+    = let ((f1:f2:fs):fss) = splice (length ps + 1) fresh
+          freshTyVar   = TyVar (Phi (RefVar f2)) f1
+          conTy        = fst $ inst fs (fromJust $ lookup con env) -- NOTE: fst introduces after type-change of inst
+          (ts, es, cs) = unzip3 (zipWith (patternType' env) fss ps)
+          patTy        = foldr (TyFun (Phi RefLambda)) freshTyVar ts
+       in (freshTyVar, concat es, (conTy =: patTy) : concat cs)
+
+patternType' env fresh (PInfixApp phi p1 con p2)
+    = patternType' env fresh (PApp phi con [p1, p2])
+
+patternType' env (f:fresh) (PTuple _ ps)
+    = let fs           = splice (length ps) fresh
+          (ts, es, cs) = unzip3 (zipWith (patternType' env) fs ps)
+       in (TyTuple (Phi (RefVar f)) Boxed ts, concat es, concat cs)
+
+patternType' env (f1:f2:fresh) (PList _ ps)
+    = let fss          = splice (length ps) fresh
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/PrintType.hs

@@ -1,0 +1,16 @@
+module PrintType where
+
+import Data.List (intersperse)
+
+import Language.Haskell.Exts.Annotated
+
+import Refinement
+
+printType :: Type Phi -> String
+printType (TyFun phi a b  )      = "(" ++ printType a ++ " -> " ++ printType b ++ ")@" ++ show phi
+printType (TyTuple phi boxed ts) = "(" ++ concat (intersperse "," (map printType ts)) ++ ")@" ++ show phi
+printType (TyList phi a   )      = "[" ++ printType a ++ "]@" ++ show phi
+printType (TyVar phi name )      = prettyPrint name -- ++ "@" ++ show phi
+printType (TyCon phi qname)      = prettyPrint qname ++ "@" ++ show phi
+printType (TyForall (UC { uc = cs }) Nothing Nothing t) = "forall. " ++ show cs ++ " => " ++ printType t
+printType x                      = error ("printType: " ++ show x)

### 2015-09-01 deprecated/pattern-match-analysis/src/Refinement.hs

@@ -1,0 +1,152 @@
+module Refinement where
+
+import Data.List
+import Data.Map (Map)
+import qualified Data.Map as Map
+import qualified Data.Set as Set
+
+import Language.Haskell.Exts.Annotated hiding (Cons)
+
+import Abstract.Unit
+import Abstract.Bool
+import Abstract.Int
+import Abstract.List
+
+import Latex
+
+-- | Annotations
+
+data Phi
+    = Phi { phi :: Refinement }
+    | NoPhi
+    | UC { uc :: [RefConstr] } -- unsolved constraints
+    deriving (Eq, Ord)
+    
+instance Show Phi where
+    show (Phi r) = show r
+    show (NoPhi) = "X"
+
+-- | Refinements
+
+data Refinement
+    = RefVar (Name Phi)
+    | RefUnit AbsUnit
+    | RefBool AbsBool
+    | RefInt  AbsInt
+    | RefList (AbsList (Name Phi))
+    | RefLambda         -- perhaps just use top?
+    | RefTop            -- for now: chars, tuples, lists, etc.
+    deriving (Eq, Ord)
+
+instance Show Refinement where
+    show (RefVar    name   ) = prettyPrint name
+    show (RefUnit   absunit) = show absunit
+    show (RefBool   absbool) = show absbool
+    show (RefInt    absint ) = show absint
+    show (RefList   abslist) = show abslist
+    show (RefLambda        ) = "λ"
+    show (RefTop           ) = "T"
+
+instance Latex Refinement where
+    latex (RefVar    name  ) = prettyPrint name
+    latex (RefInt    absint) = undefined -- latex absint
+    latex (RefLambda       ) = "\\lambda"
+    latex (RefTop          ) = "\\top"
+    
+-- | Constraints
+
+data RefConstr
+    = Refinement :<: Refinement
+    deriving (Eq, Ord)
+
+instance Show RefConstr where
+    show (t1 :<: t2) = show t1 ++ " ⊆ " ++ show t2
+    
+-- | Substitutions (ONLY USED IN Main for sanity checking)
+
+type RefSubst = Map (Name Phi) (Refinement, Refinement)
+
+infixr 0 $@^!
+
+($@^!) :: RefSubst -> Refinement -> Refinement
+sigma $@^! ref@(RefVar name)
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/SamplePatterns.hs

@@ -1,0 +1,95 @@
+module Main where
+
+data Either a b
+    = Left a
+    | Right b
+    
+data Pair a b
+    = Pair { fst :: a, snd :: b }
+    
+-- | Multiple patterns
+
+main = \x -> case x of
+                Left [a]     -> 1
+                Right (a, 4) -> 2
+    
+-- | Single patterns
+
+-- * PIrrPat
+{-
+main = \x -> case x of
+                ~(a:as) -> 1
+-}
+-- * PWildcard
+{-
+main = \x -> case x of
+                _ -> 1
+-}
+
+-- * PAsPat
+{-
+main = \x -> case x of
+                a@(Left b) -> a
+-}
+
+-- * PRec
+{-
+main = \x -> case x of
+                Pair { fst = i, snd = b } -> 1
+-}
+
+-- * PList
+{-
+main = \x -> case x of
+                [1,a] -> a
+-}
+{-
+main = \x -> case x of
+                [1] -> 2
+-}
+{-
+main = \x -> case x of
+                Left [a,b] -> a
+-}
+{-
+main = \x -> case x of
+                Left [a] -> a
+-}
+
+-- * PTuple
+{-
+main = \x -> case x of
+                (a,b) -> (b,a)
+-}
+
+-- * PLit, PNeg
+{-
+main = \x -> case x of
+                "a" -> "b"
+-}
+{-
+main = \x -> case x of
+                'a' -> 'b'
+-}
+{-
+main = \x -> case x of
+                -1 -> -2
+-}
+{-
+main = \x -> case x of
+                1 -> 2
+-}
+
+-- * PVar, PApp, PInfixApp
+{-
+main = \x -> case x of
+                () -> ()
+-}
+{-
+main = \x -> case x of
+                (a:as) -> 1
+-}
+{-
+main = \x -> case x of
+                Left a -> a
+-}

### 2015-09-01 deprecated/pattern-match-analysis/src/Solver.hs

@@ -1,0 +1,263 @@
+module Solver where
+
+import Prelude hiding (lookup)
+import Data.Map as Map hiding (foldr, map, singleton)
+import qualified Data.List as List
+import Data.Set (Set, singleton)
+import qualified Data.Set as Set
+
+import Language.Haskell.Exts.Annotated
+
+import Common
+import PrintType
+import Refinement
+
+import Util
+import Util.Set
+
+import Abstract.Unit
+import Abstract.Bool
+import Abstract.Int
+import Abstract.List
+
+-- | Dependency analysis
+
+-- * Dependendent constraints for the worklist
+
+data DependencyGraph
+    = DependencyGraph { lowerDep :: Map (Name Phi) (Set.Set RefConstr)
+                      , upperDep :: Map (Name Phi) (Set.Set RefConstr)
+                      }
+
+instance Show DependencyGraph where
+    show (DependencyGraph ld ud) = "Lower: {" ++ concatMap showDep (toList ld) ++ "}\n"
+                                   ++ "Upper: {" ++ concatMap showDep (toList ud) ++ "}"
+                                    where showDep (Ident _ k, v) = "[" ++ k ++ " ~> " ++ showSet v ++ "]"
+                                    
+type DependencyGraph' = Map (Name Phi) (Set.Set (Name Phi))
+                       
+dependencyAnalysis :: [RefConstr] -> DependencyGraph
+dependencyAnalysis = foldr f (DependencyGraph { lowerDep = Map.empty, upperDep = Map.empty })
+    where f :: RefConstr -> DependencyGraph -> DependencyGraph
+          f c@(a :<: b) (DependencyGraph { lowerDep = ld, upperDep = ud })
+            = DependencyGraph { lowerDep = foldr (\s ldr -> if List.null (getVars b) then ldr else insertWith Set.union s (Set.singleton c) ldr) ld (getVars a)
+                              , upperDep = foldr (\s udr -> if List.null (getVars a) then udr else insertWith Set.union s (Set.singleton c) udr) ud (getVars b)
+                              }
+                              
+-- * Full dependency analysis for constraint set partitioning, building on worklist and previous dependency-analysis (see solveRT below for details on the implementation)
+
+dependencyAnalysis' :: [RefConstr] -> DependencyGraph'
+dependencyAnalysis' rcs = worklist (g (dependencyAnalysis rcs)) empty rcs
+    where
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/TypeInference.hs

@@ -1,0 +1,392 @@
+-- {-# LANGAUGE ScopedTypeVariables #-}
+{-# LANGUAGE StandaloneDeriving  #-}
+
+-- TODO: desugar TyList into TyApp (TyCon []) (TyVar ...)
+
+module TypeInference where
+
+import Prelude
+
+import Data.Graph
+import Data.Map hiding (lookup, map, toList)
+import Data.Set (toList)
+
+import Util
+import Util.List
+
+import Language.Haskell.Exts.Annotated
+
+import Common
+import PatternTyping
+import Refinement
+import Solver
+import qualified AbstractInterpretation as AI
+
+import Abstract.Unit
+import Abstract.Bool
+import Abstract.Int
+import Abstract.List hiding (Cons)
+
+deriving instance Show a => Show (SCC a)
+
+-- | Result of type inference
+
+data R = R { _ty :: Type Phi            -- annotated type
+           , _cs :: [Constr]            -- constraints on simple types
+           , _rc :: [RefConstr]         -- constraints on refinement variables
+        -- , _ae :: Exp (Type Phi)      -- type-annotated AST
+           }
+    
+-- | Create a fresh variable
+
+makeVar :: Name Phi -> Name Phi -> Type Phi
+makeVar t phi = TyVar (Phi (RefVar phi)) t
+    
+-- | Type inference
+
+typeInfer :: (Ord l, Show l) => Env -> Fresh -> Exp l -> R
+
+-- * Variables
+typeInfer gamma (f1:f2:f3:f4:f5:fresh) (Var _ name)
+    | Just ty <- lookup (stripQNameAnn name) gamma
+       = let (tyInst, rcInst) = inst fresh ty
+          in R { _ty = tyInst
+               , _cs = []
+               , _rc = rcInst
+               }
+    | UnQual NoPhi (Ident NoPhi "fst") <- stripQNameAnn name
+        = let tv1 = TyFun (Phi (RefVar f4)) (TyTuple (Phi (RefVar f3)) Boxed [TyVar NoPhi f1, TyVar NoPhi f2]) (TyVar NoPhi f1)
+           in R { _ty = tv1
+                , _cs = []
+                , _rc = [RefVar f3 :<: RefTop, RefLambda :<: RefVar f4]
+                }
[…]

### 2015-09-01 deprecated/pattern-match-analysis/src/Util.hs

@@ -1,0 +1,12 @@
+module Util where
+
+-- | Fixed point
+
+fix f x = fix' (iterate f x)
+    where fix' (a1:as@(a2:_)) | a1 == a2  = a1
+                              | otherwise = fix' as
+
+-- | Show
+
+listPerLine :: Show a => [a] -> String
+listPerLine = concatMap ((++ "\n") . show)

### 2015-09-01 deprecated/pattern-match-analysis/src/Util/Set.hs

@@ -1,0 +1,8 @@
+module Util.Set where
+
+import Data.List (intersperse)
+import Data.Set
+
+showSet :: Show a => Set a -> String
+showSet = surround "{" "}" . concat . intersperse ", " . fmap show . toList
+    where surround l r m = l ++ m ++ r

### 2015-09-01 deprecated/pattern-match-analysis/thesis/build.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+lhs2TeX thesis.tex > thesis.lhs2TeX.tex
+pdflatex thesis.lhs2TeX.tex && bibtex thesis.lhs2TeX && makeindex thesis.lhs2TeX.idx && pdflatex thesis.lhs2TeX.tex && pdflatex thesis.lhs2TeX.tex && mv thesis.lhs2TeX.pdf thesis.pdf && evince thesis.pdf

### 2015-09-01 exception-analysis/examples/Risers.hs
exception-analysis

@@ -1,0 +1,9 @@
+module Main where
+
+risers :: [Int] -> [[Int]]
+risers [ ] = [ ]
+risers [x] = [[x]]
+risers (x : y : etc) = if x <= y then (x : s) : ss else [x] : (s : ss)
+    where (s : ss) = risers (y : etc)
+    
+main = putStrLn . show $ risers [1,6,3,7,3,6,5]

### 2015-09-01 exception-analysis/prototype/pma-v2/Deprecated/DF.hs

@@ -1,0 +1,955 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+module Deprecated.DF where
+
+import Control.Monad
+import Control.Monad.State
+
+import qualified Data.Graph    as G
+import qualified Data.List     as L
+import qualified Data.Map      as M
+import qualified Data.Map.Util as M
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+import qualified Data.Tree     as T
+
+import Common
+
+-- | Abstract Syntax
+
+data Expr'  -- refinement annotated expression
+    = Var'        Name              Ref
+    | Con'        Con               Ref
+    | Abs'        Name  Expr'       Ref
+    | Fix'        Name  Name  Expr' Ref
+    | App'        Expr' Expr'       Ref
+    | Let'        Name  Expr' Expr' Ref
+    | If'   Lbl   Expr' Expr' Expr' Ref
+    -- Operators
+    | Op'         Op    Expr' Expr' Ref
+    -- Pairs
+    | Pair'       Expr' Expr'       Ref
+    | Fst'        Expr'             Ref
+    | Snd'        Expr'             Ref
+    -- Lists
+    | Nil'                          Ref
+    | Cons'       Expr' Expr'       Ref
+    | Case' Lbl   Expr' (Maybe Expr') (Maybe (Name, Name, Expr')) Ref
+    deriving (Eq, Ord, Show)
+    
+-- | Static Semantics
+
+data Ty
+    = TyVar Name
+    | TyCon TyCon
+    | TyFun Ty Ref Ty Ref
+    -- Pairs
+    | TyPair Ty Ref Ty Ref
+    -- Lists
+    | TyList Ty Ref
+    deriving (Eq, Ord, Show)
+
+data TyCon
+    = TyBool
+    | TyInt
+    deriving (Eq, Ord, Show)
+
+data Ref
+    = RefUnif Name
+    | Ref (S.Set Ref')
+    deriving (Eq, Ord, Show)
+
+data Ref'
+    = RefVar Name
+    | RefTop
+    | RefBool Bool
+    | RefInt  Int'
+    -- Lists
+    | RefList List'
+    deriving (Eq, Ord, Show)
+    
+data Int'
+    = Neg
+    | Zero
+    | Pos
+    deriving (Eq, Ord, Show)
+    
+data List'
+    = E
+    | NE
[…]

### 2015-09-01 exception-analysis/prototype/pma-v2/Deprecated/Exn.hs

@@ -1,0 +1,896 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+module Deprecated.Exn where
+
+import Control.Monad
+import Control.Monad.State
+
+import qualified Data.Graph      as G
+import qualified Data.Graph.Util as G
+import qualified Data.List       as L
+import qualified Data.Map        as M
+import qualified Data.Map.Util   as M
+import qualified Data.Set        as S
+import qualified Data.Set.Util   as S
+import qualified Data.Tree       as T
+
+import Common
+
+-- | Abstract Syntax
+
+data Expr'  -- exception annotated expression
+    = Var'        Name              Exn
+    | Con'        Con               Exn
+    | Abs'        Name  Expr'       Exn
+    | Fix'        Name  Name  Expr' Exn
+    | App'        Expr' Expr'       Exn
+    | Let'        Name  Expr' Expr' Exn
+    | If'   Lbl   Expr' Expr' Expr' Exn
+    -- Operators
+    | Op'         Op    Expr' Expr' Exn
+    -- Pairs
+    | Pair'       Expr' Expr'       Exn
+    | Fst'        Expr'             Exn
+    | Snd'        Expr'             Exn
+    -- Lists
+    | Nil'                          Exn
+    | Cons'       Expr' Expr'       Exn
+    | Case' Lbl   Expr' (Maybe Expr') (Maybe (Name, Name, Expr')) Exn
+    deriving (Eq, Ord, Show)
+    
+-- | Static Semantics
+
+data Ty
+    = TyVar Name
+    | TyCon TyCon
+    | TyFun Ty Exn Ty Exn
+    -- Pairs
+    | TyPair Ty Exn Ty Exn
+    -- Lists
+    | TyList Ty Exn
+    deriving (Eq, Ord, Show)
+
+data TyCon
+    = TyBool
+    | TyInt
+    deriving (Eq, Ord, Show)
+
+data Exn
+    = ExnUnif Name
+    | Exn (S.Set Exn')
+    deriving (Eq, Ord, Show)
+
+data Exn'
+    = ExnVar Name
+    | ExnCrash Lbl
+    deriving (Eq, Ord, Show)
+    
+type Constr = S.Set Constr'
+
+data Constr'
+    = Exn :<: Exn
+    deriving (Eq, Ord, Show)
+
[…]

### 2015-09-01 exception-analysis/prototype/pma-v2/Deprecated/Verif.hs

@@ -1,0 +1,1055 @@
+{-# LANGUAGE FlexibleInstances     #-}
+{-# LANGUAGE TypeSynonymInstances  #-}
+
+module Deprecated.Verif where
+
+import Control.Monad
+import Control.Monad.State
+
+import qualified Data.Graph     as G
+import qualified Data.List      as L
+import qualified Data.Map       as M
+import qualified Data.Map.Util  as M
+import           Data.Maybe
+import qualified Data.Set       as S
+import qualified Data.Set.Util  as S
+import qualified Data.Tree      as T
+
+import           Common
+import qualified Deprecated.Exn as Exn
+import qualified Deprecated.DF  as DF
+
+-- | Abstract Syntax
+
+data Expr'
+    = Var'  Name                    Exn.Exn DF.Ref
+    | Con'  Con                     Exn.Exn DF.Ref
+    | Abs'  Name  Expr'             Exn.Exn DF.Ref
+    | Fix'  Name  Name  Expr'       Exn.Exn DF.Ref
+    | App'  Expr' Expr'             Exn.Exn DF.Ref
+    | Let'  Name  Expr' Expr'       Exn.Exn DF.Ref
+    | If'   Lbl   Expr' Expr' Expr' Exn.Exn DF.Ref
+    -- Operators
+    | Op'   Op    Expr' Expr'       Exn.Exn DF.Ref
+    -- Pairs
+    | Pair' Expr' Expr'             Exn.Exn DF.Ref
+    | Fst'  Expr'                   Exn.Exn DF.Ref
+    | Snd'  Expr'                   Exn.Exn DF.Ref
+    -- Lists
+    | Nil'                          Exn.Exn DF.Ref
+    | Cons' Expr' Expr'             Exn.Exn DF.Ref
+    | Case' Lbl   Expr' (Maybe Expr') (Maybe (Name, Name, Expr')) Exn.Exn DF.Ref
+    deriving (Eq, Ord, Show)
+    
+getExn :: Expr' -> Exn.Exn
+getExn (Var'    _     exn _) = exn
+getExn (Con'    _     exn _) = exn
+getExn (Abs'    _ _   exn _) = exn
+getExn (Fix'    _ _ _ exn _) = exn
+getExn (App'    _ _   exn _) = exn
+getExn (Let'    _ _ _ exn _) = exn
+getExn (If'   _ _ _ _ exn _) = exn
[…]

### 2015-09-01 exception-analysis/prototype/pma-v3/Data/Tuple/Util.hs

@@ -1,0 +1,10 @@
+module Data.Tuple.Util where
+
+fst3 :: (a, b, c) -> a
+fst3 (x, y, z) = x
+
+snd3 :: (a, b, c) -> b
+snd3 (x, y, z) = y
+
+trd3 :: (a, b, c) -> c
+trd3 (x, y, z) = z

### 2015-09-01 exception-analysis/prototype/pma-v3/Semantics/BigStep.hs

@@ -1,0 +1,472 @@
+{-# LANGUAGE FlexibleInstances    #-}
+{-# LANGUAGE TypeSynonymInstances #-}
+{-# LANGUAGE TupleSections        #-}
+
+module Semantics.BigStep where
+
+-- ||| Launchbury-style natural semantics (call-by-need)
+
+import Control.Applicative
+import Control.Monad.State
+
+import Data.Traversable
+
+import qualified Data.List     as L
+import qualified Data.Map      as M
+import qualified Data.Set      as S
+import qualified Data.Set.Util as S
+
+import Syntax
+import Ref
+import Exn
+
+-- | Normalisation
+
+-- * Check if bound variables are distinct
+
+boundVariablesDistinct :: Expr () -> Bool
+boundVariablesDistinct = maybe False (const True) . boundVariablesDistinct'
+    where boundVariablesDistinct' :: Expr () -> Maybe (S.Set Name)
+          boundVariablesDistinct' (Var x ())
+            = return (S.singleton x)
+          boundVariablesDistinct' (Con _ ())
+            = return S.empty
+          boundVariablesDistinct' (Abs x e ())
+            = do xs <- boundVariablesDistinct' e
+                 if x `S.member` xs then
+                    notDistinct
+                 else
+                    return (x `S.insert` xs)
+          boundVariablesDistinct' (Fix f e ())
+            = do xs <- boundVariablesDistinct' e
+                 if f `S.member` xs then
+                    notDistinct
+                 else
+                    return (f `S.insert` xs)
+          boundVariablesDistinct' (App e1 e2 ())
+            = do xs1 <- boundVariablesDistinct' e1
+                 xs2 <- boundVariablesDistinct' e2
+                 if xs1 `S.overlap` xs2 then
+                    notDistinct
+                 else
+                    return (xs1 `S.union` xs2)
+          boundVariablesDistinct' (Let x e1 e2 ())
+            = do xs1 <- boundVariablesDistinct' e1
+                 xs2 <- boundVariablesDistinct' e2
+                 if x `S.member` xs1 || x `S.member` xs2 || xs1 `S.overlap` xs2 then
+                    notDistinct
+                 else
+                    return (x `S.insert` (xs1 `S.union` xs2))
[…]

### 2015-09-01 exception-analysis/prototype/pma-v3/Semantics/SmallStep.hs

@@ -1,0 +1,184 @@
+module Semantics.SmallStep where
+
+-- ||| Stuctured operational semantics with environment
+-- !!! Currenly call-by-value instead of name/need
+
+import qualified Data.Map as M
+
+import Syntax
+import Ref
+import Exn
+
+-- | Values
+
+isValue :: Expr a -> Bool
+isValue (Con _ _)
+    = True
+isValue (Nil _)
+    = True
+isValue (Close _ _)
+    = True
+isValue _
+    = False
+
+-- | Environment
+
+newtype Env = Env (M.Map Name (Expr Env)) deriving Show
+    -- for CBV we assume that the values are values (no pun intended)
+
+-- | One-step reduction relation
+
+reduce :: Env -> Expr Env -> Expr Env
+-- Var
+reduce (Env env) (Var x _)
+    = env M.! x
+-- Abs
+reduce env e@(Abs _ _ _)
+    = Close e env 
+-- Fix
+reduce env e@(Fix _ _ _)
+    = Close e env
+-- App
+reduce env (App e1 e2 z) | not (isValue e1)
+    = let e1' = reduce env e1
+       in App e1' e2 z
+reduce env (App (Close (Abs x e1 _) (Env env1)) e2 _)
+    = Bind (Env (M.insert x e2 env1)) e1
+reduce env (App (Close e@(Fix f e1 z) (Env env1)) e2 _)
+    = Bind (Env (M.insert f (Close e (Env env1)) env1)) e1
+-- Let
+reduce (Env env) (Let x e1 e2 _)
+    = Bind (Env (M.insert x e1 env)) e2
+-- If
+reduce env (If l e1 e2 e3 z) | not (isValue e1)
+    = let e1' = reduce env e1
+       in If l e1' e2 e3 z
+reduce env (If _l (Con (Bool True) _) e2 e3 _)
+    = e2
+reduce env (If _l (Con (Bool False) _) e2 e3 _)
+    = e3
+-- Op
+reduce env (Op op e1 e2 z) | not (isValue e1)
+    = let e1' = reduce env e1
+       in Op op e1' e2 z
+reduce env (Op op v1 e2 z) | not (isValue e2)
+    = let e2' = reduce env e2
+       in Op op v1 e2' z
[…]

### 2015-09-03 deprecated/program-analysis-2/sample/In.hs
remove duplicate

@@ -1,431 +1,0 @@
--- {-# LANGUAGE NoImplicitPrelude #-}
-{-# LANGUAGE NoMonomorphismRestriction #-}
-
-module Test where
-{-
-data Bool
-    = True
-    | False
--}
-{-
-data Either a b
-    = Left a
-    | Right b
--}
-{-
-data Pair a b
-    = Pair { fst :: a, snd :: b }
--}
-
--- | Patterns
-
--- * Real
-{- head
-main = \x -> case x of
-                (a:as) -> as
--}
-
--- * Complex
-{-
-main = \x -> case x of
-                Left [a]     -> [a,True]
-                Right (a, 4) -> [a,a]
--}
-{-
-main = \x -> case x of
-                Left [a]     -> 1
-                Right (a, 4) -> 2
--}
-
--- | Either
-
--- * Case-statement
-
-{-
-main = \x -> case x of
-                Left  a -> Right a
-                Right b -> Left  b
--}
-{-
-main = \x -> case x of
-                Left  a -> a
-                Right b -> b
--}
-{-
-main = \x -> case x of
-                Left  a -> 10
-                Right b -> 20
--}      
-{-
-main = case Left True of
-        Left  a -> 10
-        Right b -> 20
--}
-
--- * App
-
--- main = \x -> Left x
-
--- * Type error
-
--- main = \x -> if x then Left True else Left 42
-
--- * Non-app
-
--- main = \x -> if x then Left True else Left False
--- main = \x -> if x then Left True else Right 42
--- main = \x -> if x then Left True else Right False
--- main = Left True
-
--- | Undefined
-
--- main = \x -> error x
--- main = error
--- main = undefined
-
--- | Let-bindings
-
--- * Mutually recursive
-{-
-main = let f = g
-           g = h
-           h = f
-        in h
--}
-{-
-main = let f = g
-           g = f
-        in f
--}
-
--- * Simultaneous
-{-
-main = let a = b
-           b = 42
-        in b
--}
-{-
-main = let a = b
-           b = 42
-        in a
--}
-{-
-main = let a = 0
-           b = 1
-        in b
--}
-{-
-main = let a = 0
-           b = 1
-        in a
---}
--- * Recursion & Generalization
-{-
-main = let di = \a -> let b = di a in a
-        in (di True, di 4, di)
--}
-{-
-main = let di = \a -> di a
-        in (di True, di 4, di)
--}
-
--- * Recursive
-
-{- [Int] -> Int
[…]

### 2015-09-03 deprecated/program-analysis-2/src/Common.hs

@@ -1,299 +1,0 @@
-{-# LANGUAGE MultiParamTypeClasses #-}
-{-# LANGUAGE FlexibleInstances     #-}
-
-module Common  where
-
-import Data.List (intersperse)
-import qualified Data.Map as Map
-import Data.Map (Map, insertWith)
-import Data.Set (Set, (\\), delete, empty, member, singleton, toList, union, unions)
-import qualified Data.Set as Set
-
-import Util.List
-
-import Language.Haskell.Exts.Annotated
-
-import PrintType
-import Refinement
-
-import Control.Monad
-import TCMonad
-
--- | Basic types
-
-tyBool ref = TyCon (Phi ref) (UnQual NoPhi (Ident NoPhi "Bool"))
-
--- | Annotation-related
-
--- FIXME: Change type of Env to [(forall l. Name l, Type Phi)] instead?
-
-stripNameAnn :: Name l -> Name Phi
-stripNameAnn (Ident  _ s) = Ident  NoPhi s
-stripNameAnn (Symbol _ s) = Symbol NoPhi s
-
-unstripNameAnn :: Name Phi -> Name l
-unstripNameAnn (Ident  NoPhi s) = Ident (error "unstripNameAnn: Ident")  s
-unstripNameAnn (Symbol NoPhi s) = Ident (error "unstripNameAnn: Symbol") s
-
-stripQNameAnn :: QName l -> QName Phi
-stripQNameAnn (Qual    _ _ _ ) = error "stripQNameAnn: qualified name"
-stripQNameAnn (UnQual  _ name) = UnQual  NoPhi (stripNameAnn name)
-stripQNameAnn (Special _ con ) = Special NoPhi (stripSpecialConAnn con)
-
-stripSpecialConAnn :: SpecialCon l -> SpecialCon Phi
-stripSpecialConAnn (UnitCon  _            ) = UnitCon  NoPhi
-stripSpecialConAnn (ListCon  _            ) = ListCon  NoPhi
-stripSpecialConAnn (TupleCon _ boxed arity) = TupleCon NoPhi boxed arity
-stripSpecialConAnn (Cons     _            ) = Cons     NoPhi
-
-stripTyVarBindAnn :: TyVarBind l -> TyVarBind Phi
-stripTyVarBindAnn (UnkindedVar _ name) = UnkindedVar NoPhi (stripNameAnn name)
-
-stripTypeAnn :: Type l -> Type Phi
-stripTypeAnn = error "stripTypeAnn"
-
[…]

### 2015-09-03 deprecated/program-analysis-2/src/Main.hs

@@ -1,58 +1,0 @@
-module Main where
-
-import Data.List (intersperse, nub)
-import Data.Set (toList)
-
-import Util
-import Util.List
-
-import Language.Haskell.Exts.Annotated
-
-import Common
-import TCMonad
-import TypeInference
-import Refinement
-import PrintType
-
-import Solver.Unification
-                               
--- | Main
-
-main = do let fileName = "../sample/In.hs"
-          contents <- readFile fileName
-          let parseResult = parseModuleWithMode defaultParseMode contents
-          case parseResult of
-            ParseOk hModule            
-                -> do putStrLn "==============================================="
-                
-                      let ( R { _ty = ty, _cs = c, _rc = rc }, l)
-                            = analyzeMain hModule
-
-                      putStrLn $ listPerLine $ l
-                      
-                      putStrLn "==============================================="
-
-            ParseFailed srcLoc message
-                -> do print srcLoc
-                      print message
-                      
-showConstrSet rcs = "(" ++ concat (intersperse ", " (map (init . show) rcs)) ++ ")"
-
--- * Analysis
-
-analyzeMain :: (Ord l, Show l) => Module l -> (R, [String])
-analyzeMain (Module _ _ _ _ topLevel)
-    = let (patBinds, dataDecls) = splitTopLevel topLevel
-          constructorEnv        = [] --concatMap dataDeclType dataDecls
-          topLevelBindings      = Let undefined (BDecls undefined patBinds) (Var undefined (UnQual undefined (Ident undefined "main")))
-       in let (x, st) = runTC (typeInfer ({-builtinConstructors ++ -}constructorEnv) topLevelBindings)
-           in (x, TCMonad.log st)
-
--- * Split top-level declarations into bindings and data type declarations
-
-splitTopLevel :: [Decl l] -> ([Decl l], [Decl l])
-splitTopLevel [] = ([],[])
-splitTopLevel (d@(DataDecl _ _ _ _ _ _):decls)
-    = let (ps, ds) = splitTopLevel decls in (ps, d:ds)
-splitTopLevel (p@(PatBind _ _ _ _ _):decls)
-    = let (ps, ds) = splitTopLevel decls in (p:ps, ds)

### 2015-09-03 deprecated/program-analysis-2/src/PrintType.hs

@@ -1,16 +1,0 @@
-module PrintType where
-
-import Data.List (intersperse)
-
-import Language.Haskell.Exts.Annotated
-
-import Refinement
-
-printType :: Type Phi -> String
-printType (TyFun phi a b  )      = "(" ++ printType a ++ " -> " ++ printType b ++ ")@" ++ show phi
-printType (TyTuple phi boxed ts) = "(" ++ concat (intersperse "," (map printType ts)) ++ ")@" ++ show phi
-printType (TyList phi a   )      = "[" ++ printType a ++ "]@" ++ show phi
-printType (TyVar phi name )      = prettyPrint name -- ++ "@" ++ show phi
-printType (TyCon phi qname)      = prettyPrint qname ++ "@" ++ show phi
-printType (TyForall (UC { uc = cs }) (Just vbs) Nothing t) = "forall " ++ unwords (map prettyPrint vbs) ++ ". " ++ show cs ++ " => " ++ printType t
-printType x                      = error ("printType: " ++ show x)

### 2015-09-03 deprecated/program-analysis-2/src/Refinement.hs

@@ -1,50 +1,0 @@
-module Refinement where
-
-import           Data.List
-import           Data.Map (Map)
-import qualified Data.Map as Map
-import qualified Data.Set as Set
-
-import Language.Haskell.Exts.Annotated hiding (Cons)
-
--- | Annotations
-
-data Phi
-    = Phi { phi :: Refinement }
-    | NoPhi
-    | UC { uc :: [RefConstr] } -- unsolved constraints
-    deriving (Eq, Ord)
-    
-instance Show Phi where
-    show (Phi r) = show r
-    show (NoPhi) = "X"
-
--- | Refinements
-
-data Refinement
-    = RefVar (Name Phi)
-    | RefCon
-    deriving (Eq, Ord)
-
-instance Show Refinement where
-    show (RefVar    name   ) = prettyPrint name
-    show (RefCon           ) = "RefCon"
-  
--- | Constraints
-
-data RefConstr
-    = Refinement :<: Refinement
-    deriving (Eq, Ord)
-
-instance Show RefConstr where
-    show (t1 :<: t2) = show t1 ++ " ⊆ " ++ show t2
-    
--- * Unification variable renaming for instantiation
-
-infixr 0 $@*^
-
-($@*^) :: Map (Name Phi) (Name Phi) -> Refinement -> Refinement
-sigma $@*^ (RefVar name)
-    = RefVar (Map.findWithDefault name name sigma)
-sigma $@*^ x
-    = x

### 2015-09-03 deprecated/program-analysis-2/src/Solver/Unification.hs

@@ -1,67 +1,0 @@
-module Solver.Unification
-    (unify)
-where
-
-import Data.Map as Map
-import Data.Set as Set
-
-import Language.Haskell.Exts.Annotated
-
-import Common
-import Refinement
-
--- | Unification | -------------------------------------------------------------
-
--- TODO: factor out the recursion (scheme)
--- TODO: Switch occCheck and otherwise?
-
-unify :: [Constr] -> Subst
-
-unify []
-    = Subst { tySubst  = Map.empty
-            , annSubst = Map.empty }
-unify ((s :=: t)       : c')
-    | s == t       = unify c'
-unify ((s@(TyVar NoPhi x) :=: t) : c')
-    | occCheck s t = let u = Subst { tySubst  = Map.singleton x t
-                                   , annSubst = Map.empty         }
-                      in unify (u `applySubst` c') $. u
-    | otherwise    = error "unify: occurs check"
-unify ((s :=: t@(TyVar NoPhi x)) : c')
-    | occCheck t s = let u = Subst { tySubst  = Map.singleton x s
-                                   , annSubst = Map.empty         }
-                      in unify (u `applySubst` c') $. u
-    | otherwise    = error "unify: occurs check"
-unify ((s@(TyFun sr s1 s2) :=: t@(TyFun tr t1 t2)) : c')
-    = let u = uniphi sr tr
-       in unify (u `applySubst` ((s1 :=: t1) : (s2 :=: t2) : c')) $. u
-unify ((s@(TyList sr s1) :=: t@(TyList tr t1)) : c')
-    = let u = uniphi sr tr
-       in unify (u `applySubst` ((s1 :=: t1) : c')) $. u
-unify ((s@(TyTuple sr Boxed ss) :=: t@(TyTuple tr Boxed ts)) : c')
-    | length ss == length ts
-        = let u = uniphi sr tr
-           in unify (u `applySubst` (zipWith (:=:) ss ts ++ c')) $. u
-    | otherwise    = error "unify: tuple arity mismatch"
-unify (((TyCon (Phi (RefVar phi)) s) :=: (TyCon (Phi (RefVar psi)) t)) : c')
-    | s == t       = let u = Subst { tySubst  = Map.empty
-                                   , annSubst = Map.singleton phi psi }
-                      in unify (u `applySubst` c') $. u
-    | otherwise    = error "unify: constructor clash"
[…]

### 2015-09-03 deprecated/program-analysis-2/src/TCMonad.hs

@@ -1,25 +1,0 @@
-module TCMonad where
-
-import Prelude hiding (log)
-
-data State = State { freshVars :: [Int]
-                   , log       :: [String]
-                   }
-    deriving Show
-newtype TC a  = TC (State -> (a, State))
-
-instance Monad TC where
-    TC c1 >>= fc2 = TC (\s0 -> let (r, s1) = c1 s0
-                                   TC c2   = fc2 r
-                                in c2 s1
-                       )
-    return k      = TC (\s -> (k, s))
-    
-freshTC :: TC Int
-freshTC = TC (\s -> let (fv:fvs) = freshVars s in (fv, State { freshVars = fvs, log = log s }))
-
-logTC :: String -> TC ()
-logTC msg = TC (\s -> ((), State { freshVars = freshVars s, log = msg : log s } ))
-
-runTC :: TC a -> (a, State)
-runTC (TC c) = c (State { freshVars = [0..], log = [] })

### 2015-09-03 deprecated/program-analysis-2/src/TypeInference.hs

@@ -1,376 +1,0 @@
--- {-# LANGAUGE ScopedTypeVariables #-}
-{-# LANGUAGE StandaloneDeriving  #-}
-
--- TODO: desugar TyList into TyApp (TyCon []) (TyVar ...)
-
-module TypeInference where
-
-import Prelude
-
-import Control.Monad
-
-import Data.Graph
-import Data.List (intersperse)
-import Data.Map hiding (lookup, map, toList)
-import Data.Set (toList)
-
-import Util
-import Util.List
-
-import Language.Haskell.Exts.Annotated
-
-import Common
-import PrintType
-import Refinement
-
-import TCMonad
-import TypeInference.Instantiation
-import TypeInference.Generalization
-
-import Solver.Unification
-
-deriving instance Show a => Show (SCC a)
-
--- | Result of type inference
-
-data R = R { _ty :: Type Phi            -- annotated type
-           , _cs :: [Constr]            -- constraints on simple types
-           , _rc :: [RefConstr]         -- constraints on refinement variables
-        -- , _ae :: Exp (Type Phi)      -- type-annotated AST
-           }
-    
--- | Create a fresh variable
-
-makeVar :: Name Phi -> Name Phi -> Type Phi
-makeVar t phi = TyVar (Phi (RefVar phi)) t
-
--- | Type inference
-
-typeInfer :: (Ord l, Show l) => Env -> Exp l -> TC R
-
-
--- * Variables
-typeInfer gamma (Var _ name)
-    | Just ty <- lookup (stripQNameAnn name) gamma
-       = do (tyInst, rcInst) <- inst ty
-            return R { _ty = tyInst
-                     , _cs = []
-                     , _rc = rcInst
-                     }
-    | UnQual NoPhi (Ident NoPhi "fst") <- stripQNameAnn name
-        = do tv1 <- freshTV
-             tv2 <- freshTV
-             rv1 <- freshRV
-             rv2 <- freshRV
-
-             let ty = TyFun (Phi rv1) (TyTuple (Phi rv2) Boxed [tv1, tv2]) tv1
-
-             return (R { _ty = ty
-                       , _cs = []
-                       , _rc = []
-                       })
-    | UnQual NoPhi (Ident NoPhi "snd") <- stripQNameAnn name
-        = do tv1 <- freshTV
-             tv2 <- freshTV
[…]

### 2015-09-03 deprecated/program-analysis-2/src/TypeInference/Generalization.hs

@@ -1,36 +1,0 @@
-module TypeInference.Generalization
-    (gen)
-where
-
-import Data.Set
-
-import Language.Haskell.Exts.Annotated
-
-import Common
-import Refinement
-
-import TCMonad
-
-import Solver.Unification
-
--- | Generalization
-
--- NOTE: Assumes tau is a type and not a type scheme (should be fine without
---       higher-ranked types.)
--- NOTE: Does not explicitly quantify over unsolved constraints (i.e., all
---       annotation variables appearing in constraints, but not mentioned in the
---       quantifier are implicitly existentially quantified.)
---       FIXME: We probably should (annotations not bound in the environment...)
--- NOTE: Currently does not rely on being in the TC monad (but perhaps a solver
---       would like to generate fresh names?)
-
-gen :: Env -> Type Phi -> [Constr] -> [RefConstr] -> TC (Type Phi, Subst)
-
-gen gamma tau cs rcs
-    = do let subst  = unify cs
-         let tau'   = subst `applySubst` tau
-         let rcs'   = subst `applySubst` rcs
-         let alphas = ftv tau' \\ ftvEnv gamma
-         let betas  = fav tau' \\ favEnv gamma
-         let sigma  = TyForall (UC { uc = rcs' }) (Just (fmap (UnkindedVar NoPhi) (toList alphas ++ toList betas))) Nothing tau'
-         return (sigma, subst)

### 2015-09-03 deprecated/program-analysis-2/src/TypeInference/Instantiation.hs

@@ -1,25 +1,0 @@
-module TypeInference.Instantiation
-    (inst)
-where
-
-import Control.Monad
-
-import Data.Map
-
-import Language.Haskell.Exts.Annotated
-
-import Common
-import Refinement
-
-import TCMonad
-
--- | Instantiation
-
-inst :: Type Phi -> TC (Type Phi, [RefConstr])
-
-inst  (TyForall (UC { uc = cs }) (Just varBinds) Nothing t)
-    = do fvs   <- replicateM (length varBinds) freshName
-         let sigma = fromList $ zipWith (\(UnkindedVar NoPhi name) f -> (name, f)) varBinds fvs
-         return (sigma $@* t, (Subst {annSubst = sigma}) `applySubst` cs)
-inst  t
-    = return (t, []) --error "inst: not a type scheme"

### 2015-09-03 deprecated/program-analysis-2/src/Util.hs

@@ -1,12 +1,0 @@
-module Util where
-
--- | Fixed point
-
-fix f x = fix' (iterate f x)
-    where fix' (a1:as@(a2:_)) | a1 == a2  = a1
-                              | otherwise = fix' as
-
--- | Show
-
-listPerLine :: [String] -> String
-listPerLine = concatMap (++ "\n")

### 2015-09-03 deprecated/program-analysis-2/src/Util/List.hs

@@ -1,24 +1,0 @@
-module Util.List where
-
-{-# INLINE splice3 #-}
-splice3 :: [a] -> ([a],[a],[a])
-splice3 []           = ([],[],[])
-splice3 [x]          = ([x],[],[])
-splice3 [x,y]        = ([x],[y],[])
-splice3 [x,y,z]      = ([x],[y],[x])
-splice3 (x:y:z:xyzs) = let (xs,ys,zs) = splice3 xyzs in (x:xs,y:ys,z:zs)
-
-{-# INLINE splice #-}
-splice :: Int -> [a] -> [[a]]
-splice n xs = let (xs', xss) = splitAt n xs
-               in zipWith' (:) xs' (splice n xss ++ repeat [])
-
-{-# INLINE zipWith' #-}
-zipWith' :: (a -> b -> c) -> [a] -> [b] -> [c]
-zipWith' f []     _       = []
-zipWith' f (x:xs) ~(y:ys) = x `f` y : zipWith' f xs ys
-
-{-# INLINE unzip4 #-}
-unzip4 :: [(a, b, c, d)] -> ([a], [b], [c], [d])
-unzip4 = foldr (\(a,b,c,d) ~(as,bs,cs,ds) -> (a:as,b:bs,c:cs,d:ds))
-               ([],[],[],[])

### 2015-09-03 deprecated/program-analysis-2/src/Util/Set.hs

@@ -1,8 +1,0 @@
-module Util.Set where
-
-import Data.List (intersperse)
-import Data.Set
-
-showSet :: Show a => Set a -> String
-showSet = surround "{" "}" . concat . intersperse ", " . fmap show . toList
-    where surround l r m = l ++ m ++ r

### 2015-09-03 higher-ranked-exception-types/doodles/new-lambda-union/LambdaUnion.hs.hs
new lambda union

@@ -1,0 +1,18 @@
+module Main where
+
+type Idx = Int
+
+data Name
+    = Free  Idx
+    | Bound Idx
+    | Con   Idx
+    deriving (Eq, Read)
+
+data Ty = Arr [Ty] Idx
+
+base :: Idx -> Ty
+base = Arr []
+
+data Tm = Tm [Ty] [(Either Name Tm, [Tm])]
+
+data Nf = Nf [Ty] [(       Name,    [Nf])]

### 2016-10-20 higher-ranked-dimension-analysis/doodles/ho-eq-unif/Unif/FirstOrder/BooleanRings.hs
some BR stuff

@@ -35,6 +35,10 @@ lookupWithDefault def k xs = maybe def id (lookup k xs)
 -- if we don't keep intermediate expressions in CNF. (At least asymptotically,
 -- but practically?)
 
+-- TODO: Unification in arbitrary (as opposed to free) Boolean rings. See
+--       Martin & Nipkow, Section 6. (Needs Knuth-Bendix to compute an
+--       orthogonal basis efficiently.)
+
 data BA
     = And
     | Or
@@ -53,6 +57,15 @@ data BR
 
 type BAExp = T BA () Int Int
 type BRExp = T BR () Int Int
+
+showBRExp :: BRExp -> String
+showBRExp (X x          ) = "x" ++ show x
+showBRExp (C c          ) = "C" ++ show c
+showBRExp (F Add  [p, q]) = "(" ++ showBRExp p ++ "+" ++ showBRExp q ++ ")"
+showBRExp (F Mul  [p, q]) = "(" ++ showBRExp p ++ "+" ++ showBRExp q ++ ")"
+showBRExp (F Inv  [p]   ) = "~(" ++ showBRExp p ++ ")"
+showBRExp (F Zero []    ) = "0"
+showBRExp (F One  []    ) = "1"
 
 br2ba :: BRExp -> BAExp
 br2ba (X x) = (X x)
@@ -101,16 +114,25 @@ baUnif1 t =
 
                  in Just $ map f [0 .. numVars - 1]
 
+cToBase :: Int -> BRExp -> BRExp
+cToBase _ (X x) = X x
+cToBase c (C c')
+    | c == c'   = F One  []
+    | otherwise = F Zero []
+cToBase c (F f ts) = F f (map (cToBase c) ts)
+
 -- * Solve the equation f(X) = 0 in BR.
-brUnif1 :: BRExp -> Maybe BRSubst
-brUnif1 t =
-    let numVars = numX t
-     in case DPLL.dpll (DPLL.toCNF (ba2exp (br2ba t))) of
+brUnif1 :: Int -> BRExp -> Maybe BRSubst
+brUnif1 c t =
+    let t'      = cToBase c t
+        numVars = numX t
+        numBase = numC t
+     in case DPLL.dpll (DPLL.toCNF (ba2exp (br2ba t'))) of
             []        -> Nothing
             (val : _) ->
                 let b = valuationToVector numVars val
 
                     f :: Int -> BRExp
[…]

### 2016-10-20 higher-ranked-dimension-analysis/doodles/ho-eq-unif/Unif/FirstOrder/Types.hs

@@ -51,7 +51,7 @@ vars' (F' _ ts) = concatMap vars' ts
 numX :: T f f' c Int -> Int
 numX (X  x   ) = x + 1
 numX (X' _   ) = 0
-numX (C  _   ) = error "numX: C" -- 0
+numX (C  _   ) = 0 -- error "numX: C"
 numX (F  _ ts) = maximum' 0 (map numX ts)
 numX (F' _ ts) = maximum' 0 (map numX ts)
 

### 2016-10-21 higher-ranked-exception-types/documentation/lncs.tex
esop 2017 submission

\documentclass{llncs}

%FIXME - ICFP: Haskell -> functional programming language
%FIXME - ICFP: neutered overview
%FIXME - ICFP: neutered section/lambda-union
%FIXME - ICFP: moved figures around
%FIXME - ICFP: neutered acknowledgements
%FIXME - ICFP: neutered bibliography

% Math Times
%\usepackage[T1]{fontenc}
%\usepackage{times}
%\usepackage[mtbold,mtplusscr,mtpluscal,slantedgreek,T1]{mathtime}
%\usepackage{bm}

% MTPro2
%\usepackage[lite,subscriptcorrection,slantedGreek,nofontinfo]{mtpro2}
%\usepackage[lite,subscriptcorrection,slantedGreek,nofontinfo]{mtpro2}
% slantedGreek..?

\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{stmaryrd}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[numbers]{natbib}
\usepackage{xspace}
\usepackage{hyperref}

\newcommand{\ThesisPaper}[2]{#2}

\allowdisplaybreaks[1]

%\def\labelitemi{--}

%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%
%
%
% First, let's redefine the forall, and the dot.
%
%
% This is made in such a way that after a forall, the next
% dot will be printed as a period, otherwise the formatting
% of `comp_` is used. By redefining `comp_`, as suitable
% composition operator can be chosen. Similarly, period_
% is used for the period.
%
\ReadOnlyOnce{forall.fmt}%
\makeatletter

% The HaskellResetHook is a list to which things can
% be added that reset the Haskell state to the beginning.
% This is to recover from states where the hacked intelligence
% is not sufficient.

\let\HaskellResetHook\empty
\newcommand*{\AtHaskellReset}[1]{%
  \g@addto@macro\HaskellResetHook{#1}}
\newcommand*{\HaskellReset}{\HaskellResetHook}

\global\let\hsforallread\empty

\newcommand\hsforall{\global\let\hsdot=\hsperiodonce}
\newcommand*\hsperiodonce[2]{#2\global\let\hsdot=\hscompose}
\newcommand*\hscompose[2]{#1}

\AtHaskellReset{\global\let\hsdot=\hscompose}

% In the beginning, we should reset Haskell once.
\HaskellReset

\makeatother
\EndFmtInput

\newcommand{\TODO}{\textcolor{red}{\textsc{To do.}}}
\newcommand{\TODOX}[1]{\textcolor{red}{\textsc{To do}: #1}}
\newcommand{\LORUMIPSUM}{\textcolor{red}{Lorem ipsum dolor sit amet, consectetur adipiscing elit, sed do eiusmod tempor incididunt ut labore et dolore magna aliqua. Ut enim ad minim veniam, quis nostrud exercitation ullamco laboris nisi ut aliquip ex ea commodo consequat. Duis aute irure dolor in reprehenderit in voluptate velit esse cillum dolore eu fugiat nulla pariatur. Excepteur sint occaecat cupidatat non proident, sunt in culpa qui officia deserunt mollit anim id est laborum.}}
\newcommand{\fv}[1]{\mathrm{fv}(#1)}
%\newtheorem{conjecture}{Conjecture}
%\newtheorem{corollary}{Corollary}
%\newtheorem{lemma}{Lemma}
%\newtheorem{theorem}{Theorem}

%\theoremstyle{definition}
%\newtheorem{definition}{Definition}
%\newtheorem{example}{Example}
\newtheorem{counterexample}{Counterexample}

\newcommand{\IH}[1]{\stackrel{\text{i.h.}}{#1}}
% Alignments
\newcommand{\HOR}{\quad}
\newcommand{\VER}{\\[1.5ex]}

% Inference rules
\newcommand{\Rule}[3]{\frac{#2}{#3}\mbox{\ [\textsc{#1}]}}
\newcommand{\RulX}[3]{\frac{#2}{#3}}
\newcommand{\CiteRule}[1]{\mbox{\textsc{#1}}}

\newcommand{\CaseRule}[1]{Case \textsc{#1}: }
\newcommand{\Case}[1]{Case ``#1'': }
\newcommand{\Subcase}[1]{Subcase ``#1'': }

% Environments
\newcommand{\EmptyEnv}{\emptyset}

% Substitutions
\newcommand{\Subst}[3]{#3[#1/#2]}
\newcommand{\SubstTwo}[5]{#5[#1;#3/#2;#4]}
\newcommand{\SubstS}[3]{#3[\overline{#1}/\overline{#2}]}
%%%% GENERIC FORMATTING FOR LHS2TEX CODE

\newcommand\labell[1]{\label{\currentprefix:#1}}
\newcommand\refl[1]{\ref{\currentprefix:#1}}
\newcommand\eqrefl[1]{\eqref{\currentprefix:#1}}

\newcommand{\SystemFw}{System~$\mathrm{F}_\omega$\xspace}
\newcommand{\ACIone}{\textsc{aci}\oldstylenums{1}\xspace}
\newcommand{\LambdaUnion}{\lambda^\cup}

\newcommand{\ClassLuCon}{\mathbf{Con}}
\newcommand{\ClassLuEnv}{\mathbf{Env}}
\newcommand{\ClassLuTm}{\mathbf{Tm}}
\newcommand{\ClassLuTy}{\mathbf{Ty}}
\newcommand{\ClassLuKind}{\mathbf{Kind}}            %% Lu?
\newcommand{\ClassLuVar}{\mathbf{Var}}

\newcommand{\LuAbs}[3]{\lambda #1 : #2. #3}
\newcommand{\LuTyAbs}[3]{\Lambda #1 : #2. #3}       %% Lu?
\newcommand{\LuApp}[2]{#1\ #2}
\newcommand{\LuTyApp}[2]{#1\ \left<#2\right>}       %% Lu?
\newcommand{\LuArr}[2]{{#1 \to #2}}
\newcommand{\LuARR}[2]{{#1 \Rightarrow #2}}         %% Lu?
\newcommand{\LuBase}{\text{\rotatebox[origin=c]{180}{$\star$}}}
\newcommand{\LuCON}{c}
\newcommand{\LuCon}[1][\LuCON]{\{#1\}}
\newcommand{\LuEmpty}{\emptyset}
\newcommand{\LuEnv}{\Gamma}
\newcommand{\LuOrder}[2]{#1 \prec #2}
\newcommand{\LuTm}{t}
\newcommand{\LuTy}{\tau}
\newcommand{\LuTyVar}{\alpha}                       %% Lu?
\newcommand{\LuForall}[3]{\forall #1 :: #2. #3}     %% Lu?
\newcommand{\LuKind}{\kappa}                        %% Lu?
\newcommand{\LuUnion}[2]{#1 \cup #2}
\newcommand{\LuVal}{v}
\newcommand{\LuVar}{x}
\newcommand{\LuVarY}{y}

\newcommand{\LuCtx}[1]{C\!\left[#1\right]}

\newcommand{\LuAPPLIED}[1][]{\LuVar\ \LuTm_1#1 \cdots \LuTm_n#1}

\newcommand{\JudgeLuTy}[3][\LuEnv]{#1 \vdash #2 : #3}
\newcommand{\JudgeLuSub}[3][\LuEnv]{#1 \vdash #2 \lesssim #3}
\newcommand{\JudgeLuEqv}[3][\LuEnv]{#1 \vdash #2 \cong #3}

\newcommand{\REDUCE}{\longrightarrow}
\newcommand{\Reduce}[2]{{#1 \REDUCE #2}}
\newcommand{\ReduceMulti}[2]{{#1 \REDUCE^{*} #2}}

\newcommand{\LuDen}[2][\rho]{\llbracket #2 \rrbracket_{#1}}
\newcommand{\LuDenote}[2]{\llbracket #1 \rrbracket_{\rho\left[\LuVar\mapsto #2\right]}}
\newcommand{\ClassEnv}{\mathbf{Env}}
\newcommand{\ClassTyEnv}{\mathbf{TyEnv}}
\newcommand{\ClassKiEnv}{\mathbf{KiEnv}}
\newcommand{\ClassVal}{\mathbf{Val}}
\newcommand{\ClassExnVal}{\mathbf{ExnVal}}
\newcommand{\ClassTm}{\mathbf{Tm}}
\newcommand{\ClassExnTm}{\mathbf{ExnTm}}
\newcommand{\ClassTy}{\mathbf{Ty}}
\newcommand{\ClassExnTy}{\mathbf{ExnTy}}
\newcommand{\ClassExn}{\mathbf{Exn}}
\newcommand{\ClassKind}{\mathbf{ExnKi}}
\newcommand{\ClassSubst}{\mathbf{Subst}}

\newcommand{\Throws}[2]{#1\langle #2\rangle}    %%%%%%%%%%%%% WHICH TO USE?
\newcommand{\ExnTy}{\widehat\tau}
%\newcommand{\ExnTyBool}{\widehat{\mathrm{bool}}}
\newcommand{\ExnTyBool}{\mathrm{b\widehat{oo}l}}    % FIXME: deprecate
\newcommand{\ExnTyList}[2]{\left[\Throws{#1}{#2}\right]\!}
\newcommand{\ExnTyArr}[4]{\Throws{#1}{#2} \to \Throws{#3}{#4}}
\newcommand{\ExnTyArrS}[4]{\Throws{#1}{#2} \xrightarrow{#4} #3}
\newcommand{\ExnTyArrT}[5]{\Throws{#1}{#2} \xrightarrow{#5} \Throws{#3}{#4}}
\newcommand{\ExnTyForall}[3]{\forall #1 :: #2. #3}
\newcommand{\ExnTyForallS}[3]{\forall \overline{#1 :: #2}. #3}
\newcommand{\ExnForall}[1]{\forall {#1}.}           % FIXME: deprecate

\newcommand{\Exn}{\xi}
\newcommand{\ExnAlt}{\zeta}
\newcommand{\ExnVar}{e}
\newcommand{\ExnPat}{\ExnVar^\prime\ \overline{\ExnVar_i}}
\newcommand{\ExnPatVar}{\pi}
\newcommand{\ExnCon}[1]{\{#1\}}
\newcommand{\ExnLblE}{\mathbf{E}}
\newcommand{\ExnConE}{\ExnCon{\ExnLblE}}
\newcommand{\ExnAbs}[3]{\lambda #1 : #2 . #3}
\newcommand{\ExnApp}[2]{#1\ #2}
\newcommand{\ExnUnion}[2]{#1 \cup #2}
\newcommand{\ExnEmpty}{\emptyset}

\newcommand{\ExnBool}{\mathbf{b\widehat{oo}l}}      % FIXME: \ExnTyBool
\newcommand{\ExnInt}{\mathbf{i\widehat{n}t}}        % FIXME: \ExnTyInt

\newcommand{\Kind}{\kappa}                          % FIXME: \KindVar
\newcommand{\KindEXN}{\textsc{exn}}
\newcommand{\KindDIM}{\textsc{dim}}
\newcommand{\KindArr}{\Rightarrow}

\newcommand{\KindENV}{\Delta}
\newcommand{\KindEnv}[1]{{\overline{\ExnVar_{#1} :: \Kind_{#1}}}}

% Judgements
\newcommand{\JudgeComplete}[5]{{#1 \vdash #2 : #3\ \&\ #4 \triangleright #5}}

\newcommand{\ClassVar}{\mathbf{Var}}
\newcommand{\ClassCon}{\mathbf{Con}}
\newcommand{\ClassExnVar}{\mathbf{ExnVar}}

\newcommand{\Val}{v}
\newcommand{\ExnVal}{\widehat{v}}

\newcommand{\Tm}{t}
\newcommand{\TmVar}{x}
\newcommand{\TmVal}{v}
\newcommand{\TmExnVar}{e}
\newcommand{\TmCon}[1]{c_{#1}}
\newcommand{\TmTrue}{\mathbf{true}}
\newcommand{\TmFalse}{\mathbf{false}}
\newcommand{\TmAbs}[3]{\lambda #1 : #2. #3}
\newcommand{\TmAbsExn}[4]{\lambda #1 : #2\ \&\ #3. #4}
\newcommand{\TmAnnAbs}[3]{\Lambda #1 :: #2. #3}
\newcommand{\TmAnnAbsS}[3]{\Lambda \overline{#1 :: #2}. #3}
\newcommand{\TmApp}[2]{#1\ #2}
\newcommand{\TmAnnApp}[2]{#1\ \langle #2 \rangle}
\newcommand{\TmOp}[2]{#1 \oplus #2}
\newcommand{\TmIf}[3]{\mathbf{if}\ #1\ \mathbf{then}\ #2\ \mathbf{else}\ #3}
\newcommand{\TmCrash}[2]{\lightning_{#1}^{#2}}
\newcommand{\TmSeq}[2]{#1\ \mathbf{seq}\ #2}
\newcommand{\TmFix}[1]{\mathbf{fix}\ #1}
\newcommand{\TmFIX}[3]{\mathbf{fix}\ #1 : #2. #3}
\newcommand{\TmFixExn}[4]{\mathbf{fix}\ #1 : #2\ \&\ #3. #4}
\newcommand{\TmNil}[1]{\left[\right]_{#1}}
\newcommand{\TmCons}[2]{#1 :: #2}
\newcommand{\TmCase}[5]{\mathbf{case}\ #1\ \mathbf{of}\ \{\left[\right]\mapsto #2; #3 :: #4 \mapsto #5\}}

\newcommand{\ExnLbl}{\ell}

\newcommand{\ClassType}{\mathbf{Ty}}
\newcommand{\Ty}{\tau}
\newcommand{\TyBool}{\mathbf{bool}}
\newcommand{\TyInt}{\mathbf{int}}
\newcommand{\TyReal}[1]{\mathbf{real}\langle #1 \rangle}
\newcommand{\TyList}[1]{\left[#1\right]}
\newcommand{\TyArr}[2]{#1 \to #2}

\newcommand{\TyJoin}[2]{#1 \sqcup #2}

\newcommand{\TyEnv}{\Gamma}
\newcommand{\BindTy}[3]{#1 : #2\ \&\ #3}
\newcommand{\KiEnv}{\Delta}
\newcommand{\BindKi}[2]{#1 : #2}

\newcommand{\BottomTy}[1]{\bot_{#1}}

% Judgements
\newcommand{\JudgeTy}[3]{{#1 \vdash #2 : #3}}
\newcommand{\JudgeExnTy}[5]{{#1; #2 \vdash #3 : #4\ \&\ #5}}
\newcommand{\JudgeExn}[3]{{#1 \vdash #2 : #3}}
\newcommand{\JudgeKind}[3]{{#1 \vdash #2 :: #3}}
\newcommand{\JudgeSubTy}[3]{{#1 \vdash #2 \leq #3}}
\newcommand{\JudgeSubExn}[3]{{#1 \vdash #2 \leq #3}}
\newcommand{\JudgeElab}[6]{{#1; #2 \vdash #3 \hookrightarrow #4 : #5\ \&\ #6}}
\newcommand{\JudgeElabM}[6]{{#1; #2 \vdash #3 \hookrightarrow}\\{\quad\quad\quad\quad #4 : #5\ \&\ #6}}
\newcommand{\JudgeTyElab}[3]{{#1 \vdash #2 \downarrow #3}}
\newcommand{\JudgeExnTyWff}[3]{{#2 \vdash #3\ \text{wff}}}
\newcommand{\Interp}[1]{\llbracket #1 \rrbracket}

% Judgements
\newcommand{\JudgeEval}[2]{{#1 \longrightarrow #2}}
\newcommand{\JudgeEvalT}[2]{\begin{multlined}#1 \\ \longrightarrow #2\end{multlined}}
%%%% LHS2TEX COMMANDS FOR TYPESETTING THE ALGORITHMS

%%ormat ExnForall (e) (k) (t)           = "(\ExnForall{e_{" e "} :: " k "} " t ")"

\newcommand{\Complete}{\mathcal{C}}
\newcommand{\Reconstruct}{\mathcal{R}}
\newcommand{\Solve}{\mathcal{S}}
\newcommand{\Match}{\mathcal{M}}
\newcommand{\Instantiate}{\mathcal{I}}
\newcommand{\JOIN}{\mathcal{J}}

\newcommand{\ListNil}{\left[\right]}
\newcommand{\ListCons}[2]{#1 :: #2}
\newcommand{\ListCase}[5]{\mathbf{case}\ #1\ \mathbf{of}\ \left\{ \ListNil \mapsto #2; \ListCons{#3}{#4} \mapsto #5 \right\}}

\begin{document}

%BEGIN sigplanconf-template.tex

%\special{papersize=8.5in,11in}
%\setlength{\pdfpageheight}{\paperheight}
%\setlength{\pdfpagewidth}{\paperwidth}

%\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
%\copyrightyear{20yy} 
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
%\doi{nnnnnnn.nnnnnnn}

% Uncomment one of the following two, if you are not going for the 
% traditional copyright transfer agreement.

%\exclusivelicense                 % ACM gets exclusive license to publish, 
                                  % you retain copyright

%\permissiontopublish             % ACM gets nonexclusive license to publish
                                  % (paid open-access papers, 
                                  % short abstracts)

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Higher-ranked Exception Types}
\author{Ruud Koot\\\texttt{inbox@ruudkoot.nl}}
\institute{Utrecht University}
%\titlenote{Prototype and \textsc{icfp} reviews: \url{https://www.staff.science.uu.nl/~0422819/hret/supplementary-material.tar.gz}}
% https://www.staff.science.uu.nl/~0422819/hret/supplementary-material.tar.gz
%\subtitle{Subtitle Text, if any}

%\authorinfo{Ruud Koot}
%           {Utrecht University}
%           {inbox@@ruudkoot.nl}
%\authorinfo{Ruud Koot}
%           {Utrecht University\titlenote{This material is based upon work supported by the Netherlands Organisation for Scientific Research (\textsc{nwo}) under the project \emph{Higher-ranked Polyvariance Explored} (612.001.120).}}
%           {inbox@@ruudkoot.nl}
%\authorinfo{Ruud Koot}
%           {Utrecht University}
%           {inbox@@ruudkoot.nl}
%\authorinfo{Jurriaan Hage}
%           {Utrecht University}
%           {j.hage@@uu.nl}
%\authorinfo{Ruud Koot \and Juriaan Hage}
%           {Utrecht University\titlenote{This material is based upon work supported by the Netherlands Organisation for Scientific Research (\textsc{nwo}) under the project \emph{Higher-ranked Polyvariance Explored} (612.001.120).}}
%           {inbox@@ruudkoot.nl \and j.hage@@uu.nl}

%\authorinfo{John Doe}
%           {ACME University\titlenote{This material may or may not be based upon work supported by some shady organization under a cryptic project title. We categorically deny everything.}}
%           {john.doe@@acm.org}

\maketitle

\begin{abstract}
We present a type-and-effect system that derives an exception-annotated type signature for a given term of a simply typed non-strict functional language with general recursion and a list data type. This signature declares the set of exceptional values that may be present among the values of the term, or produced by terms of function type. Higher-ranked effect polymorphism and effect operators reminiscent of \SystemFw help to achieve precision and clarity.
%
By restricting the use of higher-ranked polymorphism and operators to the effects, we conjecture the inference problem to remain decidable (in contrast to the type inference problem for \SystemFw). We give a type inference algorithm that builds on the techniques developed by \cite{HoldermansHage}.
%
The types in \SystemFw form a simply typed \mbox{$\lambda$-calculus}. Similarly, the effects in our system form a simply typed algebraic $\lambda$-calculus embellished with the \ACIone-structure of sets ($\LambdaUnion$). We briefly study this language in its own right.
\end{abstract}

%\category{D.1.1}{Programming Techniques}{Applicative (Functional) Programming}
%\category{F.3.3}{Logics and Meanings of Programs}{Studies of Program Constructs---Type structure}
%\category{D.3.3}{Programming Languages}{Language Constructs and Features---Polymorphism}
%\category{F.3.2}{Logics and Meanings of Programs}{Semantics of Programming Languages---Program analysis}

% general terms are not compulsory anymore, 
% you may leave them out
%\terms
%Languages, Design, Theory

%\keywords
%exceptions, higher-ranked polymorphism, polymorphic recursion, type-and-effect systems, type-based program analysis, type inference, type operators, unification theory

%END sigplanconf-template.tex

\section{\ThesisPaper{Motivation}{Introduction}}

%FIXME - ICFP: Haskell -> functional programming language
%FIXME - ICFP: neutered overview

An often-heard selling point of non-strict functional languages is that they provide strong and expressive type systems that make side-effects explicit. This supposedly makes software more reliable by lessening the mental burden placed on programmers. Many programmers with a background in object-oriented languages are thus quite surprised, when making the transition to a functional language, that they lose a feature their type system formerly did provide: the tracking of uncaught exceptions.

There is an excuse for why this feature is missing from the type systems of contemporary non-strict functional languages: in a strict first-order language it is sufficient to annotate each function with a single set of uncaught exceptions the function may raise; in a non-strict higher-order language the situation becomes significantly more complicated. Let us first consider the two aspects `higher-order' and `non-strict' in isolation:

\begin{description}

    \item [Higher-order functions] The set of exceptions that may be raised by a higher-order function is not given by a fixed set of exceptions, but depends on the set of exceptions that may be raised by the function that is passed as its functional argument. Higher-order functions are thus \emph{exception polymorphic}.

    \item [Non-strict evaluation] In non-strictly evaluated languages, exceptions are not a form of control flow, but a kind of value. Typically the set of values of each type is extended with an \emph{exceptional value} $\lightning$ (more commonly denoted $\bot$, but we shall not do so to avoid ambiguity), or family of exceptional values~$\lightning^\ell$. This means we do not only need to give all functions an exception-annotated function type, but give every other expression an exception-annotated type as well.

\end{description}

Now let us consider these two aspects in combination. Take as an example the \ensuremath{\Varid{map}} function:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}c<{\hspost}@{}}%
\column{6E}{@{}l@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{28}{@{}>{\hspre}l<{\hspost}@{}}%
\column{36}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{map}{}\<[6]%
\>[6]{}:{}\<[6E]%
\>[10]{}\forall \alpha\beta\hsforall \hsdot{\circ }{.}(\alpha\to \beta)\to [\mskip1.5mu \alpha\mskip1.5mu]\to [\mskip1.5mu \beta\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\Varid{map}{}\<[6]%
\>[6]{}\mathrel{=}{}\<[6E]%
\>[10]{}\lambda\Varid{f}.\lambda\Varid{xs}.\;\mathbf{case}\;\Varid{xs}\;\mathbf{of}{}\<[E]%
\\
\>[10]{}\hsindent{18}{}\<[28]%
\>[28]{}[\mskip1.5mu \mskip1.5mu]{}\<[36]%
\>[36]{}\mapsto [\mskip1.5mu \mskip1.5mu]{}\<[E]%
\\
\>[10]{}\hsindent{18}{}\<[28]%
\>[28]{}(\Varid{y}::\Varid{ys}){}\<[36]%
\>[36]{}\mapsto \Varid{f}\;\Varid{y}::\Varid{map}\;\Varid{f}\;\Varid{ys}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

We denote the exception-annotated type of a term by $\widehat\tau\ \&\ \Exn$ or~$\widehat\tau\langle\Exn\rangle$. For function types we occasionally write $\widehat\tau_1\langle\Exn_1\rangle \xrightarrow{\Exn} \widehat\tau_2\langle\Exn_2\rangle$ instead of $\left(\widehat\tau_1\langle\Exn_1\rangle \rightarrow \widehat\tau_2\langle\Exn_2\rangle\right)\!\langle\Exn\rangle$. If $\Exn$ is the empty exception set, then we sometimes omit this annotation completely.

The fully exception-polymorphic and exception-annotated type, or \emph{exception type}, of \ensuremath{\Varid{map}} is\ThesisPaper{\TODO{cramped formatting}}{}
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{6}{@{}>{\hspre}c<{\hspost}@{}}%
\column{6E}{@{}l@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{57}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{map}{}\<[6]%
\>[6]{}:{}\<[6E]%
\>[10]{}\forall \alpha\hsforall \;\beta\;\Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{3}}\hsdot{\circ }{.}(\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \Varid{e}_{\mathrm{3}}}\beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{1}} \rangle){}\<[E]%
\\
\>[10]{}\hsindent{47}{}\<[57]%
\>[57]{}\xrightarrow{\emptyset}(\forall \Varid{e}_{\mathrm{4}}\hsforall \;\Varid{e}_{\mathrm{5}}\hsdot{\circ }{.}[\mskip1.5mu \alpha\langle \Varid{e}_{\mathrm{4}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle\xrightarrow{\emptyset}[\mskip1.5mu \beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{4}}\cup \Varid{e}_{\mathrm{3}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The exception type of the first argument \ensuremath{\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \Varid{e}_{\mathrm{3}}}\beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{1}} \rangle} states that it can be instantiated with a function that accepts any exceptional value as its argument (as the exception set \ensuremath{\Varid{e}_{\mathrm{1}}} is universally quantified) and returns a possibly exceptional value. In case the return value is exceptional, then it is one from the exception set \ensuremath{\Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{1}}}. Here \ensuremath{\Varid{e}_{\mathrm{2}}} is an \emph{exception set operator}---a function that takes a number of exception sets and exception set operators, and transforms them into another exception set, for example by adding a number of new elements to them, or discarding them and returning the empty set. Furthermore, the function (closure) itself may be an exceptional value from the exception set \ensuremath{\Varid{e}_{\mathrm{3}}}.

The exception type of the second argument \ensuremath{[\mskip1.5mu \alpha\langle \Varid{e}_{\mathrm{4}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle} states that it should be a list. Any of the exceptional elements in the list must be exceptional values from the exception set \ensuremath{\Varid{e}_{\mathrm{4}}}. Any exceptional values among the constructors that form the spine of the list must be exceptional values from the exception set \ensuremath{\Varid{e}_{\mathrm{5}}}.

The result of \ensuremath{\Varid{map}} is a list with the exception type \ensuremath{[\mskip1.5mu \beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{4}}\cup \Varid{e}_{\mathrm{3}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle}. Any exceptional constructors in the spine of this list must be exceptional values from the exception set \ensuremath{\Varid{e}_{\mathrm{5}}}, the same exception set as where exceptional values in the spine of the list argument \ensuremath{\Varid{xs}} come from. By looking at the definition of \ensuremath{\Varid{map}} we can see why this is the case: \ensuremath{\Varid{map}} only produces non-exceptional constructors, but the pattern-match on the list argument \ensuremath{\Varid{xs}} propagates any exceptional values encountered there. The elements of the list are produced by the function application \ensuremath{\Varid{f}\;\Varid{y}}. Recall that \ensuremath{\Varid{f}} has the exception type \ensuremath{\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \Varid{e}_{\mathrm{3}}}\beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{1}} \rangle}. Now, one of two things can happen:
    \begin{enumerate}

        \item If \ensuremath{\Varid{f}} is an exceptional function value, then it must be one from the exception set \ensuremath{\Varid{e}_{\mathrm{3}}}. Applying the exceptional value to an argument causes the exceptional value to be propagated.

        \item Otherwise, \ensuremath{\Varid{f}} is a non-exceptional value. The argument \ensuremath{\Varid{y}} has exception type \ensuremath{\alpha\langle \Varid{e}_{\mathrm{4}} \rangle}---it is an element from the list argument \ensuremath{\Varid{xs}}---and so can only be applied to \ensuremath{\Varid{f}} if we instantiate \ensuremath{\Varid{e}_{\mathrm{1}}} to \ensuremath{\Varid{e}_{\mathrm{4}}} first. If \ensuremath{\Varid{f}\;\Varid{y}} produces an exceptional value, then it is thus one from the exception set \ensuremath{\Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{4}}}.

    \end{enumerate}
To account for both cases we need to take the union of the two exception sets, giving us a value with the exception type \mbox{\ensuremath{\beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{4}}\cup \Varid{e}_{\mathrm{3}} \rangle}}.

To get a better intuition for the behavior of these exception types and exception set operators, let us see what happens when we apply \ensuremath{\Varid{map}} to two different functions: the identity function \ensuremath{\Varid{id}} and the constant exception-valued function \ensuremath{\Varid{const}\;\lightning^\mathbf{E}}. These two functions can individually be given the exception types:
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{61}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[21]{}\Varid{id}{}\<[37]%
\>[37]{}\mathrel{=}\lambda\Varid{x}.\Varid{x}{}\<[61]%
\>[61]{}:\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \emptyset}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle{}\<[E]%
\\
\>[21]{}\Varid{const}\;\lightning^\mathbf{E}{}\<[37]%
\>[37]{}\mathrel{=}\lambda\Varid{x}.\lightning^\mathbf{E}{}\<[61]%
\>[61]{}:\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \emptyset}\beta\langle \{\mathbf{E}\} \rangle{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The term \ensuremath{\Varid{id}} merely propagates its argument to the result unchanged, so it also propagates any exceptional values unchanged. The term \ensuremath{\Varid{const}\;\lightning^\mathbf{E}} discards its argument and always returns the exceptional value \ensuremath{\lightning^\mathbf{E}}. This behavior of \ensuremath{\Varid{id}} and \ensuremath{\Varid{const}\;\lightning^\mathbf{E}} is also reflected in their exception types.

When we apply \ensuremath{\Varid{map}} to \ensuremath{\Varid{id}}, we need to unify the exception type of the formal parameter \ensuremath{\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \Varid{e}_{\mathrm{3}}}\beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{1}} \rangle} with the exception type of the actual parameter \ensuremath{\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \emptyset}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle}. This can be accomplished by instantiating \ensuremath{\Varid{e}_{\mathrm{3}}} to \ensuremath{\emptyset} and \ensuremath{\Varid{e}_{\mathrm{2}}} to \ensuremath{\lambda\Varid{x}.\Varid{x}}---as \ensuremath{(\lambda\Varid{x}.\Varid{x})\;\Varid{e}_{\mathrm{1}}} evaluates to \ensuremath{\Varid{e}_{\mathrm{1}}}---giving us the resulting exception type
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{map}\;\Varid{id}{}\<[23]%
\>[23]{}:\forall \alpha\hsforall \;\Varid{e}_{\mathrm{4}}\;\Varid{e}_{\mathrm{5}}\hsdot{\circ }{.}[\mskip1.5mu \alpha\langle \Varid{e}_{\mathrm{4}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle\xrightarrow{ \emptyset}[\mskip1.5mu \alpha\langle \Varid{e}_{\mathrm{4}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
In other words, mapping the identity function over a list propagates all exceptional values already present in the list and introduces no new exceptional values.

When we apply \ensuremath{\Varid{map}} to \ensuremath{\Varid{const}\;\lightning^\mathbf{E}} we unify the exception type of the formal parameter with \ensuremath{\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \emptyset}\beta\langle \{\mathbf{E}\} \rangle}, which can be accomplished by instantiating \ensuremath{\Varid{e}_{\mathrm{3}}} to \ensuremath{\emptyset} and \ensuremath{\Varid{e}_{\mathrm{2}}} to \ensuremath{\lambda\Varid{x}.\{\mathbf{E}\}}---as \ensuremath{(\lambda\Varid{x}.\{\mathbf{E}\})\;\Varid{e}_{\mathrm{1}}} evaluates to \ensuremath{\{\mathbf{E}\}}---giving us the exception type
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{map}\;(\Varid{const}\;\lightning^\mathbf{E}){}\<[23]%
\>[23]{}:\forall \alpha\hsforall \;\beta\;\Varid{e}_{\mathrm{4}}\;\Varid{e}_{\mathrm{5}}\hsdot{\circ }{.}[\mskip1.5mu \alpha\langle \Varid{e}_{\mathrm{4}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle\xrightarrow{ \emptyset}[\mskip1.5mu \beta\langle \{\mathbf{E}\} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
In other words, mapping the constant function with the exceptional value \ensuremath{\lightning^\mathbf{E}} as its range over a list discards all existing exceptional values from the list and produces only non-exceptional values or the exceptional value \ensuremath{\lightning^\mathbf{E}} as elements of the list.

% LNCS %include introduction/overview.lhs2tex
% LNCS %include introduction/contributions.lhs2tex
\section{The $\LambdaUnion$-calculus}\label{section-lambda-union}

%FIXME - ICFP: made the syntax of \lambda^\cup into not-a-figure
%FIXME - ICFP: neutered widening

The $\LambdaUnion$-calculus is a simply typed $\lambda$-calculus extended at the term-level with empty set and singleton set constants, and a set union operator. The $\LambdaUnion$-calculus forms the language of effects in the type-and-effect system developed in Section~\ref{section-exception-types}.

\subsection{Syntax, types and semantics}

We let $\LuVar \in \ClassVar$ range over an infinite set of variables and $c \in \ClassCon$ over a non-empty set of constants. The syntax of types $\LuTy$, terms $\LuTm$, and environments $\LuEnv$ is given by:
%\begin{figure}[h]
%\paragraph{Types}
\begin{align*}
    \LuTy \in \ClassLuTy \ ::=\ \LuBase \ |\  \LuArr{\LuTy_1}{\LuTy_2}
\end{align*}
%\paragraph{Terms}
\begin{align*}
    \LuTm \in \ClassLuTm \ ::=\    \LuVar
                         \ |\   \LuAbs{\LuVar}{\LuTy}{\LuTm}
                         \ |\   \LuApp{\LuTm_1}{\LuTm_2}
                         \ |\   \LuEmpty
                         \ |\   \LuCon
                         \ |\   \LuUnion{\LuTm_1}{\LuTm_2}
\end{align*}
%\paragraph{Environments}
\begin{align*}
    \LuEnv \in \ClassLuEnv\ ::=\ \EmptyEnv \ | \ \LuEnv, \LuVar : \LuTy
\end{align*}
%\caption{$\LambdaUnion$-calculus: syntax}
%\end{figure}

%\subsection{Typing relation}

The typing relation of the $\LambdaUnion$-calculus is an extension of the typing relation of the simply typed $\lambda$-calculus.
% ICFP \begin{figure}[h]
\begin{gather*}
    \RulX{T-Var}
         {}
         {\JudgeLuTy[\LuEnv, \LuVar : \LuTy]{\LuVar}{\LuTy}}
    \quad
    \RulX{T-Abs}
         {\JudgeLuTy[\LuEnv, \LuVar : \LuTy_1]{\LuTm}{\LuTy_2}}
         {\JudgeLuTy{\LuAbs{\LuVar}{\LuTy_1}{\LuTm}}{\LuArr{\LuTy_1}{\LuTy_2}}}
    \quad
    \RulX{T-App}
         {\JudgeLuTy{\LuTm_1}{\LuArr{\LuTy_1}{\LuTy_2}} \quad \JudgeLuTy{\LuTm_2}{\LuTy_1}}
         {\JudgeLuTy{\LuApp{\LuTm_1}{\LuTm_2}}{\LuTy_2}}
    \VER
    \RulX{T-Empty}
         {}
         {\JudgeLuTy{\LuEmpty}{\LuBase}}
    \quad
    \RulX{T-Con}
         {}
         {\JudgeLuTy{\LuCon}{\LuBase}}
    \quad
    \RulX{T-Union}
         {\JudgeLuTy{\LuTm_1}{\LuTy} \quad \JudgeLuTy{\LuTm_2}{\LuTy}}
         {\JudgeLuTy{\LuUnion{\LuTm_1}{\LuTm_2}}{\LuTy}}
\end{gather*}
% ICFP \caption{$\LambdaUnion$-calculus: type system}
% ICFP \end{figure}
The empty set and singleton set constants are of base type and the set union of two terms can only be taken if the involved terms have the same type.

%\subsection{Semantics}

In the $\LambdaUnion$-calculus, terms are interpreted as sets and types as powersets.
%\begin{figure}[h]
    \paragraph{Types and values}
    \begin{align*}
        V_\LuBase &= \mathcal{P}(\ClassCon) \\
        V_{\tau_1 \to \tau_2} &= \mathcal{P}(V_{\tau_1} \to V_{\tau_2})
    \end{align*}
    \paragraph{Environments}
    \begin{align*}
        \rho &: \ClassVar \to \bigcup \left\{V_\tau \mid \tau\ \mathrm{type}\right\}
    \end{align*}
    \paragraph{Terms}
    \begin{align*}
        \llbracket \LuVar \rrbracket_\rho &= \rho(x) \\
        \llbracket \LuAbs{\LuVar}{\LuTy}{\LuTm} \rrbracket_\rho &= \left\{ \lambda v \in V_\tau. \llbracket t \rrbracket_{\rho[\LuVar \mapsto v]} \right\} \\
        \llbracket \LuApp{\LuTm_1}{\LuTm_2} \rrbracket_\rho &= \bigcup \left\{ \varphi(\llbracket \LuTm_2 \rrbracket_\rho) \mid \varphi \in \llbracket \LuTm_1 \rrbracket_\rho \right\}\\
        \llbracket \LuEmpty \rrbracket_\rho &= \LuEmpty \\
        \llbracket \LuCon \rrbracket_\rho &= \LuCon \\
        \llbracket \LuUnion{\LuTm_1}{\LuTm_2} \rrbracket_\rho &= \llbracket \LuTm_1 \rrbracket_\rho \cup \llbracket \LuTm_2 \rrbracket_\rho
    \end{align*}
%    \caption{$\LambdaUnion$-calculus: denotational semantics}
%\end{figure}

\subsection{Subsumption and observational equivalence}

The set-structure of the $\LambdaUnion$-calculus induces a partial order on the terms.

%\begin{definition}
%We say that a term environment $\rho$ is \emph{compatible} with a type environment $\Gamma$ if $\mathrm{dom}(\rho) = \mathrm{dom}(\Gamma)$ and for all $x \in \mathrm{dom}(\rho)$ we have that $\Gamma \vdash \rho(x) : \Gamma(x)$.
%\end{definition}

\begin{definition}
Denote by $\LuCtx{}$ a \emph{context}---a $\LambdaUnion$-term with a single hole in it---and by $\LuCtx{\LuTm}$ the term obtained by replacing the hole in $\LuCtx{}$ with the term $\LuTm$.
\end{definition}

\begin{definition}\label{definition-subsumption}
    Let $\LuTm_1$ and $\LuTm_2$ be terms such that $\JudgeLuTy{\LuTm_1}{\LuTy}$ and $\JudgeLuTy{\LuTm_2}{\LuTy}$.
    We say the term $\LuTm_2$ \emph{subsumes} the term $\LuTm_1$, written $\JudgeLuSub{\LuTm_1}{\LuTm_2}$, if for any context $\LuCtx{}$ such that $\JudgeLuTy[]{\LuCtx{\LuTm_1}}{\LuBase}$ and $\JudgeLuTy[]{\LuCtx{\LuTm_2}}{\LuBase}$ we have that $\LuDen[\emptyset]{\LuCtx{\LuTm_1}} \subseteq \LuDen[\emptyset]{\LuCtx{\LuTm_2}}$.
\end{definition}

\begin{definition}
    Let $\LuTm_1$ and $\LuTm_2$ be terms such that $\JudgeLuTy{\LuTm_1}{\LuTy}$ and $\JudgeLuTy{\LuTm_2}{\LuTy}$.
    We say that terms $\LuTm_1$ and $\LuTm_2$ are \emph{observationally equivalent}, denoted $\JudgeLuEqv{\LuTm_1}{\LuTm_2}$, if
    \begin{enumerate}
        \item $\JudgeLuSub{\LuTm_1}{\LuTm_2}$ and $\JudgeLuSub{\LuTm_2}{\LuTm_1}$, or equivalently that
        \item for any context $\LuCtx{}$ such that $\JudgeLuTy[]{\LuCtx{\LuTm_1}}{\LuBase}$ and $\JudgeLuTy[]{\LuCtx{\LuTm_2}}{\LuBase}$ we have that $\LuDen[\emptyset]{\LuCtx{\LuTm_1}} = \LuDen[\emptyset]{\LuCtx{\LuTm_2}}$.
    \end{enumerate}
\end{definition}

%%include ../theorem/lu-gamma1.lhs2tex
%%include ../theorem/lu-gamma2.lhs2tex
%%include ../proof/lu-gamma2-sketch.lhs2tex

\subsection{Normalization}

To reduce $\LambdaUnion$-terms to a canonical normal form we combine the $\beta$-reduction rule of the simply typed $\lambda$-calculus with rewrite rules that deal with the associativity, commutativity, idempotence and identity (\ACIone) properties of the set union operator.

\subsubsection{$\beta$- and $\gamma$-reduction}

If a term $t$ is $\eta$-long---i.e., it cannot be $\eta$-expanded without introducing additional $\beta$-redexes---it can be written in the form
\begin{align*}
    t = \lambda x_1 \cdots x_n. f_1(t_{11}, ..., t_{1q_1}) \cup \cdots \cup f_p(t_{p1}, ..., t_{pq_p})
\end{align*}
where $f_i$ can be a free or bound variable, a singleton-set constant, or another $\eta$-long term; and $q_i$ is equal to the arity of $f_i$ (for all $1 \leq i \leq p$).
Here we have removed any empty set constants (unit elements), duplicate terms $f_i(t_{i1}, ..., t_{iq_i})$ (idempotent elements), and `forgotten' how the set union operator associates.

A \emph{normal form} $v$ of a term $t$---obtained by repeatedly applying the reduction rules
%\begin{figure}[h]
    \begin{gather*}
        \RulX{R-Beta}
             {
             }
             {(\lambda x.t_1)\ t_2 \longrightarrow t_1\left[t_2 / x\right]
             }
        \quad
        \RulX{R-Gamma$_1$}
             {
             }
             {(t_1 \cup \cdots \cup t_n) \ t \longrightarrow t_1\ t \cup \cdots \cup t_n\ t
             }
        \VER
        \RulX{R-Gamma$_2$}
             {
             }
             {(\lambda x.t_1) \cup \cdots \cup (\lambda x.t_n) \longrightarrow \lambda x.t_1 \cup \cdots \cup t_n
             }
    \end{gather*}
%\caption{$\LambdaUnion$-calculus: reduction}\label{figure-lu-reduction}
%\end{figure}
and removing any empty set constants and duplicate terms---can be written as
\begin{align*}
    v = \lambda x_1 \cdots x_n. k_1(v_{11}, ..., v_{1q_1}) \cup \cdots \cup k_p(v_{p1}, ..., v_{pq_p})
\end{align*}
where $k_i$ can be a free or bound variable, or a singleton-set constant, but not a $\lambda$-abstraction (as this would form a $\beta$-redex), nor a union (as this would form a $\gamma_1$-redex).
%For each $k_i, k_j$ with $i < j$ we must also have that $k_i < k_j$ for some total order on $\ClassVar \cup \ClassCon$.
%Not only does this imply that each term $k_i(v_{i1}, ..., v_{iq_i})$ occurs only once in $k_1(v_{11}, ..., v_{1q_1}), ..., k_p(v_{p1}, ..., v_{pq_p})$, but also the stronger condition that $k_i \neq k_j$ for all $i\neq j$.

\subsubsection{Canonical ordering}

To be able to efficiently check two normalized terms for definitional equality up to \ACIone, we also need to deal with the commutativity of the union operator. We can bring normalized terms into a fully canonical form by defining a total order on terms and using it to order unions of terms.

First, pick a strict total order $\prec$ on variables and constants. The order must be fixed and be invariant under $\alpha$-renaming of variables (for example, choose the De Bruijn index of a variable), but can otherwise be arbitrary.
We extend this order to a total order on $\beta\gamma$-normal $\eta$-long terms lexicographically:
\begin{enumerate}
\item Given two fully applied terms $k(v_1, ..., v_n)$ and $k^\prime(v_1^\prime, ..., v_m^\prime)$ we define:
\begin{align*}
    k(v_1, ..., v_n) &\prec k^\prime(v_1^\prime, ..., v_m^\prime) && \text{if $k \prec k^\prime$} \\
    k(v_1, ..., v_i, ..., v_n) &\prec k(v_1, ..., v_{i-1},v_i^\prime, ..., v_m^\prime) &&\text{if $v_i \prec v_i^\prime$}
\end{align*}
\item Given two values $\lambda x_1\cdots x_n. K_1 \cup \cdots \cup K_{i-1} \cup K_i \cup \cdots \cup K_p$ and $\lambda x_1\cdots x_n. K_1 \cup \cdots \cup K_{i-1} \cup K_i^\prime \cup \cdots \cup K_q^\prime$ that have been ordered such that $K_1 \prec \cdots \prec K_{i-1} \prec K_i \prec \cdots \prec K_p$ and $K_1 \prec \cdots \prec K_{i-1} \prec K_i^\prime \prec \cdots \prec K_q^\prime$, we define:
\begin{align*}
    &\lambda x_1\cdots x_n. K_1 \cup \cdots \cup K_{i-1} \cup K_i \cup \cdots \cup K_p \\
    &\quad\quad\quad\quad\prec \lambda x_1\cdots x_n. K_1 \cup \cdots \cup K_{i-1} \cup K_i^\prime \cup \cdots \cup K_q^\prime
\end{align*}
if $K_i \prec K_i^\prime$.
\end{enumerate}

Given a term $t$ with the types of the free variables given by the environment $\Gamma$, we denote by $\llfloor t \rrfloor_\Gamma$ the $\beta\gamma$-normal $\eta$-long and canonically ordered reduction of the term $t$.

\subsection{Pattern unification}
%Solving equations in $\lambda$-calculi is usually quite hard, but sometimes very easy.

\begin{definition}\label{definition-pattern}
A $\LambdaUnion$-term is called a \emph{pattern} if it is of the form $f(e_1, ..., e_n)$ where $f$ is a free variable and $e_1, ..., e_n$ are distinct bound variables.
\end{definition}

Note that this definition is a special case of what is usually called a \emph{pattern} in higher-order unification theory \cite{Miller1991,Dowek:2001:HUM:778522.778525}.

If $f(e_1, ..., e_n)$ is a pattern and $t$ a term, then the equation
\[ f : \LuTy_1 \to \cdots \to \LuTy_n \to \LuTy \vdash \forall e_1 : \LuTy_1, ..., e_n : \LuTy_n. f(e_1, ..., e_n) = t \]
has a unique solution given by the unifier \[ \theta = \left[f \mapsto \lambda e_1 : \LuTy_1, ..., e_n : \LuTy_n. t \right]. \]

\section{Source language}\label{section-source-language}

The type-and-effect system is applicable to a simple non-strict functional language that supports boolean, integer and list data types, as well as general recursion.
%In this section we briefly state its syntax and semantics.
%\begin{figure}[h]
\paragraph{Terms}
\begin{alignat*}{2}
    \Tm \in \ClassTm &\ ::=\   &&\ \TmVar                   %\tag{term variable}     \\
                     \ |  \ \TmCon{\Ty}              %\tag{term constant}     \\
                     \ |  \ \TmCrash{\Ty}{\ExnLbl}   %\tag{exceptional constant}\\
                     \ |  \ \TmAbs{\TmVar}{\Ty}{\Tm} %\tag{term abstraction}  \\
                     \ |  \ \TmApp{\Tm_1}{\Tm_2}     %\tag{term application}  \\
                     \ |  \ \TmFIX{\TmVar}{\Ty}{\Tm} %\tag{general recursion} \\
                     \\& &&\ \TmSeq{\Tm_1}{\Tm_2}     %\tag{forcing}           \\
                     \ |  \ \TmOp{\Tm_1}{\Tm_2}      %\tag{operator}          \\
                     \ |  \ \TmIf{\Tm_1}{\Tm_2}{\Tm_3}%\tag{conditional}      \\
                     \\& &&\ \TmNil{\Ty}              %\tag{nil constructor}   \\
                     \ |  \ \TmCons{\Tm_1}{\Tm_2}    %\tag{cons constructor}  \\
                     \ |  \ \TmCase{\Tm_1}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}
                                                            %\tag{list eliminator}
\end{alignat*}
\paragraph{Values}
\begin{alignat}{2}
\Val \in \ClassVal &\ ::=\  &&\ \TmCon{\Ty}\ |\ \TmAbs{\TmVar}{\Ty}{\Tm}\ |\ \TmFIX{\TmVar}{\Ty}{\Tm}\ |\ \TmNil{\Ty}\ |\ \TmCons{\Tm_1}{\Tm_2} \notag
\end{alignat}
\begin{alignat}{2}
\ExnVal \in \ClassExnVal &\ ::=\  &&\ \TmCrash{\Ty}{\ExnLbl}\ |\ \Val \notag
\end{alignat}
%\caption{Source language: syntax}\label{figure-source-language-syntax}
%\end{figure}

Most constructs in the source language should be familiar. The $\mathbf{seq}$-construct evaluates the term on the left to a value and then continues evaluating the term on the right.

Missing from the language is a construct to `catch' exceptional values. While this may be surprising to programmers familiar with strict languages, it is a common design decision to omit such a construct from the pure fragment of non-strict languages. The omission of such a construct allows for the introduction of a certain amount of non-determinism in the operational semantics of the language---giving more freedom to an optimizing compiler---without breaking referential transparency.

The values of the source language are stratified into non-exceptional values $\Val$ and possibly exceptional values $\ExnVal$.

\subsection{Underlying type system}

The type system of the source language is given here for reference. This is the \emph{underlying type system} with respect to the type-and-effect system that is presented in Section~\ref{section-exception-types}. We assume that any term we type in the type-and-effect system is already well-typed in the underlying type system.

%\begin{figure*}[p]
    \begin{gather*}
        \RulX{U-Var}
             {}
             {\JudgeTy{\TyEnv, \TmVar : \Ty}
                      {\TmVar}
                      {\Ty}
             }
        \quad
        \RulX{U-Con}
             {}
             {\JudgeTy{\TyEnv}
                      {\TmCon{\Ty}}
                      {\Ty}
             }
        \quad
        \RulX{U-Crash}
             {}
             {\JudgeTy{\TyEnv}
                      {\TmCrash{\Ty}{\ExnLbl}}
                      {\Ty}
             }
        \quad
        \RulX{U-Abs}
             {\JudgeTy{\TyEnv, \TmVar : \Ty_1}
                      {\Tm}
                      {\Ty_2}
             }
             {\JudgeTy{\TyEnv}
                      {\TmAbs{\TmVar}{\Ty_1}{\Tm}}
                      {\TyArr{\Ty_1}{\Ty_2}}
             }
        \\[1.5ex]
        \RulX{U-App}
             {\JudgeTy{\TyEnv}
                      {\Tm_1}
                      {\TyArr{\Ty_2}{\Ty}}
              \quad
              \JudgeTy{\TyEnv}
                      {\Tm_2}
                      {\Ty_2}
             }
             {\JudgeTy{\TyEnv}
                      {\TmApp{\Tm_1}{\Tm_2}}
                      {\Ty}
             }
        \quad
        \RulX{U-Fix}
             {\JudgeTy{\TyEnv,\TmVar:\Ty}
                      {\Tm}
                      {\Ty}
             }
             {\JudgeTy{\TyEnv}
                      {\TmFIX{\TmVar}{\Ty}{\Tm}}
                      {\Ty}
             }
        \quad
        \RulX{U-Op}
             {\JudgeTy{\TyEnv}
                      {\Tm_1}
                      {\TyInt}
              \quad
              \JudgeTy{\TyEnv}
                      {\Tm_2}
                      {\TyInt}
             }
             {\JudgeTy{\TyEnv}
                      {\TmOp{\Tm_1}{\Tm_2}}
                      {\TyBool}
             }
        \\[1.5ex]
        \RulX{U-Seq}
             {\JudgeTy{\TyEnv}
                      {\Tm_1}
                      {\Ty_1}
              \quad
              \JudgeTy{\TyEnv}
                      {\Tm_2}
                      {\Ty_2}
             }
             {\JudgeTy{\TyEnv}
                      {\TmSeq{\Tm_1}{\Tm_2}}
                      {\Ty_2}
             }
        \quad
        \RulX{U-If}
             {\JudgeTy{\TyEnv}
                      {\Tm_1}
                      {\TyBool}
              \quad
              \JudgeTy{\TyEnv}
                      {\Tm_2}
                      {\Ty}
              \quad
              \JudgeTy{\TyEnv}
                      {\Tm_3}
                      {\Ty}
             }
             {\JudgeTy{\TyEnv}
                      {\TmIf{\Tm_1}{\Tm_2}{\Tm_3}}
                      {\Ty}
             }             
        \\[1.5ex]
        \RulX{U-Nil}
             {}
             {\JudgeTy{\TyEnv}
                      {\TmNil{\Ty}}
                      {\TyList{\Ty}}
             }
        \quad
        \RulX{U-Cons}
             {\JudgeTy{\TyEnv}
                      {\Tm_1}
                      {\Ty}
              \quad
              \JudgeTy{\TyEnv}
                      {\Tm_2}
                      {\TyList{\Ty}}
             }
             {\JudgeTy{\TyEnv}
                      {\TmCons{\Tm_1}{\Tm_2}}
                      {\TyList{\Ty}}
             }
        \\[1.5ex]
        \RulX{U-Case}
             {%\begin{gathered}
                  \JudgeTy{\TyEnv}
                          {\Tm_1}
                          {\TyList{\Ty_1}}
                  \quad
                  \JudgeTy{\TyEnv}
                          {\Tm_2}
                          {\Ty}
                  \quad
                  \JudgeTy{\TyEnv, \TmVar_1 : \Ty_1, \TmVar_2 : \TyList{\Ty_1}}
                          {\Tm_3}
                          {\Ty}
              %\end{gathered}
             }
             {\JudgeTy{\TyEnv}
                      {\TmCase{\Tm_1}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}}
                      {\Ty}
             }
    \end{gather*}
%    \caption{Underlying type system ($\JudgeTy{\TyEnv}{\Tm}{\Ty}$)}
%    \label{figure-underlying-type-system}   
%\end{figure*}

\subsection{Operational semantics}

The operational semantics of the source language is given below. Note that there is a small amount of non-determinism in the order of reduction. For example, in the reduction rules for primitive operators.

We do not go so far as to have an \emph{imprecise exception semantics} \cite{PeytonJones:1999:SIE:301618.301637}. I.e., when the guard of a conditional evaluates to an exceptional value, we do not continue evaluation of the two branches in exception finding mode.

%\begin{figure*}[p]
    \begin{gather*}
        \RulX{E-App}
             {\JudgeEval{\Tm_1}
                        {\Tm_1^\prime}
             }
             {\JudgeEval{\TmApp{\Tm_1}{\Tm_2}}
                        {\TmApp{\Tm_1^\prime}{\Tm_2}}
             }
        \quad
        \RulX{E-AppAbs}
             {}
             {\JudgeEval{\TmApp{(\TmAbsExn{\TmVar}{\ExnTy}{\Exn}{\Tm_1})}{\Tm_2}}
                        {\Subst{\Tm_2}{\TmVar}{\Tm_1}}
             }
        \HOR
        \RulX{E-AppExn}
             {}
             {\JudgeEval{\TmApp{\TmCrash{}{\ExnLbl}}{\Tm_2}}
                        {\TmCrash{}{\ExnLbl}}
             }
        \VER
        \RulX{E-AnnApp}
             {\JudgeEval{\Tm}
                        {\Tm^\prime}
             }
             {\JudgeEval{\TmAnnApp{\Tm}{\Exn}}
                        {\TmAnnApp{\Tm^\prime}{\Exn}}
             }
        \HOR
        \RulX{E-AnnAppAbs}
             {}
             {\JudgeEval{\TmAnnApp{(\TmAnnAbs{\TmExnVar}{\Kind}{\Tm})}{\Exn}}
                        {\Subst{\Exn}{\TmExnVar}{\Tm}}
             }
        \VER
        \RulX{E-Fix$_1$}
             {\JudgeEval{\Tm}
                        {\Tm^\prime}
             }
             {\JudgeEval{\TmFixExn{\TmVar}{\ExnTy}{\Exn}{\Tm}}
                        {\TmFixExn{\TmVar}{\ExnTy}{\Exn}{\Tm^\prime}}
             }
        \quad
        \RulX{E-Fix$_2$}
             {}
             {\JudgeEval{\TmFixExn{\TmVar}{\ExnTy}{\Exn}{\Tm}}
                        {\Subst{\TmFixExn{\TmVar}{\ExnTy}{\Exn}{\Tm}}{\TmVar}{\Tm}}
             }
%        \HOR
%        \RulX{E-FixExn}
%             {}
%             {\JudgeEval{\TmFix{\TmCrash{}{\ExnLbl}}}
%                        {\TmCrash{}{\ExnLbl}}
%             }
        \VER
        \RulX{E-Op$_1$}
             {\JudgeEval{\Tm_1}{\Tm_1^\prime}}
             {\JudgeEval{\TmOp{\Tm_1}{\Tm_2}}{\TmOp{\Tm_1^\prime}{\Tm_2}}}
        \quad
        \RulX{E-Op$_2$}
             {\JudgeEval{\Tm_2}{\Tm_2^\prime}}
             {\JudgeEval{\TmOp{\Tm_1}{\Tm_2}}{\TmOp{\Tm_1}{\Tm_2^\prime}}}
        \HOR
        \RulX{E-Op}
             {}
             {\JudgeEval{\TmOp{\TmVal_1}{\TmVal_2}}{\Interp{\TmOp{\TmVal_1}{\TmVal_2}}}}
        \VER
        \RulX{E-OpExn$_1$}
             {}
             {\JudgeEval{\TmOp{\TmCrash{}{\ExnLbl}}{\Tm_2}}{\TmCrash{}{\ExnLbl}}}
        \quad
        \RulX{E-OpExn$_2$}
             {}
             {\JudgeEval{\TmOp{\Tm_1}{\TmCrash{}{\ExnLbl}}}{\TmCrash{}{\ExnLbl}}}
        \\[1.5ex]
        \RulX{E-Seq$_1$}
             {\JudgeEval{\Tm_1}{\Tm_1^\prime}}
             {\JudgeEval{\TmSeq{\Tm_1}{\Tm_2}}{\TmSeq{\Tm_1^\prime}{\Tm_2}}}
        \quad
        \RulX{E-Seq$_2$}
             {}
             {\JudgeEval{\TmSeq{\TmVal_1}{\Tm_2}}{\Tm_2}}
        \quad
        \RulX{E-SeqExn}
             {}
             {\JudgeEval{\TmSeq{\TmCrash{}{\ExnLbl}}{\Tm_2}}{\TmCrash{}{\ExnLbl}}}
        \\[1.5ex]
        \RulX{E-If}
             {\JudgeEval{\Tm_1}{\Tm_1^\prime}}
             {\JudgeEval{\TmIf{\Tm_1}{\Tm_2}{\Tm_3}}{\TmIf{\Tm_1^\prime}{\Tm_2}{\Tm_3}}}
        \quad
        \RulX{E-IfTrue}
             {}
             {\JudgeEval{\TmIf{\TmTrue}{\Tm_2}{\Tm_3}}{\Tm_2}}
        \\[1.5ex]
        \RulX{E-IfFalse}
             {}
             {\JudgeEval{\TmIf{\TmFalse}{\Tm_2}{\Tm_3}}{\Tm_3}}
        \quad
        \RulX{E-IfExn}
             {}
             {\JudgeEval{\TmIf{\TmCrash{}{\ExnLbl}}{\Tm_2}{\Tm_3}}{\TmCrash{}{\ExnLbl}}}
        \\[1.5ex]
        \RulX{E-Case}
             {\JudgeEval{\Tm_1}{\Tm_1^\prime}}
             {\JudgeEval{\TmCase{\Tm_1}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}}
                        {\TmCase{\Tm_1^\prime}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}}
             }
        \\[1.5ex]
        \RulX{E-CaseNil}
             {}
             {\JudgeEval{\TmCase{\TmNil{}}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}}
                        {\Tm_2}
             }
        \\[1.5ex]
        \RulX{E-CaseCons}
             {}
             {\JudgeEval{\TmCase{\TmCons{\Tm_1}{\Tm_1^\prime}}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}}
                        {\SubstTwo{\Tm_1}{\TmVar_1}{\Tm_1^\prime}{\TmVar_2}{\Tm_3}}
             }
        \\[1.5ex]
        \RulX{E-CaseExn}
             {}
             {\JudgeEval{\TmCase{\TmCrash{}{\ExnLbl}}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}}
                        {\TmCrash{}{\ExnLbl}}
             }
    \end{gather*}
%    \caption{Operational semantics ($\JudgeEval{\Tm_1}{\Tm_2}$)}
%    \label{figure-operational-semantics}
%\end{figure*}

The reduction rules on the second row apply to constructs that are introduced to the language in Section~\ref{section-exception-types}. This also holds for the additional annotations on the $\lambda$-abstraction and the $\mathbf{fix}$-operator.

\section{Exception types}\label{section-exception-types}

%\subsection{Exception types}

The syntax of well-formed exception types is given in Figure~\ref{fig-exception-type-combined}.
We let $\ExnVar$ range over an infinite set of exception set variables and $\ExnLbl$ over a finite set of exception labels.
An exception type~$\ExnTy$ is formed out of base types (booleans and integers), compound types (lists), function types, and quantifiers (ranging over exception set variables---to avoid complicating the presentation we do \emph{not} allow quantification over type variables, i.e. polymorphism in the underlying type system).

\begin{figure}[h]
    \paragraph{Kinds, types and exception annotations}
    \begin{alignat}{2}
        \Kind \in \ClassKind &\ ::=\ &&\KindEXN \tag{exception set} \\
                             &\quad|  &&\Kind_1 \KindArr \Kind_2 \tag{exception set operator}\\
        \notag\\
        \Exn, \ExnAlt \in \ClassExn  &\ ::=\ &&\ExnVar \tag{exception set variables} \\
                            &\quad|  &&\ExnAbs{\ExnVar}{\Kind}{\Exn} \tag{exception set abstraction}\\
                            &\quad|    &&\ExnApp{\Exn_1}{\Exn_2} \tag{exception set application}\\
                            &\quad|    &&\ExnEmpty \tag{empty exception set} \\
                            &\quad|    &&\ExnCon{\ExnLbl} \tag{singleton exception set} \\
                            &\quad|    &&\ExnUnion{\Exn_1}{\Exn_2} \tag{exception set union} \\
    \notag\\
        \ExnTy \in \ClassExnTy &\ ::=\ &&\ExnForall{\ExnVar :: \Kind}\ExnTy \tag{exception set quantification} \\
                     &\quad|  &&\ExnBool \tag{boolean type} \\
                     &\quad|  &&\ExnInt  \tag{integer type} \\
                     &\quad|  &&\ExnTyList{\ExnTy}{\Exn} \tag{list type} \\
                     &\quad|  &&\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2} \tag{function type}
    \end{alignat}
%    \caption{Exception types: syntax}
%    \label{fig-exception-type-syntax}
%\end{figure}
%\begin{figure}[h]
    \paragraph{Well-formedness}
    \begin{gather*}
        \Rule{W-Forall}
             {\JudgeExnTyWff{d}{\KiEnv,\ExnVar :: \Kind}{\ExnTy}}
             {\JudgeExnTyWff{d}{\KiEnv}{\ExnTyForall{\ExnVar}{\Kind}{\ExnTy}}}
        \VER
        \Rule{W-Bool}
             {}
             {\JudgeExnTyWff{d}{\KiEnv}{\ExnBool}}
        \HOR
        \Rule{W-Int}
             {}
             {\JudgeExnTyWff{d}{\KiEnv}{\ExnInt}}
        \VER
        \Rule{W-List}
             {\JudgeExnTyWff{d}{\KiEnv}{\ExnTy} \quad \JudgeKind{\KiEnv}{\Exn}{\KindEXN}}
             {\JudgeExnTyWff{d}{\KiEnv}{\ExnTyList{\ExnTy}{\Exn}}}
        \VER
        \Rule{W-Arr}
             {\begin{gathered}\JudgeExnTyWff{d}{\KiEnv}{\ExnTy_1} \  \JudgeKind{\KiEnv}{\Exn_1}{\KindEXN} \  \JudgeExnTyWff{d}{\KiEnv}{\ExnTy_2} \  \JudgeKind{\KiEnv}{\Exn_2}{\KindEXN}\end{gathered}}
             {\JudgeExnTyWff{d}{\KiEnv}{\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}}}
    \end{gather*}
     \caption{Exception types: syntax and well-formedness ($\JudgeExnTyWff{d}{\KiEnv}{\ExnTy}$)}
%    \label{fig-exception-type-well-formed}
    \label{fig-exception-type-combined}
\end{figure}

%Type signatures are denoted as $f : \ExnTy\ \&\ \Exn$, where $\ExnTy$ is the exception type of $f$ and $\Exn$ is the (top-most) effect.

For a list with exception type $\ExnTyList{\ExnTy}{\Exn}$ and effect $\ExnAlt$, the type $\ExnTy$ of the elements in the list is \emph{annotated} with an exception set expression~$\Exn$ of kind~$\KindEXN$. This expression gives a set of exceptions, from which any one may be raised when an element of the list is forced. The effect $\ExnAlt$ gives a set of exceptions, from which any one may be raised when a constructor forming the spine of the list is forced.

For a function with exception type $\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}$ and effect~$\ExnAlt$, the argument of type $\ExnTy_1$ is annotated with an exception set expression $\Exn_1$ that gives a set of exceptions that may be raised if the argument is forced in the body of the function. The result of type $\ExnTy_2$ is annotated with an exception set expression $\Exn_2$ that gives the set of exceptions that may be raised when the result of the function is forced. The effect $\ExnAlt$ gives the set of exceptions from which any one may be raised when the function closure is forced.

\begin{example}
The identity function
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}c<{\hspost}@{}}%
\column{9E}{@{}l@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[5]{}\Varid{id}{}\<[9]%
\>[9]{}:{}\<[9E]%
\>[13]{}\forall \Varid{e}\hsforall \hsdot{\circ }{.}\ExnTyArr{\ExnBool}{\Varid{e}}{\ExnBool}{\Varid{e}}\ \&\ \emptyset{}\<[E]%
\\
\>[5]{}\Varid{id}{}\<[9]%
\>[9]{}\mathrel{=}{}\<[9E]%
\>[13]{}\lambda\Varid{x}.\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    propagates any exceptional value passed to it as an argument to the result unchanged. As the identity function is constructed by a literal $\lambda$-abstraction, no exception is raised when the resulting closure is forced, hence the empty effect.
\end{example}

\begin{example}
The exceptional function value
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{22}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[5]{}\lightning^\mathbf{E}_{\TyBool \to \TyBool}:{}\<[22]%
\>[22]{}\forall \Varid{e}\hsforall \hsdot{\circ }{.}\ExnTyArr{\ExnBool}{\Varid{e}}{\ExnBool}{\emptyset}\ \&\ \{\mathbf{E}\}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    raises an exception when its closure is forced---as happens when it is applied to an argument, for example. As this function can never produce a result, it certainly cannot produce an exceptional value. So the result type is annotated with an empty exception set.
\end{example}

The exception set expressions $\Exn$ and their kinds $\Kind$ are an instance of the \mbox{$\LambdaUnion$-calculus}, where exception set expressions are terms and kinds are the types. As the constants we take the set of exception labels present in the program.
Two exception set expressions are considered equivalent if they are convertible as $\LambdaUnion$-terms, which is to say that they reduce to the same normal form.

The type system resembles \SystemFw \cite{Girard1972} in that we have quantification, abstraction and application at the type level. A key difference is that abstraction and application are restricted to the effects ($\ClassExn$) and cannot be used in the types ($\ClassExnTy$) directly.
Quantification, on the other hand, is restricted to the types, where it ranges over effects, and is not allowed to appear in the effect itself. The types thus remain predicative.

\subsection{Subtyping}

Exception types are endowed with the usual subtyping relation for type-and-effect systems. The function type is contravariant in its first argument for both the type and the effect. The subeffecting relation $\JudgeSubExn{\KiEnv}{\Exn_1}{\Exn_2}$ is the subsumption relation $\JudgeLuSub{\LuTm_1}{\LuTm_2}$ from the $\LambdaUnion$-calculus (Definition~\ref{definition-subsumption}).

%\begin{figure}[h]
    \begin{gather*}
        \Rule{S-Refl}
             {}
             {\JudgeSubTy{\KiEnv}{\ExnTy}{\ExnTy}}
        \quad
        \Rule{S-Trans}
             {\JudgeSubTy{\KiEnv}{\ExnTy_1}{\ExnTy_2}
              \quad
              \JudgeSubTy{\KiEnv}{\ExnTy_2}{\ExnTy_3}
             }
             {\JudgeSubTy{\KiEnv}{\ExnTy_1}{\ExnTy_3}}
        \\[1.5ex]
        \Rule{S-Forall}
             {\JudgeSubTy{\KiEnv, \ExnVar :: \Kind}{\ExnTy_1}{\ExnTy_2}}
             {\JudgeSubTy{\KiEnv}
                         {\ExnForall{\ExnVar :: \Kind}{\ExnTy_1}}
                         {\ExnForall{\ExnVar :: \Kind}{\ExnTy_2}}
             }
        \quad
        \Rule{S-List}
             {\JudgeSubTy{\KiEnv}{\ExnTy}{\ExnTy^\prime}
                \quad
              \JudgeSubExn{\KiEnv}{\Exn}{\Exn^\prime}
             }
             {\JudgeSubTy{\KiEnv}
                         {\ExnTyList{\ExnTy}{\Exn}}
                         {\ExnTyList{\ExnTy^\prime}{\Exn^\prime}}
             }
        \\[1.5ex]
        %\Rule{S-Bool}
        %     {}
        %     {\JudgeSubTy{\KiEnv}{\ExnBool}{\ExnBool}}
        %\quad
        %\Rule{S-Int}
        %     {}
        %     {\JudgeSubTy{\KiEnv}{\ExnInt}{\ExnInt}}
        %\\[1.5ex]
        \Rule{S-Arr}
             {\begin{gathered}
              \JudgeSubTy{\KiEnv}{\ExnTy_1^\prime}{\ExnTy_1}
                \ \ 
              \JudgeSubExn{\KiEnv}{\Exn_1^\prime}{\Exn_1}
                \ \ 
              \JudgeSubTy{\KiEnv}{\ExnTy_2}{\ExnTy_2^\prime}
                \ \ 
              \JudgeSubExn{\KiEnv}{\Exn_2}{\Exn_2^\prime}
              \end{gathered}
             }
             {\JudgeSubTy{\KiEnv}
                         {\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}}
                         {\ExnTyArr{\ExnTy_1^\prime}{\Exn_1^\prime}{\ExnTy_2^\prime}{\Exn_2^\prime}}
             }
    \end{gather*}
%\caption{Exception types: subtyping relation ($\JudgeSubTy{\KiEnv}{\ExnTy_1}{\ExnTy_2}$)}\label{figure-subtyping}
%\end{figure}

\subsection{Conservative types}

Any program that is typeable in the underlying type system should also have an exception type: the exception type system is a \emph{conservative extension} of the underlying type system. Like type systems for strictness or control flow analysis---and unlike type systems for information flow security or dimensional analysis---we do not want to reject any program that is well-typed in the underlying type system, but merely provide more insight into its behavior.

If we furthermore want the type system to be modular---allowing type checking and inference to work on individual modules instead of whole programs---we cannot and need not make any assumptions about the exception types of the arguments that are applied to any function, as the function may be called from outside the module with an argument that also comes from outside the module and which we cannot know anything about.

For base and compound types that stand in an argument position their effect and any nested annotations must thus be able to be instantiated to any arbitrary exception set expression. They must therefore be exception set variables that have been universally quantified.

These observations lead to the following definition of \emph{conservative exception types}:\footnote{\cite{HoldermansHage} call pattern types \emph{fully parametric} and conservative types \emph{fully flexible}.}

\begin{definition}
An exception set expression $\Exn$ is \emph{simple} if it is a single exception set variable $\ExnVar$, an exception set expression is a \emph{pattern} if it fits Definition~\ref{definition-pattern}, and any exception set expression is \emph{conservative}.

We lift these three judgments to exception types $\ExnTy$ in the following manner:

    \begin{itemize}

        \item If $\ExnTy = \ExnBool$ or $\ExnTy = \ExnInt$, then $\ExnTy$ is simple, a pattern and conservative.

        \item If $\ExnTy = [\Throws{\ExnTy^\prime}{\Exn}]$, then $\ExnTy$ is simple, a pattern or conservative if $\ExnTy^\prime$ and $\Exn$ are respectively simple, patterns  or conservative.

        \item If $\ExnTy = \forall \overline{\ExnVar_i :: \kappa_i}.\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}$, then $\ExnTy$ is both simple and a pattern if $\ExnTy_1$ and $\Exn_1$ are simple and $\ExnTy_2$ and $\Exn_2$ are patterns; and $\ExnTy$ is conservative if $\ExnTy_1$ and $\Exn_1$ are simple and $\ExnTy_2$ and $\Exn_2$ are conservative.

    \end{itemize}
\end{definition}

\begin{example} The function \ensuremath{\Varid{tail}} can be applied to any list, but may produce an additional exceptional value \ensuremath{\mathbf{E}}, because it is partial:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[5]{}\Varid{tail}:\forall \Varid{e}_{\mathrm{1}}\hsforall \;\Varid{e}_{\mathrm{2}}\hsdot{\circ }{.}\ExnTyArr{\ExnTyList{\ExnBool}{\Varid{e}_{\mathrm{1}}}}{\Varid{e}_{\mathrm{2}}}{\ExnTyList{\ExnBool}{\Varid{e}_{\mathrm{1}}}}{\Varid{e}_{\mathrm{2}}\cup \{\mathbf{E}\}}\ \&\ \emptyset{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The type and effect of the argument are simple, while the type and effect of the result are conservative, making the whole type conservative.

The conjunction operator \ensuremath{\mathrel{\wedge}} can be applied to any two booleans, and---operators being strict in both arguments---will propagate any exceptional values:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{11}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[5]{}\land {}\<[11]%
\>[11]{}:\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\ExnTyArr{\ExnBool}{\Varid{e}_{\mathrm{1}}}{(\forall \Varid{e}_{\mathrm{2}}\hsforall \hsdot{\circ }{.}\ExnTyArr{\ExnBool}{\Varid{e}_{\mathrm{2}}}{\ExnBool}{\Varid{e}_{\mathrm{1}}\cup \Varid{e}_{\mathrm{2}}})}{\emptyset}\ \&\ \emptyset{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Here both arguments have simple types and effects.
\end{example}

For function types that stand in an argument position (the functional parameters of a higher-order function) the situation is slightly more complicated. For the argument of this function we can inductively assume that this is a universally quantified exception set variable. The result of this function, however, is some exception set expression that depends on the exception set variables that were quantified over in the argument. We cannot simply introduce a new exception set variable here, but must introduce a Skolem function that depends on each of the universally quantified exception set variables.

\begin{example} Consider the higher-order function \ensuremath{\Varid{apply}} that applies its first argument to the second.
% apply : A e2:E. A e3:E=>E.
%           (A e1:E. bool<e1> --<e2>-> bool<e3 e2>) ->
%               A e4:E. bool<e4> -> bool<e2 + e3 e4>
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}c<{\hspost}@{}}%
\column{12E}{@{}l@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[5]{}\Varid{apply}{}\<[12]%
\>[12]{}:{}\<[12E]%
\>[17]{}\forall \Varid{e}_{\mathrm{2}}\hsforall ::\KindEXN\hsdot{\circ }{.}\forall \Varid{e}_{\mathrm{3}}\hsforall ::\KindEXN\Rightarrow \KindEXN\hsdot{\circ }{.}{}\<[E]%
\\
\>[17]{}\hsindent{2}{}\<[19]%
\>[19]{}(\forall \Varid{e}_{\mathrm{1}}\hsforall ::\KindEXN\hsdot{\circ }{.}\ExnTyArr{\ExnBool}{\Varid{e}_{\mathrm{1}}}{\ExnBool}{\Varid{e}_{\mathrm{3}}\;\Varid{e}_{\mathrm{1}}})\langle\Varid{e}_{\mathrm{2}}\rangle\rightarrow\;{}\<[E]%
\\
\>[19]{}\hsindent{2}{}\<[21]%
\>[21]{}(\forall \Varid{e}_{\mathrm{4}}\hsforall ::\KindEXN\hsdot{\circ }{.}\ExnTyArr{\ExnBool}{\Varid{e}_{\mathrm{4}}}{\ExnBool}{\Varid{e}_{\mathrm{2}}\cup \Varid{e}_{\mathrm{3}}\;\Varid{e}_{\mathrm{4}}})\langle\ExnEmpty \rangle{}\<[E]%
\\
\>[12]{}\ \&\ {}\<[12E]%
\>[17]{}\ExnEmpty {}\<[E]%
\\
\>[5]{}\Varid{apply}{}\<[12]%
\>[12]{}\mathrel{=}{}\<[12E]%
\>[15]{}\lambda\Varid{f}.\lambda\Varid{x}.\Varid{f}\;\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The first (functional) argument of \ensuremath{\Varid{apply}} has exception type \ensuremath{\forall \Varid{e}_{\mathrm{1}}\hsforall ::\KindEXN\hsdot{\circ }{.}\ExnTyArr{\ExnBool}{\Varid{e}_{\mathrm{1}}}{\ExnBool}{\Varid{e}_{\mathrm{3}}\;\Varid{e}_{\mathrm{1}}}} and effect \ensuremath{\Varid{e}_{\mathrm{2}}}. It can be instantiated with any function that accepts an argument annotated with any exception set effect, and produces a result annotated with some exception set effect depending on the exception set effect of the argument; the function closure itself may raise any exception. All functions of underlying type \ensuremath{\TyBool\to \TyBool} satisfy these constraints, so we are not really constrained at all.

As \ensuremath{\Varid{e}_{\mathrm{1}}} has been quantified over, only the exception set operator~\ensuremath{\Varid{e}_{\mathrm{3}}} and the effect \ensuremath{\Varid{e}_{\mathrm{2}}} are left free. We quantify over them outside the outer function space constructor, allowing them to appear in the annotation \ensuremath{\Varid{e}_{\mathrm{2}}\cup \Varid{e}_{\mathrm{3}}\;\Varid{e}_{\mathrm{4}}} on the result. The exception set operator \ensuremath{\Varid{e}_{\mathrm{3}}} is now applied to \ensuremath{\Varid{e}_{\mathrm{4}}}, as the term-level application \ensuremath{\Varid{f}\;\Varid{x}} instantiates the quantified exception set variable \ensuremath{\Varid{e}_{\mathrm{1}}} to \ensuremath{\Varid{e}_{\mathrm{4}}}.

(Note that the exception annotation $e_2$ on the closure---unlike the exception set operator $e_3$ on the result---does not depend on the exception variable $e_1$, the annotation on the argument. As a closure is already a value, it being exceptional or not can never depend on the argument it is later applied to.)
\end{example}

\begin{example}
    The semantics of terms in the source language is not invariant under $\eta$-conversion in the presence of exceptional values---thus neither are exception types. The term
    \savecolumns
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{42}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\lambda\Varid{x}:\TyBool.\TmCrash{\TyBool\to\TyBool}{\ExnLblE}\;\Varid{x}{}\<[42]%
\>[42]{}:\forall \Varid{e}\hsforall ::\KindEXN\hsdot{\circ }{.}\ExnTyArrT{\ExnBool}{\Varid{e}}{\ExnBool}{\{\mathbf{E}\}}{\emptyset}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    does not have the same exception type as the $\eta$-equivalent term
    \restorecolumns
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{42}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\TmCrash{\TyBool\to\TyBool}{\ExnLblE}{}\<[42]%
\>[42]{}:\forall \Varid{e}\hsforall ::\KindEXN\hsdot{\circ }{.}\ExnTyArrT{\ExnBool}{\Varid{e}}{\ExnBool}{\emptyset}{\{\mathbf{E}\}}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    They cannot be distinguished by applying them to an argument
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
\column{50}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}(\lambda\Varid{x}:\TyBool.\TmCrash{\TyBool\to\TyBool}{\ExnLblE}\;\Varid{x})\;{}\<[44]%
\>[44]{}\mathbf{true}{}\<[50]%
\>[50]{}:\ExnBool\ \&\ \{\mathbf{E}\}{}\<[E]%
\\
\>[9]{}\TmCrash{\TyBool\to\TyBool}{\ExnLblE}\;{}\<[44]%
\>[44]{}\mathbf{true}{}\<[50]%
\>[50]{}:\ExnBool\ \&\ \{\mathbf{E}\}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    but they can be distinguished by forcing the closure
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{44}{@{}>{\hspre}l<{\hspost}@{}}%
\column{49}{@{}>{\hspre}l<{\hspost}@{}}%
\column{55}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}(\lambda\Varid{x}:\TyBool.\TmCrash{\TyBool\to\TyBool}{\ExnLblE}\;\Varid{x})\;{}\<[44]%
\>[44]{}\mathbf{seq}\;{}\<[49]%
\>[49]{}\mathbf{true}{}\<[55]%
\>[55]{}:\ExnBool\ \&\ \emptyset{}\<[E]%
\\
\>[9]{}\TmCrash{\TyBool\to\TyBool}{\ExnLblE}\;{}\<[44]%
\>[44]{}\mathbf{seq}\;{}\<[49]%
\>[49]{}\mathbf{true}{}\<[55]%
\>[55]{}:\ExnBool\ \&\ \{\mathbf{E}\}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{example}

\subsection{Exception type completion}

Given an underlying type $\Ty$ we can compute the most general exception type $\ExnTy$ that erases to $\Ty$. This is done using the type completion system below, that defines a type completion relation $\JudgeComplete{\KindENV}{\Ty}{\ExnTy}{\Exn}{\KindENV'}$.

    \begin{gather*}
        \Rule{C-Bool}
             {}
             {\JudgeComplete{\KindEnv{i}}
                            {\TyBool}
                            {\ExnBool}
                            {\ExnVar\ \overline{\ExnVar_i}}
                            {\ExnVar :: \overline{\Kind_i\KindArr}\KindEXN}
             }
        \VER
        \Rule{C-Int}
             {}
             {\JudgeComplete{\KindEnv{i}}
                            {\TyInt}
                            {\ExnInt}
                            {\ExnVar\ \overline{\ExnVar_i}}
                            {\ExnVar :: \overline{\Kind_i\KindArr}\KindEXN}
             }
        \VER
        \Rule{C-List}
             {\JudgeComplete{\KindEnv{i}}
                            {\Ty}
                            {\ExnTy}
                            {\Exn}
                            {\KindEnv{j}}
             }
             {\JudgeComplete{\KindEnv{i}}
                            {\TyList{\Ty}}
                            {\ExnTyList{\ExnTy}{\Exn}}
                            {\ExnVar\ \overline{\ExnVar_i}}
                            {\ExnVar :: \overline{\Kind_i\KindArr}%
                                \KindEXN, \KindEnv{j}}
             }
        \VER
        \Rule{C-Arr}
             {\JudgeComplete{}
                            {\Ty_1}
                            {\ExnTy_1}
                            {\Exn_1}
                            {\KindEnv{j}}
              \quad\quad\quad\quad
              \JudgeComplete{\KindEnv{i},\KindEnv{j}}
                            {\Ty_2}
                            {\ExnTy_2}
                            {\Exn_2}
                            {\KindEnv{k}}
             }
             {\JudgeComplete{\KindEnv{i}}
                            {\TyArr{\Ty_1}{\Ty_2}}
                            {\ExnForall{\overline{\ExnVar_j :: \Kind_j}}\left(\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}\right)}
                            {\ExnVar\ \overline{\ExnVar_i}}
                            {\ExnVar :: \overline{\Kind_i\KindArr}%
                                \KindEXN, \KindEnv{k}}
             }
        \VER
    \end{gather*}

A judgment $\JudgeComplete{\KindEnv{i}}{\Ty}{\ExnTy}{\Exn}{\KindEnv{j}}$ is read: if the kinded exception set variables $\KindEnv{i}$ are in scope, then the underlying type $\Ty$ is completed to the exception type $\ExnTy$ and effect $\Exn$, while introducing the kinded free exception set variables $\KindEnv{j}$.
A completed exception type is always a pattern type.

\begin{example} The higher-order underlying type \[ \TyArr{\TyList{\TyArr{\TyBool}{\TyBool}}}{\TyArr{\TyList{\TyBool}}{\TyList{\TyBool}}} \] is completed to the pattern type
    \begin{align*}
        &    {\ExnTyForall{\ExnVar_2}{\KindEXN}{\ExnTyForall{\ExnVar_2^\prime}{\KindEXN}{\ExnTyForall{\ExnVar_3}{\KindEXN\KindArr\KindEXN}{}}}} \\
        & \quad \Throws{[\ExnTyForall{\ExnVar_1}{\KindEXN}{\Throws{\ExnBool}{\ExnVar_1}\xrightarrow{\ExnVar_2^\prime}\Throws{\ExnBool}{\ExnVar_3\ \ExnVar_1}}]}{\ExnVar_2} \to \\
        & \quad\quad (\ExnTyForall{\ExnVar_5}{\KindEXN}{\ExnTyForall{\ExnVar_5^\prime}{\KindEXN}{\Throws{[\Throws{\ExnBool}{\ExnVar_5^\prime}]}{\ExnVar_5}}} \xrightarrow{\ExnVar_6\ \ExnVar_2\ \ExnVar_2^\prime\ \ExnVar_3} \\
        & \quad\quad\quad\quad\quad\quad \Throws{\ExnTyList{\ExnBool}{\ExnVar_7^\prime\ \ExnVar_2\ \ExnVar_2^\prime\ \ExnVar_3\ \ExnVar_5\ \ExnVar_5^\prime}}{\ExnVar_7\ \ExnVar_2\ \ExnVar_2^\prime\ \ExnVar_3\ \ExnVar_5\ \ExnVar_5^\prime})
    \end{align*}
    with effect $\ExnVar_4$, and while introducing the free \ThesisPaper{exception set}{} variables
    \begin{align*}
        \ExnVar_4  &:: \KindEXN, \\
        \ExnVar_6 &:: \KindEXN\KindArr\KindEXN\KindArr(\KindEXN\KindArr\KindEXN)\KindArr\KindEXN, \\
        \ExnVar_7, \ExnVar_7^\prime &:: \KindEXN\KindArr\KindEXN\KindArr(\KindEXN\KindArr\KindEXN)\KindArr\KindEXN\KindArr\KindEXN\KindArr\KindEXN
    \end{align*}

Note that the types of both arguments are simple types with simple exception annotations. However, as the first argument is a functional argument, the result type of that function is still a pattern.

The exception annotation on the right-most function-space constructor is a pattern that depends on $\ExnVar_2$, $\ExnVar_2^\prime$ and $\ExnVar_3$. While we previously noted that the annotation on a function-space constructor cannot depend on the annotation belonging to the argument of that function, it is possible for a set of exceptional values that the closure may come to depend on any previous arguments of the whole function. This is more concretely demonstrated by the following function:
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}c<{\hspost}@{}}%
\column{12E}{@{}l@{}}%
\column{15}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\Varid{f}{}\<[12]%
\>[12]{}::{}\<[12E]%
\>[15]{}\ExnTyForall{\ExnVar_1,\ExnVar_2}{\KindEXN}{\Throws{\ExnBool}{\ExnVar_1}\xrightarrow{\ExnEmpty}\Throws{\ExnBool}{\ExnVar_2}\xrightarrow{\ExnVar_1}\Throws{\ExnBool}{\ExnVar_2}}{}\<[E]%
\\
\>[9]{}\Varid{f}{}\<[12]%
\>[12]{}\mathrel{=}{}\<[12E]%
\>[15]{}\lambda x : \TyBool . x\ \mathbf{seq}\ \lambda y : \TyBool. y{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\end{example}
Whether the closure that is returned after partially applying \ensuremath{\Varid{f}} to one argument is an exceptional value or not, depends on that argument \ensuremath{\Varid{x}} being exceptional or not.

\subsection{Least exception types}

Besides completing an underlying type $\Ty$ to a most general exception type, we also want to compute a least exception type~$\bot_\Ty$.
Given an effect kind $\overline{\kappa_i\Rightarrow}\KindEXN$, denote by $\emptyset_{\overline{\kappa_i\Rightarrow}\KindEXN}$ the effect $\lambda \overline{e_i :: \kappa_i}. \emptyset$.
We can construct a least exception type by first completing the type $\Ty$ to the most general exception type, and then substituting $\emptyset_{\kappa_j}$ for all free freshly introduced exception set variables $\KindEnv{j}$.

\begin{example} The least exception type \[ \bot_{\TyArr{\TyList{\TyArr{\TyBool}{\TyBool}}}{\TyArr{\TyList{\TyBool}}{\TyList{\TyBool}}}} \] is the conservative type
    \begin{align*}
        &    {\ExnTyForall{\ExnVar_2}{\KindEXN}{\ExnTyForall{\ExnVar_2^\prime}{\KindEXN}{\ExnTyForall{\ExnVar_3}{\KindEXN\KindArr\KindEXN}{}}}} \\
        & \quad \Throws{[\ExnTyForall{\ExnVar_1}{\KindEXN}{\Throws{\ExnBool}{\ExnVar_1}\xrightarrow{\ExnVar_2^\prime}\Throws{\ExnBool}{\ExnVar_3\ \ExnVar_1}}]}{\ExnVar_2} \to \\
        & \quad\quad (\ExnTyForall{\ExnVar_5}{\KindEXN}{\ExnTyForall{\ExnVar_5^\prime}{\KindEXN}{\Throws{[\Throws{\ExnBool}{\ExnVar_5^\prime}]}{\ExnVar_5}}} \xrightarrow{\ExnEmpty} \Throws{\ExnTyList{\ExnBool}{\ExnEmpty}}{\ExnEmpty})
    \end{align*}
\end{example}

\begin{figure*}[p]
    \begin{gather*}
        \Rule{T-Var}
             {}
             {\JudgeExnTy{\TyEnv, \TmVar : \ExnTy\ \&\ \Exn}
                         {\KiEnv}
                         {\TmVar}
                         {\ExnTy}
                         {\Exn}
             }
        \VER
        \Rule{T-Con}
             {}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmCon{\Ty}}
                         {\BottomTy{\Ty}}
                         {\ExnEmpty}
             }
        \HOR
        \Rule{T-Crash}
             {}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmCrash{\Ty}{\ExnLbl}}
                         {\BottomTy{\Ty}}
                         {\ExnCon{\ExnLbl}}
             }
        \VER
        \Rule{T-Abs}
             {\JudgeExnTy{\TyEnv, \TmVar : \ExnTy_1\ \&\ \Exn_1}
                         {\KiEnv}
                         {\Tm}
                         {\ExnTy_2}
                         {\Exn_2}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmAbsExn{\TmVar}{\ExnTy_1}{\Exn_1}{\Tm}}
                         {\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}}
                         {\ExnEmpty}
             }
        \VER
        \Rule{T-AnnAbs}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv, \ExnVar :: \Kind}
                         {\Tm}
                         {\ExnTy}
                         {\Exn}
              \quad
              \ExnVar \notin \fv{\TyEnv,\Exn}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmAnnAbs{\ExnVar}{\Kind}{\Tm}}
                         {\ExnTyForall{\ExnVar}{\Kind}{\ExnTy}}
                         {\Exn}
             }
        \VER
        \Rule{T-App}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_1}
                         {\ExnTyArr{\ExnTy_2}{\Exn_2}{\ExnTy}{\Exn}}
                         {\Exn}
              \quad
              \JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_2}
                         {\ExnTy_2}
                         {\Exn_2}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmApp{\Tm_1}{\Tm_2}}
                         {\ExnTy}
                         {\Exn}
             }
        \VER
        \Rule{T-AnnApp}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_1}
                         {\ExnTyForall{\ExnVar}{\Kind}{\ExnTy}}
                         {\Exn}
              \quad
              \JudgeExn{\KiEnv}
                       {\Exn_2}
                       {\Kind}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmAnnApp{\Tm_1}{\Exn_2}}
                         {\Subst{\Exn_2}{\ExnVar}{\ExnTy}}
                         {\Exn}
             }
        \VER
        \Rule{T-Fix}
             {\begin{gathered}
              \JudgeExnTy{\TyEnv,\TmVar:\ExnTy\ \&\ \Exn}
                         {\KiEnv}
                         {\Tm}
                         {\ExnTy}
                         {\Exn}
              \end{gathered}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmFixExn{\TmVar}{\ExnTy}{\Exn}{\Tm}}
                         {\ExnTy}
                         {\Exn}
             }
        \VER
        \Rule{T-Op}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_1}
                         {\ExnInt}
                         {\Exn}
              \quad
              \JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_2}
                         {\ExnInt}
                         {\Exn}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmOp{\Tm_1}{\Tm_2}}
                         {\ExnBool}
                         {\Exn}
             }
        \VER
        \Rule{T-Seq}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_1}
                         {\ExnTy_1}
                         {\Exn}
              \quad
              \JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_2}
                         {\ExnTy_2}
                         {\Exn}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmSeq{\Tm_1}{\Tm_2}}
                         {\ExnTy_2}
                         {\Exn}
             }
        \VER
        \Rule{T-If}
             {\begin{gathered}
              \JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_1}
                         {\ExnBool}
                         {\Exn}
              \quad
              \JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_2}
                         {\ExnTy}
                         {\Exn}
              \quad
              \JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_3}
                         {\ExnTy}
                         {\Exn}
              \end{gathered}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmIf{\Tm_1}{\Tm_2}{\Tm_3}}
                         {\ExnTy}
                         {\Exn}
             }             
        \VER
        \Rule{T-Nil}
             {}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmNil{\Ty}}
                         {\ExnTyList{\BottomTy{\Ty}}{\ExnEmpty}}
                         {\ExnEmpty}
             }
        \VER
        \Rule{T-Cons}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_1}
                         {\ExnTy}
                         {\Exn_1}
              \quad
              \JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm_2}
                         {\ExnTyList{\ExnTy}{\Exn_1}}
                         {\Exn_2}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmCons{\Tm_1}{\Tm_2}}
                         {\ExnTyList{\ExnTy}{\Exn_1}}
                         {\Exn_2}
             }
        \VER
        \Rule{T-Case}
             {\begin{gathered}
                  \JudgeExnTy{\TyEnv}
                             {\KiEnv}
                             {\Tm_1}
                             {\ExnTyList{\ExnTy_1}{\Exn_1}}
                             {\Exn^\prime}
                  \quad
                  \JudgeExnTy{\TyEnv}
                             {\KiEnv}
                             {\Tm_2}
                             {\ExnTy}
                             {\Exn}
                  \\
                  \JudgeExnTy{\TyEnv, \TmVar_1 : \ExnTy_1\ \&\ \Exn_1, \TmVar_2 : \ExnTyList{\ExnTy_1}{\Exn_1}\ \&\ \Exn^\prime}
                             {\KiEnv}
                             {\Tm_3}
                             {\ExnTy}
                             {\Exn}
                    \quad
                    \JudgeSubExn{\KiEnv}{\Exn^\prime}{\Exn}
              \end{gathered}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\TmCase{\Tm_1}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}}
                         {\ExnTy}
                         {\Exn}
             }
        \VER
        \Rule{T-Sub}
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm}
                         {\ExnTy^\prime}
                         {\Exn^\prime}
              \quad
              \JudgeSubTy{\KiEnv}
                         {\ExnTy^\prime}
                         {\ExnTy}
              \quad
              \JudgeSubExn{\KiEnv}
                          {\Exn^\prime}
                          {\Exn}
             }
             {\JudgeExnTy{\TyEnv}
                         {\KiEnv}
                         {\Tm}
                         {\ExnTy}
                         {\Exn}
             }
    \end{gather*}
    \caption{Declarative type system ($\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$)}\label{figure-declarative-type-system}
\end{figure*}

\begin{figure*}[p]
    \begin{gather*}
        \Rule{L-Var}
             {}
             {\JudgeElab{\TyEnv, \TmVar : \ExnTy\ \&\ \Exn}
                        {\KiEnv}
                        {\TmVar}
                        {\TmVar}
                        {\ExnTy}
                        {\Exn}
             }
        \VER
        \Rule{L-Con}
             {}
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmCon{\Ty}}
                        {\TmCon{\Ty}}
                        {\BottomTy{\Ty}}
                        {\ExnEmpty}
             }
        \HOR
        \Rule{L-Crash}
             {}
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmCrash{\Ty}{\ExnLbl}}
                        {\TmCrash{\Ty}{\ExnLbl}}
                        {\BottomTy{\Ty}}
                        {\ExnCon{\ExnLbl}}
             }
        \VER
        \Rule{L-Abs}
             {\begin{gathered}
                \JudgeTyElab{\KiEnv, \overline{\ExnVar_i :: \Kind_i}}
                            {\ExnTy_1}
                            {\Ty_1}
                \quad
                \JudgeKind{\KiEnv, \overline{\ExnVar_i :: \Kind_i}}
                          {\Exn_1}
                          {\KindEXN}
                \\
                \JudgeElab{\TyEnv, x : \ExnTy_1\ \&\ \Exn_1}
                          {\KiEnv, \overline{\ExnVar_i :: \Kind_i}}
                          {\Tm}
                          {\Tm^\prime}
                          {\ExnTy_2}
                          {\Exn_2}
              \end{gathered}
             }
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmAbs{\TmVar}{\Ty_1}{\Tm}}
                        {\TmAnnAbsS{\ExnVar_i}{\Kind_i}{\TmAbsExn{\TmVar}{\ExnTy_1}{\Exn_1}{\Tm^\prime}}}
                        {\ExnTyForallS{\ExnVar_i}{\Kind_i}{\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}}}
                        {\ExnEmpty}
             }
        \VER
        \Rule{L-App}
             {\begin{gathered}
                \JudgeSubTy{\KiEnv}
                           {\ExnTy_2}
                           {\SubstS{\ExnAlt_i}{\ExnVar_i}{\ExnTy_1}}
                \quad
                \JudgeSubExn{\KiEnv}
                            {\Exn_2}
                            {\SubstS{\ExnAlt_i}{\ExnVar_i}{\Exn_1}}
                \quad
                \overline{
                    \JudgeKind{\KiEnv}
                              {\ExnAlt_i}
                              {\Kind_i}
                }
                \\
                \JudgeElab{\TyEnv}
                          {\KiEnv}
                          {\Tm_1}
                          {\Tm_1^\prime}
                          {\ExnTyForallS{\ExnVar_i}{\Kind_i}{\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy}{\Exn}}}
                          {\Exn^\prime}
                \quad
                \JudgeElab{\TyEnv}
                          {\KiEnv}
                          {\Tm_2}
                          {\Tm_2^\prime}
                          {\ExnTy_2}
                          {\Exn_2}
              \end{gathered}
             }
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmApp{\Tm_1}{\Tm_2}}
                        {\TmApp{\TmAnnApp{\Tm_1^\prime}{\overline{\ExnAlt_i}}}{\Tm_2^\prime}}
                        {\SubstS{\ExnAlt_i}{\ExnVar_i}{\ExnTy}}
                        {\ExnUnion{\SubstS{\ExnAlt_i}{\ExnVar_i}{\Exn}}{\Exn^\prime}}
             }
        \VER
        \Rule{L-Fix}
             {\begin{gathered}
                \JudgeTyElab{\KiEnv}
                            {\ExnTy}
                            {\Ty}
                \quad
                \JudgeKind{\KiEnv}
                          {\Exn}
                          {\KindEXN}
                \quad
                \JudgeSubTy{\KiEnv}
                           {\ExnTy^\prime}
                           {\ExnTy}
                \quad
                \JudgeSubExn{\KiEnv}
                            {\Exn^\prime}
                            {\Exn}
                \\
                \JudgeElab{\TyEnv,x:\ExnTy\ \&\ \Exn}
                          {\KiEnv}
                          {\Tm}
                          {\Tm^\prime}
                          {\ExnTy^\prime}
                          {\Exn^\prime}
              \end{gathered}
             }
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmFIX{x}{\Ty}{\Tm}}
                        {\TmFixExn{x}{\ExnTy}{\Exn}{\Tm^\prime}}
                        {\ExnTy}
                        {\Exn}
             }
        \VER
        \Rule{L-Op}
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\Tm_1}
                        {\Tm_1^\prime}
                        {\ExnInt}
                        {\Exn_1}
              \quad
              \JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\Tm_2}
                        {\Tm_2^\prime}
                        {\ExnInt}
                        {\Exn_2}
             }
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmOp{\Tm_1}{\Tm_2}}
                        {\TmOp{\Tm_1^\prime}{\Tm_2^\prime}}
                        {\ExnBool}
                        {\ExnUnion{\Exn_1}{\Exn_2}}
             }
        \VER
        \Rule{L-Seq}
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\Tm_1}
                        {\Tm_1^\prime}
                        {\ExnTy_1}
                        {\Exn_1}
              \quad
              \JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\Tm_2}
                        {\Tm_2^\prime}
                        {\ExnTy_2}
                        {\Exn_2}
             }
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmSeq{\Tm_1}{\Tm_2}}
                        {\TmSeq{\Tm_1^\prime}{\Tm_2^\prime}}
                        {\ExnTy_2}
                        {\ExnUnion{\Exn_1}{\Exn_2}}
             }
        \VER
        \Rule{L-If}
             {\begin{gathered}
                  \JudgeElab{\TyEnv}
                            {\KiEnv}
                            {\Tm_1}
                            {\Tm_1^\prime}
                            {\ExnBool}
                            {\Exn_1}
                  \\
                  \JudgeElab{\TyEnv}
                            {\KiEnv}
                            {\Tm_2}
                            {\Tm_2^\prime}
                            {\ExnTy_2}
                            {\Exn_2}
                  \quad
                  \JudgeElab{\TyEnv}
                            {\KiEnv}
                            {\Tm_3}
                            {\Tm_3^\prime}
                            {\ExnTy_3}
                            {\Exn_3}
             \end{gathered}}
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmIf{\Tm_1}{\Tm_2}{\Tm_3}}
                        {\TmIf{\Tm_1^\prime}{\Tm_2^\prime}{\Tm_3^\prime}}
                        {\TyJoin{\ExnTy_2}{\ExnTy_3}}
                        {\ExnUnion{\Exn_1}{\ExnUnion{\Exn_2}{\Exn_3}}}
             }             
        \VER
        \Rule{L-Nil}
             {}
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmNil{\Ty}}
                        {\TmNil{\Ty}}
                        {\ExnTyList{\BottomTy{\Ty}}{\ExnEmpty}}
                        {\ExnEmpty}
             }
        \VER
        \Rule{L-Cons}
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\Tm_1}
                        {\Tm_1^\prime}
                        {\ExnTy_1}
                        {\Exn_1}
              \quad
              \JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\Tm_2}
                        {\Tm_2^\prime}
                        {\ExnTyList{\ExnTy_1^\prime}{\Exn_1^\prime}}
                        {\Exn_2}
             }
             {\JudgeElab{\TyEnv}
                        {\KiEnv}
                        {\TmCons{\Tm_1}{\Tm_2}}
                        {\TmCons{\Tm_1^\prime}{\Tm_2^\prime}}
                        {\ExnTyList{\TyJoin{\ExnTy_1}{\ExnTy_1^\prime}}{\ExnUnion{\Exn_1}{\Exn_1^\prime}}}
                        {\Exn_2}
             }
        \VER
        \Rule{L-Case}
             {\begin{gathered}
                  \JudgeElab{\TyEnv}
                            {\KiEnv}
                            {\Tm_1}
                            {\Tm_1^\prime}
                            {\ExnTyList{\ExnTy_1}{\Exn_1}}
                            {\Exn_1^\prime}
                  \quad
                  \JudgeElab{\TyEnv}
                            {\KiEnv}
                            {\Tm_2}
                            {\Tm_2^\prime}
                            {\ExnTy_2}
                            {\Exn_2}
                  \\
                  \JudgeElab{\TyEnv, \TmVar_1 : \ExnTy_1\ \&\ \Exn_1, \TmVar_2 : \ExnTyList{\ExnTy_1}{\Exn_1}\ \&\ \Exn_1^\prime}
                            {\KiEnv}
                            {\Tm_3}
                            {\Tm_3^\prime}
                            {\ExnTy_3}
                            {\Exn_3}
              \end{gathered}
             }
             {\begin{multlined}
                  \JudgeElabM{\TyEnv}
                             {\KiEnv}                 
                             {\TmCase{\Tm_1}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3}}
                             {\TmCase{\Tm_1^\prime}{\Tm_2^\prime}{\TmVar_1}{\TmVar_2}{\Tm_3^\prime}}
                             {\TyJoin{\ExnTy_2}{\ExnTy_3}}
                             {\ExnUnion{\Exn_1^\prime}{\ExnUnion{\Exn_2}{\Exn_3}}}
              \end{multlined}
             }
    \end{gather*}
    \caption{Syntax-directed type elaboration system ($\JudgeElab{\TyEnv}{\KiEnv}{\Tm}{\Tm^\prime}{\ExnTy}{\Exn}$)}\label{type-elaboration-system}
\end{figure*}

\subsection{Exception typing and elaboration}

In Figure~\ref{figure-declarative-type-system} we give a declarative system for deriving exception typing judgments $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$.

These judgments work on an explicitly typed language and for this purpose we extend the terms of the source language with two new term-level constructs: effect abstraction and effect application.

%\begin{figure}[h]
\paragraph{Terms}
\begin{alignat}{2}
\!  \Tm \in \ClassExnTm &\ ::=\   &&\ ...                            \notag       \\
                        &\quad|  &&\ \TmAbs{\TmVar}{\ExnTy\ \&\ \Exn}{\Tm} \tag{term abstraction} \\
                        &\quad|  &&\ \TmFixExn{\TmVar}{\ExnTy}{\Exn}{\Tm} \tag{general recursion} \\
                        &\quad|  &&\ ...                            \notag       \\
                        &\quad|  &&\ \TmAnnAbs{\ExnVar}{\Kind}{\Tm} \tag{effect abstraction}  \\
                        &\quad|  &&\ \TmAnnApp{\Tm}{\Exn}           \tag{effect application}
\end{alignat}
%\caption{Source language: extended syntax}\label{figure-source-language-syntax-extended}
%\end{figure}

As the source language is not explicitly typed, we also give a type elaboration system that given an implicitly typed term in the source language produces an explicitly typed term (Figure~\ref{type-elaboration-system}).

The auxiliary judgment $\JudgeTyElab{\KiEnv}{\ExnTy}{\Ty}$ holds for any exception type $\ExnTy$ that erases to the underlying type $\Ty$. The type $\ExnTy_1 \sqcup \ExnTy_2$ is an exception type such that $\JudgeSubTy{\KiEnv}{\ExnTy_1}{\ExnTy_1 \sqcup \ExnTy_2}$ and $\JudgeSubTy{\KiEnv}{\ExnTy_2}{\ExnTy_1 \sqcup \ExnTy_2}$.

\subsection{Presentation of exception types}

For most-general conservative exception types the location of the quantifiers is uniquely determined, we can therefore omit them from the type without introducing ambiguity. For example, the exception type of the \ensuremath{\Varid{map}} function from the introduction may be presented as:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}(\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \Varid{e}_{\mathrm{3}}}\beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{1}} \rangle)\to [\mskip1.5mu \alpha\langle \Varid{e}_{\mathrm{4}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle\to [\mskip1.5mu \beta\langle \Varid{e}_{\mathrm{2}}\;\Varid{e}_{\mathrm{4}}\cup \Varid{e}_{\mathrm{3}} \rangle\mskip1.5mu]\langle \Varid{e}_{\mathrm{5}} \rangle{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\section{Type inference}\label{section-type-inference-algorithm}

%In this section we give an inference algorithm for the exception types presented in the previous section.
A type inference algorithm is given in Figure~\ref{figure-type-inference-algorithm}.

\begin{figure}[p]
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{38}{@{}>{\hspre}c<{\hspost}@{}}%
\column{38E}{@{}l@{}}%
\column{41}{@{}>{\hspre}l<{\hspost}@{}}%
\column{42}{@{}>{\hspre}l<{\hspost}@{}}%
\column{45}{@{}>{\hspre}l<{\hspost}@{}}%
\column{46}{@{}>{\hspre}l<{\hspost}@{}}%
\column{47}{@{}>{\hspre}l<{\hspost}@{}}%
\column{48}{@{}>{\hspre}c<{\hspost}@{}}%
\column{48E}{@{}l@{}}%
\column{51}{@{}>{\hspre}l<{\hspost}@{}}%
\column{58}{@{}>{\hspre}l<{\hspost}@{}}%
\column{64}{@{}>{\hspre}l<{\hspost}@{}}%
\column{73}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[5]{}\mathcal{R}:\ClassTyEnv\times \ClassKiEnv\times \ClassTm\to \ClassExnTm\times \ClassExnTy\times \ClassExn{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmVar){}\<[38]%
\>[38]{}\mathrel{=}{}\<[38E]%
\>[41]{}\Varid{x}:\TyEnv(x){}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; c_\Ty){}\<[38]%
\>[38]{}\mathrel{=}{}\<[38E]%
\>[41]{}c_\Ty : \BottomTy{\Ty}\ \&\ \ExnEmpty{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmCrash{\Ty}{\ExnLbl}){}\<[38]%
\>[38]{}\mathrel{=}{}\<[38E]%
\>[41]{}\TmCrash{\Ty}{\ExnLbl} : \BottomTy{\Ty}\ \&\ \ExnCon{\ExnLbl}{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmAbs{\TmVar}{\Ty}{\Tm}){}\<[38]%
\>[38]{}\mathrel{=}{}\<[38E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{let}\;{}\<[17]%
\>[17]{}\ExnTy_1\ \&\ \Varid{e}\triangleright\overline{\ExnVar_i :: \Kind_i}{}\<[48]%
\>[48]{}\mathrel{=}{}\<[48E]%
\>[51]{}\mathcal{C}(\emptyset; \Ty){}\<[E]%
\\[\blanklineskip]%
\>[17]{}\Tm' : \ExnTy_2\ \&\ \Exn_2{}\<[48]%
\>[48]{}\mathrel{=}{}\<[48E]%
\>[51]{}\mathcal{R}(\TyEnv, \TmVar : \ExnTy_1\ \&\ \Varid{e}; \KiEnv,\overline{\ExnVar_i :: \Kind_i}; \Tm){}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{in}\;{}\<[17]%
\>[17]{}\TmAnnAbsS{\ExnVar_i}{\Kind_i}{\TmAbsExn{\TmVar}{\ExnTy_1}{e}{\Tm^\prime}} : \ExnForall{\overline{\ExnVar_i :: \Kind_i}}{\ExnTyArr{\ExnTy_1}{\Varid{e}}{\ExnTy_2}{\Exn_2}}\ \&\ \ExnEmpty{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmApp{\Tm_1}{\Tm_2}){}\<[38]%
\>[38]{}\mathrel{=}{}\<[38E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{let}\;{}\<[17]%
\>[17]{}\Tm'_{\mathrm{1}} : \ExnTy_1\ \&\ \Exn_1{}\<[73]%
\>[73]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{1}}){}\<[E]%
\\
\>[17]{}\Tm'_{\mathrm{2}} : \ExnTy_2\ \&\ \Exn_2{}\<[73]%
\>[73]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{2}}){}\<[E]%
\\
\>[17]{}\ExnTyArr{\ExnTy_2^\prime}{\ExnVar_2^\prime}{\ExnTy^\prime}{\Exn^\prime}\triangleright \overline{\ExnVar_i :: \Kind_i}{}\<[73]%
\>[73]{}\mathrel{=}\mathcal{I}(\ExnTy_1){}\<[E]%
\\
\>[17]{}\theta{}\<[73]%
\>[73]{}\mathrel{=}\left[\ExnVar_2^\prime\mapsto \Exn_2\right]\hsdot{\circ }{.}\mathcal{M}(\emptyset; \ExnTy_2^\prime; \ExnTy_2){}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{in}\;{}\<[17]%
\>[17]{}\TmApp{\TmAnnApp{\Tm_1^\prime}{\overline{\theta e_i}}}{\Tm_2^\prime} : \llfloor \theta \ExnTy^\prime\rrfloor_{\KiEnv}\ \&\ \llfloor \ExnUnion{\theta \Exn^\prime}{\Exn_1}\rrfloor_{\KiEnv}{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmFIX{\TmVar}{\Ty}{\Tm})\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{do}\;{}\<[17]%
\>[17]{}\Varid{i}; \ExnTy_0\ \&\ \Exn_0{}\<[42]%
\>[42]{}\leftarrow \mathrm{0}; \BottomTy{\Ty}\ \&\ \ExnEmpty{}\<[E]%
\\
\>[17]{}\mathbf{repeat}\;{}\<[25]%
\>[25]{}\Tm^\prime_{i+1} : \ExnTy_{i+1}\ \&\ \Exn_{i+1}{}\<[58]%
\>[58]{}\leftarrow \mathcal{R}(\TyEnv,\Varid{x}:\ExnTy_i\mathbin{\&}\Exn_i; \KiEnv; \Tm){}\<[E]%
\\
\>[25]{}\Varid{i}{}\<[58]%
\>[58]{}\leftarrow \Varid{i}\mathbin{+}\mathrm{1}{}\<[E]%
\\
\>[17]{}\mathbf{until}\;\ExnTy_i\ \&\ \Exn_i\equiv \ExnTy_{i-1}\ \&\ \Exn_{i-1}{}\<[E]%
\\
\>[17]{}\mathbf{return}\;\TmFixExn{\TmVar}{\ExnTy_i}{\Exn_i}{\Tm^\prime_i} : \ExnTy_i\ \&\ \Exn_i{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmSeq{\Tm_1}{\Tm_2})\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{let}\;{}\<[17]%
\>[17]{}\Tm'_{\mathrm{1}} : \ExnTy_1\ \&\ \Exn_1{}\<[45]%
\>[45]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{1}}){}\<[E]%
\\
\>[17]{}\Tm'_{\mathrm{2}} : \ExnTy_2\ \&\ \Exn_2{}\<[45]%
\>[45]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{2}}){}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{in}\;{}\<[17]%
\>[17]{}\TmSeq{\Tm^\prime_1}{\Tm^\prime_2} : \ExnTy_2\ \&\ \llfloor \ExnUnion{\Exn_1}{\Exn_2}\rrfloor_{\KiEnv}{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmOp{\Tm_1}{\Tm_2})\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{let}\;{}\<[17]%
\>[17]{}\Tm'_{\mathrm{1}} : \ExnInt\ \&\ \Exn_1{}\<[46]%
\>[46]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{1}}){}\<[E]%
\\
\>[17]{}\Tm'_{\mathrm{2}} : \ExnInt\ \&\ \Exn_2{}\<[46]%
\>[46]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{2}}){}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{in}\;{}\<[17]%
\>[17]{}\TmOp{\Tm^\prime_1}{\Tm^\prime_2} : \ExnBool\ \&\ \llfloor \ExnUnion{\Exn_1}{\Exn_2}\rrfloor_{\KiEnv}{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmIf{\Tm_1}{\Tm_2}{\Tm_3})\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{let}\;{}\<[17]%
\>[17]{}\Tm'_{\mathrm{1}} : \ExnBool\ \&\ \Exn_1{}\<[47]%
\>[47]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{1}}){}\<[E]%
\\
\>[17]{}\Tm'_{\mathrm{2}} : \ExnTy_2\ \&\ \Exn_2{}\<[47]%
\>[47]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{2}}){}\<[E]%
\\
\>[17]{}\Tm'_{\mathrm{3}} : \ExnTy_3\ \&\ \Exn_3{}\<[47]%
\>[47]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{3}}){}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{in}\;{}\<[17]%
\>[17]{}\TmIf{\Tm^\prime_1}{\Tm^\prime_2}{\Tm^\prime_3} : \llfloor \TyJoin{\ExnTy_2}{\ExnTy_3}\rrfloor_{\KiEnv}\ \&\ \llfloor \ExnUnion{\Exn_1}{\ExnUnion{\Exn_2}{\Exn_3}}\rrfloor_{\KiEnv}{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmNil{\Ty}){}\<[38]%
\>[38]{}\mathrel{=}{}\<[38E]%
\>[41]{}\TmNil{\Ty} : \ExnTyList{\BottomTy{\Ty}}{\ExnEmpty}\ \&\ \ExnEmpty{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmCons{\Tm_1}{\Tm_2})\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{let}\;{}\<[17]%
\>[17]{}\Tm'_{\mathrm{1}} : \ExnTy_1\ \&\ \Exn_1{}\<[64]%
\>[64]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{1}}){}\<[E]%
\\
\>[17]{}\Tm'_{\mathrm{2}} : \ExnTyList{\ExnTy_2}{\Exn_2^\prime}\ \&\ \Exn_2{}\<[64]%
\>[64]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{2}}){}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{in}\;{}\<[17]%
\>[17]{}\Tm'_{\mathrm{1}}::\Tm'_{\mathrm{2}} : \llfloor \ExnTyList{(\TyJoin{\ExnTy_1}{\ExnTy_2})}{\ExnUnion{\Exn_1}{\Exn_2^\prime}}\rrfloor_{\KiEnv}\ \&\ \Exn_2{}\<[E]%
\\[\blanklineskip]%
\>[5]{}\mathcal{R}(\TyEnv; \KiEnv; \TmCase{\Tm_1}{\Tm_2}{\TmVar_1}{\TmVar_2}{\Tm_3})\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{let}\;{}\<[17]%
\>[17]{}\Tm'_{\mathrm{1}} : \ExnTyList{\ExnTy_1}{\Exn_1^\prime}\ \&\ \Exn_1{}\<[64]%
\>[64]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{1}}){}\<[E]%
\\
\>[17]{}\TyEnv^\prime{}\<[64]%
\>[64]{}\mathrel{=}\TyEnv, \TmVar_{\mathrm{1}} : \ExnTy_1\ \&\ \Exn_1^\prime, \TmVar_{\mathrm{2}} : \ExnTyList{\ExnTy_1}{\Exn_1^\prime}\ \&\ \Exn_1{}\<[E]%
\\
\>[17]{}\Tm'_{\mathrm{2}} : \ExnTy_2\ \&\ \Exn_2{}\<[64]%
\>[64]{}\mathrel{=}\mathcal{R}(\TyEnv; \KiEnv; \Tm_{\mathrm{2}}){}\<[E]%
\\
\>[17]{}\Tm'_{\mathrm{3}} : \ExnTy_3\ \&\ \Exn_3{}\<[64]%
\>[64]{}\mathrel{=}\mathcal{R}(\TyEnv^\prime; \KiEnv; \Tm_{\mathrm{3}}){}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\mathbf{in}\;{}\<[17]%
\>[17]{}\TmCase{\Tm^\prime_1}{\Tm^\prime_2}{\TmVar_1}{\TmVar_2}{\Tm^\prime_3} : {}\<[E]%
\\
\>[17]{}\hsindent{28}{}\<[45]%
\>[45]{}\llfloor \TyJoin{\ExnTy_2}{\ExnTy_3}\rrfloor_{\KiEnv}\ \&\ \llfloor \ExnUnion{\Exn_1}{\ExnUnion{\Exn_2}{\Exn_3}}\rrfloor_{\KiEnv}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
\caption{Type inference algorithm ($\mathcal{R}$)}\label{figure-type-inference-algorithm}
\end{figure}

\subsection{Polymorphic abstraction}
The cases for abstraction and application are handled similarly to the corresponding cases in \cite{HoldermansHage}.
%The cases for abstraction and application in Figure~\ref{figure-type-inference-algorithm} are handled similarly to the corresponding cases in \cite{HoldermansHage}.

In the case of abstractions, we first complete the type of the bound variable to a most general exception type using the procedure \ensuremath{\mathcal{C}:\ClassKiEnv\times \ClassTy\to \ClassExnTy\times \ClassExn\times \ClassKiEnv}. This procedure is a functional interpretation of the type completion relation $\JudgeComplete{\KindENV}{\Ty}{\ExnTy}{\Exn}{\KindENV'}$, where the first two arguments $\KindENV$ and $\Ty$ are taken to be the domain and the last three arguments $\ExnTy$, $\Exn$ and $\KindENV'$ are taken to be the range. Next, we infer the exception type of the body of the abstraction under the assumption that the bound variable has the just completed exception type-and-effect $\ExnTy_1\ \&\ \ExnVar_1$. Finally we quantify over all free variables $\overline{\ExnVar_i :: \Kind_i}$ introduced by completion.

In the case of applications, we instantiate ($\mathcal{I}$) all quantified variables of the exception type of $t_1$ with fresh exception variables. Next we use the auxiliary procedure $\mathcal{M}$ to find a matching substitution between the exception types of the formal and the actual parameters.

\begin{figure}[ht]
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{23}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{38}{@{}>{\hspre}l<{\hspost}@{}}%
\column{51}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\mathcal{M}:{}\<[19]%
\>[19]{}\ClassKiEnv\times \ClassExnTy\times \ClassExnTy\to \ClassSubst{}\<[E]%
\\
\>[9]{}\mathcal{M}({}\<[16]%
\>[16]{}\KindENV; {}\<[23]%
\>[23]{}\ExnBool; {}\<[37]%
\>[37]{}\ExnBool){}\<[51]%
\>[51]{}\mathrel{=}\emptyset{}\<[E]%
\\
\>[9]{}\mathcal{M}({}\<[16]%
\>[16]{}\KindENV; {}\<[23]%
\>[23]{}\ExnInt; {}\<[37]%
\>[37]{}\ExnInt){}\<[51]%
\>[51]{}\mathrel{=}\emptyset{}\<[E]%
\\
\>[9]{}\mathcal{M}({}\<[16]%
\>[16]{}\KindENV; {}\<[23]%
\>[23]{}\ExnTyList{\ExnTy^\prime}{\ExnPat}; {}\<[37]%
\>[37]{}\ExnTyList{\ExnTy}{\Exn}){}\<[E]%
\\
\>[23]{}\hsindent{2}{}\<[25]%
\>[25]{}\mathrel{=}\left[\Varid{e'}\mapsto \lambda \overline{e_i :: \KindENV_{e_i}}.\Exn\right]\hsdot{\circ }{.}\mathcal{M}(\KindENV; \ExnTy^\prime; \ExnTy){}\<[E]%
\\
\>[9]{}\mathcal{M}({}\<[16]%
\>[16]{}\KindENV; {}\<[23]%
\>[23]{}\ExnTyArr{\ExnTy_1}{\ExnVar}{\ExnTy_2^\prime}{\ExnPat}; {}\<[38]%
\>[38]{}\ExnTyArr{\ExnTy_1}{\ExnVar}{\ExnTy_2}{\Exn}){}\<[E]%
\\
\>[23]{}\hsindent{2}{}\<[25]%
\>[25]{}\mathrel{=}\left[\Varid{e'}\mapsto \lambda \overline{e_i :: \KindENV_{e_i}}.\Exn\right]\hsdot{\circ }{.}\mathcal{M}(\KindENV; \ExnTy_2^\prime; \ExnTy_2){}\<[E]%
\\
\>[9]{}\mathcal{M}({}\<[16]%
\>[16]{}\KindENV; {}\<[23]%
\>[23]{}\ExnTyForall{\ExnVar}{\Kind}{\ExnTy^\prime}; {}\<[37]%
\>[37]{}\ExnTyForall{\ExnVar}{\Kind}{\ExnTy}){}\<[51]%
\>[51]{}\mathrel{=}\mathcal{M}(\KindENV,\Varid{e}::\Kind; \ExnTy^\prime; \ExnTy){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    \caption{Exception type matching ($\mathcal{M}$)}\label{figure-matching}
\end{figure}

The interesting cases of exception type matching are the cases for list and function types, where we perform pattern unification on the exception annotations. The produced substitution $\theta$ covers all variables $\overline{\ExnVar_i :: \Kind_i}$ freshly introduced by the instantiation procedure~$\mathcal{I}$. Finally, we apply the substitution $\theta$ to the exception type~$\ExnTy^\prime$ and effect~$\Exn^\prime$ of the result of $t_1$.

\subsection{Polymorphic recursion}

The $\mathbf{fix}$-construct abstracts over a variable that is of an exception polymorphic type. The algorithm handles this case with a Kleene--Mycroft iteration---which we conjecture to always converge.\footnote{\cite{HoldermansHage} note that $\lambda$-bound polymorpism gives us $\mathbf{fix}$-bound polymorphism ``for free.'' We believe this statement to be overly optimistic. While the highly polymorphic nature of these types do effectively force us to also handle polymorphic recursion, the inference step is arguably as complicated as the case for polymorphic abstraction.}

\begin{example}[Dussart--Henglein--Mossin]
    Consider the term
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}c<{\hspost}@{}}%
\column{12E}{@{}l@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\Varid{f}{}\<[12]%
\>[12]{}:{}\<[12E]%
\>[16]{}\TyBool\to \TyBool\to \TyBool{}\<[E]%
\\
\>[9]{}\Varid{f}{}\<[12]%
\>[12]{}\mathrel{=}{}\<[12E]%
\>[16]{}\mathbf{fix}\ \Varid{f'} : \TyBool\to \TyBool\to \TyBool. {}\<[E]%
\\
\>[16]{}\hsindent{5}{}\<[21]%
\>[21]{}\lambda\Varid{x}:\TyBool.\lambda\Varid{y}:\TyBool.\;\mathbf{if}\;\Varid{x}\;\mathbf{then}\;\mathbf{true}\;\mathbf{else}\;\Varid{f'}\;\Varid{y}\;\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    Algorithm $\mathcal{R}$ infers the exception type and elaborated term
%    \begin{code}
%        dhm  ::  forall e_1:EXN. ExnTyArrS ExnTyBool e_1 (forall e_2:EXN. ExnTyArrT ExnTyBool e_2 ExnTyBool (e_1 # e_2) ExnEmpty) ExnEmpty
%        dhm  =   FIX (f) (forall e_1:EXN. ExnTyArrS ExnTyBool e_1 (forall e_2:EXN. ExnTyArrT ExnTyBool e_2 ExnTyBool (e_1 # e_2) ExnEmpty) ExnEmpty)
%                    (LLAMBDA (e_1 : EXN) (LAMBDA (x :: ExnTyBool :&: e_1) (LLAMBDA (e_2 : EXN) (PLAMBDA (y :: ExnTyBool :&: e_2))))
%                        if x then True else f (Angled e_2) y (Angled e_1) x
%    \end{code}    
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{12}{@{}>{\hspre}c<{\hspost}@{}}%
\column{12E}{@{}l@{}}%
\column{16}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\Varid{f}{}\<[12]%
\>[12]{}:{}\<[12E]%
\>[16]{}\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\ExnTyArrS{\ExnBool}{\Varid{e}_{\mathrm{1}}}{\forall \Varid{e}_{\mathrm{2}}\hsforall \hsdot{\circ }{.}\ExnTyArrT{\ExnBool}{\Varid{e}_{\mathrm{2}}}{\ExnBool}{\Varid{e}_{\mathrm{1}}\cup \Varid{e}_{\mathrm{2}}}{\ExnEmpty}}{\ExnEmpty}{}\<[E]%
\\
\>[9]{}\Varid{f}{}\<[12]%
\>[12]{}\mathrel{=}{}\<[12E]%
\>[16]{}\mathbf{fix}\ \Varid{f'} : \forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\ExnTyArrS{\ExnBool}{\Varid{e}_{\mathrm{1}}}{\forall \Varid{e}_{\mathrm{2}}\hsforall \hsdot{\circ }{.}\ExnTyArrT{\ExnBool}{\Varid{e}_{\mathrm{2}}}{\ExnBool}{\Varid{e}_{\mathrm{1}}\cup \Varid{e}_{\mathrm{2}}}{\ExnEmpty}}{\ExnEmpty}. {}\<[E]%
\\
\>[16]{}\hsindent{5}{}\<[21]%
\>[21]{}\Lambda\Varid{e}_{\mathrm{1}}::\KindEXN.\lambda\Varid{x}:\ExnBool\ \&\ \Varid{e}_{\mathrm{1}}.\Lambda\Varid{e}_{\mathrm{2}}::\KindEXN.\lambda\Varid{y}:\ExnBool\ \&\ \Varid{e}_{\mathrm{2}}.{}\<[E]%
\\
\>[21]{}\hsindent{4}{}\<[25]%
\>[25]{}\mathbf{if}\;\Varid{x}\;\mathbf{then}\;\mathbf{true}\;\mathbf{else}\;\Varid{f'}\;\langle \Varid{e}_{\mathrm{2}}\rangle\;\Varid{y}\;\langle \Varid{e}_{\mathrm{1}}\rangle\;\Varid{x}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    Let us convince ourselves that the elaborated term is type-correct. 
    \begin{align*}
        x &: \ExnBool\ \&\ e_1 \\
        y &: \ExnBool\ \&\ e_2 \\
        \TmTrue &: \ExnBool\ \&\ \emptyset \\
        f' \langle e_2\rangle\ y\ \langle e_1\rangle\ x &: \ExnBool\ \&\ e_2 \cup e_1
    \end{align*}
   Therefore, \[ \TmIf{x}{\TmTrue}{f' \langle e_2\rangle\ y\ \langle e_1\rangle\ x} : \ExnBool \sqcup \ExnBool\ \&\ e_1 \cup \emptyset \cup e_2 \cup e_1 \]
    By commutativity and idempotence of the union operator and the empty set being the unit, this reduces to
    \[ \TmIf{x}{\TmTrue}{f' \langle e_2\rangle\ y\ \langle e_1\rangle\ x} : \ExnBool\ \&\ e_1 \cup e_2 \]

    Type checking is easier than type inference, however. To infer the type of the recursive definition \ensuremath{\Varid{f}} we have to ``guess'' a type for it. How do we guess this type? We first try the least exception type $\bot_{\TyBool\to\TyBool\to\TyBool}$:
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\ExnTyArrS{\ExnBool}{\Varid{e}_{\mathrm{1}}}{\forall \Varid{e}_{\mathrm{2}}\hsforall \hsdot{\circ }{.}\ExnTyArrT{\ExnBool}{\Varid{e}_{\mathrm{2}}}{\ExnBool}{\ExnEmpty}{\ExnEmpty}}{\ExnEmpty}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    If we continue inferring the type with this guess, then we end up with a larger type than the guess:
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\ExnTyArrS{\ExnBool}{\Varid{e}_{\mathrm{1}}}{\forall \Varid{e}_{\mathrm{2}}\hsforall \hsdot{\circ }{.}\ExnTyArrT{\ExnBool}{\Varid{e}_{\mathrm{2}}}{\ExnBool}{\Varid{e}_{\mathrm{1}}}{\ExnEmpty}}{\ExnEmpty}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    We try inferring the type again, but now start with this type as our guess instead of the least type. We end up with an even larger type:
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\forall \Varid{e}_{\mathrm{1}}\hsforall \hsdot{\circ }{.}\ExnTyArrS{\ExnBool}{\Varid{e}_{\mathrm{1}}}{\forall \Varid{e}_{\mathrm{2}}\hsforall \hsdot{\circ }{.}\ExnTyArrT{\ExnBool}{\Varid{e}_{\mathrm{2}}}{\ExnBool}{\Varid{e}_{\mathrm{1}}\cup \Varid{e}_{\mathrm{2}}}{\ExnEmpty}}{\ExnEmpty}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    Finally, if we take this type as our guess, we obtain the same type and conclude we have reached a fixed point.
\end{example}

\ThesisPaper{TODOX{Example: Glynn, Stuckey, Sulzman}}{}

\subsection{Least upper bounds}

The remaining cases of the algorithm are all relatively straightforward. Several of the cases (\textbf{if-then-else}, \textbf{case-of} and the list-consing constructor) require the least upper bound of two exception types to be computed. The fact that exception types and annotations occurring in argument positions of function types are always simple makes this easy, as they must be equal up to $\alpha$-renaming \cite{HoldermansHage}. This allows us to treat those arguments invariantly instead of contravariantly, obviating the need to also compute greatest lower bounds of exception types and annotations.

\begin{figure}[ht]
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{14}{@{}>{\hspre}c<{\hspost}@{}}%
\column{14E}{@{}l@{}}%
\column{18}{@{}>{\hspre}l<{\hspost}@{}}%
\column{53}{@{}>{\hspre}c<{\hspost}@{}}%
\column{53E}{@{}l@{}}%
\column{61}{@{}>{\hspre}l<{\hspost}@{}}%
\column{82}{@{}>{\hspre}l<{\hspost}@{}}%
\column{105}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\sqcup {}\<[14]%
\>[14]{}:{}\<[14E]%
\>[18]{}\ClassExnTy\times \ClassExnTy\to \ClassExnTy{}\<[E]%
\\
\>[9]{}\ExnBool{}\<[53]%
\>[53]{}\sqcup {}\<[53E]%
\>[61]{}\ExnBool{}\<[105]%
\>[105]{}\mathrel{=}\ExnBool{}\<[E]%
\\
\>[9]{}\ExnInt{}\<[53]%
\>[53]{}\sqcup {}\<[53E]%
\>[61]{}\ExnInt{}\<[105]%
\>[105]{}\mathrel{=}\ExnInt{}\<[E]%
\\
\>[9]{}\ExnTyList{\ExnTy}{\Exn}{}\<[53]%
\>[53]{}\sqcup {}\<[53E]%
\>[61]{}\ExnTyList{\ExnTy^\prime}{\Exn^\prime}{}\<[105]%
\>[105]{}\mathrel{=}\ExnTyList{(\ExnTy\sqcup \ExnTy^\prime)}{\ExnUnion{\Exn}{\Exn^\prime}}{}\<[E]%
\\
\>[9]{}\ExnTyArr{\ExnTy_1}{\Varid{e}}{\ExnTy_2}{\Exn}{}\<[53]%
\>[53]{}\sqcup {}\<[53E]%
\>[61]{}\ExnTyArr{\ExnTy_1}{\Varid{e}}{{}\<[82]%
\>[82]{}\ExnTy_2^\prime}{\Exn^\prime}{}\<[105]%
\>[105]{}\mathrel{=}\ExnTyArr{\ExnTy_1}{\Varid{e}}{(\ExnTy_2\sqcup \ExnTy_2^\prime)}{\Exn\cup \Exn^\prime}{}\<[E]%
\\
\>[9]{}\ExnForall{\Varid{e}::\Kind}{\ExnTy}{}\<[53]%
\>[53]{}\sqcup {}\<[53E]%
\>[61]{}\ExnForall{\Varid{e}::\Kind}{\ExnTy^\prime}{}\<[105]%
\>[105]{}\mathrel{=}\ExnForall{\Varid{e}::\Kind}{(\ExnTy\sqcup \ExnTy^\prime)}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
    \caption{Exception types: least upper bounds ($\sqcup$)}\label{figure-lub}
\end{figure}

\subsection{Complexity}
There are three aspects that affect the run-time complexity of the algorithm: the complexity of the underlying type system, reduction of the effects, and the fixpoint-iteration in the inference step of the $\mathbf{fix}$-construct.
We have a simply typed underlying type system, but if we would extend this to full Hindley--Milner, then it is possible for types to become exponentially larger than terms \cite{Mairson:1989:DMT:96709.96748,Kfoury:1990:MTD:101856.101870}. The effects are $\LambdaUnion$-terms, which contains the simply typed $\lambda$-calculus as a special case. Reduction of terms in the simply typed $\lambda$-calculus is non-elementary recursive \cite{stat79b}. It is also easy to find an artificial family of terms that requires at least a linear number of iterations to converge to a fixpoint. For these reasons we do not believe the algorithm to have an attractive theoretical bound on time-complexity.

Anecdotal evidence suggests that the practical time-complexity is acceptable, however. Hindley--Milner has almost linear complexity in non-pathological cases. Types do not grow larger than the terms. The same seems to hold for the effects. Reduction of effects takes a small number of steps, as does the convergence of the fixpoint-iteration.
In cases where the exception annotation does become too large, a widening rule could be applied.

\section{Related work}\label{section-related-work}

\subsection{Higher-ranked polymorphism in type-and-effect systems}

\paragraph{Effect polymorphism} For plain type systems, Hindley--Milner's \ensuremath{\mathbf{let}}-bound polymorphism generally provides a good compromise between expressiveness of the type system and complexity of the inference algorithm \cite{Hindley1969,Milner78atheory,Damas:1982:PTF:582153.582176}.
Type systems were extended with effects---including \ensuremath{\mathbf{let}}-bound effect-polymorphism---by \cite{Lucassen:1988:PES:73560.73564,Jouvelot:1991:ART:99583.99623}; and \cite{JFP:1457464,Talpin:1994:TED:191349.184660}.
In type-and-effect systems it has long been recognized that $\mathbf{fix}$-bound polymorphism (polymorphic recursion) \emph{in the effects} is often beneficial or even necessary for achieving precise analysis results. For example, in type-and-effect systems for regions \cite{Tofte:1994:ITC:174675.177855}, dimensions \cite{Kennedy:1994:DT:645390.651419,Rittri:1994:STT:194679.194683,Rittri:1995:DIU:224164.224197}, binding times \cite{Dussart:1995:PRS:647163.717680}, and exceptions \cite{Glynn:2002:EAN:581478.581488,Koot:2015:TEA:2678015.2682542}.

Inferring principal types in a type system with polymorphic recursion is equivalent to solving the undecidable semi-unification problem \cite{Mycroft1984,Kfoury:1990:USP:100216.100279,Kfoury:1993:TRP:169701.169687,Henglein:1993:TIP:169701.169692}.
When restricted to polymorphic recursion in the effects, the problem often becomes decidable again. In \cite{Tofte:1994:ITC:174675.177855} this is a conjecture based on empirical observation. \cite{Rittri:1995:DIU:224164.224197} gives a semi-unification procedure based on the general semi-unification semi-algorithm by \cite{Baaz93APTCC} and proves it terminates in the special case of semi-unification in Abelian groups. \cite{Dussart:1995:PRS:647163.717680} use a constraint-based algorithm. They show that all variables that do not occur free in the context or type can be eliminated from the constraint set by a constraint reduction step during each Kleene--Mycroft iteration. As at most $n^2$ subeffecting constraints can be formed over $n$ free variables, the whole procedure must terminate. By not restarting the Kleene--Mycroft iteration from bottom, their algorithm runs in polynomial time---even in the presence of nested fixpoints.

The extension to polymorphic effect-abstraction ($\lambda$-bound, higher-ranked effect polymorphism) remained less well-studied, possibly because it is of limited use without the simultaneous introduction of effect operators---in contrast to the situation of higher-ranked polymorphism in plain type systems.

\paragraph{Effect operators} \cite{KennedyThesis} presents a type system that ensures the dimensional consistency of an ML-like language extended with units of measure ($\mathrm{ML}_\delta$). This language has predicative prenex dimension polymorphism. Kennedy gives an Algorithm~$\mathcal{W}$-like type inference procedure that uses equational unification to deal with the Abelian group (\textsc{ag}) structure of dimension expressions. Also described are two explicitly typed variants of the language: a System F-like language with higher-ranked dimension polymorphism ($\Lambda_\delta$), and a System~$\mathrm{F}_\omega$-like language that extends $\Lambda_\delta$ with dimension operators ($\Lambda_{\delta\omega}$). \cite{KennedyThesis} notes that this language can type strictly more programs than the language without dimension operators:
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}c<{\hspost}@{}}%
\column{21E}{@{}l@{}}%
\column{25}{@{}>{\hspre}l<{\hspost}@{}}%
\column{29}{@{}>{\hspre}l<{\hspost}@{}}%
\column{33}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\Varid{twice}{}\<[21]%
\>[21]{}:{}\<[21E]%
\>[25]{}\forall \Conid{F}\hsforall ::\KindDIM\Rightarrow \KindDIM\hsdot{\circ }{.}{}\<[E]%
\\
\>[25]{}\hsindent{4}{}\<[29]%
\>[29]{}(\forall \Varid{d}\hsforall ::\KindDIM\hsdot{\circ }{.}\TyReal{\Varid{d}}\to \TyReal{\Conid{F}\;\Varid{d}})\to {}\<[E]%
\\
\>[29]{}\hsindent{4}{}\<[33]%
\>[33]{}(\forall \Varid{d}\hsforall ::\KindDIM\hsdot{\circ }{.}\TyReal{\Varid{d}}\to \TyReal{\Conid{F}\;(\Conid{F}\;\Varid{d})}){}\<[E]%
\\
\>[9]{}\Varid{twice}{}\<[21]%
\>[21]{}\mathrel{=}{}\<[21E]%
\>[25]{}\Lambda\Conid{F}::\KindDIM\Rightarrow \KindDIM.{}\<[E]%
\\
\>[25]{}\hsindent{4}{}\<[29]%
\>[29]{}\lambda\Varid{f}:(\forall \Varid{d}\hsforall ::\KindDIM\hsdot{\circ }{.}\TyReal{\Varid{d}}\to \TyReal{\Conid{F}\;\Varid{d}}).{}\<[E]%
\\
\>[29]{}\hsindent{4}{}\<[33]%
\>[33]{}\Lambda\Varid{d}::\KindDIM.\lambda\Varid{x}:\TyReal{\Varid{d}}.\Varid{f}\;\langle \Conid{F}\;\Varid{d}\rangle\;(\Varid{f}\;\langle \Varid{d}\rangle\;\Varid{x}){}\<[E]%
\\[\blanklineskip]%
\>[9]{}\Varid{square}{}\<[21]%
\>[21]{}:{}\<[21E]%
\>[25]{}\forall \Varid{d}\hsforall ::\KindDIM\hsdot{\circ }{.}\TyReal{\Varid{d}}\to \TyReal{\Varid{d}^\mathrm{2}}{}\<[E]%
\\
\>[9]{}\Varid{square}{}\<[21]%
\>[21]{}\mathrel{=}{}\<[21E]%
\>[25]{}\Lambda\Varid{d}::\KindDIM.\lambda\Varid{x}:\TyReal{\Varid{d}}.\Varid{x}^\mathrm{2}{}\<[E]%
\\[\blanklineskip]%
\>[9]{}\Varid{fourth}{}\<[21]%
\>[21]{}:{}\<[21E]%
\>[25]{}\forall \Varid{d}\hsforall ::\KindDIM\hsdot{\circ }{.}\TyReal{\Varid{d}}\to \TyReal{\Varid{d}^\mathrm{4}}{}\<[E]%
\\
\>[9]{}\Varid{fourth}{}\<[21]%
\>[21]{}\mathrel{=}{}\<[21E]%
\>[25]{}\Varid{twice}\;\langle \Lambda\Varid{d}::\KindDIM.\Varid{d}^\mathrm{2}\rangle\;\Varid{square}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Without dimension operators the type of the higher-order function \ensuremath{\Varid{twice}} would not allow the application of the function \ensuremath{\Varid{square}} at the two distinct types \ensuremath{\forall \Varid{d}\hsforall ::\KindDIM\hsdot{\circ }{.}\TyReal{\Varid{d}}\to \TyReal{\Varid{d}^\mathrm{2}}} and \ensuremath{\forall \Varid{d}\hsforall ::\KindDIM\hsdot{\circ }{.}\TyReal{\Varid{d}^\mathrm{2}}\to \TyReal{\Varid{d}^\mathrm{4}}} when invoked from the function \ensuremath{\Varid{fourth}}.

The language $\Lambda_{\delta\omega}$ bears a striking resemblance to the language in Section~\ref{section-exception-types}: the empty and singleton exception sets constants, and the exception set union operator have been replaced with a unit dimension, and dimension product and inverse operators---as dimensions have an \textsc{ag} structure, whereas exception sets have an \ACIone structure; in the dimension type system the annotation is placed only on the real number base type instead of on the compound types, and there is no effect. No type inference algorithm is given for this language, however.

\cite{Faxen:1996:PPF:646912.710963} presents a type system for flow analysis that uses constrained type schemes in the style of \cite{Aiken:1993:TIC:165180.165188}, and has $\lambda$-bound polymorphism (but no type operators) in the style of System~F. To make the inference algorithm terminate for recursive programs the size of the name supply needs to be bounded, leading to imprecision. \cite{SmithWang} present a similar framework, but one that can be instantiated with variants of either $k$-\textsc{cfa} \cite{Shivers:1991:CAH:124950} or \textsc{cpa} \cite{Agesen:1995:CPA:646153.679533} to ensure termination.

\cite{HoldermansHage} design a \SystemFw-like type system for flow analysis for a strict language that has both polymorphic abstraction and effect operators. Our type inference algorithm builds on their techniques. A key difference is that they work with a constraint-based type system and a constraint solver, while we replace these with reduction of terms in an algebraic $\lambda$-calculus. This difference expresses itself particularly in how the case of (polymorphic) recursion is handled. We believe our approach will scale more easily to analyses that are either not conservative extensions of the underlying type system, or require more expressive effects (see Section~\ref{section-future-research}).

\subsection{$\LambdaUnion$-calculus}

\cite{DBLP:conf/lics/Breazu-Tannen88a}, \cite{Okada1989}, and \cite{BREAZUTANNEN19913} prove that if a simply typed $\lambda$-calculus is extended with a many-sorted algebraic rewrite system $R$ (by introducing the symbols of the algebraic theory as higher-order constants in the $\lambda$-calculus), then the combined rewrite system $\beta\eta R$ is confluent and strongly normalizing if $R$ is confluent and strongly normalizing.

\cite{Revesz:1992:LEL:131057.131060} introduced an untyped $\lambda$-calculus with applicative lists. A model is given by \cite{Durfee97amodel}. This calculus satisfies the equations
    \begin{align}
        \langle t_1, ..., t_n \rangle\ t^\prime &= \langle t_1\ t^\prime, ..., t_n\ t^\prime \rangle \tag{$\gamma_1$} \\
        \lambda x. \langle t_1, ... t_n \rangle &= \langle \lambda x. t_1, ..., \lambda x. t_n \rangle \tag{$\gamma_2$}
    \end{align}
similar to our typed $\LambdaUnion$-calculus.

\subsection{Exception analyses}

Several exception analyses have been described in the literature; these primarily target the detection of uncaught exceptions in ML.
The exception analysis by \cite{kyi} is based on abstract interpretation.
\cite{Guzman94anextended} and \cite{Fahndrich:1998:TDE:893957} describe type-based exception analyses. \cite{Leroy:2000:TAU:349214.349230} presents a row-based type system for exception analysis that contains a data-flow analysis component targeted towards tracking value-carrying exceptions.

\cite{Glynn:2002:EAN:581478.581488} developed the first exception analysis for a non-strict language; a type-based analysis using Boolean constraints. \cite{Koot:2015:TEA:2678015.2682542} present a constraint-based type system for exception analysis of a non-strict language, where the exception-flow could depend on the data-flow using conditional constraints. This increases the accuracy in the presence of exceptions raised by pattern-matching failures.
\section{Further research}\label{section-future-research}

%At least two opportunities for further research present itself:

\paragraph{Can we infer types for Kennedy's higher-ranked $\Lambda_{\delta\omega}$?} One problem that immediately presents itself is that this type system is not a conservative extension of the underlying type system: programs can be rejected because they, while being type correct in the underlying type system, may still be dimensionally inconsistent.
Unlike the system in this paper, the annotations on function arguments will no longer be of the simple form (patterns) required for the straightforward matching step in the type inference algorithm. Instead, we suspect we have to solve a higher-order equational (pre)unification problem, which is only semi-decidable.
\cite{Snyder:1990:HOE:648229.752474}, \cite{NipkowQian1991} and \cite{QIAN1996401} do give us semi-algorithms for solving such problems.

\paragraph{Can we further improve the precision of exception types?}
\cite{Koot:2015:TEA:2678015.2682542} argue
%It is argued by \cite{Koot:2015:TEA:2678015.2682542}
that an accurate exception typing system for non-strict languages should also take the data flow of the program into account, as many exceptions that can be raised in non-strict languages are caused by incomplete case-analyses during pattern-matching. The canonical example is the \ensuremath{\Varid{risers}} function---which splits a list into monotonically increasing subsegments; for example, \ensuremath{\Varid{risers}\;[\mskip1.5mu \mathrm{1},\mathrm{3},\mathrm{5},\mathrm{1},\mathrm{2}\mskip1.5mu]} evaluates to \ensuremath{[\mskip1.5mu [\mskip1.5mu \mathrm{1},\mathrm{3},\mathrm{5}\mskip1.5mu],[\mskip1.5mu \mathrm{1},\mathrm{2}\mskip1.5mu]\mskip1.5mu]}---by \cite{Mitchell:2008:PBE:1411286.1411293}:

    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{34}{@{}>{\hspre}c<{\hspost}@{}}%
\column{34E}{@{}l@{}}%
\column{37}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\Varid{risers}{}\<[17]%
\>[17]{}:[\mskip1.5mu \TyInt\mskip1.5mu]\to [\mskip1.5mu [\mskip1.5mu \TyInt\mskip1.5mu]\mskip1.5mu]{}\<[E]%
\\
\>[9]{}\Varid{risers}\;{}\<[17]%
\>[17]{}[\mskip1.5mu \mskip1.5mu]{}\<[34]%
\>[34]{}\mathrel{=}{}\<[34E]%
\>[37]{}[\mskip1.5mu \mskip1.5mu]{}\<[E]%
\\
\>[9]{}\Varid{risers}\;{}\<[17]%
\>[17]{}[\mskip1.5mu \Varid{x}\mskip1.5mu]{}\<[34]%
\>[34]{}\mathrel{=}{}\<[34E]%
\>[37]{}[\mskip1.5mu [\mskip1.5mu \Varid{x}\mskip1.5mu]\mskip1.5mu]{}\<[E]%
\\
\>[9]{}\Varid{risers}\;(\Varid{x}_{\mathrm{1}}::\Varid{x}_{\mathrm{2}}::\Varid{xs}){}\<[34]%
\>[34]{}\mathrel{=}{}\<[34E]%
\\
\>[9]{}\hsindent{4}{}\<[13]%
\>[13]{}\mathbf{if}\;\Varid{x}_{\mathrm{1}}\leq \Varid{x}_{\mathrm{2}}\;\mathbf{then}\;(\Varid{x}_{\mathrm{1}}::\Varid{y})::\Varid{ys}\;\mathbf{else}\;[\mskip1.5mu \Varid{x}_{\mathrm{1}}\mskip1.5mu]::(\Varid{y}::\Varid{ys}){}\<[E]%
\\
\>[13]{}\hsindent{4}{}\<[17]%
\>[17]{}\mathbf{where}\;(\Varid{y}::\Varid{ys})\mathrel{=}\Varid{risers}\;(\Varid{x}_{\mathrm{2}}::\Varid{xs}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The inference algorithm in Figure~\ref{figure-type-inference-algorithm} assigns \ensuremath{\Varid{risers}} the type
    % (∀e1::E.(∀e2::E.([int⟨e2⟩]⟨e1⟩→[[int⟨e2⟩]⟨∅⟩]⟨({risers}∪(e1∪e2))⟩))) & ∅    
    \begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[9]{}\forall \Varid{e}_{\mathrm{1}}\hsforall ::\KindEXN\hsdot{\circ }{.}\forall \Varid{e}_{\mathrm{2}}\hsforall ::\KindEXN\hsdot{\circ }{.}{}\<[E]%
\\
\>[9]{}\hsindent{4}{}\<[13]%
\>[13]{}\ExnTyArr{\ExnTyList{\ExnInt}{\Varid{e}_{\mathrm{2}}}}{\Varid{e}_{\mathrm{1}}}{\ExnTyList{\ExnTyList{\ExnInt}{\Varid{e}_{\mathrm{2}}}}{\ExnEmpty}}{\Varid{e}_{\mathrm{1}}\cup \Varid{e}_{\mathrm{2}}\cup \{\mathbf{E}\}}\ \&\ \emptyset{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
where \ensuremath{\mathbf{E}} is the label of the exception raised when the pattern-match in the \ensuremath{\mathbf{where}}-clause fails.\footnote{This exception is left implicit in the above program, but becomes explicit when the code is desugared into our core language.} However, the pattern-match happens on the result of the recursive call \ensuremath{\Varid{risers}\;(\Varid{x}_{\mathrm{2}}::\Varid{xs})}. When \ensuremath{\Varid{risers}} is given a non-empty list (such as \ensuremath{\Varid{x}_{\mathrm{2}}::\Varid{xs}}) as an argument, it always returns a non-empty list as its result. The pattern-match can thus never fail, and the exception labelled \ensuremath{\mathbf{E}} can thus never be raised.

%In our previous work we demonstrated
\cite{Koot:2015:TEA:2678015.2682542} demonstrate
how this exception can be elided by having the exception flow depend on the data flow. The $\LambdaUnion$-calculus terms that form the effect annotations cannot express this dependence, however.
%In our earlier work we used
\cite{Koot:2015:TEA:2678015.2682542} use
a slightly ad hoc form of conditional constraints to model this dependence. We believe that extending a $\lambda$-calculus with an equational theory of Boolean rings may form the basis of a more principled approach. Boolean rings have already been successfully used to design type systems for strictness analysis \cite{Wright}, records \cite{LIX-RR-96-09} and exception tracking \cite{Benton:2007:SEA:1190315.1190320}.

\paragraph{Metatheory}
We have not yet worked out the metatheory of the type system presented in this paper. Of particular interest are the (syntactic) soundness, completeness and totality of the inference step for recursive definitions. We expect that soundness and completeness can be shown by a similar argument as in \cite{Mycroft1984} and \cite{Dussart:1995:PRS:647163.717680}.

We conjectured the totality of our inference algorithm. We have a good reason to do so: we only expect the fixpoint iteration to diverge if no fixpoint exists---that is to say, the program is type incorrect. Assuming the program is well-typed in the underlying type system, there are no type incorrect programs in our exception typing system, however.

To show the fixpoint iteration is guaranteed to terminate in their binding-time analysis, \cite{Dussart:1995:PRS:647163.717680} note that only a finite number of type constraints and therefore constrained type schemes can be formed over a finite number of variables (after constraints have been simplified). As it is still possible to form an infinite number of $\LambdaUnion$-normal forms over a finite number of variables, such an argument is not going to work directly.
\section{Conclusion}

We show that it is feasible to extend non-strict higher-order languages with exception-annotated types, as is already done in some strict first-order languages. We argue that higher-ranked exception polymorphic types with exception set operators \emph{\`a la} System~$F_\omega$ are not only more accurate, but are also more readable when presented to the programmer \emph{vis-\`a-vis} constrained type schemes: the exception terms in the annotations more closely mirror what is happening at the term level than constraint sets do.

\section*{Acknowledgements}
We would like to thank Jurriaan Hage and the members of the Software Technology reading club at Utrecht University for their comments on earlier drafts of this paper. Vincent van Oostrom and Femke van Raamsdonk provided some helpful pointers to related literature.

%\clearpage

%\bibliographystyle{plainnat}
\bibliographystyle{splncs03}
\renewcommand{\bibname}{}   %% LNCS HACK
\small                      %% LNCS HACK
\bibliography{higher-ranked-exception-types-neutered}

%BEGIN sigplanconf-template.tex

% We recommend abbrvnat bibliography style.
%\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

%\begin{thebibliography}{}
%\softraggedright
%
%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...
%
%\end{thebibliography}

%END sigplanconf-template.tex

\clearpage

\appendix

\section{Prototype}

A prototype implementation of the inference algorithm is available from \url{https://www.staff.science.uu.nl/~0422819/hret/supplementary-material.tar.gz}.

\end{document}

\section{Metatheory}

\subsection{$\LambdaUnion$-calculus}

\begin{lemma}
    The terms $\LuApp{(\LuUnion{\LuTm_1}{\LuTm_2})}{\LuTm}$ and $\LuUnion{\LuApp{\LuTm_1}{\LuTm}}{\LuApp{\LuTm_2}{\LuTm}}$ are equivalent.
\end{lemma}
\begin{proof}
    \begin{align*}
        &\LuDen{\LuApp{(\LuUnion{\LuTm_1}{\LuTm_2})}{\LuTm}} \\
        &= \bigcup\left\{\varphi(\LuDen{\LuTm})\mid\varphi\in\LuDen{\LuUnion{\LuTm_1}{\LuTm_2}}\right\} \\
        &= \bigcup\left\{\varphi(\LuDen{\LuTm})\mid\varphi\in\LuDen{\LuTm_1}\cup\LuDen{\LuTm_2}\right\} \\
        &= \bigcup\left\{\varphi(\LuDen{\LuTm})\mid\varphi\in\LuDen{\LuTm_1}\right\} \cup \bigcup\left\{\varphi(\LuDen{\LuTm})\mid\varphi\in\LuDen{\LuTm_2}\right\} \\
        &= \LuDen{\LuApp{\LuTm_1}{\LuTm}} \cup \LuDen{\LuApp{\LuTm_2}{\LuTm}} \\
        &= \LuDen{\LuUnion{(\LuApp{\LuTm_1}{\LuTm})}{(\LuApp{\LuTm_2}{\LuTm})}} \qedhere
    \end{align*}
\end{proof}

\begin{lemma}
    The terms $\LuUnion{\left(\LuAbs{\LuVar}{\LuTy}{\LuTm_1}\right)}{\left(\LuAbs{\LuVar}{\LuTy}{\LuTm_2}\right)}$ and $\LuAbs{\LuVar}{\LuTy}{\LuUnion{\LuTm_1}{\LuTm_2}}$ are extensionally equivalent.
\end{lemma}
\begin{proof}
    We show that \[\llbracket \LuApp{\left(\LuUnion{\left(\LuAbs{\LuVar}{\LuTy}{\LuTm_1}\right)}{\left(\LuAbs{\LuVar}{\LuTy}{\LuTm_2}\right)}\right)}{\LuTm_3} \rrbracket_\rho = \llbracket \LuApp{\left(\LuAbs{\LuVar}{\LuTy}{\LuUnion{\LuTm_1}{\LuTm_2}}\right)}{\LuTm_3} \rrbracket_\rho \] for all suitable $\rho$ and $\LuTm_3$.
    \begin{align*}
        & \llbracket \LuApp{\left(\LuUnion{\left(\LuAbs{\LuVar}{\LuTy}{\LuTm_1}\right)}{\left(\LuAbs{\LuVar}{\LuTy}{\LuTm_2}\right)}\right)}{\LuTm_3} \rrbracket_\rho \\
        &= \bigcup \left\{ \varphi(\llbracket \LuTm_3 \rrbracket_\rho) \mid \varphi \in \llbracket \LuUnion{\left(\LuAbs{\LuVar}{\LuTy}{\LuTm_1}\right)}{\left(\LuAbs{\LuVar}{\LuTy}{\LuTm_2}\right)} \rrbracket_\rho \right\} \\
        &= \bigcup \left\{ \varphi(\llbracket \LuTm_3 \rrbracket_\rho) \mid \varphi \in \llbracket \LuAbs{\LuVar}{\LuTy}{\LuTm_1} \rrbracket_\rho \cup \llbracket \LuAbs{\LuVar}{\LuTy}{\LuTm_2} \rrbracket_\rho \right\} \\
        %&= \bigcup \left\{ \varphi(\llbracket \LuTm_3 \rrbracket_\rho) \mid \varphi \in \left\{ \lambda v \in V_\tau. \llbracket \LuTm_1 \rrbracket_{\rho\left[\LuVar \mapsto v\right]}, \lambda v \in V_\tau. \llbracket \LuTm_2 \rrbracket_{\rho\left[\LuVar \mapsto v\right]}\right\} \right\} \\
        &= \bigcup \left\{ \varphi(\llbracket \LuTm_3 \rrbracket_\rho) \mid \varphi \in \left\{ \lambda v \in V_\tau. \llbracket \LuTm_i \rrbracket_{\rho\left[\LuVar \mapsto v\right]} \mid i \in \left\{1,2\right\} \right\} \right\} \\
        &= \bigcup \left\{ \llbracket \LuTm_1 \rrbracket_{\rho\left[\LuVar \mapsto \llbracket \LuTm_3 \rrbracket_\rho\right]}, \llbracket \LuTm_2 \rrbracket_{\rho\left[\LuVar \mapsto \llbracket \LuTm_3 \rrbracket_\rho\right]} \right\} \\
        &= \llbracket \LuTm_1 \rrbracket_{\rho\left[\LuVar \mapsto \llbracket \LuTm_3 \rrbracket_\rho\right]} \cup \llbracket \LuTm_2 \rrbracket_{\rho\left[\LuVar \mapsto \llbracket \LuTm_3 \rrbracket_\rho\right]} \\
        &= \bigcup \left\{ \llbracket \LuTm_1 \rrbracket_{\rho\left[\LuVar \mapsto \llbracket \LuTm_3 \rrbracket_\rho\right]} \cup \llbracket \LuTm_2 \rrbracket_{\rho\left[\LuVar \mapsto \llbracket \LuTm_3 \rrbracket_\rho\right]} \right\} \\
        &= \bigcup \left\{ \llbracket \LuUnion{\LuTm_1}{\LuTm_2} \rrbracket_{\rho\left[\LuVar \mapsto \llbracket \LuTm_3 \rrbracket_\rho\right]} \right\} \\
        &= \bigcup \left\{ \varphi(\llbracket \LuTm_3 \rrbracket_\rho) \mid \varphi \in \left\{ \lambda v \in V_\tau. \llbracket \LuUnion{\LuTm_1}{\LuTm_2} \rrbracket_{\rho\left[\LuVar \mapsto v\right]} \right\} \right\} \\
        &= \bigcup \left\{ \varphi(\llbracket \LuTm_3 \rrbracket_\rho) \mid \varphi \in \llbracket \LuAbs{\LuVar}{\LuTy}{\LuUnion{\LuTm_1}{\LuTm_2}} \rrbracket_\rho \right\} \\
        &= \llbracket \LuApp{\left(\LuAbs{\LuVar}{\LuTy}{\LuUnion{\LuTm_1}{\LuTm_2}}\right)}{\LuTm_3} \rrbracket_\rho \qedhere
    \end{align*}
\end{proof}

\subsection{Declarative type system}

\begin{lemma}[Canonical forms]\label{thm-canonical-forms}\ 
    \begin{enumerate}
        \item If $\ExnVal$ is a possibly exceptional value of type $\ExnBool$, then $\ExnVal$ is either $\TmTrue$, $\TmFalse$, or~$\TmCrash{}{\ExnLbl}$.
        \item If $\ExnVal$ is a possibly exceptional value of type $\ExnInt$, then $\ExnVal$ is either some integer $n$, or an exceptional value $\TmCrash{}{\ExnLbl}$.
        \item If $\ExnVal$ is a possibly exceptional value of type $\ExnTyList{\ExnTy}{\Exn}$, then $\ExnVal$ is either $\TmNil{}$, $\TmCons{\Tm}{\Tm^\prime}$, or~$\TmCrash{}{\ExnLbl}$.
        \item If $\ExnVal$ is a possibly exceptional value of type $\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}$, then $\ExnVal$ is either $\TmAbsExn{x}{\ExnTy_1}{\Exn_1}{\Tm^\prime}$ or $\TmCrash{}{\ExnLbl}$.
        \item If $\ExnVal$ is a possibly exceptional value of type $\ExnTyForall{\ExnVar}{\Kind}{\ExnTy}$, then $\ExnVal$ is $\TmAnnAbs{\ExnVar}{\Kind}{\Tm}$
    \end{enumerate}
\end{lemma}
\begin{proof}
    For each part, inspect all forms of $\ExnVal$ and discard the unwanted cases by inversion of the typing relation. Note that $\BottomTy{\Ty}$ cannot give us a type of the form $\ExnTyForall{\ExnVar}{\Kind}{\ExnTy}$.
\end{proof}
\TODO: Say something about \CiteRule{T-Sub}?

\begin{theorem}[Progress]\label{thm-progress}
    If $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$ with $\Tm$ a closed term, then $\Tm$ is either a possibly exceptional value $\ExnVal$ or there is a closed term $\Tm^\prime$ such that $\JudgeEval{\Tm}{\Tm^\prime}$.
\end{theorem}
\begin{proof}
By induction on the typing derivation $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$.

    The case \CiteRule{T-Var} can be discarded, as a variable is not a closed term. The cases \CiteRule{T-Con}, \CiteRule{T-Crash}, \CiteRule{T-Abs}, \CiteRule{T-AnnAbs}, \CiteRule{T-Nil} and \CiteRule{T-Cons} are immediate as they are values.

    \CaseRule{T-App} We can immediately apply the induction hypothesis to $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm_1}{\ExnTyArr{\ExnTy_2}{\Exn_2}{\ExnTy}{\Exn}}{\Exn}$, giving us either a $\Tm_1^\prime$ such that $\JudgeEval{\Tm_1}{\Tm_1^\prime}$ or that $\Tm_1 = \ExnVal$. In the former case we can make progress using \CiteRule{E-App}. In the latter case the canonical forms lemma tells us that either $\Tm_1 = \TmAbsExn{x}{\ExnTy_2}{\Exn_2}{\Tm_1^\prime}$ or $\Tm_1 = \TmCrash{}{\ExnLbl}$, in which case we can make progress using \CiteRule{E-AppAbs} or \CiteRule{E-AppExn}, respectively.

    The remaining cases follow by analogous reasoning.
\end{proof}

\begin{figure}[ht]
    \begin{align*}
        \Subst{\Exn}{\ExnVar}{\ExnVar} &\equiv \Exn \\
        \Subst{\Exn}{\ExnVar}{\ExnVar^\prime} &\equiv \ExnVar^\prime &&\text{if $\ExnVar \neq \ExnVar^\prime$}\\
        \Subst{\Exn}{\ExnVar}{\ExnCon{\ExnLbl}} &\equiv \ExnCon{\ExnLbl} \\
        \Subst{\Exn}{\ExnVar}{\ExnEmpty} &\equiv \ExnEmpty \\
        \Subst{\Exn}{\ExnVar}{\left(\ExnAbs{\ExnVar^\prime}{\Kind}{\Exn^\prime}\right)} &\equiv \ExnAbs{\ExnVar^\prime}{\Kind}{\Subst{\Exn}{\ExnVar}{\Exn^\prime}} &&\text{if $\ExnVar \neq \ExnVar^\prime$ and $\ExnVar^\prime \notin \fv{\Exn}$}\\
        \Subst{\Exn}{\ExnVar}{\left(\ExnApp{\ExnVar_1}{\ExnVar_2}\right)} &\equiv \ExnApp{\left(\Subst{\Exn}{\ExnVar}{\ExnVar_1}\right)}{\left(\Subst{\Exn}{\ExnVar}{\ExnVar_2}\right)} \\
        \Subst{\Exn}{\ExnVar}{\left(\ExnUnion{\ExnVar_1}{\ExnVar_2}\right)} &\equiv \ExnUnion{\Subst{\Exn}{\ExnVar}{\ExnVar_1}}{\Subst{\Exn}{\ExnVar}{\ExnVar_2}}
    \end{align*}
    \caption{Annotation substitution}
\end{figure}

\begin{figure}[ht]
    \begin{align*}
        \Subst{\Tm}{x}{x} &\equiv \Tm \\
        \Subst{\Tm}{x}{x^\prime} &\equiv x^\prime &&\text{if $x \neq x^\prime$} \\
        \Subst{\Tm}{x}{\TmCon{\Ty}} &\equiv \TmCon{\Ty} \\
        \Subst{\Tm}{x}{\left(\TmAbs{x^\prime}{\ExnTy}{\Tm^\prime}\right)} &\equiv \TmAbs{x^\prime}{\ExnTy}{\Subst{\Tm}{x}{\Tm^\prime}} &&\text{if $x \neq x^\prime$ and $x^\prime \notin \fv{\Tm}$}\\
        & \dots
    \end{align*}
    \caption{Term substitution}
\end{figure}

\begin{lemma}[Annotation substitution]\label{thm-annotation-substitution}\ 
    \begin{enumerate}
        \item If $\JudgeExn{\KiEnv,\BindKi{\ExnVar}{\Kind^\prime}}{\Exn}{\Kind}$ and $\JudgeExn{\KiEnv}{\Exn^\prime}{\Kind^\prime}$ then $\JudgeExn{\KiEnv}{\Subst{\Exn^\prime}{\ExnVar}{\Exn}}{\Kind}$.
        \item If $\JudgeSubExn{\KiEnv,\BindKi{\ExnVar}{\Kind^\prime}}{\Exn_1}{\Exn_2}$ and $\JudgeKind{\KiEnv}{\Exn^\prime}{\Kind^\prime}$ then $\JudgeSubExn{\KiEnv}{\Subst{\Exn^\prime}{\ExnVar}{\Exn_1}}{\Subst{\Exn^\prime}{\ExnVar}{\Exn_2}}$.
        \item If $\JudgeSubTy{\KiEnv,\BindKi{\ExnVar}{\Kind^\prime}}{\ExnTy_1}{\ExnTy_2}$ and $\JudgeKind{\KiEnv}{\Exn^\prime}{\Kind^\prime}$ then $\JudgeSubExn{\KiEnv}{\Subst{\Exn^\prime}{\ExnVar}{\ExnTy_1}}{\Subst{\Exn^\prime}{\ExnVar}{\ExnTy_2}}$.
        \item If $\JudgeExnTy{\TyEnv}{\KiEnv,\BindKi{\ExnVar}{\Kind^\prime}}{\Tm}{\ExnTy}{\Exn}$ and $\JudgeExn{\KiEnv}{\Exn^\prime}{\Kind^\prime}$ then $\JudgeExnTy{\TyEnv}{\KiEnv}{\Subst{\Exn^\prime}{\ExnVar}{\Tm}}{\Subst{\Exn^\prime}{\ExnVar}{\ExnTy}}{\Exn}$.
    \end{enumerate}
\end{lemma}

\TODO: In part 4, either we need the assumption $e \notin \fv{\Exn}$ (which seems to be satisfied everywhere we want to apply this lemma), or we also need to apply the substitution to $\Exn$ (is this expected or not in a type-and-effect system)? \CiteRule{T-Fix} seems to be to only rule where an exception variable can flow from $\ExnTy$ to $\Exn$.... 
\begin{proof}
1. By induction on the derivation of $\JudgeExn{\KiEnv,\BindKi{\ExnVar}{\Kind^\prime}}{\Exn}{\Kind}$.
The cases \CiteRule{A-Var}, \CiteRule{A-Abs} and \CiteRule{A-App} are analogous to the respective cases in the proof of term substitution below.
In the case \CiteRule{A-Con} one can strengthen the assumption $\JudgeExn{\KiEnv,\BindKi{\ExnVar}{\Kind^\prime}}{\ExnCon{\ExnLbl}}{\KindEXN}$ to $\JudgeExn{\KiEnv}{\ExnCon{\ExnLbl}}{\KindEXN}$ as $\ExnVar \notin \fv{\ExnCon{\ExnLbl}}$, the result is then immediate; similarly for \CiteRule{A-Empty}. The case \CiteRule{A-Union} goes analogous to \CiteRule{A-App}.

2. \TODO

3. \TODO

4. By induction on the derivation of $\JudgeExnTy{\TyEnv}{\KiEnv,\BindKi{\ExnVar}{\Kind^\prime}}{\Tm}{\ExnTy}{\Exn}$. Most cases can be discarded by a straightforward application of the induction hypothesis; we show only the interesting case.

    \def\currentprefix{annotation-substitution:T-AnnApp}
    \CaseRule{T-AnnApp} \TODO

\TODO

\end{proof}

\begin{lemma}[Term substitution]\label{thm-term-substitution}
If $\JudgeExnTy{\TyEnv,\BindTy{\TmVar}{\ExnTy^\prime}{\Exn^\prime}}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$ and $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm^\prime}{\ExnTy^\prime}{\Exn^\prime}$ then $\JudgeExnTy{\TyEnv}{\KiEnv}{\Subst{\Tm^\prime}{\TmVar}{\Tm}}{\ExnTy}{\Exn}$.
\end{lemma}
\begin{proof}
By induction on the derivation of $\JudgeExnTy{\TyEnv,\BindTy{\TmVar}{\ExnTy^\prime}{\Exn}}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$.

    \def\currentprefix{term-substitution:T-App}
    \CaseRule{T-Var} We either have $\Tm = \TmVar$ or $\Tm = \TmVar^\prime$ with $\TmVar \neq \TmVar^\prime$.
    In the first case we need to show that $\JudgeExnTy{\TyEnv}{\KiEnv}{\Subst{\Tm^\prime}{\TmVar}{\TmVar}}{\ExnTy}{\Exn}$, which by definition of substitution is equal to $\JudgeExnTy{\TyEnv}{\KiEnv}{\TmVar}{\ExnTy}{\Exn}$, but this is one of our assumptions.
    In the second case we need to show that $\JudgeExnTy{\TyEnv,\BindTy{\TmVar^\prime}{\ExnTy}{\Exn}}{\KiEnv}{\Subst{\Tm}{\TmVar}{\TmVar^\prime}}{\ExnTy}{\Exn}$, which by definition of substitution is equal to $\JudgeExnTy{\TyEnv,\BindTy{\TmVar^\prime}{\ExnTy}{\Exn}}{\KiEnv}{\TmVar^\prime}{\ExnTy}{\Exn}$. This follows immediately from \CiteRule{T-Var}.

    \def\currentprefix{term-substitution:T-Abs}
    \CaseRule{T-Abs} Our assumptions are
        \begin{align}
            & \JudgeExnTy{\TyEnv,\BindTy{x}{\ExnTy^\prime}{\Exn^\prime},\BindTy{y}{\ExnTy_1}{\Exn_1}}{\KiEnv}{\Tm}{\ExnTy_2}{\Exn_2} \labell{ass1} \\
            & \JudgeExnTy{\TyEnv}{\KiEnv}{\Tm^\prime}{\ExnTy^\prime}{\Exn^\prime}. \labell{ass2}
        \end{align}
    By the Barendregt convention we may assume that $y \neq x$ and $y \notin \fv{\Tm^\prime}$. We need to show that $\JudgeExnTy{\TyEnv}{\KiEnv}{\Subst{\Tm^\prime}{x}{(\TmAbsExn{y}{\ExnTy_1}{\Exn_1}{\Tm})}}{\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}}{\ExnEmpty}$, which by definition of substitution is equal to
        \begin{align}
            & \JudgeExnTy{\TyEnv}{\KiEnv}{\TmAbsExn{y}{\ExnTy_1}{\Exn_1}{\Subst{\Tm^\prime}{x}{\Tm}}}{\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}}{\ExnEmpty}. \labell{ts}
        \end{align}
    We weaken \eqrefl{ass2} to $\JudgeExnTy{\TyEnv,\BindTy{y}{\ExnTy_1}{\Exn_1}}{\KiEnv}{\Tm^\prime}{\ExnTy^\prime}{\Exn^\prime}$ and apply the induction hypothesis on this and \eqrefl{ass1} to obtain
        \begin{align}
            & \JudgeExnTy{\TyEnv,\BindTy{y}{\ExnTy_1}{\Exn_1}}{\KiEnv}{\Subst{\Tm^\prime}{x}{\Tm}}{\ExnTy_2}{\Exn_2}. \labell{ih}
        \end{align}
    The desired result \eqrefl{ts} can be constructed from \eqrefl{ih} using \CiteRule{T-Abs}.

    \def\currentprefix{term-substitution:T-AnnAbs}
    \CaseRule{T-AnnAbs} Our assumptions are $\JudgeExnTy{\TyEnv,\BindTy{x}{\ExnTy^\prime}{\Exn^\prime}}{\KiEnv,\BindKi{\ExnVar}{\Kind}}{\Tm}{\ExnTy}{\Exn}$ and $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm^\prime}{\ExnTy^\prime}{\Exn^\prime}$.
    By the Barendregt convention we may assume that $\ExnVar \notin \fv{\Tm^\prime}$.
    We need to show that $\JudgeExnTy{\TyEnv}{\KiEnv}{\Subst{\Tm^\prime}{x}{\left(\TmAnnAbs{\ExnVar}{\Kind}{\Tm}\right)}}{\ExnTy}{\Exn}$, which is equal to $\JudgeExnTy{\TyEnv}{\KiEnv}{\TmAnnAbs{\ExnVar}{\Kind}{\Subst{\Tm^\prime}{\Kind}{\Tm}}}{\ExnTy}{\Exn}$ by definition of substitution.
    By applying the induction hypothesis we obtain $\JudgeExnTy{\TyEnv}{\KiEnv,\BindKi{\ExnVar}{\Kind}}{\Subst{\Tm^\prime}{x}{\Tm}}{\ExnTy}{\Exn}$.
    The desired result can be constructed using \CiteRule{T-AnnAbs}.

    \def\currentprefix{term-substitution:T-App}
    \CaseRule{T-App} Our assumptions are
        \begin{align}
           & \JudgeExnTy{\TyEnv,\BindTy{x}{\ExnTy^\prime}{\Exn^\prime}}{\KiEnv}{\Tm_1}{\ExnTyArr{\ExnTy_2}{\Exn_2}{\ExnTy}{\Exn}}{\Exn} \labell{ass1} \\
           & \JudgeExnTy{\TyEnv,\BindTy{x}{\ExnTy^\prime}{\Exn^\prime}}{\KiEnv}{\Tm_2}{\ExnTy_2}{\Exn_2}. \labell{ass2}
        \end{align}
    We need to show that $\JudgeExnTy{\TyEnv}{\KiEnv}{\Subst{\Tm^\prime}{x}{(\TmApp{\Tm_1}{\Tm_2})}}{\ExnTy}{\Exn}$, which by definition of substitution is equal to
        \begin{align}
            & \JudgeExnTy{\TyEnv}{\KiEnv}{\TmApp{\left(\Subst{\Tm^\prime}{x}{\Tm_1}\right)}{\left(\Subst{\Tm^\prime}{x}{\Tm_2}\right)}}{\ExnTy}{\Exn}. \labell{ts}
        \end{align}
    By applying the induction hypothesis to \eqrefl{ass1} respectively \eqrefl{ass2} we obtain
        \begin{align}
            & \JudgeExnTy{\TyEnv}{\KiEnv}{\Subst{\Tm^\prime}{x}{\Tm_1}}{\ExnTyArr{\ExnTy_2}{\Exn_2}{\ExnTy}{\Exn}}{\Exn} \labell{ass1ih} \\
            & \JudgeExnTy{\TyEnv}{\KiEnv}{\Subst{\Tm^\prime}{x}{\Tm_2}}{\ExnTy_2}{\Exn_2}. \labell{ass2ih}
        \end{align}
    The desired result \eqrefl{ts} can be constructed by applying \CiteRule{T-App} to \eqrefl{ass1ih} and \eqrefl{ass2ih}.

    All other cases are either immediate or analogous to the case of \CiteRule{T-App}.
\end{proof}

\begin{lemma}[Inversion]\label{thm-inversion}\ 
    \begin{enumerate}
        \item If $\JudgeExnTy{\TyEnv}{\KiEnv}{\TmAbsExn{x}{\ExnTy}{\Exn}{\Tm}}{\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}}{\Exn_3}$, then
            \begin{itemize}
                \item $\JudgeExnTy{\TyEnv,\BindTy{x}{\ExnTy}{\Exn}}{\KiEnv}{\Tm}{\ExnTy^\prime}{\Exn^\prime}$,
                \item $\JudgeSubTy{\KiEnv}{\ExnTy_1}{\ExnTy}$ and $\JudgeSubExn{\KiEnv}{\Exn_1}{\Exn}$,
                \item $\JudgeSubTy{\KiEnv}{\ExnTy^\prime}{\ExnTy_2}$ and $\JudgeSubExn{\KiEnv}{\Exn^\prime}{\Exn_2}$.
            \end{itemize}
        \item If $\JudgeExnTy{\TyEnv}{\KiEnv}{\TmAnnAbs{\ExnVar}{\Kind}{\Tm}}{\ExnTyForall{\ExnVar}{\Kind}{\ExnTy}}{\Exn}$, then
            \begin{itemize}
                \item $\JudgeExnTy{\TyEnv}{\KiEnv,\BindKi{\ExnVar}{\Kind}}{\Tm}{\ExnTy^\prime}{\Exn^\prime}$,
                \item $\JudgeSubTy{\KiEnv,\BindKi{\ExnVar}{\Kind}}{\ExnTy^\prime}{\ExnTy}$,
                \item $\JudgeSubExn{\KiEnv}{\Exn^\prime}{\Exn}$.
                \item \TODO\ $e \notin \fv{\Exn}$ and/or $e \notin \fv{\Exn^\prime}$.
            \end{itemize}
    \end{enumerate}
\end{lemma}
\begin{proof}
1. By induction on the typing derivation.

    \CaseRule{T-Abs} We have $\ExnTy = \ExnTy_1$, $\Exn = \Exn_1$ and take $\ExnTy^\prime = \ExnTy_2$, $\Exn^\prime = \Exn_2$, the result then follows immediately from the assumption $\JudgeExnTy{\TyEnv,\BindTy{x}{\ExnTy}{\Exn}}{\KiEnv}{\Tm}{\ExnTy_2}{\Exn_2}$ and reflexivity of the subtyping and subeffecting relations.

    \def\currentprefix{inversion:T-Sub}
    \CaseRule{T-Sub} We are given the additional assumptions
        \begin{align}
            & \JudgeExnTy{\TyEnv}{\KiEnv}{\TmAbsExn{x}{\ExnTy}{\Exn}{\Tm}}{\ExnTyArr{\ExnTy_1^\prime}{\Exn_1^\prime}{\ExnTy_2^\prime}{\Exn_2^\prime}}{\Exn_3^\prime}, \labell{ass1} \\
            & \JudgeSubExn{\KiEnv}{\ExnTyArr{\ExnTy_1^\prime}{\Exn_1^\prime}{\ExnTy_2^\prime}{\Exn_2^\prime}}{\ExnTyArr{\ExnTy_1}{\Exn_1}{\ExnTy_2}{\Exn_2}}, \labell{ass2}\\
            & \JudgeSubExn{\KiEnv}{\Exn_3^\prime}{\Exn_3}.
        \end{align}
    Applying the induction hypothesis to \eqrefl{ass1} gives us
        \begin{align}
            & \JudgeExnTy{\TyEnv,\BindTy{x}{\ExnTy}{\Exn}}{\KiEnv}{\Tm}{\ExnTy_2^{\prime\prime}}{\Exn_2^{\prime\prime}}, \labell{ih1}\\
            & \JudgeSubTy{\KiEnv}{\ExnTy_1^\prime}{\ExnTy}, \quad \JudgeSubExn{\KiEnv}{\Exn_1^\prime}{\Exn}, \labell{ih2}\\
            & \JudgeSubTy{\KiEnv}{\ExnTy_2^{\prime\prime}}{\ExnTy_2^\prime}, \quad \JudgeSubExn{\KiEnv}{\Exn_2^{\prime\prime}}{\Exn_2^\prime}. \labell{ih3}
        \end{align}
    Inversion of the subtyping relation on \eqrefl{ass2} gives us
        \begin{align}
            & \JudgeSubTy{\KiEnv}{\ExnTy_1^\prime}{\ExnTy}, \quad \JudgeSubExn{\KiEnv}{\Exn_1^\prime}{\Exn}, \labell{inv1}\\
            & \JudgeSubTy{\KiEnv}{\ExnTy_2^{\prime\prime}}{\ExnTy_2^\prime}, \quad \JudgeSubExn{\KiEnv}{\Exn_2^{\prime\prime}}{\Exn_2^\prime}. \labell{inv2}
        \end{align}
    The result follows from \eqrefl{ih1} and combining \eqrefl{inv1} with \eqrefl{ih2} and \eqrefl{ih3} with \eqrefl{inv2} using the transitivity of the subtyping and subeffecting relations.

2. By induction on the typing derivation.

    \CaseRule{T-AnnAbs} We need to show that $\JudgeExnTy{\TyEnv}{\KiEnv,\BindKi{\ExnVar}{\Kind}}{\Tm}{\ExnTy}{\Exn}$, which is one of our assumptions, and that $\JudgeSubTy{\KiEnv,\BindKi{\ExnVar}{\Kind}}{\ExnTy}{\ExnTy}$ and $\JudgeSubExn{\KiEnv}{\Exn}{\Exn}$; this follows from the reflexivity of the subtyping, respectively subeffecting, relation (noting that $\ExnVar \notin \fv{\Exn}$).

    \CaseRule{T-Sub} Similar to the case \CiteRule{T-Sub} in part 1.

\end{proof}

\begin{theorem}[Preservation]\label{thm-preservation}
    If $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$ and $\JudgeEval{\Tm}{\Tm^\prime}$, then $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm^\prime}{\ExnTy}{\Exn}$.
\end{theorem}
\begin{proof}
    By induction on the typing derivation $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$.

    The cases for \CiteRule{T-Var}, \CiteRule{T-Con}, \CiteRule{T-Crash}, \CiteRule{T-Abs}, \CiteRule{T-AnnAbs}, \CiteRule{T-Nil}, and \CiteRule{T-Cons} can be discarded immediately, as they have no applicable evaluation rules.

    \TODO
\end{proof}

\subsection{Syntax-directed type elaboration}

\subsection{Type inference algorithm}

\begin{theorem}[Syntactic soundness]\label{thm-syntactic-soundness}
    If $\Reconstruct\ \TyEnv\ \KiEnv\ \Tm = \langle \ExnTy; \Exn \rangle$, then $\JudgeExnTy{\TyEnv}{\KiEnv}{\Tm}{\ExnTy}{\Exn}$.
\end{theorem}
\begin{proof}
    By induction on the term $\Tm$.

    \TODO
\end{proof}

\begin{theorem}[Termination]\label{thm-termination}
    $\Reconstruct\ \TyEnv\ \KiEnv\ \Tm$ terminates.
\end{theorem}
\begin{proof}
    By induction on the term $\Tm$.

    \TODO
\end{proof}

\clearpage

\section{TODO}

\begin{itemize}

    \item naming and ordering of quantified exception set variables in the examples is inconsistent.

    \item notation of type signatures in the examples is inconsisten.

    \item standard polyrec examples (DHM + GSM)

    \item polyrec does not ``come for free''

    \item type inference: we have a fixpoint a la DHM

    \item widening

    \item ``algebraic'' effects?

    \item unexpected decidablity

    \item ack: Vincent + Femke; Andrew + Jeremy + Stephanie + Andres; ST-RC

    \item check wiki and folder for notes

    \item exception type of twice (and other h-o funs)

    \item no slanted-greek for lambda and Lambda

    \item typeset System $F_\omega$ correctly

    \item roll Metatheory into earlier sections, add new section Analysis (also add to Overview)

    \item Untracked exceptions can break information flow security.

    \item Elaborte in the subsection ``Contributions''. Mention prototype?

    \item : vs ::

    \item add function composition as an additional example

\end{itemize}

\subsection*{Abstract}

    \begin{itemize}

        \item title: Higher-order effect types

    \end{itemize}

\subsection{Introduction}

    \begin{itemize}

        \item In the final example (map (const undefined)) no non-exceptional values will be \emph{actually} present in the resulting list

        \item Why not \ensuremath{\alpha\langle \Varid{e}_{\mathrm{1}} \rangle\xrightarrow{ \Varid{e}_{\mathrm{3}}}\beta\langle \Varid{e}_{\mathrm{2}} \rangle}?! Give some examples why higher-rankedness is needed. The example on the poster/\ensuremath{\Varid{map}} isn't sufficient. Postpone to a later section?

        \item Need an example that clearly demonstrates why HRP is \emph{needed} (May be difficult without also constructing a complete subtyping-based analysis?)

    \end{itemize}

\subsection{The $\LambdaUnion$-calculus}

    \begin{itemize}

        \item \TODOX{We can make union only work on base types (as we not longer \emph{need} to distribute unions over applications)? Only need during widening, now...}

        \item \TODOX{Prove more Lemmas about reduction rules (esp. $\gamma_1$)}

        \item Add reduction rules for $\emptyset$ and idempotence to the Figure. (Do wee need the bars above the rules?)

        \item Prove semantics is ACI1. We have a different unit for each type!
        \item $\mathcal{P}(V_{\tau_1} \to V_{\tau_2}) \simeq V_{\tau_1} \to \mathcal{P}(V_{\tau_2})$? Cardinallity suggests not: $2^{(\beta^\alpha)} \neq (2^\beta)^\alpha$.
        \item If we don't distribute unions over applications, can we ever get them deep inside terms?
        \item If we don't \emph{and} the outermost lambdas are not there because is always of kind star, can we get non-trivial terms? I.e. something other than $e_1(e_{11},...,e_{1n_1}) \cup \cdots \cup e_k(e_{k1},...,e_{kn_k}$ (note: $e$ and not $t$ as arugments).

    \end{itemize}

    \subsubsection{Widening}

        \begin{itemize}

            \item Does this give us any bounds on the complexity? (Or do the fact that we have arguments prevent this?)

            \item Footnote about \emph{narrowing}.

            \item Note that this is the reason $\cup$ needs to be higher-order.

        \end{itemize}

\subsection{Source language}

    \begin{itemize}

        \item We either need to omit the type annotations on $\TmCrash{\Ty}{\ExnLbl}$, or add them to $\TmIf{}{}{}$ and $\TmCase{}{}{}{}{}$.

        \item We do not have a rule \CiteRule{E-AnnAppExn}. Check that the canonical forms lemma gives us that terms of universally quantified type cannot be exceptional values.

        \item Replace the arrow with another one? (This one clashes with the reduction relation from $\LambdaUnion$.)

        \item Define the meaning of the double brackets in the redecution rule E-Op.

        \item Let-bindings can be defined in terms of abstractions (because HRP)

    \end{itemize}

\subsection{Exception types}

    \begin{itemize}

        \item ``Type signatures are denoted as ...'' there are several other ways used to write this

        \item The syntax of environments is omitted from the ``syntax'' figure

        \item Merge the figures with syntax and well-formedness?

        \item Elaborate on well-formedness

        \item Prove that $\xleftrightarrow{*} \iff \simeq$.

        \item \TODOX{Rename stuff in \CiteRule{T-App} in the elaboration system (now subtype/-effect of the result instead of the argument and clashes with the indices enumerated over by $i$!}

        \item \TODOX{\CiteRule{T-AnnAbs}: $e \notin \mathrm{fv}(\Gamma)$}

        \item $\ExnVar \in \ClassExnVar$

        \item Well-formedness of exception types: embed conservativity / full-flexibility?

        \item Can we roll \textsc{Univ} and \textsc{Arr} into a single construct: $\ExnTyForall{\ExnVar}{\Kind}{\ExnTyArr{\ExnTy_1}{\ExnVar}{\ExnTy_2}{\Exn(\ExnVar)}}$? Still need to deal with the well-formedness of $\ExnTy_1$... Also may need to quantify over more than one variable simultaneously...

    \end{itemize}

\subsubsection{Subtyping}

    \begin{itemize}

        \item Is \CiteRule{S-Refl} an admissable/derivable rule, or should we drop \CiteRule{S-Bool} and \CiteRule{S-Int}?

        \item Possibly useful lemma: $\ExnTy_1 = \ExnTy_2 \iff \ExnTy_1 \leq \ExnTy_2 \land \ExnTy_2 \leq \ExnTy_1$.

    \end{itemize}

\subsubsection{Conservative types}

    \begin{itemize}

        \item \TODOX{Atomicity: \ensuremath{\Varid{e}_{\mathrm{1}}\cup \Varid{e}_{\mathrm{2}}\to \ExnTyList{\Varid{e}_{\mathrm{1}}}{\Varid{e}_{\mathrm{2}}}} is not useful, because no introspection}

        \item \TODO check all examples types against prototype

        \item \TODO properly typeset example types

        \item \TODO Skolemization and explicit existential quantification over unification variables?

    \end{itemize}

\subsubsection{Declarative type system}

    \begin{itemize}

        \item \TODOX{Exception type erasure relation}

        \item \TODOX{Least upper bounds (declaratively, as subtypes)}

    \end{itemize}

    \begin{itemize}

        \item In T-AnnAbs, $\Gamma, \xi = \Delta$?

        \item In T-Abs and T-AnnAbs, should the term-level term-abstraction also have an explicit effect annotation?

        \item In T-AnnAbs, might need a side condition stating that $\ExnVar$ is not free in $\KiEnv$.

        \item In T-App, note the double occurence of $\Exn$ when typing $\Tm_1$. Is subeffecting sufficient here? Also note that we do \emph{not} expect an exception variable in the left-hand side annotation of the function space constructor.

        \item In T-AnnApp, note the substitution. We need a substitution lemma for annotations.

        \item In T-Fix, the might be some universal quantifiers in our way. Do annotation applications in $\Tm$ take care of this, already? Perhaps we do need to change $\TmFix{\Tm}$ into a binding construct to resolve this? Also, there is some implicit subeffecting going on between the annotations and effect.

        \item In T-Case, note the use of explicit subeffecting. Can this be done using implicit subeffecting?

        \item For T-Sub, should we introduce a term-level coercion, as in Dussart--Henglein--Mossin? We now do shape-conformant subtyping, is subeffecting sufficient?

        \item Do we need additional kinding judgements in some of the rules? Can we merge the kinding judgement with the subtyping and/or -effecting judgement? Kind-preserving substitutions.

    \end{itemize}

\subsubsection{Type elaboration system}

    \begin{itemize}

        \item In \CiteRule{T-App} and {T-Fix}, note that there are substitutions in the premises of the rules. Are these inductive? (Probably, as these premises are not ``recursive'' ones.)

    \end{itemize}

\subsection{Type inference}

    \begin{itemize}

        \item Complexity: reduction corresponds to agressive constraint simplification

        \item alternative (faster?) version of Kleene-Mycroft

        \item In R-App and R-Fix: check that the fresh variables generated by \ensuremath{\mathcal{I}(\cdot )} are subsituted away by the substitution \ensuremath{\theta} created by \ensuremath{\mathcal{M}(\cdot ; \cdot ; \cdot )}. Also, we don't need those variables in the algorithm if we don't generate the elaborated term.

        \item In R-Fix we could get rid of the auxillary underlying type function if the fixpoint construct was replaced with a binding variant with an explicit type annotation.

        \item For R-Fix, make sure the way we handle fixpoints of exceptional value in a manner that is sound w.r.t. to the operational semantics we are going to give to this.

        \item Note that we do not construct the elaborated term, as it is not useful other than for metatheoretic purposes.

        \item Lemma: The algorithm maintains the invariant that exception types and exceptions are in normal form.

        \item Typesetting issues \ensuremath{\llfloor \cdot \rrfloor_{\cdot }}

    \end{itemize}

\subsection{Related work}

    \begin{itemize}

        \item linear-algebriac lambda-calculi (Arrighi and Dowek, Vaux)

        \item More differences between \cite{HoldermansHage} (e.g. data types)?

        \item Christian Mossin. ``Exact flow types'' (intersection types, also non-elementary recursive by Statman)

        \item algebraic lambda calculus: higher-order $\to$ second-order?

    \end{itemize}

\subsection{Future research}

    \begin{itemize}

        \item higher-ranked algebraic effect types, Koka

    \end{itemize}
